### 2011-06-21 paper.tex
Got the basic skeleton of the paper going

\documentclass{sig-alternate}

\begin{document}

\title{A Catchy Title}
\numberofauthors{2}
\author{
\alignauthor
Ben Trovato\titlenote{Dr.~Trovato insisted his name be first.}\\
       \affaddr{Institute for Clarity in Documentation}\\
       \affaddr{1932 Wallamaloo Lane}\\
       \affaddr{Wallamaloo, New Zealand}\\
       \email{trovato@corporation.com}
% 2nd. author
\alignauthor
G.K.M. Tobin\titlenote{The secretary disavows
any knowledge of this author's actions.}\\
       \affaddr{Institute for Clarity in Documentation}\\
       \affaddr{P.O. Box 1212}\\
       \affaddr{Dublin, Ohio 43017-6221}\\
       \email{webmaster@marysville-ohio.com}
}
\additionalauthors{Additional authors: John Smith (The Th{\o}rv{\"a}ld Group,
email: {\texttt{jsmith@affiliation.org}}) and Julius P.~Kumquat
(The Kumquat Consortium, email: {\texttt{jpkumquat@consortium.net}}).}
\date{30 July 1999}

\maketitle
\begin{abstract}
A cool abstract
\end{abstract}

\category{H.4}{Information Systems Applications}{Miscellaneous}
\terms{Theory}
\keywords{ACM proceedings, \LaTeX, text tagging}

\section{Introduction}

\bibliographystyle{abbrv}
\bibliography{sigproc}  % sigproc.bib is the name of the Bibliography in this case

\end{document}

### 2011-06-21 paper.tex
Added in abstract and myself as an author

@@
 
-\title{A Catchy Title}
-\numberofauthors{2}
+\title{Chemoinformatics: The Computer Science of Chemistry}
+\numberofauthors{3}
 \author{
@@
        \email{webmaster@marysville-ohio.com}
+% 3rd author
+\alignauthor
+Rajarshi Guha\\
+\affaddr{NIH Center for Translational Therapeutics}\\
+\affaddr{9800 Medical Center Drive}\\
+\affaddr{Rockvile, MD 20850}\\
+\email{guhar@mail.nih.gov}
 }
+
 \additionalauthors{Additional authors: John Smith (The Th{\o}rv{\"a}ld Group,
@@
 \begin{abstract}
-A cool abstract
+  One of the most prominent success stories in all of science over the last decade has
+  been the advance of bioinformatics: interdisciplinary collaboration between
+  computer scientists and molecular biologists led to the sequencing the human genome
+  (among many other accomplishments), because researchers connected biological
+  technologies to the theory of string algorithms. However, despite this great
+  success, few computer scientists are familiar with a related (and older!)
+  discipline -- chemoinformatics, the use of computers to discover new molecules with
+  desired behavior, such as new pharmaceuticals or industrial solvents. Until
+  recently, the data and techniques of chemoinformatics have been closely guarded
+  secrets of companies whose financial success depended on being the first to produce
+  the new "miracle molecule". Only within the last decade -- and largely because of
+  chemists volunteering their time for an Open Science "movement" -- do researchers
+  now have access to freely available software packages and databases of tens of
+  millions of chemicals. As a result, chemists now confront hundreds of unsolved
+  algorithmic problems that could not have been tackled a decade ago, but whose
+  solutions are critical to research ranging from determining the behavior of small
+  molecules in biological pathways, to finding a cure for malaria.
 \end{abstract}
@@
 \terms{Theory}
-\keywords{ACM proceedings, \LaTeX, text tagging}
+\keywords{cheminformatics, graphs, chemistry}
 
-\section{Introduction}
+\section{Is Cheminformatics the New Bioinformatics?}
 
 \bibliographystyle{abbrv}
-\bibliography{sigproc}  % sigproc.bib is the name of the Bibliography in this case
+\bibliography{paper}
 

### 2011-06-21 paper.tex
Added first section. Also added a place holder figure - can we use PDF?

@@
 
-\bibliographystyle{abbrv}
+Novel technologies in the life sciences churn out information at an ever
+increasing rate – public data stores such as at the European Bioinformatics
+Institute (EBI) contain on the order of 10 Petabytes of biological information,
+a long way from the few kilobytes of the first sequenced genome in 1977.
+However, while biological information has been available publicly for decades,
+the same has not been true for chemical information until very recently. Only
+2004 saw a large public compound repository, PubChem, entering the public
+domain, followed more recently by other databases as discussed later in this
+article. In the same vein, open source software and the publication of
+algorithms have only recently become a main focus of the field \cite{faulon2010}.
+
+So why should we actually care about chemical information being made public --
+and how does this relate to the field of computer science?
+
+Both of these questions have relatively straightforward answers. We should care
+about chemical information being made public since the drugs we take are the
+result of intensive and costly research, and the more information is shared (a
+process sometimes very hard to achieve in pharmaceutical companies) the more
+easy it should be to develop novel treatments drugs. How chemical information
+relates to computer science becomes clear when realizing the amount of data
+available currently – 74 million different molecules with annotations are
+currently stored in public databases such as PubChem with a large amount of
+annotations, and the design of data structures for chemicals, as well as
+subsequent data mining of this information, are areas where expertise in
+computer science is dearly needed. (For a recent, detailed introduction to the
+field written for computer scientists see \cite{brown2009} .) Cheminformatics comprises
+different areas, which broadly speaking can be divided into three fields:
+capturing data (using lab notebooks or potentially using formats such as the
+Chemical Markup Language for publications); storing data (designing database
+schemas, devising ontologies) and mining data (such as for bioactivity
+prediction, which might involve algorithms based on graphs). However, what is a
+major difference to the bioinformatics field is the type of information being
+analyzed: While bioinformatics often deals with sequences, the domain of
+cheminformatics is chemical structures. In case of the former, information can
+very frequently be represented as one-dimensional strings, which are relatively
+easy to handle computationally. On the other hand, chemical structures may
+possess rings, branches as well as multiple valid representations of even the
+same structure, potentially giving rise to ambiguities. Hence, chemical
+structure is often thought of as being more difficult to standardize than
+information from other domains. As an illustration, different representations of
+the same chemical structure are shown in Figure 1. The top left hand corner
+shows a structure a chemist would draw it – every ‘corner’ representing a carbon
+atom (which can also be represented as a C), every O denoting oxygen, every N
+nitrogen, every H hydrogen and so on. This is the standard format chemists use
+when exchanging information with their peers. However, this is not directly
+amenable to data mining -- different atoms might be collapsed into one node, sets
+of atoms might have the same relevant properties assigned. Hence, the top right
+hand corner of Figure 1 shows an annotated graph representation, taking into
+account that atoms, though being of different elemental atom type, can have
+similar properties -- e.g. both oxygen and nitrogen carry an 'On' label in this
+graph.
+
+\begin{figure}
+\centering
+\psfig{file=placeholder.ps}
+\caption{The many different facets of representing chemical structures. Shown
+are a chemist’s usually representation of a molecule (top left) and a
+graph-based representation amenable to data mining (top right). These
+representations are supplemented by a 3D representation of the same molecule
+(bottom left) that is closer to reality and a representation using surface
+properties (bottom right), which is often more relevant for the properties the
+molecule exhibits in an experimental setup (such as when measuring its
+solubility, or bioactivity).}
+\end{figure}
+
+On this graph, and employing large databases of thousands of molecules, graph
+mining algorithms can be applied to discover patterns in large-scale datasets
+and a large number of successful applications of this concept have been
+published \cite{wegner2006,horst2009}. Still, this second representation is far
+from being the end of the story. Molecules are 3D entities as represented in the
+bottom left hand corner of the figure, and the precise 3D shape is relevant in
+many situations (while not in others), but this is information that is lost when
+only considering the graph of a molecule. To even extend this concept, in many
+cases actually the surface properties of theq molecule are what appears to be
+relevant in e.g. biological systems, as shown in the bottom right hand corner.
+Hence, while graph mining on chemicals is certainly a promising avenue to
+explore, one needs to keep in mind that not all molecular properties are encoded
+in the graph (such as rotational states, or different charge states etc.) and
+thus often other 'descriptors' need to be used to capture the essentials of a
+chemical system \cite{bender2004}.
+
+From the above it becomes clear that the nature of chemical problems is often a
+complex mixture of combinatorial and continuous problems. If we oversimplify the
+nature of chemical structures and their properties we can not describe the world
+around us in sufficient details. On the other hand, if we try encoding all
+details the complexity to collect, store, and mine data increases dramatically
+(e.g. all possible 3D structures and their properties, all metabolites of a
+structure in humans in different tissues). For the field of cheminformatics to
+flourish, we need closer collaboration between chemists (or, more generally,
+life scientists) and computer scientists – where the former needs to be able to
+pose the problem in a way relevant for practical applications; and where the
+latter is able to devise ways of capturing, storing, and analyzing chemical data
+by bridging the gap between risk in generalizable encodings and complexity
+(space and time). The goal is supporting a better chemical decision making by
+reducing the risks by 1. storing and integrating data in maintainable ways, 2.
+providing enough open standards and tools to allow software engineering concepts
+to help bridging silos, and 3. mining the many chemical property spaces in a
+time and space efficient way. Many aspects pertaining to those areas will be
+discussed in the following sections.
+
+\section{Bridging Cheminformatics \& Computer Science}
+\label{sec:bridg-chem-}
+
+\subsection{Databases \& Ontologies}
+\label{sec:datab--ontol}
+
+\bibliographystyle{srt,abbrv}  
 \bibliography{paper}

### 2011-06-21 paper.tex
Put in all current sections. Some references are put in

@@
 
-\begin{figure}
-\centering
-\psfig{file=placeholder.ps}
-\caption{The many different facets of representing chemical structures. Shown
-are a chemist’s usually representation of a molecule (top left) and a
-graph-based representation amenable to data mining (top right). These
-representations are supplemented by a 3D representation of the same molecule
-(bottom left) that is closer to reality and a representation using surface
-properties (bottom right), which is often more relevant for the properties the
-molecule exhibits in an experimental setup (such as when measuring its
-solubility, or bioactivity).}
-\end{figure}
+% \begin{figure}
+% \centering
+% \psfig{file=placeholder.ps}
+% \caption{The many different facets of representing chemical structures. Shown
+% are a chemist’s usually representation of a molecule (top left) and a
+% graph-based representation amenable to data mining (top right). These
+% representations are supplemented by a 3D representation of the same molecule
+% (bottom left) that is closer to reality and a representation using surface
+% properties (bottom right), which is often more relevant for the properties the
+% molecule exhibits in an experimental setup (such as when measuring its
+% solubility, or bioactivity).}
+% \end{figure}
 
@@
 
-\bibliographystyle{srt,abbrv}  
+Most cheminformatics applications rely on large databases of chemical structures
+and their properties. Organising and maintaining such databases, searching based
+on features of chemical structures, and clustering similar structures together,
+are essential utilities to enable many scientific applications. Yet, each one of
+those challenges poses its unique computer science complexities.
+
+There is a considerable progress in open chemical and bioactivity databases like
+ChEMBL and NCBI-PubChem, which are both freely available. Still, the integration
+of chemical databases with each other remains a challenge, not only due to the
+normalization problems in the chemical and bioactivity landscape, but also due
+to the sheer data volume of them. As mentioned earlier, we have to be able to
+search in complex data, for example multi-labelled molecular graphs, where
+labels can be different tautomeric states, different protonation states, or many
+variations of implicit chemical expert knowledge we need to capture in a
+risk-bounded machine encoded way.
+
+Chemical graphs can differ based on the manner in which the structures are
+drawn. Checking whether two graphs represent the same chemical is an expensive
+operation (the next section will discuss this in-depth), thus, a representation
+of the structure that is unique and invariant with respect to how the structure
+is drawn is needed for identification. An early effort in this direction was the
+Simplified Molecular Input Line Specification (SMILES). However, several
+different implementations became available. After years of research and various
+commercial products, is there today not a single public standard for a
+line-notation of molecular labelled graphs, which is available to a broad
+audience (academic and commercial). The very same problem does also exist for
+bioassay ontologies and the lack of options to normalize the property spaces of
+chemical compounds, e.g. biological assays or clinical data outcome
+(biobanking).
+
+With each database using their own algorithm for molecular encoding, little
+facilitated the automated exchange of data between different databases. As more
+and more data became available online, structure-based identification became a
+pressing need, and the IUPAC International Chemical Identifier (InChI) code was
+developed. The InChI is a non-proprietary, structured textual identifier for
+chemical entities [1]. InChI identifiers are not intended to be read and
+understood by humans, but are useful for computational matching of chemical
+entities. For rapid database lookups, the InChIKey is a hashed key for the
+InChI, which has an invariant length of 14 characters. Figure 1 illustrates the
+SMILES, InChI and InChIKey for lipoic acid.
+
+Figure 1: The standard chemical graph illustration together with the SMILES,
+InChI and InChIKey codes for lipoic acid. Note that the SMILES is much more
+human-readable than the others.
+
+The InChI is now widely used for matching identical chemical structures, but it
+is still limited in a few scenarios. For example, it cannot differentiatebetween
+certain types of stereoisomers: Figure 2 illustrates two stereoisomers for which
+the generated InChI is the same.
+
+Figure 2: The generated InChI is the same for the stereoisomers cisplatin and
+transplatin.
+
+InChI or other identity-mapping algorithms allow for exact searching (within the
+limitations of the algorithms), but two other requirements for scientific
+discovery based on chemical databases are substructure searching and similarity
+searching. In substructure searching, the database is searched for a specified
+wholly contained part of the structures in the database; in similarity
+searching, the database is searched for structures similar to a provided search
+structure. Chemical search packages are often implemented and optimised for a
+given database technology, for example, the OrChem package is an open source
+chemical search package for the Oracle database \cite{rijnbeek2009}.
+
+Graph substructure matching is a variant of graph isomorphism detection,which is
+well known to be computationally intractable [3]. To execute a graph isomorphism
+search across a full chemical database of thousands or even millions of
+structures is completely infeasible. For that reason, complex heuristics must be
+used to drastically filter the candidate search space. Structure-based
+fingerprints are commonly used for this purpose. Fingerprints encode
+characteristic features of a given chemical structure in a boolean array, or
+bitmap. The fingerprint is created by an algorithm which generates bitmap
+patterns for each atom in the structure; then each atom and its nearest
+neighbours, including the bonds between them; then each group of atoms connected
+by paths of increasing lengths up to the maximum implemented for the algorithm
+(commonly around eight). The bitmap patterns are added together into the final
+fingerprint. If all bits in a query fingerprint are also present in the target
+fingerprint of a stored database structure, only then is the structure subjected
+to the computationally expensive subgraph matching algorithm. Bit operations are
+very fast and independent of the number of atoms in a structure due to the fixed
+length of the fingerprint.
+
+Fingerprints also facilitate the rapid calculation of quantitative
+structure-based similarity measures. An example measure of similarity is the
+Tanimoto coefficient, calculated as the ratio $T(a,b) = \frac{c}{(a+b−c)}$,
+where c is the count of bits on in the same position in both the two
+fingerprints, a is the count of bits on in the first structure, and b is the
+count of bits on in the second structure. The Tanimoto coefficient varies in the
+range 0.0 -- 1.0, with a score of 1.0 indicating that the two structures are very
+similar (i.e. their fingerprints are the same). Structure-based similarity is of
+limited use in grouping of chemicals based on non-structural features, such as
+shared bioactivity profiles. For this purpose, a large-scale and flexible
+classification structure is needed for chemicals and their linked biological
+information. One such answer to this need is provided in the form of ontologies.
+An ontology is a formal specification of entities and their relationships in a
+domain of interest. It is structured around an underlying logical representation
+such as Description Logic [baaderdl2007]. The logic-based representation allows
+automated reasoning to derive new knowledge (inferences) from the knowledge
+encoded, making the representation both compact and immensely powerful.
+Ontologies are already in widespread use across the biomedical domain. A
+similarity measure based on the information encoded in ontologies is termed
+‘semantic similarity, and such measures have already been applied to enhance
+chemical classification in several problem areas [couto2010]. Many open
+questionsstill remain as to the best way to combine the information encoded in
+chemical structures with chemical ontologies [hastingsowled2010].
+
+\subsection{Structure Enumeration}
+\label{sec:struct-enum}
+
+Enumerating molecules is a combinatorial problem that has fascinated chemists,
+computer scientists and mathematicians alike for more than a century. It is one
+important risk-reduction element in still being able to encode closely-related
+molecules and for keeping the space complexity of databases under (algorithmic)
+control. Indeed, while trying to solve the problem of counting the isomers of
+paraffin structures or counting substituted aromatic compounds, fundamental
+principles in graph theory and combinatorics were developed by Cayley, Polya and
+others. Even terms that are widely-used nowadays like graph and tree were
+originally coined in a chemistry context (Chapters 1 and 8 in [1]). More than
+four decades ago, Joshua Lederberg, Carl Djerassi, Edward Feigenbaum and others
+from Stanford University started developing codes to explicitly enumerate
+molecules. Their project named DENDRAL was funded by NASA with the goal of
+developing an automated system to elucidate compounds from spectral data. While
+studying this challenging problem, important concepts in computer science were
+developed. DENDRAL is generally quoted in artificial intelligence textbooks as a
+pioneer project that produced the first expert system.
+
+Enumerating molecules is not only an interesting academic exercise but has
+practical utility as well. The foremost application of enumeration is structure
+elucidation of natural compounds such as metabolites, geochemicals, and
+petrochemicals. Ideally, the wishful bench chemist collects experimental data
+from an unknown compound, the data is fed into a code, and the resulting
+structure is returned. Although that streamlined process may not yield a unique
+solution, there are commercial software products such as MOLGEN
+(http://molgen.de) that can, for instance, list all structures matching a given
+molecular formula, or mass spectrometry (MS) data. Another important application
+of enumeration is molecular design. Here the problem is to design compounds
+(drugs, for instance) that optimize some physical, chemical, or biological
+property or activity. Although less prolific and more recent than structure
+elucidation, molecular design has introduced some novel solutions to molecular
+enumeration. Finally, with the advent of combinatorial chemistry, molecular
+enumeration has taken a central role as it allows computational chemists to
+construct virtual libraries, test hypotheses, and provides guidance in the
+design of optimal combinatorial experiments.
+
+The major difficulty with enumeration is that the in silico representation of
+molecules, the so-called molecular graphs, are labeled object (i.e. labeled
+graphs), while the atoms of a molecule are of course not uniquely labeled. The
+mathematical concept to tackle this problem is to consider orbits of labeled
+molecular graphs under the operation of the symmetric group. When enumerating
+molecular structures, one has thus to solve the so–called isomorphism problem
+The good news is that isomorphism can be solved in polynomial time for molecular
+graphs due to the fact that molecules belong to a restricted class of graphs
+known as bounded valence graphs; however, that same restriction prevents
+applying for molecular graph the polynomial delay enumeration results obtained
+for general graphs (Chapter 8 in [1]). While the complexity of molecular graph
+enumeration remains an open problem it has nonetheless been shown that molecules
+can be sampled efficiently [2]. Sampling procedures based on the Metropolis or
+the Genetic Algorithms have been developed and used to elucidate natural
+compounds from NMR data.
+
+The underlying computational complexity and potential intractability of
+molecular enumeration may not be a bottleneck after all, as computational
+chemists have developed and successfully used enumeration tools to generate
+large chemical libraries like GDC-13 comprising almost a trillion chemical
+structures (http://www.dcb-server.unibe.ch/groups/reymond/gdb/home.html). Yet,
+the current enumeration software products do not generally produce stereoisomers
+[3] and tautomers [4], these require specific enumeration procedures, which are
+still the subject of investigations by the chemoinformatics community. GDB-13
+along with other databases such as PubChem (NCBI/NIH) or ZINC (UCSF) are
+valuable resources to search for new drugs. For instance the well known
+ligand-based design approach is making use of these databases along with
+structure similarity tools to discover new compounds binding to biological
+targets (protein or nucleic acid). Compounds retrieved through ligand-based
+design, however, must already be present in databases, which is a limitation of
+the technique. To palliate that limitation algorithms have recently been
+developed to enumerate all compounds in the chemical space matching circular [5]
+or path fingerprints [6], these fingerprints being popular tools to measure
+similarity. Enumerating molecules from fingerprints, or more generally
+enumerating molecules from structure-activity molecular descriptors, is still
+very much into development. At stake is the design of new chemicals or drugs
+that have not yet been synthesized.
+
+Another area where enumeration plays a role is in the generation of chemical
+reaction networks. The problem here consists of enumerating all the possible
+compounds that can be produced by applying reaction rules to a set of initial
+molecules. By reversing the reactions rules one can also find sets of starting
+compounds necessary for producing a given target. This latter process is known
+in chemistry as retrosynthesis since the works of E.J. Corey who received the
+Nobel Price in Chemistry in 1990. Designing new drugs or chemicals,
+understanding the kinetics of combustion and petroleum refining, studying the
+dynamics of metabolic networks, applying metabolic engineering and synthetic
+biology to produce heterologous compounds in microorganisms, all are
+applications that require to enumerate reaction networks. As reviewed in Chapter
+11 in [1] several network enumeration techniques have been developed. However,
+these techniques generally suffer from a combinatorial explosion of intermediate
+compounds being produced. One way to limit the number of compounds generated is
+to simulate the dynamics of the network while it is being constructed and remove
+compounds having a low concentration. Following that idea, methods have been
+developed based on the Gillespie Stochastic Simulation Algorithm (SSA) to
+compute on-the-fly species concentrations. Chemical reaction network enumeration
+and sampling is an active field of research particularly in the context of
+metabolism, either to study biodegradation, or to propose metabolic engineering
+strategies to biosynthesize compounds of commercial interest. The difficulty
+with metabolic network design is that in addition to network generation based on
+reactions, one also needs to verify that there are enzymatic sequences enabling
+reactions catalyses. That additional task requires encompassing both chemical
+and sequence information and the development of tools that are at the interface
+between chemoinformatics and bioinformatics.
+
+\subsection{Activity Mining \& Prediction}
+\label{sec:activity-mining-}
+
+The basis of predictive modeling in cheminformatics is that chemical structure
+implies biological activity. Thus, the goal of any modeling approach is to
+capture and characterize the correlations between structural features and the
+observed biological activities. The goal is to describe cause-effect
+relationships with the lowest possible risk, and also being able to describe the
+risk of being wrong (confidence) when using such models for decision making. In
+many cases, the activity of a small molecule is due to its interaction with a
+receptor. Traditional QSAR approaches ignore receptor interactions, but focus
+only on small molecule features, and therefore loose valuable information. As a
+result, techniques such as proteochemometric modeling methods have been designed
+to take into account both ligand and receptor structures.
+
+The first step in the predictive modeling of biological activities is to compare
+compounds. Either we transform molecular graphs into a numerical vector
+representation (a.k.a., descriptors or features) of chemical structures, or we
+use algorithms for comparing molecular graphs directly (molecule kernel). In the
+first case thousands of descriptors (many of which are equivalent) have been
+described [Todeschini, Handbook of Molecular Descriptors] and a variety of tools
+are available for their calculation. Given a set of molecules, observed
+activities and a descriptor vector, the problem can be considered a traditional
+statistical modeling problem. In the second case no explicit vector
+representation is required, which prevents the molecular graph to vector
+transformation challenge (loss-less encoding risk). On the other hand imposes
+the implicit kernel encoding that we have to compare all compounds against all
+compounds, which imposes large-scale mining challenge (time and space complexity
+risk). Note that the public PubChem dataset alone contains a double-digit
+million number of compounds.
+
+First, we must consider the goal of a predictive model in a cheminformatics
+setting. In some cases, pure predictive ability is desired - such as in a
+virtual screening setting. For such cases, the complexity of the modeling
+approach is immaterial, as it leads to accurate and generalizable results. Here
+a large-scale mining capacity is crucial. On the other hand, when working with
+chemists and biologist the interpretability of an model and risk reduction
+arguments are paramount, since we have to translate better predicted in-silico
+compounds into real-world compounds. In such cases, black box methods can be
+less useful, though attempts have been made to interpret vectorial models built
+using neural networks, SVMs or random forests. Related to this, there is the
+never ending debate on the choice of descriptors, e.g. by applying feature
+selection.
+
+Second, the concept of model applicability - when is the prediction for a new
+object reliable (confidence)? - has grown in importance with the increasing use
+of predictive models in regulatory settings. A variety of methods have been
+developed that attempt to characterize the “domain” of a model. It does not only
+allow if models are applicable, but can also be used for deciding if additional
+biological experiments are required for reducing the risk on certain compound
+classes.
+
+One of the key challenges that faces predictive modeling is the fact that small
+molecules are not static and do not exist in isolation. Traditionally,
+predictive models have focused on a single structure for a small molecule and
+have ignored the role of the receptor (in receptor-mediated scenarios). Yet,
+small molecules can exist in multiple tautomeric forms and usually in multiple
+conformations. As a result, enhancing the accuracy of predictions ideally will
+require that such aspects be taking into account and, as far as possible, take
+into account the receptor simultaneously. Multi-conformer modeling has been
+addressed in the 4D-QSAR methodology described by Hopfinger and co-workers.
+Techniques such as multiple-instance learning could also be applied to the
+multi-conformer problem. More recently, simultaneous modeling of receptor and
+ligand has been addressed by the proteochemometric approach (WHO FIRST STARTED
+THIS?).
+
+With the advent of high-throughput screening technologies, large libraries of
+compounds can be screened against multiple targets in an efficient manner. Such
+panel assays provide a broad, systems-level view of small molecule activities.
+Models developed on such data afford us the opportunity to identify targets,
+characterize off-target effects and so on. However, most approaches to this
+problem are rather simplistic since they develop multiple individual models.
+Instead one could imagine an approach that takes into account the correlations
+in observed activities against multiple targets within a single model. Such an
+approach should lead to more robust predictions for panel assays. Finally, does
+this also better reflect clinical relevant compound profiles [SIDER database]
+and a ‘personalized medicine’ concept (e.g. drug-drug interaction profiles) [The
+Decision Tree].
+
+\subsection{Knowledge Management}
+\label{sec:knowledge-management}
+
+The relevance of scientific knowledge management is increasing not only for
+reducing the risk in current research, but also for enabling new
+collaboration/innovation opportunities with internal/external and fast growing
+number of external/public partners. In fact, in cheminformatics and chemistry,
+scientists have already switched many years ago from paper lab-notebooks to
+online collaboration platforms, called ELN (Electronic Lab Notebook). So, why
+have chemists and companies already adopted to 'Enterprise 2.0', an social
+online collaboration culture? Many other disciplines are still busy with solving
+ROI (return-of-investment), risk reduction, opportunity creation, change
+culture, and technical collaboration challenges for scientists and scientific
+data.
+
+Before the millenium (<year 2000) many chemists were still using paper
+lab-notebooks, multiple compound registration, and search tools. Compared to
+today there were many pitfalls and drawbacks with this. The overall architecture
+was too inflexible to adopt to fast changing data standards, scientists spend
+too much time with administrative work, chemical data quality was inconsistent,
+an alignment with other working groups was inefficient (e.g. for running
+analytical experiments), a legal dispute required to find data in paper
+lab-notebooks, data synchronization between systems caused lag times and access
+delays, and local silos hampered an efficient global collaboration. Now,
+chemists prefer ELNs, because they can search in data of colleagues. This does
+not only allow to find experts faster, but allows also access to highly trusted
+data with approved quality workflows. Since the data allows also access to
+analytical and legal data directly every new ELN release is aiming at increasing
+the integration of scientific and business workflows for reducing the risk due
+to irrational or bounded-rational decision making and increasing decision making
+efficiency.
+
+ELNs are typically much stronger accepted and used within the cheminformatics
+domain than in the bioinformatics domain, which is picking up speed lately. The
+question is why? In chemistry every experiment requires a lot of expert
+knowledge of chemists. Therefore, approaches for reducing risks by communicating
+with fellow chemists about similar reactions or compounds is critical. This
+requires capturing the reaction details, compound details, and time consuming
+analysis ensuring quality data, which can be searched by others.
+
+It is worth mentioning that chemistry is traditionally a field where many people
+have been used to huge bookshelves of reactions and compounds. Chemists changed
+not only to online ELN collaboration for getting more efficient, but also
+because it allowed collaboration with trusted colleagues. So, third party paper
+catalogs and academic publications they used previously proofed being
+inefficient, and even more critical, they did not take compounds of trusted
+colleagues into account. Overall the change management to ELNs, similar to
+'Enterprise 2.0', required to deliver on the promise to increase
+interconnectivity, serve as silo-breaker, and to improve collaboration
+opportunities, and it did … many years ago, already.
+
+The future still leaves many challenges ahead. Using external/public databases
+with chemical and bioactivity remains a challenge due to differences in
+ontologies, synchronization and maintenance challenges, and efficient
+substructure and similarity search within complex data types, aka compounds. A
+collaboration with external parties, e.g. contract research, poses again
+problems with duplicate checking for compounds (e.g. tautomeric or salt
+structures), efficient data synchronization, most often a double-registration is
+required for fulfilling the data quality processes on both sides. If external
+partners are small they often do not even have sufficient IT resources
+themselves and rely on external services themselves. Cloud services could help
+not only for providing a service infrastructure for all parties being involved,
+but also provide the required private/public access toolbox. Furthermore are
+there still many data sources not being indexed properly, since chemistry
+patents are often cryptic and image/text mining remains a puzzle. One could
+argue that the image/text mining in scientific journals is less challenging, in
+fact, it remains a huge challenge and many chemists do not trust automated
+non-expert validated mining sources. The closed CAS (Chemical Abstract Service)
+is a highly trusted source, and public chemistry and bioactivity space has to
+improve quality and interconnectivity. Efforts like Chembl and PubChem-BioAssay
+are on the right track. Still, improving data quality and standards between
+public and closed sources will be absolutely critical for ensuring a constant
+growth, usage, and collaboration scenarios between private and public parties.
+Only when the technical and scientific questions are solved we can ensure that
+also legal questions about contribution metrics will get enough room to grow.
+
+\section{The Role of Open Source \& Open Data}
+\label{sec:role-open-source}
+
+\subsection{Open Source Chemical Toolkits}
+
+A key success of the open source movement is that experts can team up for
+collaborating in their expert areas. Especially in the chemistry domain, with a
+strong expert system history, this is even more relevant due to its complex data
+types and relation to many outcome variables, like biological activities,
+toxicological effects, influence to diseases, or effects on the environment.
+
+Open Babel and the Chemistry Development Kit (CDK) are two of the most widely
+used Open Source chemical toolkits. Such toolkits are used, for example, to
+read/write chemical file formats, manipulate chemical structures, measure
+similarity of molecules, search for substructures, and generate 2D diagrams. The
+underlying algorithms used include graph algorithms (e.g. maximal common
+substructure, canonical labelling of graphs), geometrical methods (e.g. Kabsch
+alignment) and vector manipulation (e.g. converting coordinates in various
+systems, 3D structure generation). As well as being of direct use to practising
+cheminformaticians in their day-to-day work, many applications have been
+developed that rely on these toolkits to handle chemical data; for example, the
+molecular viewer Avogadro (http://avogadro.openmolecules.net) uses Open Babel,
+while the molecular workbench Bioclipse \cite{Bioclipse2} uses the CDK.
+
+Open Babel (http://openbabel.org) is composed of a set of command-line
+applications, a GUI, and a C++ library with SWIG-generated bindings in several
+languages. It runs cross-platform and uses CMake and CTest for building and
+testing. One of its unique features is support for a large number of chemical
+file formats and its babel command-line application is widely used for
+conversion between them. The origin of Open Babel is OELib, a chemical toolkit
+released under the GPL by OpenEye Scientific Software. In 2001, OpenEye decided
+to discontinue work on OELib and to focus instead on an internal rewrite of the
+library which would not be open source (OEChem). At that point, OELib was
+renamed to Open Babel and since then it has been maintained and developed by a
+community led by Dr Geoffrey Hutchison (University of Pittsburgh).
+
+The Chemistry Development Kit (http://cdk.sf.net, \cite{Steinbeck2006}) project
+originated in 2000 as a common base for two existing Java applications Jmol and
+JChemPaint. The initial code bases inherited from both these project and
+CompChem. It soon outgrew this initial goal and is now a fully-featured
+cheminformatics toolkit in its own right. The CDK is a Java library and uses Ant
+and JUnit for building and testing, as well as PMD and OpenJavaDocCheck for code
+quality control. A particular strength of the toolkit is its support for a large
+number of so-called “molecular descriptors” (numerical representations of
+chemical structures used in machine learning to link chemical structure to a
+physical or biological property). The project was initially led by Dr Christoph
+Steinbeck (EMBL-EBI, UK) and Dr Egon Willighagen (Karolinska Institutet,
+Sweden), and more recently Dr Rajarshi Guha (NIH, USA) joined as third, current
+project leader.
+
+Both the CDK and Open Babel are very open to new contributors to the projects.
+They both have an open mailing list, without any further private lists.
+Likewise, issue tracking systems are fully open, and ongoing and future
+development is openly discussed. Contributors are encouraged to get involved
+with any of the following areas: development of new features, improvement of
+existing features, performance optimisation, fixing bugs, documentation, website
+development, reporting bugs, and helping users on the mailing list. Anyone
+interested should email the respective developer’s mailing list (either
+cdk-devel@lists.sf.net or openbabel-devel@lists.sf.net) with their interests and
+background, and a mentor will be assigned to help them join the project.
+
+Contributions as such are made to both projects on a regular basis, both from
+industries more interested in the use and application of these libraries, as
+well as from research groups in academia. And example here is the contribution
+of a ring search algorithm replacing the earlier Figueras algorithm by
+mathematicians from the TU M\"{u}nchen in Germand (\cite{Berger2004}). The former is
+maintained in the CDK, as the project also aims to provide an educational
+resource.
+
+Challenges faced by the CDK and OpenBabel libraries are abundant. Many of these
+originate from the complexity of the graph theory used in chemistry, which is an
+approximate but insufficient representation to capture chemical reality.
+Existing areas of interest are enumeration of colored graphs, taking into
+account symmetry, and symmetry detection itself, which not merely derives from
+the chemical graph, but typically also from the 3D geometry of the chemical.
+However, more challenging is that chemical representation of a molecule really
+needs to capture non-static graphs, in order to take into account delocalisation
+and tautomeric phenomena molecules can practically undergo.
+
+\bibliographystyle{abbrv}  
 \bibliography{paper}

### 2011-06-21 paper.tex
Fixed a variety of non ascii symbols

@@
 relates to computer science becomes clear when realizing the amount of data
-available currently – 74 million different molecules with annotations are
+available currently -- 74 million different molecules with annotations are
 currently stored in public databases such as PubChem with a large amount of
@@
 computer science is dearly needed. (For a recent, detailed introduction to the
-field written for computer scientists see \cite{brown2009} .) Cheminformatics comprises
-different areas, which broadly speaking can be divided into three fields:
-capturing data (using lab notebooks or potentially using formats such as the
-Chemical Markup Language for publications); storing data (designing database
+field written for computer scientists see \cite{brown2009}). Cheminformatics
+comprises different areas, which broadly speaking can be divided into three
+fields: capturing data (using lab notebooks or potentially using formats such as
+the Chemical Markup Language for publications); storing data (designing database
 schemas, devising ontologies) and mining data (such as for bioactivity
@@
 the same chemical structure are shown in Figure 1. The top left hand corner
-shows a structure a chemist would draw it – every ‘corner’ representing a carbon
+shows a structure a chemist would draw it -- every 'corner' representing a carbon
 atom (which can also be represented as a C), every O denoting oxygen, every N
@@
 when exchanging information with their peers. However, this is not directly
-amenable to data mining -- different atoms might be collapsed into one node, sets
-of atoms might have the same relevant properties assigned. Hence, the top right
-hand corner of Figure 1 shows an annotated graph representation, taking into
-account that atoms, though being of different elemental atom type, can have
+amenable to data mining -- different atoms might be collapsed into one node,
+sets of atoms might have the same relevant properties assigned. Hence, the top
+right hand corner of Figure 1 shows an annotated graph representation, taking
+into account that atoms, though being of different elemental atom type, can have
 similar properties -- e.g. both oxygen and nitrogen carry an 'On' label in this
@@
 this also better reflect clinical relevant compound profiles [SIDER database]
-and a ‘personalized medicine’ concept (e.g. drug-drug interaction profiles) [The
+and a 'personalized medicine' concept (e.g. drug-drug interaction profiles) [The
 Decision Tree].

### 2011-06-21 paper.tex
some fixes here and there
Added some text to highight the "stort" were trying present

@@
 Novel technologies in the life sciences churn out information at an ever
-increasing rate – public data stores such as at the European Bioinformatics
-Institute (EBI) contain on the order of 10 Petabytes of biological information,
-a long way from the few kilobytes of the first sequenced genome in 1977.
+increasing rate -- public data stores such as at the European Bioinformatics
+Institute (EBI) contain on the order of 10 PB of biological information.
 However, while biological information has been available publicly for decades,
-the same has not been true for chemical information until very recently. Only
-2004 saw a large public compound repository, PubChem, entering the public
-domain, followed more recently by other databases as discussed later in this
-article. In the same vein, open source software and the publication of
-algorithms have only recently become a main focus of the field \cite{faulon2010}.
+the same has not been true for chemical information until very recently. It was
+not until 2004 that a large public small molecule structure repository (Pubchem)
+became freely available. This was followed by other databases as discussed later
+in this article. In the same vein, open source software and the publication of
+algorithms have only recently become a main focus of the field
+\cite{faulon2010}.
 
@@
 computer science is dearly needed. (For a recent, detailed introduction to the
-field written for computer scientists see \cite{brown2009}). Cheminformatics
-comprises different areas, which broadly speaking can be divided into three
-fields: capturing data (using lab notebooks or potentially using formats such as
-the Chemical Markup Language for publications); storing data (designing database
-schemas, devising ontologies) and mining data (such as for bioactivity
-prediction, which might involve algorithms based on graphs). However, what is a
-major difference to the bioinformatics field is the type of information being
-analyzed: While bioinformatics often deals with sequences, the domain of
-cheminformatics is chemical structures. In case of the former, information can
-very frequently be represented as one-dimensional strings, which are relatively
-easy to handle computationally. On the other hand, chemical structures may
-possess rings, branches as well as multiple valid representations of even the
-same structure, potentially giving rise to ambiguities. Hence, chemical
-structure is often thought of as being more difficult to standardize than
-information from other domains. As an illustration, different representations of
-the same chemical structure are shown in Figure 1. The top left hand corner
-shows a structure a chemist would draw it -- every 'corner' representing a carbon
-atom (which can also be represented as a C), every O denoting oxygen, every N
-nitrogen, every H hydrogen and so on. This is the standard format chemists use
-when exchanging information with their peers. However, this is not directly
-amenable to data mining -- different atoms might be collapsed into one node,
-sets of atoms might have the same relevant properties assigned. Hence, the top
-right hand corner of Figure 1 shows an annotated graph representation, taking
-into account that atoms, though being of different elemental atom type, can have
-similar properties -- e.g. both oxygen and nitrogen carry an 'On' label in this
-graph.
+field written for computer scientists see \cite{brown2009}).
+
+Cheminformatics comprises different areas, which broadly speaking can be divided
+into three fields: capturing data (using lab notebooks or potentially using
+formats such as the Chemical Markup Language for publications); storing data
+(designing database schemas, devising ontologies) and mining data (such as for
+bioactivity prediction, which might involve algorithms based on graphs).
+However, what is a major difference to the bioinformatics field is the type of
+information being analyzed: While bioinformatics often deals with sequences, the
+domain of cheminformatics is chemical structures. In case of the former,
+information can very frequently be represented as one-dimensional strings, which
+are relatively easy to handle computationally. On the other hand, chemical
+structures may possess rings, branches as well as multiple valid representations
+of even the same structure, potentially giving rise to ambiguities. Hence,
+chemical structure is often thought of as being more difficult to standardize
+than information from other domains. As an illustration, different
+representations of the same chemical structure are shown in Figure 1. The top
+left hand corner shows a structure a chemist would draw it -- every 'corner'
+representing a carbon atom (which can also be represented as a C), every O
+denoting oxygen, every N nitrogen, every H hydrogen and so on. This is the
+standard format chemists use when exchanging information with their peers.
+However, this is not directly amenable to data mining -- different atoms might
+be collapsed into one node, sets of atoms might have the same relevant
+properties assigned. Hence, the top right hand corner of Figure 1 shows an
+annotated graph representation, taking into account that atoms, though being of
+different elemental atom type, can have similar properties -- e.g. both oxygen
+and nitrogen carry an 'On' label in this graph.
 
@@
 \section{Bridging Cheminformatics \& Computer Science}
-\label{sec:bridg-chem-}
+
+The motivation for this article is to highlight a variety of topics in
+cheminformatics and present the computational and algorithmic problems that are
+addressed i the field. The hope is that we will be able to highlight areas where modern CS
+research could contribute to these problems.
+
+While cheminformatics has a broad range of applicability, ranging from
+agrochemical research and drug discovery to the design of novel materials, we
+have chosen to use drug discovery as the context of this article. The remainder
+of this article is structured around the concept of ``risk minimization'' in
+drug discovery -- how does one minimize the chances of a small molecule failing
+as a drug candidate and how do cheminformatics methods play a key role in this
+efforts.
 
 \subsection{Databases \& Ontologies}
-\label{sec:datab--ontol}
 
@@
 
+See e.g. role for the EBI and genome campus,
+https://registration.hinxton.wellcome.ac.uk/display_info.asp?id=246 See also
+impact of open chemical data, ChEMBL to future, e.g. ELIXIR
+http://www.elixir-europe.org/page.php
+
 \subsection{Open Source Chemical Toolkits}
@@
 
+\section{Conclusions}
+\label{sec:conclusions}
+
 \bibliographystyle{abbrv}  

### 2011-06-22 paper.tex
reworked the open source and open data section

@@
   (among many other accomplishments), because researchers connected biological
-  technologies to the theory of string algorithms. However, despite this great
+  concepts to the theory of string algorithms. However, despite this great
   success, few computer scientists are familiar with a related (and older!)
-  discipline -- chemoinformatics, the use of computers to discover new molecules with
-  desired behavior, such as new pharmaceuticals or industrial solvents. Until
+  discipline -- chemoinformatics, the use of computers to represent
+  small molecules structures and analyze their properties based on aspects
+  of their structures. Cheminformatics has a wide applicability
+  including the discovery of novel drug candidates to optimizing
+  certain physicochemical properties of small molecules. Until
   recently, the data and techniques of chemoinformatics have been closely guarded
@@
   now have access to freely available software packages and databases of tens of
-  millions of chemicals. As a result, chemists now confront hundreds of unsolved
+  millions of chemicals. As a result, chemists now confront a variety
+  of unsolved
   algorithmic problems that could not have been tackled a decade ago, but whose
@@
 The motivation for this article is to highlight a variety of topics in
-cheminformatics and present the computational and algorithmic problems that are
-addressed i the field. The hope is that we will be able to highlight areas where modern CS
-research could contribute to these problems.
+cheminformatics and present the computational and algorithmic problems
+that are addressed in the field. The hope is that we will be able to
+highlight areas where modern CS research could contribute to these
+problems.
 
 While cheminformatics has a broad range of applicability, ranging from
-agrochemical research and drug discovery to the design of novel materials, we
-have chosen to use drug discovery as the context of this article. The remainder
-of this article is structured around the concept of ``risk minimization'' in
-drug discovery -- how does one minimize the chances of a small molecule failing
-as a drug candidate and how do cheminformatics methods play a key role in this
-efforts.
+agrochemical research and drug discovery to the design of novel
+materials, we have chosen to use drug discovery as the context of this
+article. The remainder of this article is structured around the
+concept of ``risk minimization'' in drug discovery -- how does one
+minimize the chances of a small molecule failing as a drug candidate
+and how do cheminformatics methods play a key role in this efforts.
 
@@
 
-See e.g. role for the EBI and genome campus,
-https://registration.hinxton.wellcome.ac.uk/display_info.asp?id=246 See also
-impact of open chemical data, ChEMBL to future, e.g. ELIXIR
-http://www.elixir-europe.org/page.php
+A key aspect of cheminformatics is that much of its development
+requries access to collections of molecular structures (and usually,
+associated bioactivity data). As has been noted, it is only recently
+that large structure and bioactivity data collections have become
+available. As a result, this enables much more robust validation of
+cheminformatics methodologies as well as large scale benchmarking of
+algorithms. The latter is especially relevant for the plethora of
+data mining techniques that are emoloyed in cheminformatics. At this
+point, the nature of Open Data focuses primarily on structure and
+actvity data types. On the other hand, there is a distinct lack of
+textual data (such as journal articles) that are Open in nature. While
+PubMed abstract serve as a proxy for journal articles, text mining
+methods in cheminformatics are hindered by this state of affairs.
+Interestingly, patent information is publicly accessible (cf. Google
+Patents) and there have been efforts to mine these for chemical
+information using Open tools.
+
+Obviously freely accessible data is important for research; but
+equally important are the tools to work with chemical data -- more
+specifically, chemical structures. Chemical toolkits are the workhorses
+of cheminformatics research and a variety of toolkits are available,
+both commericial (which, in many cases, provide free academic
+licenses) and Open Source. In this article we do not debate the
+arguments for and against Open Source; rather, we focus on Open Source
+tools since this ensures that a reader will be able to explore
+cheminformatics problems with a minimum of encumbrances.
 
@@
 
-A key success of the open source movement is that experts can team up for
-collaborating in their expert areas. Especially in the chemistry domain, with a
-strong expert system history, this is even more relevant due to its complex data
-types and relation to many outcome variables, like biological activities,
-toxicological effects, influence to diseases, or effects on the environment.
+A number of toolkits are available including the Chemistry Development
+Kit (CDK), OpenBabel, RDKit and Indigo. These toolkits are written in
+Java or C++ (though toolkits written in the latter invariably have
+bindings to a variety of languages). Such toolkits are used, for
+example, to read/write chemical file formats, manipulate chemical
+structures, measure similarity of molecules, search for substructures,
+and generate 2D diagrams. The underlying algorithms used include graph
+algorithms (e.g. maximal common substructure, canonical labelling of
+graphs), geometrical methods (e.g. Kabsch alignment) and vector
+manipulation (e.g. converting coordinates in various systems, 3D
+structure generation). As well as being of direct use to practising
+cheminformaticians in their day-to-day work, many applications have
+been developed that rely on these toolkits to handle chemical data;
+for example, the molecular viewer Avogadro
+(http://avogadro.openmolecules.net) uses Open Babel, while the
+molecular workbench Bioclipse \cite{Bioclipse2} uses the CDK.
 
-Open Babel and the Chemistry Development Kit (CDK) are two of the most widely
-used Open Source chemical toolkits. Such toolkits are used, for example, to
-read/write chemical file formats, manipulate chemical structures, measure
-similarity of molecules, search for substructures, and generate 2D diagrams. The
-underlying algorithms used include graph algorithms (e.g. maximal common
-substructure, canonical labelling of graphs), geometrical methods (e.g. Kabsch
-alignment) and vector manipulation (e.g. converting coordinates in various
-systems, 3D structure generation). As well as being of direct use to practising
-cheminformaticians in their day-to-day work, many applications have been
-developed that rely on these toolkits to handle chemical data; for example, the
-molecular viewer Avogadro (http://avogadro.openmolecules.net) uses Open Babel,
-while the molecular workbench Bioclipse \cite{Bioclipse2} uses the CDK.
+The various toolkits have many features in common and at the same time
+have certain distinguishing features. For example, the CDK implements
+a large collection of molecular descriptors. On the other hand,
+OpenBabel has robust support for 3D structure generation and
+optimization and supports the interconversion of a huge number of
+chemical file formats.
 
-Open Babel (http://openbabel.org) is composed of a set of command-line
-applications, a GUI, and a C++ library with SWIG-generated bindings in several
-languages. It runs cross-platform and uses CMake and CTest for building and
-testing. One of its unique features is support for a large number of chemical
-file formats and its babel command-line application is widely used for
-conversion between them. The origin of Open Babel is OELib, a chemical toolkit
-released under the GPL by OpenEye Scientific Software. In 2001, OpenEye decided
-to discontinue work on OELib and to focus instead on an internal rewrite of the
-library which would not be open source (OEChem). At that point, OELib was
-renamed to Open Babel and since then it has been maintained and developed by a
-community led by Dr Geoffrey Hutchison (University of Pittsburgh).
+While algorithmic challenges are certainly faced by toolkits, software
+engineering challenges probably affect the use and development of the
+toolkits on a daily basis. Examples include verifying code coverage,
+thread safety, documentation quality and so on. Some projects such as
+the CDK employ a variety of tools to address some of these issues.
+Examples include the use of PMD to analyze code quality and JUnit to
+perform unit testing. At the level of implementation, there are still
+inefficiencies in some toolkits. For example the CDK employs a JJTree
+parser to implement the SMARTS specification, but employs a hand built
+parser to handle SMILES. The latter is essentially a complex tangle of
+\emph{if \ldots then} statements and is not easily maintainable.
 
-The Chemistry Development Kit (http://cdk.sf.net, \cite{Steinbeck2006}) project
-originated in 2000 as a common base for two existing Java applications Jmol and
-JChemPaint. The initial code bases inherited from both these project and
-CompChem. It soon outgrew this initial goal and is now a fully-featured
-cheminformatics toolkit in its own right. The CDK is a Java library and uses Ant
-and JUnit for building and testing, as well as PMD and OpenJavaDocCheck for code
-quality control. A particular strength of the toolkit is its support for a large
-number of so-called “molecular descriptors” (numerical representations of
-chemical structures used in machine learning to link chemical structure to a
-physical or biological property). The project was initially led by Dr Christoph
-Steinbeck (EMBL-EBI, UK) and Dr Egon Willighagen (Karolinska Institutet,
-Sweden), and more recently Dr Rajarshi Guha (NIH, USA) joined as third, current
-project leader.
+It is also important to note that certain challenges faced by nearly
+all cheminformatics toolkit stem from the graph representation of
+chemistry. This representation, while succinct and amenable to
+computation, is only an approximation of reality -- and edges cases
+abound. Existing areas of interest are enumeration of colored graphs,
+taking into account symmetry, and symmetry detection itself, which not
+merely derives from the chemical graph, but typically also from the 3D
+geometry of the chemical. However, more challenging is that chemical
+representation of a molecule really needs to capture non-static
+graphs, in order to take into account delocalisation and tautomeric
+phenomena molecules can practically undergo. It is important to
+realize that though chemical representations have increased in
+sophistication over the years, many chiral molecules cannot be
+satisfactorily handled by current representation systems.
 
-Both the CDK and Open Babel are very open to new contributors to the projects.
-They both have an open mailing list, without any further private lists.
-Likewise, issue tracking systems are fully open, and ongoing and future
-development is openly discussed. Contributors are encouraged to get involved
-with any of the following areas: development of new features, improvement of
-existing features, performance optimisation, fixing bugs, documentation, website
-development, reporting bugs, and helping users on the mailing list. Anyone
-interested should email the respective developer’s mailing list (either
-cdk-devel@lists.sf.net or openbabel-devel@lists.sf.net) with their interests and
-background, and a mentor will be assigned to help them join the project.
-
-Contributions as such are made to both projects on a regular basis, both from
-industries more interested in the use and application of these libraries, as
-well as from research groups in academia. And example here is the contribution
-of a ring search algorithm replacing the earlier Figueras algorithm by
-mathematicians from the TU M\"{u}nchen in Germand (\cite{Berger2004}). The former is
-maintained in the CDK, as the project also aims to provide an educational
-resource.
-
-Challenges faced by the CDK and OpenBabel libraries are abundant. Many of these
-originate from the complexity of the graph theory used in chemistry, which is an
-approximate but insufficient representation to capture chemical reality.
-Existing areas of interest are enumeration of colored graphs, taking into
-account symmetry, and symmetry detection itself, which not merely derives from
-the chemical graph, but typically also from the 3D geometry of the chemical.
-However, more challenging is that chemical representation of a molecule really
-needs to capture non-static graphs, in order to take into account delocalisation
-and tautomeric phenomena molecules can practically undergo.
+Given that there is ample scope for software engineering (algorithm
+implementation, code quality and analysis, build systems etc), both
+the CDK and OpenBabel are open to new contributions. Both are driven
+via public, active mailing lists and public issue tracking systems.
+Development questions are actively and openly discussed to arrive at a
+consensus and contributors are encouraged to get involved with any of
+the following areas: development of new features, improvement of
+existing features, performance optimisation, fixing bugs,
+documentation, website development, reporting bugs, and helping users
+on the mailing list. Anyone interested should email the respective
+developer’s mailing list (either cdk-devel@lists.sf.net or
+openbabel-devel@lists.sf.net) with their interests and background, and
+a mentor will be assigned to help them join the project.
 

### 2011-06-29 paper.tex
Added in Aarons updates

@@
 
-\title{Chemoinformatics: The Computer Science of Chemistry}
-\numberofauthors{3}
+\title{Cheminformatics: The Computer Science of Chemical Discovery}
+\numberofauthors{9}
 \author{
 \alignauthor
-Ben Trovato\titlenote{Dr.~Trovato insisted his name be first.}\\
-       \affaddr{Institute for Clarity in Documentation}\\
-       \affaddr{1932 Wallamaloo Lane}\\
-       \affaddr{Wallamaloo, New Zealand}\\
-       \email{trovato@corporation.com}
+Joerg Kurt Wegner\\
+       \affaddr{Tibotec BVBA}\\
+       \affaddr{Turnhoutseweg 30}\\
+       \affaddr{2340 Beerse Turnhout, Belgium}\\
+       \email{jwegner@its.jnj.com}
 % 2nd. author
 \alignauthor
-G.K.M. Tobin\titlenote{The secretary disavows
-any knowledge of this author's actions.}\\
-       \affaddr{Institute for Clarity in Documentation}\\
-       \affaddr{P.O. Box 1212}\\
-       \affaddr{Dublin, Ohio 43017-6221}\\
-       \email{webmaster@marysville-ohio.com}
+Aaron Sterling\\
+       \affaddr{Department of Computer Science}\\
+       \affaddr{Iowa State University}\\
+       \affaddr{Ames, Iowa, USA}\\
+       \email{sterling@iastate.edu}
 % 3rd author
@@
 
-\additionalauthors{Additional authors: John Smith (The Th{\o}rv{\"a}ld Group,
-email: {\texttt{jsmith@affiliation.org}}) and Julius P.~Kumquat
-(The Kumquat Consortium, email: {\texttt{jpkumquat@consortium.net}}).}
-\date{30 July 1999}
+\additionalauthors{Additional authors:
+Andreas Bender (University of Cambridge, email: {\texttt{bender.andreas@gmail.com}}),
+Jean-Loup Faulon (University of Evry, email: {\texttt{Jean-Loup.Faulon@issb.genopole.fr}}),
+Janna Hastings (European Bioinformatics Institute, Cambridge, UK, email: {\texttt{janna.hastings@gmail.com}}),
+Noel O'Boyle (University College, Cork, Ireland, email: {\texttt{baoilleach@gmail.com}}),
+John Overington (European Bioinformatics Institute, Cambridge, UK, email: {\texttt{jpo@ebi.ac.uk}}),
+Herman Van Vlijmen (Tibotec, Beerse, Belgium, email: {\texttt{hvvlijme@its.jnj.com}}), and
+Egon Willighagen (Karolinska Institutet, Stockholm, Sweden, email: {\texttt{baoilleach@gmail.com}})
+.}
+\date{25 June 2011}
 
@@
   success, few computer scientists are familiar with a related (and older!)
-  discipline -- chemoinformatics, the use of computers to represent
+  discipline---cheminformatics, the use of computers to represent
   small molecules structures and analyze their properties based on aspects
@@
   secrets of companies whose financial success depended on being the first to produce
-  the new "miracle molecule". Only within the last decade -- and largely because of
-  chemists volunteering their time for an Open Science "movement" -- do researchers
+  the new ``miracle molecule.'' Only within the last decade---and largely because of
+  chemists volunteering their time for an Open Science ``movement''---do researchers
   now have access to freely available software packages and databases of tens of
-  millions of chemicals. As a result, chemists now confront a variety
-  of unsolved
+  millions of chemicals. As a result, chemists now confront a variety of unsolved
   algorithmic problems that could not have been tackled a decade ago, but whose
@@
 \category{H.4}{Information Systems Applications}{Miscellaneous}
-\terms{Theory}
-\keywords{cheminformatics, graphs, chemistry}
+\terms{
+\textit{Tautomerism} - Tautomers are isomers of organic compounds that readily interconvert by a chemical reaction called tautomerization.
+\textit{Protonation} - In chemistry, protonation is the addition of a proton (H+) to an atom, molecule, or ion.
+}
+\keywords{cheminformatics, graphs, chemistry, databases, knowledge management, open source, open data}
 
@@
 Novel technologies in the life sciences churn out information at an ever
-increasing rate -- public data stores such as at the European Bioinformatics
-Institute (EBI) contain on the order of 10 PB of biological information.
+increasing rate---public data stores such as at the European Bioinformatics
+Institute (EBI) contain on the order of 10 petabytes of biological information.
 However, while biological information has been available publicly for decades,
@@
 not until 2004 that a large public small molecule structure repository (Pubchem)
-became freely available. This was followed by other databases as discussed later
+became freely available. This was followed by other databases, as discussed later
 in this article. In the same vein, open source software and the publication of
@@
 
-So why should we actually care about chemical information being made public --
-and how does this relate to the field of computer science?
+So why should we actually care about chemical information being made public---and how does this relate to the field of computer science?
 
-Both of these questions have relatively straightforward answers. We should care
-about chemical information being made public since the drugs we take are the
+We should care about chemical information being made public since the drugs we take are the
 result of intensive and costly research, and the more information is shared (a
@@
 relates to computer science becomes clear when realizing the amount of data
-available currently -- 74 million different molecules with annotations are
-currently stored in public databases such as PubChem with a large amount of
+available currently---74 million different molecules are
+currently stored in public databases (such as PubChem) with a large amount of
 annotations, and the design of data structures for chemicals, as well as
 subsequent data mining of this information, are areas where expertise in
-computer science is dearly needed. (For a recent, detailed introduction to the
-field written for computer scientists see \cite{brown2009}).
+computer science is dearly needed. (For a recent, detailed introduction to cheminformatics, written for computer scientists, see \cite{brown2009}).
 
 Cheminformatics comprises different areas, which broadly speaking can be divided
-into three fields: capturing data (using lab notebooks or potentially using
-formats such as the Chemical Markup Language for publications); storing data
-(designing database schemas, devising ontologies) and mining data (such as for
+into three fields: \emph{capturing data} (using lab notebooks or potentially using
+formats such as the Chemical Markup Language for publications); \emph{storing data}
+(designing database schemas, devising ontologies) and \emph{mining data} (such as for
 bioactivity prediction, which might involve algorithms based on graphs).
-However, what is a major difference to the bioinformatics field is the type of
+However, a major difference from the bioinformatics field is the type of
 information being analyzed: While bioinformatics often deals with sequences, the
@@
 information can very frequently be represented as one-dimensional strings, which
-are relatively easy to handle computationally. On the other hand, chemical
+are relatively easy to handle computationally. In the case of the latter, chemical
 structures may possess rings, branches as well as multiple valid representations
 of even the same structure, potentially giving rise to ambiguities. Hence,
-chemical structure is often thought of as being more difficult to standardize
-than information from other domains. As an illustration, different
+chemical structure is often thought to be more difficult to standardize.
+
+To illustrate the difficulty of standardizing chemical structures, different
 representations of the same chemical structure are shown in Figure 1. The top
-left hand corner shows a structure a chemist would draw it -- every 'corner'
+left hand corner shows a structure a chemist would draw it---every `corner'
 representing a carbon atom (which can also be represented as a C), every O
-denoting oxygen, every N nitrogen, every H hydrogen and so on. This is the
+denoting oxygen, every N nitrogen, every H hydrogen, and so on. This is the
 standard format chemists use when exchanging information with their peers.
-However, this is not directly amenable to data mining -- different atoms might
+However, this format is not directly amenable to data mining---different atoms might
 be collapsed into one node, sets of atoms might have the same relevant
@@
 annotated graph representation, taking into account that atoms, though being of
-different elemental atom type, can have similar properties -- e.g. both oxygen
-and nitrogen carry an 'On' label in this graph.
+different elemental atom type, can have similar properties---\emph{e.g.}, both oxygen
+and nitrogen carry an `On' label in this graph.
 
@@
 
-On this graph, and employing large databases of thousands of molecules, graph
-mining algorithms can be applied to discover patterns in large-scale datasets
-and a large number of successful applications of this concept have been
-published \cite{wegner2006,horst2009}. Still, this second representation is far
-from being the end of the story. Molecules are 3D entities as represented in the
+Graph mining algorithms can be applied on this type of annotated graph (employing large databases of thousands of molecules),  to discover patterns in large-scale datasets.
+Many successful applications of this concept have been
+published \cite{wegner2006,horst2009}. Still, this ``annotated graph'' representation is far
+from being the end of the story. Molecules are three-dimensional entities, as represented in the
 bottom left hand corner of the figure, and the precise 3D shape is relevant in
-many situations (while not in others), but this is information that is lost when
-only considering the graph of a molecule. To even extend this concept, in many
-cases actually the surface properties of theq molecule are what appears to be
-relevant in e.g. biological systems, as shown in the bottom right hand corner.
+many situations (while not in others).  Nevertheless, the 3D information is lost when
+one only considers the graph of a molecule. To extend the concept of three-dimensionality further, in many
+biological systems the \emph{surface properties} of the molecule are what appears to be
+relevant, as shown in the bottom right hand corner.
 Hence, while graph mining on chemicals is certainly a promising avenue to
 explore, one needs to keep in mind that not all molecular properties are encoded
-in the graph (such as rotational states, or different charge states etc.) and
-thus often other 'descriptors' need to be used to capture the essentials of a
+in the graph (for example, neither rotational states, nor different charge states are encoded) and
+thus often other `descriptors' need to be used to capture the essentials of a
 chemical system \cite{bender2004}.
@@
 complex mixture of combinatorial and continuous problems. If we oversimplify the
-nature of chemical structures and their properties we can not describe the world
-around us in sufficient details. On the other hand, if we try encoding all
-details the complexity to collect, store, and mine data increases dramatically
-(e.g. all possible 3D structures and their properties, all metabolites of a
-structure in humans in different tissues). For the field of cheminformatics to
+nature of chemical structures and their properties, we can not describe the world
+around us in sufficient detail. On the other hand, if we try to encode every
+detail (\emph{e.g.}, all possible 3D structures and their properties, all metabolites of a
+structure in humans in different tissues), the complexity to collect, store, and mine data increases dramatically. For the field of cheminformatics to
 flourish, we need closer collaboration between chemists (or, more generally,
-life scientists) and computer scientists – where the former needs to be able to
-pose the problem in a way relevant for practical applications; and where the
-latter is able to devise ways of capturing, storing, and analyzing chemical data
+life scientists) and computer scientists---where the former need to be able to
+pose their problems in a way relevant for practical applications; and where the
+latter are able to devise ways of capturing, storing, and analyzing chemical data
 by bridging the gap between risk in generalizable encodings and complexity
-(space and time). The goal is supporting a better chemical decision making by
-reducing the risks by 1. storing and integrating data in maintainable ways, 2.
-providing enough open standards and tools to allow software engineering concepts
-to help bridging silos, and 3. mining the many chemical property spaces in a
-time and space efficient way. Many aspects pertaining to those areas will be
-discussed in the following sections.
+(space and time). The goals: to support better chemical decision making by 1. storing and integrating data in maintainable ways, 2.
+providing enough open standards and tools so software engineering concepts
+allow applications and data to be used across heterogeneous platforms, and 3.
+mining the many chemical property spaces in a time- and space-efficient way.
+We will discuss all three goals in the following sections.
 
@@
 
-The motivation for this article is to highlight a variety of topics in
+Our motivation for this article is to highlight a variety of topics in
 cheminformatics and present the computational and algorithmic problems
-that are addressed in the field. The hope is that we will be able to
+that are addressed in the field. We hope to
 highlight areas where modern CS research could contribute to these
@@
 article. The remainder of this article is structured around the
-concept of ``risk minimization'' in drug discovery -- how does one
+concept of ``risk minimization'' in drug discovery---how does one
 minimize the chances of a small molecule failing as a drug candidate
-and how do cheminformatics methods play a key role in this efforts.
+and how do cheminformatics methods play a key role in these efforts.
 
@@
 ChEMBL and NCBI-PubChem, which are both freely available. Still, the integration
-of chemical databases with each other remains a challenge, not only due to the
+of chemical databases with one other remains a challenge, not only due to the
 normalization problems in the chemical and bioactivity landscape, but also due
-to the sheer data volume of them. As mentioned earlier, we have to be able to
+to sheer data volume. As mentioned earlier, we have to be able to
 search in complex data, for example multi-labelled molecular graphs, where
-labels can be different tautomeric states, different protonation states, or many
-variations of implicit chemical expert knowledge we need to capture in a
-risk-bounded machine encoded way.
+graph labels can be different. Still, we need to be able to capture
+molecular graphs in a risk-bounded machine encoded way.
 
 Chemical graphs can differ based on the manner in which the structures are
-drawn. Checking whether two graphs represent the same chemical is an expensive
-operation (the next section will discuss this in-depth), thus, a representation
+drawn. Also chemical behaviors can result in multiple encodings of ``the same'' molecule and its
+tautomeric states, different protonation states, or many
+variations of implicit chemical expert knowledge. Checking whether two graphs represent the same chemical is an expensive
+operation (the next section will discuss this in-depth); thus, a representation
 of the structure that is unique and invariant with respect to how the structure
@@
 Simplified Molecular Input Line Specification (SMILES). However, several
-different implementations became available. After years of research and various
-commercial products, is there today not a single public standard for a
-line-notation of molecular labelled graphs, which is available to a broad
-audience (academic and commercial). The very same problem does also exist for
-bioassay ontologies and the lack of options to normalize the property spaces of
-chemical compounds, e.g. biological assays or clinical data outcome
-(biobanking).
+different SMILES implementations became available, with no universal standard of interpretation. After years of research and various
+commercial products, there is today no single public standard for a
+line-notation of molecular labeled graphs, which is available to a broad
+audience (academic and commercial). %The very same problem does also exist for
+%bioassay ontologies and the lack of options to normalize the property spaces of
+%chemical compounds, e.g. biological assays or clinical data outcome
+%(biobanking).
 
-With each database using their own algorithm for molecular encoding, little
-facilitated the automated exchange of data between different databases. As more
+With each database using their own algorithm for molecular encoding, the automated exchange of data between different databases was not facilitated. As more
 and more data became available online, structure-based identification became a
@@
 
-Figure 1: The standard chemical graph illustration together with the SMILES,
-InChI and InChIKey codes for lipoic acid. Note that the SMILES is much more
-human-readable than the others.
+%Figure 1: The standard chemical graph illustration together with the SMILES,
+%InChI and InChIKey codes for lipoic acid. Note that the SMILES is much more
+%human-readable than the others.
 
 The InChI is now widely used for matching identical chemical structures, but it
-is still limited in a few scenarios. For example, it cannot differentiatebetween
+is still limited. For example, it cannot differentiate between
 certain types of stereoisomers: Figure 2 illustrates two stereoisomers for which
@@
 
-Figure 2: The generated InChI is the same for the stereoisomers cisplatin and
-transplatin.
+%Figure 2: The generated InChI is the same for the stereoisomers cisplatin and
+%transplatin.
 
@@
 limitations of the algorithms), but two other requirements for scientific
-discovery based on chemical databases are substructure searching and similarity
-searching. In substructure searching, the database is searched for a specified
+discovery based on chemical databases are \emph{substructure searching} and \emph{similarity
+searching}. In substructure searching, the database is searched for a specified
 wholly contained part of the structures in the database; in similarity
@@
 
-Graph substructure matching is a variant of graph isomorphism detection,which is
+Graph substructure matching is a variant of \emph{graph isomorphism detection},which is
 well known to be computationally intractable [3]. To execute a graph isomorphism
@@
 used to drastically filter the candidate search space. Structure-based
-fingerprints are commonly used for this purpose. Fingerprints encode
+\emph{fingerprints} are commonly used for this purpose. Fingerprints encode
 characteristic features of a given chemical structure in a boolean array, or
@@
 structure-based similarity measures. An example measure of similarity is the
-Tanimoto coefficient, calculated as the ratio $T(a,b) = \frac{c}{(a+b−c)}$,
+\emph{Tanimoto coefficient}, calculated as the ratio $T(a,b) = \frac{c}{(a+b−c)}$,
 where c is the count of bits on in the same position in both the two
@@
 classification structure is needed for chemicals and their linked biological
-information. One such answer to this need is provided in the form of ontologies.
+information. One such answer to this need is provided in the form of \emph{ontologies}.
 An ontology is a formal specification of entities and their relationships in a
@@
 automated reasoning to derive new knowledge (inferences) from the knowledge
-encoded, making the representation both compact and immensely powerful.
+encoded, making the representation both compact and powerful.
 Ontologies are already in widespread use across the biomedical domain. A
 similarity measure based on the information encoded in ontologies is termed
-‘semantic similarity, and such measures have already been applied to enhance
+\emph{semantic similarity}, and such measures have already been applied to enhance
 chemical classification in several problem areas [couto2010]. Many open
-questionsstill remain as to the best way to combine the information encoded in
-chemical structures with chemical ontologies [hastingsowled2010].
+questions still remain about the best way to combine the information encoded in
+chemical structures with chemical ontologies \cite{hastingsowled2010}.
 
@@
 from Stanford University started developing codes to explicitly enumerate
-molecules. Their project named DENDRAL was funded by NASA with the goal of
+molecules. Their project, named DENDRAL, was funded by NASA with the goal of
 developing an automated system to elucidate compounds from spectral data. While
@@
 Enumerating molecules is not only an interesting academic exercise but has
-practical utility as well. The foremost application of enumeration is structure
-elucidation of natural compounds such as metabolites, geochemicals, and
+practical utility as well. The foremost application of enumeration is \emph{structure
+elucidation} of natural compounds such as metabolites, geochemicals, and
 petrochemicals. Ideally, the wishful bench chemist collects experimental data
@@
 molecular formula, or mass spectrometry (MS) data. Another important application
-of enumeration is molecular design. Here the problem is to design compounds
+of enumeration is \emph{molecular design}. Here the problem is to design compounds
 (drugs, for instance) that optimize some physical, chemical, or biological
@@
 
-The major difficulty with enumeration is that the in silico representation of
+The major difficulty with enumeration is that the \emph{in silico} representation of
 molecules, the so-called molecular graphs, are labeled object (i.e. labeled
@@
 molecular graphs under the operation of the symmetric group. When enumerating
-molecular structures, one has thus to solve the so–called isomorphism problem
+molecular structures, one has thus to solve the so--called isomorphism problem.
 The good news is that isomorphism can be solved in polynomial time for molecular
 graphs due to the fact that molecules belong to a restricted class of graphs
-known as bounded valence graphs; however, that same restriction prevents
-applying for molecular graph the polynomial delay enumeration results obtained
+known as \emph{bounded valence graphs}; however, that same restriction prevents
+applying to a molecular graph the polynomial delay enumeration results obtained
 for general graphs (Chapter 8 in [1]). While the complexity of molecular graph
@@
 large chemical libraries like GDC-13 comprising almost a trillion chemical
-structures (http://www.dcb-server.unibe.ch/groups/reymond/gdb/home.html). Yet,
+structures\footnote{http://www.dcb-server.unibe.ch/groups/reymond/gdb/home.html}. Yet,
 the current enumeration software products do not generally produce stereoisomers
-[3] and tautomers [4], these require specific enumeration procedures, which are
-still the subject of investigations by the chemoinformatics community. GDB-13
-along with other databases such as PubChem (NCBI/NIH) or ZINC (UCSF) are
+[3] or tautomers [4]; these require specific enumeration procedures, which are
+still the subject of investigations by the cheminformatics community.
+
+GDB-13 along with other databases such as PubChem (NCBI/NIH) or ZINC (UCSF) are
 valuable resources to search for new drugs. For instance the well known
@@
 design, however, must already be present in databases, which is a limitation of
-the technique. To palliate that limitation algorithms have recently been
+the technique. To palliate that limitation, algorithms have recently been
 developed to enumerate all compounds in the chemical space matching circular [5]
@@
 enumerating molecules from structure-activity molecular descriptors, is still
-very much into development. At stake is the design of new chemicals or drugs
+very much under development. At stake is the design of new chemicals or drugs
 that have not yet been synthesized.
 
-Another area where enumeration plays a role is in the generation of chemical
-reaction networks. The problem here consists of enumerating all the possible
+Another area where enumeration plays a role is in the \emph{generation of chemical
+reaction networks}. The problem here consists of enumerating all the possible
 compounds that can be produced by applying reaction rules to a set of initial
-molecules. By reversing the reactions rules one can also find sets of starting
+molecules. By reversing the reaction rules one can also find sets of starting
 compounds necessary for producing a given target. This latter process is known
-in chemistry as retrosynthesis since the works of E.J. Corey who received the
-Nobel Price in Chemistry in 1990. Designing new drugs or chemicals,
+in chemistry as \emph{retrosynthesis}, ever since the works of E.J. Corey, who received the
+Nobel Prize in Chemistry in 1990. Designing new drugs or chemicals,
 understanding the kinetics of combustion and petroleum refining, studying the
@@
 biology to produce heterologous compounds in microorganisms, all are
-applications that require to enumerate reaction networks. As reviewed in Chapter
+applications that require the enumeration of reaction networks. As reviewed in Chapter
 11 in [1] several network enumeration techniques have been developed. However,
@@
 to simulate the dynamics of the network while it is being constructed and remove
-compounds having a low concentration. Following that idea, methods have been
+compounds of low concentration. Following that idea, methods have been
 developed based on the Gillespie Stochastic Simulation Algorithm (SSA) to
 compute on-the-fly species concentrations. Chemical reaction network enumeration
-and sampling is an active field of research particularly in the context of
+and sampling is an active field of research, particularly in the context of
 metabolism, either to study biodegradation, or to propose metabolic engineering
@@
 
-The basis of predictive modeling in cheminformatics is that chemical structure
-implies biological activity. Thus, the goal of any modeling approach is to
+The basis of predictive modeling in cheminformatics is that \emph{chemical structure
+implies biological activity}. Thus, the goal of any modeling approach is to
 capture and characterize the correlations between structural features and the
 observed biological activities. The goal is to describe cause-effect
-relationships with the lowest possible risk, and also being able to describe the
-risk of being wrong (confidence) when using such models for decision making. In
+relationships with the lowest possible risk, and also to describe the
+likelihood of error (confidence) when using such models for decision making. In
 many cases, the activity of a small molecule is due to its interaction with a
-receptor. Traditional QSAR approaches ignore receptor interactions, but focus
+receptor. Traditional QSAR approaches ignore receptor interactions, focusing
 only on small molecule features, and therefore loose valuable information. As a
-result, techniques such as proteochemometric modeling methods have been designed
+result, techniques such as proteochemometric modeling have been designed
 to take into account both ligand and receptor structures.
@@
 compounds. Either we transform molecular graphs into a numerical vector
-representation (a.k.a., descriptors or features) of chemical structures, or we
-use algorithms for comparing molecular graphs directly (molecule kernel). In the
+representation (a.k.a. \emph{descriptors} or \emph{features}) of chemical structures, or we
+use algorithms for comparing molecular graphs directly (\emph{molecule kernel}). In the
 first case thousands of descriptors (many of which are equivalent) have been
@@
 compounds, which imposes large-scale mining challenge (time and space complexity
-risk). Note that the public PubChem dataset alone contains a double-digit
-million number of compounds.
+risk). Note that the public PubChem dataset alone contains over 30 million chemical compounds and over 70 million other substances.
 
 First, we must consider the goal of a predictive model in a cheminformatics
-setting. In some cases, pure predictive ability is desired - such as in a
+setting. In some cases, pure predictive ability is desired---such as in a
 virtual screening setting. For such cases, the complexity of the modeling
@@
 a large-scale mining capacity is crucial. On the other hand, when working with
-chemists and biologist the interpretability of an model and risk reduction
-arguments are paramount, since we have to translate better predicted in-silico
+chemists and biologists, the interpretability of a model and risk reduction
+arguments are paramount, since we have to translate better-predicted \emph{in silico}
 compounds into real-world compounds. In such cases, black box methods can be
 less useful, though attempts have been made to interpret vectorial models built
-using neural networks, SVMs or random forests. Related to this, there is the
-never ending debate on the choice of descriptors, e.g. by applying feature
+using neural networks, SVMs or random forests. Related to this, there is a
+never-ending debate on the choice of descriptors, \emph{e.g.}, by applying feature
 selection.
 
-Second, the concept of model applicability - when is the prediction for a new
-object reliable (confidence)? - has grown in importance with the increasing use
+Second, the concept of model applicability---when is the prediction for a new
+object reliable (confidence)?---has grown in importance with the increasing use
 of predictive models in regulatory settings. A variety of methods have been
-developed that attempt to characterize the “domain” of a model. It does not only
-allow if models are applicable, but can also be used for deciding if additional
+developed that attempt to characterize the domain of a model. These methods not only
+determine whether models are applicable, but can also be used for deciding if additional
 biological experiments are required for reducing the risk on certain compound
@@
 
-One of the key challenges that faces predictive modeling is the fact that small
+One of the key challenges faced by predictive modeling is the fact that small
 molecules are not static and do not exist in isolation. Traditionally,
@@
 addressed in the 4D-QSAR methodology described by Hopfinger and co-workers.
-Techniques such as multiple-instance learning could also be applied to the
+Computer science techniques such as \emph{multiple-instance learning} could also be applied to the
 multi-conformer problem. More recently, simultaneous modeling of receptor and
-ligand has been addressed by the proteochemometric approach (WHO FIRST STARTED
-THIS?).
+ligand has been addressed by the proteochemometric approach as first reported by Lapinsh et al.
 
@@
 number of external/public partners. In fact, in cheminformatics and chemistry,
-scientists have already switched many years ago from paper lab-notebooks to
-online collaboration platforms, called ELN (Electronic Lab Notebook). So, why
-have chemists and companies already adopted to 'Enterprise 2.0', an social
+scientists already switched many years ago from paper lab-notebooks to
+online collaboration platforms, called ELNs (Electronic Lab Notebooks). So, why
+have chemists and companies already adopted to ``Enterprise 2.0,'' a social
 online collaboration culture? Many other disciplines are still busy with solving
@@
 
-Before the millenium (<year 2000) many chemists were still using paper
+Before the millenium (<year 2000), many chemists were still using paper
 lab-notebooks, multiple compound registration, and search tools. Compared to
 today there were many pitfalls and drawbacks with this. The overall architecture
-was too inflexible to adopt to fast changing data standards, scientists spend
+was too inflexible to adopt to fast changing data standards, scientists spent
 too much time with administrative work, chemical data quality was inconsistent,
@@
 delays, and local silos hampered an efficient global collaboration. Now,
-chemists prefer ELNs, because they can search in data of colleagues. This does
+chemists prefer ELNs, because they can run searches in the data of colleagues. This does
 not only allow to find experts faster, but allows also access to highly trusted
@@
 
-ELNs are typically much stronger accepted and used within the cheminformatics
+ELNs are typically much more accepted and used within the cheminformatics
 domain than in the bioinformatics domain, which is picking up speed lately. The
@@
 because it allowed collaboration with trusted colleagues. So, third party paper
-catalogs and academic publications they used previously proofed being
-inefficient, and even more critical, they did not take compounds of trusted
-colleagues into account. Overall the change management to ELNs, similar to
-'Enterprise 2.0', required to deliver on the promise to increase
+catalogs and academic publications proved to be
+inefficient, and even more critically, they did not take compounds of trusted
+colleagues into account. Overall the change of management to ELNs, similar to
+``Enterprise 2.0,'' required to deliver on the promise to increase
 interconnectivity, serve as silo-breaker, and to improve collaboration
-opportunities, and it did … many years ago, already.
+opportunities, and it performed these functions many years ago, already.
 
 The future still leaves many challenges ahead. Using external/public databases
-with chemical and bioactivity remains a challenge due to differences in
-ontologies, synchronization and maintenance challenges, and efficient
+with chemical and bioactivity data remains a challenge due to differences in
+ontologies, synchronization and maintenance, and efficient
 substructure and similarity search within complex data types, aka compounds. A
-collaboration with external parties, e.g. contract research, poses again
+collaboration with external parties, e.g. contract research, poses other
 problems with duplicate checking for compounds (e.g. tautomeric or salt
@@
 themselves and rely on external services themselves. Cloud services could help
-not only for providing a service infrastructure for all parties being involved,
-but also provide the required private/public access toolbox. Furthermore are
-there still many data sources not being indexed properly, since chemistry
+not only to provide a service infrastructure for all parties involved,
+but also to provide the required private/public access toolbox. Furthermore
+there are still many data sources not being indexed properly, since chemistry
 patents are often cryptic and image/text mining remains a puzzle. One could
-argue that the image/text mining in scientific journals is less challenging, in
+argue that the image/text mining in scientific journals is less challenging; in
 fact, it remains a huge challenge and many chemists do not trust automated
@@
 growth, usage, and collaboration scenarios between private and public parties.
-Only when the technical and scientific questions are solved we can ensure that
-also legal questions about contribution metrics will get enough room to grow.
+Only when the technical and scientific questions are solved can we ensure that
+legal questions about contribution metrics will get enough room to grow.
 
@@
 A key aspect of cheminformatics is that much of its development
-requries access to collections of molecular structures (and usually,
+requries access to collections of molecular structures (and, usually,
 associated bioactivity data). As has been noted, it is only recently
@@
 available. As a result, this enables much more robust validation of
-cheminformatics methodologies as well as large scale benchmarking of
-algorithms. The latter is especially relevant for the plethora of
+cheminformatics methodologies as well as large scale \emph{benchmarking of
+algorithms}. The latter is especially relevant for the plethora of
 data mining techniques that are emoloyed in cheminformatics. At this
@@
 textual data (such as journal articles) that are Open in nature. While
-PubMed abstract serve as a proxy for journal articles, text mining
+PubMed abstracts serve as a proxy for journal articles, text mining
 methods in cheminformatics are hindered by this state of affairs.
@@
 Obviously freely accessible data is important for research; but
-equally important are the tools to work with chemical data -- more
+equally important are the tools to work with chemical data---more
 specifically, chemical structures. Chemical toolkits are the workhorses
@@
 arguments for and against Open Source; rather, we focus on Open Source
-tools since this ensures that a reader will be able to explore
+tools, since this ensures that a reader will be able to explore
 cheminformatics problems with a minimum of encumbrances.
@@
 chemistry. This representation, while succinct and amenable to
-computation, is only an approximation of reality -- and edges cases
+computation, is only an approximation of reality---and edges cases
 abound. Existing areas of interest are enumeration of colored graphs,
@@
 on the mailing list. Anyone interested should email the respective
-developer’s mailing list (either cdk-devel@lists.sf.net or
+developer's mailing list (either cdk-devel@lists.sf.net or
 openbabel-devel@lists.sf.net) with their interests and background, and
@@
 \label{sec:conclusions}
+Cheminformatics, the computer science of chemical discovery, is an industrial and academic discipline with roots dating back to the 1960s, with the creation of DENDRAL, the first expert system of any kind.
+Much of its development from the 1960s until the new millenium was performed in secret, because of competition in the pharmaceutical industry.
+While chemists were among the first scientists to embrace web-based tools of knowledge management, research into fundamental algorithmic questions of structure enumeration and database management (and a host of other topics) have suffered, because of limited academic access to either tools or data.
+As we enter the second decade of the 21st century, however, this landscape is rapidly shifting.
+There now exist free-of-charge open-source cheminformatics toolkits, and open databases with tens of millions of compounds.
+We believe the prospects for academic and industrial collaboration between chemists and computer scientists are quite bright, and we would like to encourage more computer scientists to participate in cheminformatics research.
+There is an admittedly steep learning curve, as the structural issues faced by chemists are often too detailed and subtle to admit as clean an abstraction as ``algorithms on a string,'' the way many bioinformatics can be abstracted into a CS framework.
+Nevertheless, because of the changing landscape, there are many research questions we are only now in position to try to answer: for a theorist, graph-theoretic questions of three-dimensional enumeration; for a database designer, more effective search algorithms and ontologies; for a practical programmer, expansion of the many open-source cheminformatics projects.
+If more chemists think algorithmically, and more computer scientists think ``chemically,'' we may discover ``miracle drugs'' more consistently than ever before.
+
+\section*{Acknowledgements}
+We would like to thank Danny Verbinnen for sharing his insights into ELN's.
+This article was written while A.
+Sterling was visiting the Department of Electrical Engineering and Computer Science, Northwestern University, supported in part by NSF grant CCF-1049899.
 

### 2011-07-05 paper.tex
Added in Joergs edits

@@
 \terms{
-\textit{Tautomerism} - Tautomers are isomers of organic compounds that readily interconvert by a chemical reaction called tautomerization.
-\textit{Protonation} - In chemistry, protonation is the addition of a proton (H+) to an atom, molecule, or ion.
+\begin{itemize}
+\item \textit{Tautomerism} - Tautomers are isomers (same molecular formula but different structural formula) of organic compounds that readily interconvert by a chemical reaction called tautomerization.
+\item \textit{Protonation and protonation state} - In chemistry, protonation is the addition of a proton (H+) to an atom, molecule, or ion.
+The different states (different number of H+ or on different atoms) are called protonation states for a molecular compound (e.g.
+de-protonated, mono-protonated, bi-protonated, or any variations of those states).
+\item \textit{BioAssay} - A system measuring the biological activity of a chemical compound on a particular biological assay (with all its parameters).
+\item \textit{Metabolites} - Metabolites are the intermediates and products of metabolism. The term metabolite is usually restricted to small molecules (chemical compounds).
+\item \textit{Mass spectrometry} - Mass spectrometry (MS) is an analytical technique that measures the mass-to-charge ratio of charged particles.
+It is used for determining masses of particles, for determining the elemental composition of a sample or molecule, and for elucidating the chemical structures of molecules, such as peptides and other chemical compounds.
+\item \textit{Ligand (coordination chemistry)} - an atom, ion, or functional group that donates one or more of its electrons through a coordinate covalent bond to one or more central atoms or ions
+\item \textit{Ligand (biochemistry)} - a substance that binds to a protein
+\item \textit{Receptor (biochemistry)} - in biochemistry, a protein molecule that receives and responds to a neurotransmitter, or other substance
+\item \textit{Proteochemometric modeling} - combining data mining for chemical compounds (ligands) and proteins within the same machine learning framework 
+\item \textit{BioDegradation} - is the chemical dissolution of materials by bacteria or other biological means.
+\item \textit{Reaction catalyses} - Catalysis is the change in rate of a chemical reaction due to the participation of a substance called a catalyst.
+\item \textit{Conformation and conformational isomerism} - In chemistry, conformational isomerism is a form of stereoisomerism in which the isomers can be interconverted exclusively by rotations about formally single bonds.
+Single possible states of compounds are called conformers.
+\item \textit{SMARTS} - SMiles ARbitrary Target Specification (SMARTS) is a language for specifying substructural patterns in molecules
+\end{itemize}
 }
@@
 practical utility as well. The foremost application of enumeration is \emph{structure
-elucidation} of natural compounds such as metabolites, geochemicals, and
-petrochemicals. Ideally, the wishful bench chemist collects experimental data
+elucidation} of natural compounds such as metabolites, for naming just one example. 
+Ideally, the wishful bench chemist collects experimental data
 from an unknown compound, the data is fed into a code, and the resulting
@@
 similarity. Enumerating molecules from fingerprints, or more generally
-enumerating molecules from structure-activity molecular descriptors, is still
+enumerating molecules from molecular descriptors, is still
 very much under development. At stake is the design of new chemicals or drugs
@@
 with metabolic network design is that in addition to network generation based on
-reactions, one also needs to verify that there are enzymatic sequences enabling
+reactions, one also needs to verify that there are enzymatic events possible for enabling
 reactions catalyses. That additional task requires encompassing both chemical
@@
 \label{sec:conclusions}
-Cheminformatics, the computer science of chemical discovery, is an industrial and academic discipline with roots dating back to the 1960s, with the creation of DENDRAL, the first expert system of any kind.
-Much of its development from the 1960s until the new millenium was performed in secret, because of competition in the pharmaceutical industry.
-While chemists were among the first scientists to embrace web-based tools of knowledge management, research into fundamental algorithmic questions of structure enumeration and database management (and a host of other topics) have suffered, because of limited academic access to either tools or data.
+Cheminformatics, the computer science of chemical discovery, is an industrial 
+and academic discipline with roots dating back to the 1960s, with the creation of 
+DENDRAL, the first expert system of any kind.  Much of its development from the 1960s 
+until the new millenium was performed in secret, because of competition in the pharmaceutical industry.  
+While chemists were among the first scientists to embrace web-based tools of knowledge management, 
+research into fundamental algorithmic questions of structure enumeration and database management 
+(and a host of other topics) have suffered, because of limited academic access to either tools or data.  
 As we enter the second decade of the 21st century, however, this landscape is rapidly shifting.  
-There now exist free-of-charge open-source cheminformatics toolkits, and open databases with tens of millions of compounds.
-We believe the prospects for academic and industrial collaboration between chemists and computer scientists are quite bright, and we would like to encourage more computer scientists to participate in cheminformatics research.
-There is an admittedly steep learning curve, as the structural issues faced by chemists are often too detailed and subtle to admit as clean an abstraction as ``algorithms on a string,'' the way many bioinformatics can be abstracted into a CS framework.
-Nevertheless, because of the changing landscape, there are many research questions we are only now in position to try to answer: for a theorist, graph-theoretic questions of three-dimensional enumeration; for a database designer, more effective search algorithms and ontologies; for a practical programmer, expansion of the many open-source cheminformatics projects.
-If more chemists think algorithmically, and more computer scientists think ``chemically,'' we may discover ``miracle drugs'' more consistently than ever before.
+There now exist free-of-charge open-source cheminformatics toolkits, and open databases with tens 
+of millions of compounds.  We believe the prospects for academic and industrial collaboration between 
+chemists and computer scientists are quite bright, and we would like to encourage more computer 
+scientists to participate in cheminformatics research.  There is an admittedly steep learning curve, 
+as the structural issues faced by chemists are often too detailed and subtle to admit as clean an abstraction as
+ ``algorithms on a string,'' the way many bioinformatics can be abstracted into a CS framework. 
+Nevertheless, because of the changing landscape, there are many research questions we are only now in 
+position to try to answer: for a theorist, graph-theoretic questions of three-dimensional enumeration; 
+for a database designer, more effective search algorithms and ontologies; for a practical programmer, 
+expansion of the many open-source cheminformatics projects.  
+If more chemists think algorithmically, and more computer scientists think ``chemically,'' 
+%we may discover ``miracle drugs'' more consistently than ever before.
+we are much better positioned taking not only ``simple'' chemicals into account (only one identifier and isomer possible per molecule), 
+but also their complex relationships, transformations, 
+and combinatorial challenges into account. This will allow us to normalize, search, and use all chemical information, 
+around us and within us.
 

### 2011-07-05 paper.tex
some reformatting of paragraphs

@@
 
-So why should we actually care about chemical information being made public---and how does this relate to the field of computer science?
+So why should we actually care about chemical information being made
+public---and how does this relate to the field of computer science?
 
-We should care about chemical information being made public since the drugs we take are the
-result of intensive and costly research, and the more information is shared (a
-process sometimes very hard to achieve in pharmaceutical companies) the more
-easy it should be to develop novel treatments drugs. How chemical information
-relates to computer science becomes clear when realizing the amount of data
-available currently---74 million different molecules are
-currently stored in public databases (such as PubChem) with a large amount of
-annotations, and the design of data structures for chemicals, as well as
-subsequent data mining of this information, are areas where expertise in
-computer science is dearly needed. (For a recent, detailed introduction to cheminformatics, written for computer scientists, see \cite{brown2009}).
+We should care about chemical information being made public since the
+drugs we take are the result of intensive and costly research, and the
+more information is shared (a process sometimes very hard to achieve
+in pharmaceutical companies) the more easy it should be to develop
+novel treatments drugs. How chemical information relates to computer
+science becomes clear when realizing the amount of data available
+currently---74 million different molecules are currently stored in
+public databases (such as PubChem) with a large amount of annotations,
+and the design of data structures for chemicals, as well as subsequent
+data mining of this information, are areas where expertise in computer
+science is dearly needed. (For a recent, detailed introduction to
+cheminformatics, written for computer scientists, see
+\cite{brown2009}).
 
@@
 
-Graph mining algorithms can be applied on this type of annotated graph (employing large databases of thousands of molecules),  to discover patterns in large-scale datasets.
-Many successful applications of this concept have been
-published \cite{wegner2006,horst2009}. Still, this ``annotated graph'' representation is far
-from being the end of the story. Molecules are three-dimensional entities, as represented in the
-bottom left hand corner of the figure, and the precise 3D shape is relevant in
-many situations (while not in others).  Nevertheless, the 3D information is lost when
-one only considers the graph of a molecule. To extend the concept of three-dimensionality further, in many
-biological systems the \emph{surface properties} of the molecule are what appears to be
-relevant, as shown in the bottom right hand corner.
-Hence, while graph mining on chemicals is certainly a promising avenue to
-explore, one needs to keep in mind that not all molecular properties are encoded
-in the graph (for example, neither rotational states, nor different charge states are encoded) and
-thus often other `descriptors' need to be used to capture the essentials of a
-chemical system \cite{bender2004}.
+Graph mining algorithms can be applied on this type of annotated graph
+(employing large databases of thousands of molecules), to discover
+patterns in large-scale datasets.  Many successful applications of
+this concept have been published \cite{wegner2006,horst2009}. Still,
+this ``annotated graph'' representation is far from being the end of
+the story. Molecules are three-dimensional entities, as represented in
+the bottom left hand corner of the figure, and the precise 3D shape is
+relevant in many situations (while not in others).  Nevertheless, the
+3D information is lost when one only considers the graph of a
+molecule. To extend the concept of three-dimensionality further, in
+many biological systems the \emph{surface properties} of the molecule
+are what appears to be relevant, as shown in the bottom right hand
+corner.  Hence, while graph mining on chemicals is certainly a
+promising avenue to explore, one needs to keep in mind that not all
+molecular properties are encoded in the graph (for example, neither
+rotational states, nor different charge states are encoded) and thus
+often other `descriptors' need to be used to capture the essentials of
+a chemical system \cite{bender2004}.
 
-From the above it becomes clear that the nature of chemical problems is often a
-complex mixture of combinatorial and continuous problems. If we oversimplify the
-nature of chemical structures and their properties, we can not describe the world
-around us in sufficient detail. On the other hand, if we try to encode every
-detail (\emph{e.g.}, all possible 3D structures and their properties, all metabolites of a
-structure in humans in different tissues), the complexity to collect, store, and mine data increases dramatically. For the field of cheminformatics to
-flourish, we need closer collaboration between chemists (or, more generally,
-life scientists) and computer scientists---where the former need to be able to
-pose their problems in a way relevant for practical applications; and where the
-latter are able to devise ways of capturing, storing, and analyzing chemical data
-by bridging the gap between risk in generalizable encodings and complexity
-(space and time). The goals: to support better chemical decision making by 1. storing and integrating data in maintainable ways, 2.
-providing enough open standards and tools so software engineering concepts
-allow applications and data to be used across heterogeneous platforms, and 3.
-mining the many chemical property spaces in a time- and space-efficient way.
-We will discuss all three goals in the following sections.
+From the above it becomes clear that the nature of chemical problems
+is often a complex mixture of combinatorial and continuous
+problems. If we oversimplify the nature of chemical structures and
+their properties, we can not describe the world around us in
+sufficient detail. On the other hand, if we try to encode every detail
+(\emph{e.g.}, all possible 3D structures and their properties, all
+metabolites of a structure in humans in different tissues), the
+complexity to collect, store, and mine data increases
+dramatically. For the field of cheminformatics to flourish, we need
+closer collaboration between chemists (or, more generally, life
+scientists) and computer scientists---where the former need to be able
+to pose their problems in a way relevant for practical applications;
+and where the latter are able to devise ways of capturing, storing,
+and analyzing chemical data by bridging the gap between risk in
+generalizable encodings and complexity (space and time). The goals: to
+support better chemical decision making by 1. storing and integrating
+data in maintainable ways, 2.  providing enough open standards and
+tools so software engineering concepts allow applications and data to
+be used across heterogeneous platforms, and 3. mining the many
+chemical property spaces in a time- and space-efficient way. We will
+discuss all three goals in the following sections.
 
@@
 
-Chemical graphs can differ based on the manner in which the structures are
-drawn. Also chemical behaviors can result in multiple encodings of ``the same'' molecule and its
-tautomeric states, different protonation states, or many
-variations of implicit chemical expert knowledge. Checking whether two graphs represent the same chemical is an expensive
-operation (the next section will discuss this in-depth); thus, a representation
-of the structure that is unique and invariant with respect to how the structure
-is drawn is needed for identification. An early effort in this direction was the
-Simplified Molecular Input Line Specification (SMILES). However, several
-different SMILES implementations became available, with no universal standard of interpretation. After years of research and various
+Chemical graphs can differ based on the manner in which the structures
+are drawn. Also chemical behaviors can result in multiple encodings of
+``the same'' molecule and its tautomeric states, different protonation
+states, or many variations of implicit chemical expert
+knowledge. Checking whether two graphs represent the same chemical is
+an expensive operation (the next section will discuss this in-depth);
+thus, a representation of the structure that is unique and invariant
+with respect to how the structure is drawn is needed for
+identification. An early effort in this direction was the Simplified
+Molecular Input Line Specification (SMILES). However, several
+different SMILES implementations became available, with no universal
+standard of interpretation. After years of research and various
 commercial products, there is today no single public standard for a
-line-notation of molecular labeled graphs, which is available to a broad
-audience (academic and commercial). %The very same problem does also exist for
+line-notation of molecular labeled graphs, which is available to a
+broad audience (academic and
+commercial). %The very same problem does also exist for
 %bioassay ontologies and the lack of options to normalize the property spaces of
@@
 
-With each database using their own algorithm for molecular encoding, the automated exchange of data between different databases was not facilitated. As more
-and more data became available online, structure-based identification became a
-pressing need, and the IUPAC International Chemical Identifier (InChI) code was
-developed. The InChI is a non-proprietary, structured textual identifier for
-chemical entities [1]. InChI identifiers are not intended to be read and
-understood by humans, but are useful for computational matching of chemical
-entities. For rapid database lookups, the InChIKey is a hashed key for the
-InChI, which has an invariant length of 14 characters. Figure 1 illustrates the
-SMILES, InChI and InChIKey for lipoic acid.
+With each database using their own algorithm for molecular encoding,
+the automated exchange of data between different databases was not
+facilitated. As more and more data became available online,
+structure-based identification became a pressing need, and the IUPAC
+International Chemical Identifier (InChI) code was developed. The
+InChI is a non-proprietary, structured textual identifier for chemical
+entities [1]. InChI identifiers are not intended to be read and
+understood by humans, but are useful for computational matching of
+chemical entities. For rapid database lookups, the InChIKey is a
+hashed key for the InChI, which has an invariant length of 14
+characters. Figure 1 illustrates the SMILES, InChI and InChIKey for
+lipoic acid.
 
@@
 
-The underlying computational complexity and potential intractability of
-molecular enumeration may not be a bottleneck after all, as computational
-chemists have developed and successfully used enumeration tools to generate
-large chemical libraries like GDC-13 comprising almost a trillion chemical
+The underlying computational complexity and potential intractability
+of molecular enumeration may not be a bottleneck after all, as
+computational chemists have developed and successfully used
+enumeration tools to generate large chemical libraries like GDC-13
+comprising almost a trillion chemical
 structures\footnote{http://www.dcb-server.unibe.ch/groups/reymond/gdb/home.html}. Yet,
-the current enumeration software products do not generally produce stereoisomers
-[3] or tautomers [4]; these require specific enumeration procedures, which are
-still the subject of investigations by the cheminformatics community.
+the current enumeration software products do not generally produce
+stereoisomers [3] or tautomers [4]; these require specific enumeration
+procedures, which are still the subject of investigations by the
+cheminformatics community.
 
-GDB-13 along with other databases such as PubChem (NCBI/NIH) or ZINC (UCSF) are
-valuable resources to search for new drugs. For instance the well known
-ligand-based design approach is making use of these databases along with
-structure similarity tools to discover new compounds binding to biological
-targets (protein or nucleic acid). Compounds retrieved through ligand-based
-design, however, must already be present in databases, which is a limitation of
-the technique. To palliate that limitation, algorithms have recently been
-developed to enumerate all compounds in the chemical space matching circular [5]
-or path fingerprints [6], these fingerprints being popular tools to measure
-similarity. Enumerating molecules from fingerprints, or more generally
-enumerating molecules from molecular descriptors, is still
-very much under development. At stake is the design of new chemicals or drugs
-that have not yet been synthesized.
+GDB-13 along with other databases such as PubChem (NCBI/NIH) or ZINC
+(UCSF) are valuable resources to search for new drugs. For instance
+the well known ligand-based design approach is making use of these
+databases along with structure similarity tools to discover new
+compounds binding to biological targets (protein or nucleic
+acid). Compounds retrieved through ligand-based design, however, must
+already be present in databases, which is a limitation of the
+technique. To palliate that limitation, algorithms have recently been
+developed to enumerate all compounds in the chemical space matching
+circular [5] or path fingerprints [6], these fingerprints being
+popular tools to measure similarity. Enumerating molecules from
+fingerprints, or more generally enumerating molecules from molecular
+descriptors, is still very much under development. At stake is the
+design of new chemicals or drugs that have not yet been synthesized.
 
-Another area where enumeration plays a role is in the \emph{generation of chemical
-reaction networks}. The problem here consists of enumerating all the possible
-compounds that can be produced by applying reaction rules to a set of initial
-molecules. By reversing the reaction rules one can also find sets of starting
-compounds necessary for producing a given target. This latter process is known
-in chemistry as \emph{retrosynthesis}, ever since the works of E.J. Corey, who received the
-Nobel Prize in Chemistry in 1990. Designing new drugs or chemicals,
-understanding the kinetics of combustion and petroleum refining, studying the
-dynamics of metabolic networks, applying metabolic engineering and synthetic
-biology to produce heterologous compounds in microorganisms, all are
-applications that require the enumeration of reaction networks. As reviewed in Chapter
-11 in [1] several network enumeration techniques have been developed. However,
-these techniques generally suffer from a combinatorial explosion of intermediate
-compounds being produced. One way to limit the number of compounds generated is
-to simulate the dynamics of the network while it is being constructed and remove
+Another area where enumeration plays a role is in the \emph{generation
+  of chemical reaction networks}. The problem here consists of
+enumerating all the possible compounds that can be produced by
+applying reaction rules to a set of initial molecules. By reversing
+the reaction rules one can also find sets of starting compounds
+necessary for producing a given target. This latter process is known
+in chemistry as \emph{retrosynthesis}, ever since the works of
+E.J. Corey, who received the Nobel Prize in Chemistry in
+1990. Designing new drugs or chemicals, understanding the kinetics of
+combustion and petroleum refining, studying the dynamics of metabolic
+networks, applying metabolic engineering and synthetic biology to
+produce heterologous compounds in microorganisms, all are applications
+that require the enumeration of reaction networks. As reviewed in
+Chapter 11 in [1] several network enumeration techniques have been
+developed. However, these techniques generally suffer from a
+combinatorial explosion of intermediate compounds being produced. One
+way to limit the number of compounds generated is to simulate the
+dynamics of the network while it is being constructed and remove
 compounds of low concentration. Following that idea, methods have been
-developed based on the Gillespie Stochastic Simulation Algorithm (SSA) to
-compute on-the-fly species concentrations. Chemical reaction network enumeration
-and sampling is an active field of research, particularly in the context of
-metabolism, either to study biodegradation, or to propose metabolic engineering
-strategies to biosynthesize compounds of commercial interest. The difficulty
-with metabolic network design is that in addition to network generation based on
-reactions, one also needs to verify that there are enzymatic events possible for enabling
-reactions catalyses. That additional task requires encompassing both chemical
-and sequence information and the development of tools that are at the interface
-between chemoinformatics and bioinformatics.
+developed based on the Gillespie Stochastic Simulation Algorithm (SSA)
+to compute on-the-fly species concentrations. Chemical reaction
+network enumeration and sampling is an active field of research,
+particularly in the context of metabolism, either to study
+biodegradation, or to propose metabolic engineering strategies to
+biosynthesize compounds of commercial interest. The difficulty with
+metabolic network design is that in addition to network generation
+based on reactions, one also needs to verify that there are enzymatic
+events possible for enabling reactions catalyses. That additional task
+requires encompassing both chemical and sequence information and the
+development of tools that are at the interface between
+chemoinformatics and bioinformatics.
 
@@
 
-The first step in the predictive modeling of biological activities is to compare
-compounds. Either we transform molecular graphs into a numerical vector
-representation (a.k.a. \emph{descriptors} or \emph{features}) of chemical structures, or we
-use algorithms for comparing molecular graphs directly (\emph{molecule kernel}). In the
-first case thousands of descriptors (many of which are equivalent) have been
-described [Todeschini, Handbook of Molecular Descriptors] and a variety of tools
-are available for their calculation. Given a set of molecules, observed
-activities and a descriptor vector, the problem can be considered a traditional
-statistical modeling problem. In the second case no explicit vector
-representation is required, which prevents the molecular graph to vector
-transformation challenge (loss-less encoding risk). On the other hand imposes
-the implicit kernel encoding that we have to compare all compounds against all
-compounds, which imposes large-scale mining challenge (time and space complexity
-risk). Note that the public PubChem dataset alone contains over 30 million chemical compounds and over 70 million other substances.
+The first step in the predictive modeling of biological activities is
+to compare compounds. Either we transform molecular graphs into a
+numerical vector representation (a.k.a. \emph{descriptors} or
+\emph{features}) of chemical structures, or we use algorithms for
+comparing molecular graphs directly (\emph{molecule kernel}). In the
+first case thousands of descriptors (many of which are equivalent)
+have been described [Todeschini, Handbook of Molecular Descriptors]
+and a variety of tools are available for their calculation. Given a
+set of molecules, observed activities and a descriptor vector, the
+problem can be considered a traditional statistical modeling
+problem. In the second case no explicit vector representation is
+required, which prevents the molecular graph to vector transformation
+challenge (loss-less encoding risk). On the other hand imposes the
+implicit kernel encoding that we have to compare all compounds against
+all compounds, which imposes large-scale mining challenge (time and
+space complexity risk). Note that the public PubChem dataset alone
+contains over 30 million chemical compounds and over 70 million other
+substances.
 
-First, we must consider the goal of a predictive model in a cheminformatics
-setting. In some cases, pure predictive ability is desired---such as in a
-virtual screening setting. For such cases, the complexity of the modeling
-approach is immaterial, as it leads to accurate and generalizable results. Here
-a large-scale mining capacity is crucial. On the other hand, when working with
-chemists and biologists, the interpretability of a model and risk reduction
-arguments are paramount, since we have to translate better-predicted \emph{in silico}
-compounds into real-world compounds. In such cases, black box methods can be
-less useful, though attempts have been made to interpret vectorial models built
-using neural networks, SVMs or random forests. Related to this, there is a
-never-ending debate on the choice of descriptors, \emph{e.g.}, by applying feature
-selection.
+First, we must consider the goal of a predictive model in a
+cheminformatics setting. In some cases, pure predictive ability is
+desired---such as in a virtual screening setting. For such cases, the
+complexity of the modeling approach is immaterial, as it leads to
+accurate and generalizable results. Here a large-scale mining capacity
+is crucial. On the other hand, when working with chemists and
+biologists, the interpretability of a model and risk reduction
+arguments are paramount, since we have to translate better-predicted
+\emph{in silico} compounds into real-world compounds. In such cases,
+black box methods can be less useful, though attempts have been made
+to interpret vectorial models built using neural networks, SVMs or
+random forests. Related to this, there is a never-ending debate on the
+choice of descriptors, \emph{e.g.}, by applying feature selection.
 
-Second, the concept of model applicability---when is the prediction for a new
-object reliable (confidence)?---has grown in importance with the increasing use
-of predictive models in regulatory settings. A variety of methods have been
-developed that attempt to characterize the domain of a model. These methods not only
-determine whether models are applicable, but can also be used for deciding if additional
-biological experiments are required for reducing the risk on certain compound
-classes.
+Second, the concept of model applicability---when is the prediction
+for a new object reliable (confidence)?---has grown in importance with
+the increasing use of predictive models in regulatory settings. A
+variety of methods have been developed that attempt to characterize
+the domain of a model. These methods not only determine whether models
+are applicable, but can also be used for deciding if additional
+biological experiments are required for reducing the risk on certain
+compound classes.
 
-One of the key challenges faced by predictive modeling is the fact that small
-molecules are not static and do not exist in isolation. Traditionally,
-predictive models have focused on a single structure for a small molecule and
-have ignored the role of the receptor (in receptor-mediated scenarios). Yet,
-small molecules can exist in multiple tautomeric forms and usually in multiple
-conformations. As a result, enhancing the accuracy of predictions ideally will
-require that such aspects be taking into account and, as far as possible, take
-into account the receptor simultaneously. Multi-conformer modeling has been
-addressed in the 4D-QSAR methodology described by Hopfinger and co-workers.
-Computer science techniques such as \emph{multiple-instance learning} could also be applied to the
-multi-conformer problem. More recently, simultaneous modeling of receptor and
-ligand has been addressed by the proteochemometric approach as first reported by Lapinsh et al.
+One of the key challenges faced by predictive modeling is the fact
+that small molecules are not static and do not exist in
+isolation. Traditionally, predictive models have focused on a single
+structure for a small molecule and have ignored the role of the
+receptor (in receptor-mediated scenarios). Yet, small molecules can
+exist in multiple tautomeric forms and usually in multiple
+conformations. As a result, enhancing the accuracy of predictions
+ideally will require that such aspects be taking into account and, as
+far as possible, take into account the receptor
+simultaneously. Multi-conformer modeling has been addressed in the
+4D-QSAR methodology described by Hopfinger and co-workers.  Computer
+science techniques such as \emph{multiple-instance learning} could
+also be applied to the multi-conformer problem. More recently,
+simultaneous modeling of receptor and ligand has been addressed by the
+proteochemometric approach as first reported by Lapinsh et al.
 
@@
 \label{sec:conclusions}
-Cheminformatics, the computer science of chemical discovery, is an industrial 
-and academic discipline with roots dating back to the 1960s, with the creation of 
-DENDRAL, the first expert system of any kind.  Much of its development from the 1960s 
-until the new millenium was performed in secret, because of competition in the pharmaceutical industry.  
-While chemists were among the first scientists to embrace web-based tools of knowledge management, 
-research into fundamental algorithmic questions of structure enumeration and database management 
-(and a host of other topics) have suffered, because of limited academic access to either tools or data.  
-As we enter the second decade of the 21st century, however, this landscape is rapidly shifting.  
-There now exist free-of-charge open-source cheminformatics toolkits, and open databases with tens 
-of millions of compounds.  We believe the prospects for academic and industrial collaboration between 
-chemists and computer scientists are quite bright, and we would like to encourage more computer 
-scientists to participate in cheminformatics research.  There is an admittedly steep learning curve, 
-as the structural issues faced by chemists are often too detailed and subtle to admit as clean an abstraction as
- ``algorithms on a string,'' the way many bioinformatics can be abstracted into a CS framework. 
-Nevertheless, because of the changing landscape, there are many research questions we are only now in 
-position to try to answer: for a theorist, graph-theoretic questions of three-dimensional enumeration; 
-for a database designer, more effective search algorithms and ontologies; for a practical programmer, 
-expansion of the many open-source cheminformatics projects.  
-If more chemists think algorithmically, and more computer scientists think ``chemically,'' 
+Cheminformatics, the computer science of chemical discovery, is an
+industrial and academic discipline with roots dating back to the
+1960s, with the creation of DENDRAL, the first expert system of any
+kind.  Much of its development from the 1960s until the new millenium
+was performed in secret, because of competition in the pharmaceutical
+industry.  While chemists were among the first scientists to embrace
+web-based tools of knowledge management, research into fundamental
+algorithmic questions of structure enumeration and database management
+(and a host of other topics) have suffered, because of limited
+academic access to either tools or data.  As we enter the second
+decade of the 21st century, however, this landscape is rapidly
+shifting.  There now exist free-of-charge open-source cheminformatics
+toolkits, and open databases with tens of millions of compounds.  We
+believe the prospects for academic and industrial collaboration
+between chemists and computer scientists are quite bright, and we
+would like to encourage more computer scientists to participate in
+cheminformatics research.  There is an admittedly steep learning
+curve, as the structural issues faced by chemists are often too
+detailed and subtle to admit as clean an abstraction as ``algorithms
+on a string,'' the way many bioinformatics can be abstracted into a CS
+framework.  Nevertheless, because of the changing landscape, there are
+many research questions we are only now in position to try to answer:
+for a theorist, graph-theoretic questions of three-dimensional
+enumeration; for a database designer, more effective search algorithms
+and ontologies; for a practical programmer, expansion of the many
+open-source cheminformatics projects.  If more chemists think
+algorithmically, and more computer scientists think ``chemically,''
 %we may discover ``miracle drugs'' more consistently than ever before.
-we are much better positioned taking not only ``simple'' chemicals into account (only one identifier and isomer possible per molecule), 
-but also their complex relationships, transformations, 
-and combinatorial challenges into account. This will allow us to normalize, search, and use all chemical information, 
-around us and within us.
+we are much better positioned taking not only ``simple'' chemicals
+into account (only one identifier and isomer possible per molecule),
+but also their complex relationships, transformations, and
+combinatorial challenges into account. This will allow us to
+normalize, search, and use all chemical information, around us and
+within us.
 
 \section*{Acknowledgements}
-We would like to thank Danny Verbinnen for sharing his insights into ELN's.
-This article was written while A.
-Sterling was visiting the Department of Electrical Engineering and Computer Science, Northwestern University, supported in part by NSF grant CCF-1049899.
+We would like to thank Danny Verbinnen for sharing his insights into
+ELN's.  This article was written while A. Sterling was visiting the
+Department of Electrical Engineering and Computer Science,
+Northwestern University, supported in part by NSF grant CCF-1049899.
 

### 2011-07-05 paper.tex

@@
 Herman Van Vlijmen (Tibotec, Beerse, Belgium, email: {\texttt{hvvlijme@its.jnj.com}}), and
-Egon Willighagen (Karolinska Institutet, Stockholm, Sweden, email: {\texttt{baoilleach@gmail.com}})
+Egon Willighagen (Karolinska Institutet, Stockholm, Sweden, email: {\texttt{egon.willighagen@ki.se}})
 .}

### 2011-07-07 paper.tex
Initial commit. Minor typos and spellcheck

--- 2011-07-05 paper.tex
@@
 \affaddr{9800 Medical Center Drive}\\
-\affaddr{Rockvile, MD 20850}\\
+\affaddr{Rockville, MD 20850}\\
 \email{guhar@mail.nih.gov}
@@
 Janna Hastings (European Bioinformatics Institute, Cambridge, UK, email: {\texttt{janna.hastings@gmail.com}}),
-Noel O'Boyle (University College, Cork, Ireland, email: {\texttt{baoilleach@gmail.com}}),
+Noel O'Boyle (University College Cork, Cork, Ireland, email: {\texttt{baoilleach@gmail.com}}),
 John Overington (European Bioinformatics Institute, Cambridge, UK, email: {\texttt{jpo@ebi.ac.uk}}),
 Herman Van Vlijmen (Tibotec, Beerse, Belgium, email: {\texttt{hvvlijme@its.jnj.com}}), and
-Egon Willighagen (Karolinska Institutet, Stockholm, Sweden, email: {\texttt{baoilleach@gmail.com}})
+Egon Willighagen (Karolinska Institutet, Stockholm, Sweden, email: {\texttt{egon.willighagen@ki.se}})
 .}
@@
 \item \textit{Tautomerism} - Tautomers are isomers (same molecular formula but different structural formula) of organic compounds that readily interconvert by a chemical reaction called tautomerization.
-\item \textit{Protonation and protonation state} - In chemistry, protonation is the addition of a proton (H+) to an atom, molecule, or ion.
-The different states (different number of H+ or on different atoms) are called protonation states for a molecular compound (e.g.
+\item \textit{Protonation and protonation state} - In chemistry, protonation is the addition of a proton (H$^+$) to an atom, molecule, or ion.
+The different states (different number of H$^+$ or on different atoms) are called protonation states for a molecular compound (e.g.
 de-protonated, mono-protonated, bi-protonated, or any variations of those states).
@@
 It is used for determining masses of particles, for determining the elemental composition of a sample or molecule, and for elucidating the chemical structures of molecules, such as peptides and other chemical compounds.
-\item \textit{Ligand (coordination chemistry)} - an atom, ion, or functional group that donates one or more of its electrons through a coordinate covalent bond to one or more central atoms or ions
-\item \textit{Ligand (biochemistry)} - a substance that binds to a protein
-\item \textit{Receptor (biochemistry)} - in biochemistry, a protein molecule that receives and responds to a neurotransmitter, or other substance
-\item \textit{Proteochemometric modeling} - combining data mining for chemical compounds (ligands) and proteins within the same machine learning framework 
+\item \textit{Ligand (coordination chemistry)} - an atom, ion, or functional group that donates one or more of its electrons through a coordinate covalent bond to one or more central atoms or ions.
+\item \textit{Ligand (biochemistry)} - a substance that binds to a protein.
+\item \textit{Receptor (biochemistry)} - in biochemistry, a protein molecule that receives and responds to a neurotransmitter, or other substance.
+\item \textit{Proteochemometric modeling} - combining data mining for chemical compounds (ligands) and proteins within the same machine learning framework.
 \item \textit{BioDegradation} - is the chemical dissolution of materials by bacteria or other biological means.
@@
 Single possible states of compounds are called conformers.
-\item \textit{SMARTS} - SMiles ARbitrary Target Specification (SMARTS) is a language for specifying substructural patterns in molecules
+\item \textit{SMARTS} - SMiles ARbitrary Target Specification (SMARTS) is a language for specifying substructure patterns in molecules
 \end{itemize}
@@
 the same has not been true for chemical information until very recently. It was
-not until 2004 that a large public small molecule structure repository (Pubchem)
+not until 2004 that a large public small molecule structure repository (PubChem)
 became freely available. This was followed by other databases, as discussed later
@@
 Most cheminformatics applications rely on large databases of chemical structures
-and their properties. Organising and maintaining such databases, searching based
+and their properties. Organizing and maintaining such databases, searching based
 on features of chemical structures, and clustering similar structures together,
@@
 searching, the database is searched for structures similar to a provided search
-structure. Chemical search packages are often implemented and optimised for a
+structure. Chemical search packages are often implemented and optimized for a
 given database technology, for example, the OrChem package is an open source
@@
 \emph{fingerprints} are commonly used for this purpose. Fingerprints encode
-characteristic features of a given chemical structure in a boolean array, or
+characteristic features of a given chemical structure in a Boolean array, or
 bitmap. The fingerprint is created by an algorithm which generates bitmap
 patterns for each atom in the structure; then each atom and its nearest
-neighbours, including the bonds between them; then each group of atoms connected
+neighbors, including the bonds between them; then each group of atoms connected
 by paths of increasing lengths up to the maximum implemented for the algorithm
@@
 domain of interest. It is structured around an underlying logical representation
-such as Description Logic [baaderdl2007]. The logic-based representation allows
+such as Description Logic ~\cite{baaderdl2007}. The logic-based representation allows
 automated reasoning to derive new knowledge (inferences) from the knowledge
@@
 \emph{semantic similarity}, and such measures have already been applied to enhance
-chemical classification in several problem areas [couto2010]. Many open
+chemical classification in several problem areas \cite{couto2010}. Many open
 questions still remain about the best way to combine the information encoded in
@@
 first case thousands of descriptors (many of which are equivalent)
-have been described [Todeschini, Handbook of Molecular Descriptors]
+have been described \cite{todeschini2000},
 and a variety of tools are available for their calculation. Given a
@@
 simultaneous modeling of receptor and ligand has been addressed by the
-proteochemometric approach as first reported by Lapinsh et al.
+proteochemometric approach as first reported by Lapinsh
+\textit{et al.} \cite{lapinsh2001}.
 
@@
 approach should lead to more robust predictions for panel assays. Finally, does
-this also better reflect clinical relevant compound profiles [SIDER database]
+this also better reflect clinical relevant compound profiles \cite{kuhn2010}
 and a 'personalized medicine' concept (e.g. drug-drug interaction profiles) [The
@@
 is a highly trusted source, and public chemistry and bioactivity space has to
-improve quality and interconnectivity. Efforts like Chembl and PubChem-BioAssay
+improve quality and interconnectivity. Efforts like ChEMBL and PubChem-BioAssay
 are on the right track. Still, improving data quality and standards between
@@
 A key aspect of cheminformatics is that much of its development
-requries access to collections of molecular structures (and, usually,
+requires access to collections of molecular structures (and, usually,
 associated bioactivity data). As has been noted, it is only recently
@@
 algorithms}. The latter is especially relevant for the plethora of
-data mining techniques that are emoloyed in cheminformatics. At this
+data mining techniques that are employed in cheminformatics. At this
 point, the nature of Open Data focuses primarily on structure and
-actvity data types. On the other hand, there is a distinct lack of
+activity data types. On the other hand, there is a distinct lack of
 textual data (such as journal articles) that are Open in nature. While
@@
 of cheminformatics research and a variety of toolkits are available,
-both commericial (which, in many cases, provide free academic
+both commercial (which, in many cases, provide free academic
 licenses) and Open Source. In this article we do not debate the
@@
 A number of toolkits are available including the Chemistry Development
-Kit (CDK), OpenBabel, RDKit and Indigo. These toolkits are written in
+Kit (CDK) \cite{steinbeck2003}, OpenBabel, RDKit and Indigo. These toolkits are written in
 Java or C++ (though toolkits written in the latter invariably have
@@
 the following areas: development of new features, improvement of
-existing features, performance optimisation, fixing bugs,
+existing features, performance optimization, fixing bugs,
 documentation, website development, reporting bugs, and helping users

### 2011-07-07 paper.tex
Initial commit. Minor typos and spellcheck

@@
 \item \textit{Tautomerism} - Tautomers are isomers (same molecular formula but different structural formula) of organic compounds that readily interconvert by a chemical reaction called tautomerization.
-\item \textit{Protonation and protonation state} - In chemistry, protonation is the addition of a proton (H$^+$) to an atom, molecule, or ion.
-The different states (different number of H$^+$ or on different atoms) are called protonation states for a molecular compound (e.g.
+\item \textit{Protonation and protonation state} - In chemistry, protonation is the addition of a proton (H+) to an atom, molecule, or ion.
+The different states (different number of H+ or on different atoms) are called protonation states for a molecular compound (e.g.
 de-protonated, mono-protonated, bi-protonated, or any variations of those states).
@@
 It is used for determining masses of particles, for determining the elemental composition of a sample or molecule, and for elucidating the chemical structures of molecules, such as peptides and other chemical compounds.
-\item \textit{Ligand (coordination chemistry)} - an atom, ion, or functional group that donates one or more of its electrons through a coordinate covalent bond to one or more central atoms or ions.
-\item \textit{Ligand (biochemistry)} - a substance that binds to a protein.
-\item \textit{Receptor (biochemistry)} - in biochemistry, a protein molecule that receives and responds to a neurotransmitter, or other substance.
-\item \textit{Proteochemometric modeling} - combining data mining for chemical compounds (ligands) and proteins within the same machine learning framework.
+\item \textit{Ligand (coordination chemistry)} - an atom, ion, or functional group that donates one or more of its electrons through a coordinate covalent bond to one or more central atoms or ions
+\item \textit{Ligand (biochemistry)} - a substance that binds to a protein
+\item \textit{Receptor (biochemistry)} - in biochemistry, a protein molecule that receives and responds to a neurotransmitter, or other substance
+\item \textit{Proteochemometric modeling} - combining data mining for chemical compounds (ligands) and proteins within the same machine learning framework 
 \item \textit{BioDegradation} - is the chemical dissolution of materials by bacteria or other biological means.
@@
 domain of interest. It is structured around an underlying logical representation
-such as Description Logic ~\cite{baaderdl2007}. The logic-based representation allows
+such as Description Logic [baaderdl2007]. The logic-based representation allows
 automated reasoning to derive new knowledge (inferences) from the knowledge
@@
 \emph{semantic similarity}, and such measures have already been applied to enhance
-chemical classification in several problem areas \cite{couto2010}. Many open
+chemical classification in several problem areas [couto2010]. Many open
 questions still remain about the best way to combine the information encoded in
@@
 first case thousands of descriptors (many of which are equivalent)
-have been described \cite{todeschini2000},
+have been described [Todeschini, Handbook of Molecular Descriptors]
 and a variety of tools are available for their calculation. Given a
@@
 simultaneous modeling of receptor and ligand has been addressed by the
-proteochemometric approach as first reported by Lapinsh
-\textit{et al.} \cite{lapinsh2001}.
+proteochemometric approach as first reported by Lapinsh et al.
 
@@
 approach should lead to more robust predictions for panel assays. Finally, does
-this also better reflect clinical relevant compound profiles \cite{kuhn2010}
+this also better reflect clinical relevant compound profiles [SIDER database]
 and a 'personalized medicine' concept (e.g. drug-drug interaction profiles) [The
@@
 A number of toolkits are available including the Chemistry Development
-Kit (CDK) \cite{steinbeck2003}, OpenBabel, RDKit and Indigo. These toolkits are written in
+Kit (CDK), OpenBabel, RDKit and Indigo. These toolkits are written in
 Java or C++ (though toolkits written in the latter invariably have

### 2011-07-11 paper.tex
Edits up to section 2.3

@@
   of their structures. Cheminformatics has a wide applicability
-  including the discovery of novel drug candidates to optimizing
+  from the discovery of novel drug candidates to optimizing
   certain physicochemical properties of small molecules. Until
@@
 The different states (different number of H+ or on different atoms) are called protonation states for a molecular compound (e.g.
-de-protonated, mono-protonated, bi-protonated, or any variations of those states).
-\item \textit{BioAssay} - A system measuring the biological activity of a chemical compound on a particular biological assay (with all its parameters).
+de-protonated, mono-protonated, di-protonated, or any variations of those states).
+\item \textit{Bioassay} - A system measuring the biological activity of a chemical compound on a particular biological assay (with all its parameters).
 \item \textit{Metabolites} - Metabolites are the intermediates and products of metabolism. The term metabolite is usually restricted to small molecules (chemical compounds).
@@
 \item \textit{Proteochemometric modeling} - combining data mining for chemical compounds (ligands) and proteins within the same machine learning framework 
-\item \textit{BioDegradation} - is the chemical dissolution of materials by bacteria or other biological means.
-\item \textit{Reaction catalyses} - Catalysis is the change in rate of a chemical reaction due to the participation of a substance called a catalyst.
+\item \textit{Biodegradation} - is the chemical dissolution of materials by bacteria or other biological means.
+\item \textit{Reaction catalysis} - Catalysis is the change in rate of a chemical reaction due to the participation of a substance called a catalyst.
 \item \textit{Conformation and conformational isomerism} - In chemistry, conformational isomerism is a form of stereoisomerism in which the isomers can be interconverted exclusively by rotations about formally single bonds.
@@
 
-We should care about chemical information being made public since the
+Making chemical information public is important because the
 drugs we take are the result of intensive and costly research, and the
 more information is shared (a process sometimes very hard to achieve
-in pharmaceutical companies) the more easy it should be to develop
-novel treatments drugs. How chemical information relates to computer
+in pharmaceutical companies) the easier it should be to develop
+novel treatments. How chemical information relates to computer
 science becomes clear when realizing the amount of data available
@@
 
-Graph mining algorithms can be applied on this type of annotated graph
+Graph mining algorithms can be applied to this type of annotated graph
 (employing large databases of thousands of molecules), to discover
@@
 the bottom left hand corner of the figure, and the precise 3D shape is
-relevant in many situations (while not in others).  Nevertheless, the
+relevant in many situations (although not in others).  Nevertheless, the
 3D information is lost when one only considers the graph of a
@@
 many biological systems the \emph{surface properties} of the molecule
-are what appears to be relevant, as shown in the bottom right hand
+are what appear to be relevant, as shown in the bottom right hand
 corner.  Hence, while graph mining on chemicals is certainly a
@@
 and analyzing chemical data by bridging the gap between risk in
-generalizable encodings and complexity (space and time). The goals: to
+generalizable encodings and complexity (space and time). The goals of cheminformatics are to
 support better chemical decision making by 1. storing and integrating
@@
 Our motivation for this article is to highlight a variety of topics in
-cheminformatics and present the computational and algorithmic problems
-that are addressed in the field. We hope to
-highlight areas where modern CS research could contribute to these
-problems.
+cheminformatics where modern CS research could contribute, and
+to present the computational and algorithmic problems
+that are addressed in the field.
 
@@
 are essential utilities to enable many scientific applications. Yet, each one of
-those challenges poses its unique computer science complexities.
+those areas poses its unique computer science challenges.
 
@@
 ChEMBL and NCBI-PubChem, which are both freely available. Still, the integration
-of chemical databases with one other remains a challenge, not only due to the
+of chemical databases with each other remains a challenge, not only due to the
 normalization problems in the chemical and bioactivity landscape, but also due
@@
 identification. An early effort in this direction was the Simplified
-Molecular Input Line Specification (SMILES). However, several
-different SMILES implementations became available, with no universal
-standard of interpretation. After years of research and various
+Molecular Input Line Specification (SMILES), and specifically the
+generation of a single canonical SMILES for a particular structure. However,
+with time several
+different SMILES implementations became available with no universal
+standard of generating a canonical SMILES. After years of research and various
 commercial products, there is today no single public standard for a
@@
 structure-based similarity measures. An example measure of similarity is the
-\emph{Tanimoto coefficient}, calculated as the ratio $T(a,b) = \frac{c}{(a+b−c)}$,
+\emph{Tanimoto coefficient}, calculated as the ratio $T(a,b) = \frac{c}{(a+b-c)}$,
 where c is the count of bits on in the same position in both the two
@@
 The major difficulty with enumeration is that the \emph{in silico} representation of
-molecules, the so-called molecular graphs, are labeled object (i.e. labeled
+molecules, the so-called molecular graphs, are labeled objects (i.e. labeled
 graphs), while the atoms of a molecule are of course not uniquely labeled. The
@@
 computational chemists have developed and successfully used
-enumeration tools to generate large chemical libraries like GDC-13
+enumeration tools to generate large chemical libraries like GDB-13
 comprising almost a trillion chemical
@@
 (UCSF) are valuable resources to search for new drugs. For instance
-the well known ligand-based design approach is making use of these
+the well known ligand-based design approach uses these
 databases along with structure similarity tools to discover new
@@
 already be present in databases, which is a limitation of the
-technique. To palliate that limitation, algorithms have recently been
+technique. To overcome that limitation, algorithms have recently been
 developed to enumerate all compounds in the chemical space matching

### 2011-07-11 paper.tex
Edits up to section 2.3

@@
 \item \textit{Tautomerism} - Tautomers are isomers (same molecular formula but different structural formula) of organic compounds that readily interconvert by a chemical reaction called tautomerization.
-\item \textit{Protonation and protonation state} - In chemistry, protonation is the addition of a proton (H+) to an atom, molecule, or ion.
-The different states (different number of H+ or on different atoms) are called protonation states for a molecular compound (e.g.
+\item \textit{Protonation and protonation state} - In chemistry, protonation is the addition of a proton (H$^+$) to an atom, molecule, or ion.
+The different states (different number of H$^+$ or on different atoms) are called protonation states for a molecular compound (e.g.
 de-protonated, mono-protonated, di-protonated, or any variations of those states).
@@
 It is used for determining masses of particles, for determining the elemental composition of a sample or molecule, and for elucidating the chemical structures of molecules, such as peptides and other chemical compounds.
-\item \textit{Ligand (coordination chemistry)} - an atom, ion, or functional group that donates one or more of its electrons through a coordinate covalent bond to one or more central atoms or ions
-\item \textit{Ligand (biochemistry)} - a substance that binds to a protein
-\item \textit{Receptor (biochemistry)} - in biochemistry, a protein molecule that receives and responds to a neurotransmitter, or other substance
-\item \textit{Proteochemometric modeling} - combining data mining for chemical compounds (ligands) and proteins within the same machine learning framework 
+\item \textit{Ligand (coordination chemistry)} - an atom, ion, or functional group that donates one or more of its electrons through a coordinate covalent bond to one or more central atoms or ions.
+\item \textit{Ligand (biochemistry)} - a substance that binds to a protein.
+\item \textit{Receptor (biochemistry)} - in biochemistry, a protein molecule that receives and responds to a neurotransmitter, or other substance.
+\item \textit{Proteochemometric modeling} - combining data mining for chemical compounds (ligands) and proteins within the same machine learning framework.
 \item \textit{Biodegradation} - is the chemical dissolution of materials by bacteria or other biological means.

### 2011-07-12 paper.tex
Final edits

--- 2011-07-11 paper.tex
@@
 
-Before the millenium (<year 2000), many chemists were still using paper
+Before the year 2000, many chemists were still using paper
 lab-notebooks, multiple compound registration, and search tools. Compared to
@@
 chemists prefer ELNs, because they can run searches in the data of colleagues. This does
-not only allow to find experts faster, but allows also access to highly trusted
-data with approved quality workflows. Since the data allows also access to
+not only allow to find experts faster, but also allows access to highly trusted
+data with approved quality workflows. Since this also gives access to
 analytical and legal data directly every new ELN release is aiming at increasing
@@
 ontologies, synchronization and maintenance, and efficient
-substructure and similarity search within complex data types, aka compounds. A
+substructure and similarity search within complex data types
+(compounds). A
 collaboration with external parties, e.g. contract research, poses other
@@
 
-A number of toolkits are available including the Chemistry Development
-Kit (CDK), OpenBabel, RDKit and Indigo. These toolkits are written in
+The extent of open-source software in cheminformatics has somewhat
+lagged behind related fields such as bioinformatics, partly
+due to the existence of a lucrative pharmaceutical market for
+proprietary software. However, in recent years there has been a
+rise in the amount of open-source cheminformatics software. One
+contribution to this has been the Blue Obelisk group
+(http://blueobelisk.org) of chemists, who have worked together to
+create interoperable open-source chemistry software (among other
+activities). Now there are a number of open-source toolkits available
+including the Chemistry Development
+Kit (CDK), Open Babel, RDKit and Indigo. These toolkits are written in
 Java or C++ (though toolkits written in the latter invariably have
@@
 a large collection of molecular descriptors. On the other hand,
-OpenBabel has robust support for 3D structure generation and
+Open Babel has robust support for 3D structure generation and
 optimization and supports the interconversion of a huge number of
@@
 implementation, code quality and analysis, build systems etc), both
-the CDK and OpenBabel are open to new contributions. Both are driven
+the CDK and Open Babel are open to new contributions. Both are driven
 via public, active mailing lists and public issue tracking systems.
@@
 detailed and subtle to admit as clean an abstraction as ``algorithms
-on a string,'' the way many bioinformatics can be abstracted into a CS
+on a string,'' the way many bioinformatics algorithms can be abstracted into a CS
 framework.  Nevertheless, because of the changing landscape, there are

### 2011-07-13 paper.tex
Superscript + (it would be ironic if this IT paper did not even get superscript right)

--- 2011-07-11 paper.tex
@@
 \affaddr{9800 Medical Center Drive}\\
-\affaddr{Rockville, MD 20850}\\
+\affaddr{Rockvile, MD 20850}\\
 \email{guhar@mail.nih.gov}
@@
 Janna Hastings (European Bioinformatics Institute, Cambridge, UK, email: {\texttt{janna.hastings@gmail.com}}),
-Noel O'Boyle (University College Cork, Cork, Ireland, email: {\texttt{baoilleach@gmail.com}}),
+Noel O'Boyle (University College, Cork, Ireland, email: {\texttt{baoilleach@gmail.com}}),
 John Overington (European Bioinformatics Institute, Cambridge, UK, email: {\texttt{jpo@ebi.ac.uk}}),
 Herman Van Vlijmen (Tibotec, Beerse, Belgium, email: {\texttt{hvvlijme@its.jnj.com}}), and
-Egon Willighagen (Karolinska Institutet, Stockholm, Sweden, email: {\texttt{egon.willighagen@ki.se}})
+Egon Willighagen (Karolinska Institutet, Stockholm, Sweden, email: {\texttt{baoilleach@gmail.com}})
 .}
@@
   of their structures. Cheminformatics has a wide applicability
-  from the discovery of novel drug candidates to optimizing
+  including the discovery of novel drug candidates to optimizing
   certain physicochemical properties of small molecules. Until
@@
 \item \textit{Tautomerism} - Tautomers are isomers (same molecular formula but different structural formula) of organic compounds that readily interconvert by a chemical reaction called tautomerization.
-\item \textit{Protonation and protonation state} - In chemistry, protonation is the addition of a proton (H+) to an atom, molecule, or ion.
-The different states (different number of H+ or on different atoms) are called protonation states for a molecular compound (e.g.
-de-protonated, mono-protonated, di-protonated, or any variations of those states).
-\item \textit{Bioassay} - A system measuring the biological activity of a chemical compound on a particular biological assay (with all its parameters).
+\item \textit{Protonation and protonation state} - In chemistry, protonation is the addition of a proton (H$^+$) to an atom, molecule, or ion.
+The different states (different number of H$^+$ or on different atoms) are called protonation states for a molecular compound (e.g.
+de-protonated, mono-protonated, bi-protonated, or any variations of those states).
+\item \textit{BioAssay} - A system measuring the biological activity of a chemical compound on a particular biological assay (with all its parameters).
 \item \textit{Metabolites} - Metabolites are the intermediates and products of metabolism. The term metabolite is usually restricted to small molecules (chemical compounds).
@@
 \item \textit{Proteochemometric modeling} - combining data mining for chemical compounds (ligands) and proteins within the same machine learning framework 
-\item \textit{Biodegradation} - is the chemical dissolution of materials by bacteria or other biological means.
-\item \textit{Reaction catalysis} - Catalysis is the change in rate of a chemical reaction due to the participation of a substance called a catalyst.
+\item \textit{BioDegradation} - is the chemical dissolution of materials by bacteria or other biological means.
+\item \textit{Reaction catalyses} - Catalysis is the change in rate of a chemical reaction due to the participation of a substance called a catalyst.
 \item \textit{Conformation and conformational isomerism} - In chemistry, conformational isomerism is a form of stereoisomerism in which the isomers can be interconverted exclusively by rotations about formally single bonds.
 Single possible states of compounds are called conformers.
-\item \textit{SMARTS} - SMiles ARbitrary Target Specification (SMARTS) is a language for specifying substructure patterns in molecules
+\item \textit{SMARTS} - SMiles ARbitrary Target Specification (SMARTS) is a language for specifying substructural patterns in molecules
 \end{itemize}
@@
 the same has not been true for chemical information until very recently. It was
-not until 2004 that a large public small molecule structure repository (PubChem)
+not until 2004 that a large public small molecule structure repository (Pubchem)
 became freely available. This was followed by other databases, as discussed later
@@
 
-Making chemical information public is important because the
+We should care about chemical information being made public since the
 drugs we take are the result of intensive and costly research, and the
 more information is shared (a process sometimes very hard to achieve
-in pharmaceutical companies) the easier it should be to develop
-novel treatments. How chemical information relates to computer
+in pharmaceutical companies) the more easy it should be to develop
+novel treatments drugs. How chemical information relates to computer
 science becomes clear when realizing the amount of data available
@@
 
-Graph mining algorithms can be applied to this type of annotated graph
+Graph mining algorithms can be applied on this type of annotated graph
 (employing large databases of thousands of molecules), to discover
@@
 the bottom left hand corner of the figure, and the precise 3D shape is
-relevant in many situations (although not in others).  Nevertheless, the
+relevant in many situations (while not in others).  Nevertheless, the
 3D information is lost when one only considers the graph of a
@@
 many biological systems the \emph{surface properties} of the molecule
-are what appear to be relevant, as shown in the bottom right hand
+are what appears to be relevant, as shown in the bottom right hand
 corner.  Hence, while graph mining on chemicals is certainly a
@@
 and analyzing chemical data by bridging the gap between risk in
-generalizable encodings and complexity (space and time). The goals of cheminformatics are to
+generalizable encodings and complexity (space and time). The goals: to
 support better chemical decision making by 1. storing and integrating
@@
 Our motivation for this article is to highlight a variety of topics in
-cheminformatics where modern CS research could contribute, and
-to present the computational and algorithmic problems
-that are addressed in the field.
+cheminformatics and present the computational and algorithmic problems
+that are addressed in the field. We hope to
+highlight areas where modern CS research could contribute to these
+problems.
 
@@
 Most cheminformatics applications rely on large databases of chemical structures
-and their properties. Organizing and maintaining such databases, searching based
+and their properties. Organising and maintaining such databases, searching based
 on features of chemical structures, and clustering similar structures together,
 are essential utilities to enable many scientific applications. Yet, each one of
-those areas poses its unique computer science challenges.
+those challenges poses its unique computer science complexities.
 
@@
 ChEMBL and NCBI-PubChem, which are both freely available. Still, the integration
-of chemical databases with each other remains a challenge, not only due to the
+of chemical databases with one other remains a challenge, not only due to the
 normalization problems in the chemical and bioactivity landscape, but also due
@@
 identification. An early effort in this direction was the Simplified
-Molecular Input Line Specification (SMILES), and specifically the
-generation of a single canonical SMILES for a particular structure. However,
-with time several
-different SMILES implementations became available with no universal
-standard of generating a canonical SMILES. After years of research and various
+Molecular Input Line Specification (SMILES). However, several
+different SMILES implementations became available, with no universal
+standard of interpretation. After years of research and various
 commercial products, there is today no single public standard for a
@@
 searching, the database is searched for structures similar to a provided search
-structure. Chemical search packages are often implemented and optimized for a
+structure. Chemical search packages are often implemented and optimised for a
 given database technology, for example, the OrChem package is an open source
@@
 \emph{fingerprints} are commonly used for this purpose. Fingerprints encode
-characteristic features of a given chemical structure in a Boolean array, or
+characteristic features of a given chemical structure in a boolean array, or
 bitmap. The fingerprint is created by an algorithm which generates bitmap
 patterns for each atom in the structure; then each atom and its nearest
-neighbors, including the bonds between them; then each group of atoms connected
+neighbours, including the bonds between them; then each group of atoms connected
 by paths of increasing lengths up to the maximum implemented for the algorithm
@@
 structure-based similarity measures. An example measure of similarity is the
-\emph{Tanimoto coefficient}, calculated as the ratio $T(a,b) = \frac{c}{(a+b-c)}$,
+\emph{Tanimoto coefficient}, calculated as the ratio $T(a,b) = \frac{c}{(a+b−c)}$,
 where c is the count of bits on in the same position in both the two
@@
 The major difficulty with enumeration is that the \emph{in silico} representation of
-molecules, the so-called molecular graphs, are labeled objects (i.e. labeled
+molecules, the so-called molecular graphs, are labeled object (i.e. labeled
 graphs), while the atoms of a molecule are of course not uniquely labeled. The
@@
 computational chemists have developed and successfully used
-enumeration tools to generate large chemical libraries like GDB-13
+enumeration tools to generate large chemical libraries like GDC-13
 comprising almost a trillion chemical
@@
 (UCSF) are valuable resources to search for new drugs. For instance
-the well known ligand-based design approach uses these
+the well known ligand-based design approach is making use of these
 databases along with structure similarity tools to discover new
@@
 already be present in databases, which is a limitation of the
-technique. To overcome that limitation, algorithms have recently been
+technique. To palliate that limitation, algorithms have recently been
 developed to enumerate all compounds in the chemical space matching
@@
 is a highly trusted source, and public chemistry and bioactivity space has to
-improve quality and interconnectivity. Efforts like ChEMBL and PubChem-BioAssay
+improve quality and interconnectivity. Efforts like Chembl and PubChem-BioAssay
 are on the right track. Still, improving data quality and standards between
@@
 A key aspect of cheminformatics is that much of its development
-requires access to collections of molecular structures (and, usually,
+requries access to collections of molecular structures (and, usually,
 associated bioactivity data). As has been noted, it is only recently
@@
 algorithms}. The latter is especially relevant for the plethora of
-data mining techniques that are employed in cheminformatics. At this
+data mining techniques that are emoloyed in cheminformatics. At this
 point, the nature of Open Data focuses primarily on structure and
-activity data types. On the other hand, there is a distinct lack of
+actvity data types. On the other hand, there is a distinct lack of
 textual data (such as journal articles) that are Open in nature. While
@@
 of cheminformatics research and a variety of toolkits are available,
-both commercial (which, in many cases, provide free academic
+both commericial (which, in many cases, provide free academic
 licenses) and Open Source. In this article we do not debate the
@@
 the following areas: development of new features, improvement of
-existing features, performance optimization, fixing bugs,
+existing features, performance optimisation, fixing bugs,
 documentation, website development, reporting bugs, and helping users

### 2011-07-13 paper.tex
Removed duplicate acronym, and added missing period.

--- 2011-07-13 paper.tex
@@
 Single possible states of compounds are called conformers.
-\item \textit{SMARTS} - SMiles ARbitrary Target Specification (SMARTS) is a language for specifying substructural patterns in molecules
+\item \textit{SMARTS} - SMiles ARbitrary Target Specification is a language for specifying substructural patterns. in molecules
 \end{itemize}

### 2011-07-13 paper.tex
Work on missing references

@@
 Single possible states of compounds are called conformers.
-\item \textit{SMARTS} - SMiles ARbitrary Target Specification is a language for specifying substructural patterns. in molecules
+\item \textit{SMARTS} - SMiles ARbitrary Target Specification (SMARTS) is a language for specifying substructural patterns in molecules
 \end{itemize}

### 2011-07-19 paper.tex

@@
 Janna Hastings (European Bioinformatics Institute, Cambridge, UK, email: {\texttt{janna.hastings@gmail.com}}),
-Noel O'Boyle (University College, Cork, Ireland, email: {\texttt{baoilleach@gmail.com}}),
+Noel O'Boyle (University College Cork, Cork, Ireland, email: {\texttt{baoilleach@gmail.com}}),
 John Overington (European Bioinformatics Institute, Cambridge, UK, email: {\texttt{jpo@ebi.ac.uk}}),
@@
 The different states (different number of H$^+$ or on different atoms) are called protonation states for a molecular compound (e.g.
-de-protonated, mono-protonated, bi-protonated, or any variations of those states).
-\item \textit{BioAssay} - A system measuring the biological activity of a chemical compound on a particular biological assay (with all its parameters).
+de-protonated, mono-protonated, di-protonated, or any variations of those states).
+\item \textit{Bioassay} - A system measuring the biological activity of a chemical compound on a particular biological assay (with all its parameters).
 \item \textit{Metabolites} - Metabolites are the intermediates and products of metabolism. The term metabolite is usually restricted to small molecules (chemical compounds).
@@
 \item \textit{Proteochemometric modeling} - combining data mining for chemical compounds (ligands) and proteins within the same machine learning framework.
-\item \textit{BioDegradation} - is the chemical dissolution of materials by bacteria or other biological means.
-\item \textit{Reaction catalyses} - Catalysis is the change in rate of a chemical reaction due to the participation of a substance called a catalyst.
+\item \textit{Biodegradation} - is the chemical dissolution of materials by bacteria or other biological means.
+\item \textit{Reaction catalysis} - Catalysis is the change in rate of a chemical reaction due to the participation of a substance called a catalyst.
 \item \textit{Conformation and conformational isomerism} - In chemistry, conformational isomerism is a form of stereoisomerism in which the isomers can be interconverted exclusively by rotations about formally single bonds.
 Single possible states of compounds are called conformers.
-\item \textit{SMARTS} - SMiles ARbitrary Target Specification is a language for specifying substructural patterns. in molecules
+\item \textit{SMARTS} - SMiles ARbitrary Target Specification (SMARTS) is a language for specifying substructure patterns in molecules
 \end{itemize}
@@
 Most cheminformatics applications rely on large databases of chemical structures
-and their properties. Organising and maintaining such databases, searching based
+and their properties. Organizing and maintaining such databases, searching based
 on features of chemical structures, and clustering similar structures together,
 are essential utilities to enable many scientific applications. Yet, each one of
-those challenges poses its unique computer science complexities.
+those areas poses its unique computer science challenges.
 
@@
 
-A number of toolkits are available including the Chemistry Development
-Kit (CDK) \cite{steinbeck2003}, OpenBabel, RDKit and Indigo. These toolkits are written in
+The extent of open-source software in cheminformatics has somewhat
+lagged behind related fields such as bioinformatics, partly
+due to the existence of a lucrative pharmaceutical market for
+proprietary software. However, in recent years there has been a
+rise in the amount of open-source cheminformatics software. One
+contribution to this has been the Blue Obelisk group
+(http://blueobelisk.org) of chemists, who have worked together to
+create interoperable open-source chemistry software (among other
+activities). Now there are a number of open-source toolkits available
+including the Chemistry Development
+Kit (CDK) \cite{steinbeck2003}, Open Babel, RDKit and Indigo. These toolkits are written in
 Java or C++ (though toolkits written in the latter invariably have

### 2011-07-21 paper.tex
Edited paper.tex via GitHub

@@
 \begin{abstract}
-  One of the most prominent success stories in all of science over the last decade has
+  One of the most prominent success stories in all the sciences over the last decade has
   been the advance of bioinformatics: interdisciplinary collaboration between
-  computer scientists and molecular biologists led to the sequencing the human genome
-  (among many other accomplishments), because researchers connected biological
+  computer scientists and molecular biologists led to the sequencing of the human genome
+  and many other accomplishments, since researchers connected biological
   concepts to the theory of string algorithms. However, despite this great
@@
   discipline---cheminformatics, the use of computers to represent
-  small molecules structures and analyze their properties based on aspects
-  of their structures. Cheminformatics has a wide applicability
-  from the discovery of novel drug candidates to optimizing
+  the structures of small molecules and analyze their properties. 
+  Cheminformatics has a wide applicability,
+  from the discovery of novel drug candidates to the optimization of
   certain physicochemical properties of small molecules. Until
-  recently, the data and techniques of chemoinformatics have been closely guarded
+  recently, the data and techniques of cheminformatics have been closely guarded
   secrets of companies whose financial success depended on being the first to produce
   the new ``miracle molecule.'' Only within the last decade---and largely because of
-  chemists volunteering their time for an Open Science ``movement''---do researchers
-  now have access to freely available software packages and databases of tens of
-  millions of chemicals. As a result, chemists now confront a variety of unsolved
+  chemists volunteering their time for an Open Science ``movement''---have researchers
+  gained access to freely available software packages and databases of tens of
+  millions of chemicals. Academic chemists now confront a variety of unsolved
   algorithmic problems that could not have been tackled a decade ago, but whose
@@
 Single possible states of compounds are called conformers.
-\item \textit{SMARTS} - SMiles ARbitrary Target Specification (SMARTS) is a language for specifying substructure patterns in molecules
+\item \textit{SMARTS} - SMiles ARbitrary Target Specification \\(SMARTS) is a language for specifying substructure patterns in molecules
 \end{itemize}
@@
 
-\section{Is Cheminformatics the New Bioinformatics?}
+\section{Is Cheminformatics the New \\Bioinformatics?}
 
@@
 increasing rate---public data stores such as at the European Bioinformatics
-Institute (EBI) contain on the order of 10 petabytes of biological information.
-However, while biological information has been available publicly for decades,
+Institute (EBI) contain in the order of 10 petabytes of biological information.
+However, while biological information has been publicly available for decades,
 the same has not been true for chemical information until very recently. It was
@@
 Making chemical information public is important because the
-drugs we take are the result of intensive and costly research, and the
-more information is shared (a process sometimes very hard to achieve
-in pharmaceutical companies) the easier it should be to develop
-novel treatments. How chemical information relates to computer
-science becomes clear when realizing the amount of data available
+drugs we take are the result of intensive and costly research, 
+which is facilitated by the extent of information available. The
+more information is shared (a process very hard to achieve
+in pharmaceutical companies), the more information is available to each 
+researcher, facilitating the development of
+novel treatments. The relevance of computer
+science in this research becomes clear in considering the quantity of data available
 currently---74 million different molecules are currently stored in
-public databases (such as PubChem) with a large amount of annotations,
-and the design of data structures for chemicals, as well as subsequent
+public databases (such as PubChem), together with an even larger number of annotations.
+The design of data structures for this chemical information, as well as subsequent
 data mining of this information, are areas where expertise in computer
-science is dearly needed. (For a recent, detailed introduction to
+science is urgently needed. (For a recent, detailed introduction to
 cheminformatics, written for computer scientists, see
@@
 bioactivity prediction, which might involve algorithms based on graphs).
-However, a major difference from the bioinformatics field is the type of
+However, a major difference from bioinformatics lies in the type of
 information being analyzed: While bioinformatics often deals with sequences, the
-domain of cheminformatics is chemical structures. In case of the former,
+domain of cheminformatics is chemical structures. In the case of the former,
 information can very frequently be represented as one-dimensional strings, which
@@
 structures may possess rings, branches as well as multiple valid representations
-of even the same structure, potentially giving rise to ambiguities. Hence,
-chemical structure is often thought to be more difficult to standardize.
+of the same structure, potentially giving rise to ambiguities. Hence,
+chemical structure is often reported to be more difficult to standardize.
 
@@
 However, this format is not directly amenable to data mining---different atoms might
-be collapsed into one node, sets of atoms might have the same relevant
+be collapsed into one node, and sets of atoms might have the same relevant
 properties assigned. Hence, the top right hand corner of Figure 1 shows an
@@
 metabolites of a structure in humans in different tissues), the
-complexity to collect, store, and mine data increases
+complexity of collecting, storing, and mining these data increases
 dramatically. For the field of cheminformatics to flourish, we need
@@
 and where the latter are able to devise ways of capturing, storing,
-and analyzing chemical data by bridging the gap between risk in
+and analyzing chemical data which achieve optimal balances between the different risks in 
 generalizable encodings and complexity (space and time). The goals of cheminformatics are to
@@
 ChEMBL and NCBI-PubChem, which are both freely available. Still, the integration
-of chemical databases with each other remains a challenge, not only due to the
+of data from disparate chemical databasesremains a challenge, not only due to the
 normalization problems in the chemical and bioactivity landscape, but also due
@@
 different SMILES implementations became available with no universal
-standard of generating a canonical SMILES. After years of research and various
-commercial products, there is today no single public standard for a
-line-notation of molecular labeled graphs, which is available to a
-broad audience (academic and
-commercial). %The very same problem does also exist for
+standard of generating a canonical SMILES. 
+%After years of research and various
+%commercial products, there is today no single public standard for a
+%line-notation of molecular labeled graphs, which is available to a
+%broad audience (academic and
+%commercial). %The very same problem does also exist for
 %bioassay ontologies and the lack of options to normalize the property spaces of
@@
 With each database using their own algorithm for molecular encoding,
-the automated exchange of data between different databases was not
-facilitated. As more and more data became available online,
+automated exchange of data between different databases was hindered. 
+As more and more data became available online,
 structure-based identification became a pressing need, and the IUPAC
@@
 InChI is a non-proprietary, structured textual identifier for chemical
-entities [1]. InChI identifiers are not intended to be read and
+entities \cite{inchi}. InChI identifiers are not intended to be read and
 understood by humans, but are useful for computational matching of
@@
 hashed key for the InChI, which has an invariant length of 14
-characters. Figure 1 illustrates the SMILES, InChI and InChIKey for
+characters. Figure \ref{smiles} illustrates the SMILES, InChI and InChIKey for
 lipoic acid.
 
-%Figure 1: The standard chemical graph illustration together with the SMILES,
+%Figure {smiles}: The standard chemical graph illustration together with the SMILES,
 %InChI and InChIKey codes for lipoic acid. Note that the SMILES is much more
@@
 is still limited. For example, it cannot differentiate between
-certain types of stereoisomers: Figure 2 illustrates two stereoisomers for which
+certain types of stereoisomers: Figure \ref{cistrans} illustrates two stereoisomers for which
 the generated InChI is the same.
 
-%Figure 2: The generated InChI is the same for the stereoisomers cisplatin and
+%Figure {cistrans}: The generated InChI is the same for the stereoisomers cisplatin and
 %transplatin.
@@
 search across a full chemical database of thousands or even millions of
-structures is completely infeasible. For that reason, complex heuristics must be
+structures is infeasible. For that reason, complex heuristics must be
 used to drastically filter the candidate search space. Structure-based
@@
 \emph{Tanimoto coefficient}, calculated as the ratio $T(a,b) = \frac{c}{(a+b-c)}$,
-where c is the count of bits on in the same position in both the two
-fingerprints, a is the count of bits on in the first structure, and b is the
+where $c$ is the count of bits on in the same position in both the two
+fingerprints, $a$ is the count of bits on in the first structure, and $b$ is the
 count of bits on in the second structure. The Tanimoto coefficient varies in the
@@
 domain of interest. It is structured around an underlying logical representation
-such as Description Logic ~\cite{baaderdl2007}. The logic-based representation allows
+such as Description Logics~\cite{baaderdl2007}. The logic-based representation allows
 automated reasoning to derive new knowledge (inferences) from the knowledge
@@
 practical utility as well. The foremost application of enumeration is \emph{structure
-elucidation} of natural compounds such as metabolites, for naming just one example. 
+elucidation} of natural compounds such as metabolites, to name just one example. 
 Ideally, the wishful bench chemist collects experimental data
@@
 enumeration has taken a central role as it allows computational chemists to
-construct virtual libraries, test hypotheses, and provides guidance in the
+construct virtual libraries and test hypotheses, and it provides guidance in the
 design of optimal combinatorial experiments.
@@
 computational chemists have developed and successfully used
-enumeration tools to generate large chemical libraries like GDB-13
+enumeration tools to generate large chemical libraries such as GDB-13,
 comprising almost a trillion chemical
@@
 that require the enumeration of reaction networks. As reviewed in
-Chapter 11 in [1] several network enumeration techniques have been
+Chapter 11 in [1], several network enumeration techniques have been
 developed. However, these techniques generally suffer from a
@@
 receptor. Traditional QSAR approaches ignore receptor interactions, focusing
-only on small molecule features, and therefore loose valuable information. As a
+only on small molecule features, and therefore lose valuable information. As a
 result, techniques such as proteochemometric modeling have been designed
@@
 problem. In the second case no explicit vector representation is
-required, which prevents the molecular graph to vector transformation
-challenge (loss-less encoding risk). On the other hand imposes the
-implicit kernel encoding that we have to compare all compounds against
-all compounds, which imposes large-scale mining challenge (time and
-space complexity risk). Note that the public PubChem dataset alone
+required, which prevents the challenge of transforming the molecular 
+graph into a vector without losing information. On the other hand, the 
+implicit kernel encoding imposes the challenge that we have to compare all compounds against
+all compounds, which imposes large-scale mining challenge, which risks overrunning available
+processing time and space. Note that the public PubChem dataset alone
 contains over 30 million chemical compounds and over 70 million other
@@
 conformations. As a result, enhancing the accuracy of predictions
-ideally will require that such aspects be taking into account and, as
-far as possible, take into account the receptor
-simultaneously. Multi-conformer modeling has been addressed in the
+ideally will require that multiple structural forms be taken into 
+account simultaneously with receptor aspects. 
+Multi-conformer modeling has been addressed in the
 4D-QSAR methodology described by Hopfinger and co-workers.  Computer
@@
 in observed activities against multiple targets within a single model. Such an
-approach should lead to more robust predictions for panel assays. Finally, does
-this also better reflect clinical relevant compound profiles \cite{kuhn2010}
-and a 'personalized medicine' concept (e.g. drug-drug interaction profiles) [The
+approach should lead to more robust predictions for panel assays. Finally, 
+this also better reflects clinically relevant compound profiles \cite{kuhn2010}
+and a `personalized medicine' concept (e.g. drug-drug interaction profiles) [The
 Decision Tree].
@@
 reducing the risk in current research, but also for enabling new
-collaboration/innovation opportunities with internal/external and fast growing
+collaboration/innovation opportunities with internal partners and a rapidly growing
 number of external/public partners. In fact, in cheminformatics and chemistry,
@@
 online collaboration platforms, called ELNs (Electronic Lab Notebooks). So, why
-have chemists and companies already adopted to ``Enterprise 2.0,'' a social
+have chemists and companies already adopted ``Enterprise 2.0,'' a social
 online collaboration culture? Many other disciplines are still busy with solving
@@
 domain than in the bioinformatics domain, which is picking up speed lately. The
-question is why? In chemistry every experiment requires a lot of expert
-knowledge of chemists. Therefore, approaches for reducing risks by communicating
+question is why? In chemistry, every experiment requires a lot of expert
+knowledge. Therefore, approaches for reducing risks by communicating
 with fellow chemists about similar reactions or compounds is critical. This
 requires capturing the reaction details, compound details, and time consuming
-analysis ensuring quality data, which can be searched by others.
+analysis details which ensure quality data, such that these data can be searched by others.
 
@@
 have been used to huge bookshelves of reactions and compounds. Chemists changed
-not only to online ELN collaboration for getting more efficient, but also
-because it allowed collaboration with trusted colleagues. So, third party paper
+to online ELN collaboration not only to increase efficiency, but also
+because it allowed collaboration with trusted colleagues. Third party paper
 catalogs and academic publications proved to be
@@
 colleagues into account. Overall the change of management to ELNs, similar to
-``Enterprise 2.0,'' required to deliver on the promise to increase
+``Enterprise 2.0,'' required delivering on the promise to increase
 interconnectivity, serve as silo-breaker, and to improve collaboration
-opportunities, and it performed these functions many years ago, already.
+opportunities, and indeed it already performed these functions many years ago.
 
-The future still leaves many challenges ahead. Using external/public databases
+However, the future still leaves many challenges ahead. Using external/public databases
 with chemical and bioactivity data remains a challenge due to differences in
@@
 problems with duplicate checking for compounds (e.g. tautomeric or salt
-structures), efficient data synchronization, most often a double-registration is
+structures) and efficient data synchronization, and most often a double-registration is
 required for fulfilling the data quality processes on both sides. If external
 partners are small they often do not even have sufficient IT resources
-themselves and rely on external services themselves. Cloud services could help
+themselves and thus rely on external services. Cloud services could help
 not only to provide a service infrastructure for all parties involved,
-but also to provide the required private/public access toolbox. Furthermore
+but also to provide the required private/public access toolbox. Furthermore,
 there are still many data sources not being indexed properly, since chemistry
@@
 is a highly trusted source, and public chemistry and bioactivity space has to
-improve quality and interconnectivity. Efforts like ChEMBL and PubChem-BioAssay
+improve quality and interconnectivity to compare with it. Efforts like ChEMBL and PubChem-BioAssay
 are on the right track. Still, improving data quality and standards between
-public and closed sources will be absolutely critical for ensuring a constant
+public and closed sources will be absolutely critical for ensuring constant
 growth, usage, and collaboration scenarios between private and public parties.
@@
 It is also important to note that certain challenges faced by nearly
-all cheminformatics toolkit stem from the graph representation of
+all cheminformatics toolkits stem from the graph representation of
 chemistry. This representation, while succinct and amenable to
-computation, is only an approximation of reality---and edges cases
+computation, is only an approximation of reality---and edge cases
 abound. Existing areas of interest are enumeration of colored graphs,
@@
 representation of a molecule really needs to capture non-static
-graphs, in order to take into account delocalisation and tautomeric
-phenomena molecules can practically undergo. It is important to
-realize that though chemical representations have increased in
+graphs, in order to take into account delocalisation and the tautomeric
+phenomena molecules can undergo in practice. It is important to
+realize that although chemical representations have increased in
 sophistication over the years, many chiral molecules cannot be
@@
 on the mailing list. Anyone interested should email the respective
-developer's mailing list (either cdk-devel@lists.sf.net or
+developers' mailing list (either cdk-devel@lists.sf.net or
 openbabel-devel@lists.sf.net) with their interests and background, and
@@
 believe the prospects for academic and industrial collaboration
-between chemists and computer scientists are quite bright, and we
+between chemists and computer scientists are bright, and we
 would like to encourage more computer scientists to participate in
@@
 but also their complex relationships, transformations, and
-combinatorial challenges into account. This will allow us to
+combinatorial challenges. This will allow us to
 normalize, search, and use all chemical information, around us and

### 2011-07-23 paper.tex
Added a subsection outlining the document

--- 2011-07-19 paper.tex
@@
 not until 2004 that a large public small molecule structure repository (PubChem)
-became freely available. This was followed by other databases, as discussed later
+became freely available. This was followed by other databases, as discuss later
 in this article. In the same vein, open source software and the publication of
@@
 
-Making chemical information public is important because the
-drugs we take are the result of intensive and costly research, and the
-more information is shared (a process sometimes very hard to achieve
-in pharmaceutical companies) the easier it should be to develop
-novel treatments. How chemical information relates to computer
-science becomes clear when realizing the amount of data available
+Making chemical information public is important because the drugs we
+take are the result of intensive and costly research, and the more
+information is shared (a process sometimes very hard to achieve in
+pharmaceutical companies) the easier it should be to develop novel
+treatments. How chemical information relates to computer science
+becomes clear when realizing the amount of data available
 currently---74 million different molecules are currently stored in
-public databases (such as PubChem) with a large amount of annotations,
-and the design of data structures for chemicals, as well as subsequent
-data mining of this information, are areas where expertise in computer
+various public databases (such as PubChem) with a large amount of
+annotations, and the design of data structures for the representation
+and manipulation of chemical structure, as well as subsequent data
+mining of this information, are areas where expertise in computer
 science is dearly needed. (For a recent, detailed introduction to
@@
 the bottom left hand corner of the figure, and the precise 3D shape is
-relevant in many situations (although not in others).  Nevertheless, the
-3D information is lost when one only considers the graph of a
+relevant in many situations (although not in others).  Nevertheless,
+the 3D information is lost when one only considers the graph of a
 molecule. To extend the concept of three-dimensionality further, in
@@
 promising avenue to explore, one needs to keep in mind that not all
-molecular properties are encoded in the graph (for example, neither
-rotational states, nor different charge states are encoded) and thus
-often other `descriptors' need to be used to capture the essentials of
-a chemical system \cite{bender2004}.
+molecular properties are encoded in the graph and thus often other
+`descriptors' need to be used to capture the essentials of a chemical
+system \cite{bender2004}.
 
-From the above it becomes clear that the nature of chemical problems
-is often a complex mixture of combinatorial and continuous
-problems. If we oversimplify the nature of chemical structures and
-their properties, we can not describe the world around us in
-sufficient detail. On the other hand, if we try to encode every detail
-(\emph{e.g.}, all possible 3D structures and their properties, all
-metabolites of a structure in humans in different tissues), the
-complexity to collect, store, and mine data increases
+From the above it becomes clear that the nature of many problems in
+cheminformatics require that a complex mixture of combinatorial and
+continuous techniques.  If we oversimplify the nature of chemical
+structures and their properties, we can not describe the world around
+us in sufficient detail. On the other hand, if we try to encode every
+detail (\emph{e.g.}, all possible 3D structures and their properties,
+all metabolites of a structure in humans in different tissues), the
+complexity of collecting, storing, and mining data increases
 dramatically. For the field of cheminformatics to flourish, we need
@@
 and analyzing chemical data by bridging the gap between risk in
-generalizable encodings and complexity (space and time). The goals of cheminformatics are to
-support better chemical decision making by 1. storing and integrating
-data in maintainable ways, 2.  providing enough open standards and
-tools so software engineering concepts allow applications and data to
-be used across heterogeneous platforms, and 3. mining the many
-chemical property spaces in a time- and space-efficient way. We will
-discuss all three goals in the following sections.
+generalizable encodings and complexity (space and time). The goals of
+cheminformatics are to support better chemical decision making by
+1. storing and integrating data in maintainable ways, 2.  providing
+enough open standards and tools so software engineering concepts allow
+applications and data to be used across heterogeneous platforms, and
+3. mining the many chemical property spaces in a time- and
+space-efficient way. We will discuss all three goals in the following
+sections.
+
+\subsection{Outline}
+\label{sec:outline}
+
+The remainder of the article discusses various topics in
+cheminformatics by placing them in the context of the relevant
+subfields of computer science, aiming to highlight the challenges that
+need to be addressed and possible entry points for computer science
+research. Section \ref{sec:databases} discusses issues related to the
+storage of chemical structures in relational databases and how
+ontologies allow us to query such collections beyond the standard
+graph based subgraph matching approaches. While databases usually
+contain explicit molecules, enumeration of sctructures based on
+various constraints is key to a variety of chemical problems ranging
+from mass spectrometry to reaction planning. Section
+\ref{sec:struct-enum} discusses challenges in this area. Section
+\ref{sec:activity-mining} highlights the problem of making predictions
+of biological activity from chemical structure and tieing it all
+together, Section \ref{sec:knowledge-management} discusses the
+challenges of managing chemical structure data along with associated
+biological data in production scenarios. Finally Section
+\ref{sec:role-open-source} discusses the public sources of chemical
+information and tools available for the handling of chemical structre
+-- both of which are required for for the development of
+cheminformatics research programs.
 
@@
 \subsection{Databases \& Ontologies}
+\label{sec:databases}
 
@@
 
-InChI or other identity-mapping algorithms allow for exact searching (within the
-limitations of the algorithms), but two other requirements for scientific
-discovery based on chemical databases are \emph{substructure searching} and \emph{similarity
-searching}. In substructure searching, the database is searched for a specified
+InChI or other identity-mapping algorithms allow for exact searching
+(within the limitations of the algorithms), but two other requirements
+for scientific discovery based on chemical databases are
+\emph{substructure searching} and \emph{similarity searching}. In
+substructure searching, the database is searched for a specified
 wholly contained part of the structures in the database; in similarity
-searching, the database is searched for structures similar to a provided search
-structure. Chemical search packages are often implemented and optimized for a
-given database technology, for example, the OrChem package is an open source
-chemical search package for the Oracle database \cite{rijnbeek2009}.
+searching, the database is searched for structures similar to a
+provided search structure. Chemical search packages are often
+implemented and optimized for a given database technology, for
+example, the OrChem package is an open source chemical search package
+for the Oracle database \cite{rijnbeek2009}.
 
-Graph substructure matching is a variant of \emph{graph isomorphism detection},which is
-well known to be computationally intractable [3]. To execute a graph isomorphism
-search across a full chemical database of thousands or even millions of
-structures is completely infeasible. For that reason, complex heuristics must be
-used to drastically filter the candidate search space. Structure-based
-\emph{fingerprints} are commonly used for this purpose. Fingerprints encode
-characteristic features of a given chemical structure in a Boolean array, or
-bitmap. The fingerprint is created by an algorithm which generates bitmap
-patterns for each atom in the structure; then each atom and its nearest
-neighbors, including the bonds between them; then each group of atoms connected
-by paths of increasing lengths up to the maximum implemented for the algorithm
-(commonly around eight). The bitmap patterns are added together into the final
-fingerprint. If all bits in a query fingerprint are also present in the target
-fingerprint of a stored database structure, only then is the structure subjected
-to the computationally expensive subgraph matching algorithm. Bit operations are
-very fast and independent of the number of atoms in a structure due to the fixed
-length of the fingerprint.
+Graph substructure matching is a variant of \emph{graph isomorphism
+  detection},which is well known to be computationally intractable
+[3]. To execute a graph isomorphism search across a full chemical
+database of thousands or even millions of structures is completely
+infeasible. For that reason, complex heuristics must be used to
+drastically filter the candidate search space. Structure-based
+\emph{fingerprints} are commonly used for this purpose. Fingerprints
+encode characteristic features of a given chemical structure in a
+Boolean array, or bitmap. The fingerprint is created by an algorithm
+which generates bitmap patterns for each atom in the structure; then
+each atom and its nearest neighbors, including the bonds between them;
+then each group of atoms connected by paths of increasing lengths up
+to the maximum implemented for the algorithm (commonly around
+eight). The bitmap patterns are added together into the final
+fingerprint. If all bits in a query fingerprint are also present in
+the target fingerprint of a stored database structure, only then is
+the structure subjected to the computationally expensive subgraph
+matching algorithm. Bit operations are very fast and independent of
+the number of atoms in a structure due to the fixed length of the
+fingerprint.
 
 Fingerprints also facilitate the rapid calculation of quantitative
-structure-based similarity measures. An example measure of similarity is the
-\emph{Tanimoto coefficient}, calculated as the ratio $T(a,b) = \frac{c}{(a+b-c)}$,
-where c is the count of bits on in the same position in both the two
-fingerprints, a is the count of bits on in the first structure, and b is the
-count of bits on in the second structure. The Tanimoto coefficient varies in the
-range 0.0 -- 1.0, with a score of 1.0 indicating that the two structures are very
-similar (i.e. their fingerprints are the same). Structure-based similarity is of
-limited use in grouping of chemicals based on non-structural features, such as
-shared bioactivity profiles. For this purpose, a large-scale and flexible
-classification structure is needed for chemicals and their linked biological
-information. One such answer to this need is provided in the form of \emph{ontologies}.
-An ontology is a formal specification of entities and their relationships in a
-domain of interest. It is structured around an underlying logical representation
-such as Description Logic ~\cite{baaderdl2007}. The logic-based representation allows
-automated reasoning to derive new knowledge (inferences) from the knowledge
-encoded, making the representation both compact and powerful.
-Ontologies are already in widespread use across the biomedical domain. A
-similarity measure based on the information encoded in ontologies is termed
-\emph{semantic similarity}, and such measures have already been applied to enhance
-chemical classification in several problem areas \cite{couto2010}. Many open
-questions still remain about the best way to combine the information encoded in
-chemical structures with chemical ontologies \cite{hastingsowled2010}.
+structure-based similarity measures. An example measure of similarity
+is the \emph{Tanimoto coefficient}, calculated as the ratio $T(a,b) =
+\frac{c}{(a+b-c)}$, where c is the count of bits on in the same
+position in both the two fingerprints, a is the count of bits on in
+the first structure, and b is the count of bits on in the second
+structure. The Tanimoto coefficient varies in the range 0.0 -- 1.0,
+with a score of 1.0 indicating that the two structures are very
+similar (i.e. their fingerprints are the same). Structure-based
+similarity is of limited use in grouping of chemicals based on
+non-structural features, such as shared bioactivity profiles. For this
+purpose, a large-scale and flexible classification structure is needed
+for chemicals and their linked biological information. One such answer
+to this need is provided in the form of \emph{ontologies}.  An
+ontology is a formal specification of entities and their relationships
+in a domain of interest. It is structured around an underlying logical
+representation such as Description Logic ~\cite{baaderdl2007}. The
+logic-based representation allows automated reasoning to derive new
+knowledge (inferences) from the knowledge encoded, making the
+representation both compact and powerful.  Ontologies are already in
+widespread use across the biomedical domain. A similarity measure
+based on the information encoded in ontologies is termed
+\emph{semantic similarity}, and such measures have already been
+applied to enhance chemical classification in several problem areas
+\cite{couto2010}. Many open questions still remain about the best way
+to combine the information encoded in chemical structures with
+chemical ontologies \cite{hastingsowled2010}.
 
@@
 \subsection{Activity Mining \& Prediction}
-\label{sec:activity-mining-}
+\label{sec:activity-mining}
 
@@
 \subsection{Open Source Chemical Toolkits}
+\label{sec:toolkits}
 

### 2011-07-23 paper.tex
Some more minor edits

@@
 practical utility as well. The foremost application of enumeration is \emph{structure
-elucidation} of natural compounds such as metabolites, for naming just one example. 
+elucidation} of natural compounds such as metabolites. 
 Ideally, the wishful bench chemist collects experimental data
@@
 
-The basis of predictive modeling in cheminformatics is that \emph{chemical structure
-implies biological activity}. Thus, the goal of any modeling approach is to
-capture and characterize the correlations between structural features and the
-observed biological activities. The goal is to describe cause-effect
-relationships with the lowest possible risk, and also to describe the
-likelihood of error (confidence) when using such models for decision making. In
-many cases, the activity of a small molecule is due to its interaction with a
-receptor. Traditional QSAR approaches ignore receptor interactions, focusing
-only on small molecule features, and therefore loose valuable information. As a
-result, techniques such as proteochemometric modeling have been designed
-to take into account both ligand and receptor structures.
+The basis of predictive modeling in cheminformatics is that
+\emph{chemical structure implies biological activity}. Thus, the goal
+of any modeling approach is to capture and characterize the
+correlations between structural features and the observed biological
+activities. The goal is to describe cause-effect relationships with
+the lowest possible risk, and also to describe the likelihood of error
+(confidence) when using such models for decision making. In many
+cases, the activity of a small molecule is due to its interaction with
+a receptor. Traditional QSAR approaches ignore receptor interactions,
+focusing only on small molecule features, and therefore loose valuable
+information. As a result, techniques such as proteochemometric
+modeling have been designed to take into account both ligand and
+receptor structures.
 
@@
 
-With the advent of high-throughput screening technologies, large libraries of
-compounds can be screened against multiple targets in an efficient manner. Such
-panel assays provide a broad, systems-level view of small molecule activities.
-Models developed on such data afford us the opportunity to identify targets,
-characterize off-target effects and so on. However, most approaches to this
-problem are rather simplistic since they develop multiple individual models.
-Instead one could imagine an approach that takes into account the correlations
-in observed activities against multiple targets within a single model. Such an
-approach should lead to more robust predictions for panel assays. Finally, does
-this also better reflect clinical relevant compound profiles \cite{kuhn2010}
-and a 'personalized medicine' concept (e.g. drug-drug interaction profiles) [The
-Decision Tree].
+With the advent of high-throughput screening technologies, large
+libraries of compounds can be screened against multiple targets in an
+efficient manner. Such panel assays provide a broad, systems-level
+view of small molecule activities.  Models developed on such data
+afford us the opportunity to identify targets, characterize off-target
+effects and so on. However, most approaches to this problem are rather
+simplistic since they develop multiple individual models.  Instead one
+could imagine an approach that takes into account the correlations in
+observed activities against multiple targets within a single
+model. Such an approach should lead to more robust predictions for
+panel assays. Finally, does this also better reflect clinical relevant
+compound profiles \cite{kuhn2010} and a 'personalized medicine'
+concept (e.g. drug-drug interaction profiles).
 

### 2011-07-23 paper.tex
Updated refs for JLFs section

@@
 
-Enumerating molecules is a combinatorial problem that has fascinated chemists,
-computer scientists and mathematicians alike for more than a century. It is one
-important risk-reduction element in still being able to encode closely-related
-molecules and for keeping the space complexity of databases under (algorithmic)
-control. Indeed, while trying to solve the problem of counting the isomers of
-paraffin structures or counting substituted aromatic compounds, fundamental
-principles in graph theory and combinatorics were developed by Cayley, Polya and
-others. Even terms that are widely-used nowadays like graph and tree were
-originally coined in a chemistry context (Chapters 1 and 8 in [1]). More than
-four decades ago, Joshua Lederberg, Carl Djerassi, Edward Feigenbaum and others
-from Stanford University started developing codes to explicitly enumerate
-molecules. Their project, named DENDRAL, was funded by NASA with the goal of
-developing an automated system to elucidate compounds from spectral data. While
-studying this challenging problem, important concepts in computer science were
-developed. DENDRAL is generally quoted in artificial intelligence textbooks as a
-pioneer project that produced the first expert system.
+Enumerating molecules is a combinatorial problem that has fascinated
+chemists, computer scientists and mathematicians alike for more than a
+century. It is one important risk-reduction element in still being
+able to encode closely-related molecules and for keeping the space
+complexity of databases under (algorithmic) control. Indeed, while
+trying to solve the problem of counting the isomers of paraffin
+structures or counting substituted aromatic compounds, fundamental
+principles in graph theory and combinatorics were developed by Cayley,
+Polya and others. Even terms that are widely-used nowadays like graph
+and tree were originally coined in a chemistry context (Chapters 1 and
+8 in \cite{faulon2010}). More than four decades ago, Joshua Lederberg,
+Carl Djerassi, Edward Feigenbaum and others from Stanford University
+started developing codes to explicitly enumerate molecules. Their
+project, named DENDRAL, was funded by NASA with the goal of developing
+an automated system to elucidate compounds from spectral data. While
+studying this challenging problem, important concepts in computer
+science were developed. DENDRAL is generally quoted in artificial
+intelligence textbooks as a pioneer project that produced the first
+expert system.
 
@@
 
-The major difficulty with enumeration is that the \emph{in silico} representation of
-molecules, the so-called molecular graphs, are labeled objects (i.e. labeled
-graphs), while the atoms of a molecule are of course not uniquely labeled. The
-mathematical concept to tackle this problem is to consider orbits of labeled
-molecular graphs under the operation of the symmetric group. When enumerating
-molecular structures, one has thus to solve the so--called isomorphism problem.
-The good news is that isomorphism can be solved in polynomial time for molecular
-graphs due to the fact that molecules belong to a restricted class of graphs
-known as \emph{bounded valence graphs}; however, that same restriction prevents
-applying to a molecular graph the polynomial delay enumeration results obtained
-for general graphs (Chapter 8 in [1]). While the complexity of molecular graph
-enumeration remains an open problem it has nonetheless been shown that molecules
-can be sampled efficiently [2]. Sampling procedures based on the Metropolis or
-the Genetic Algorithms have been developed and used to elucidate natural
-compounds from NMR data.
+The major difficulty with enumeration is that the \emph{in silico}
+representation of molecules, the so-called molecular graphs, are
+labeled objects (i.e. labeled graphs), while the atoms of a molecule
+are of course not uniquely labeled. The mathematical concept to tackle
+this problem is to consider orbits of labeled molecular graphs under
+the operation of the symmetric group. When enumerating molecular
+structures, one has thus to solve the so--called isomorphism problem.
+The good news is that isomorphism can be solved in polynomial time for
+molecular graphs due to the fact that molecules belong to a restricted
+class of graphs known as \emph{bounded valence graphs}; however, that
+same restriction prevents applying to a molecular graph the polynomial
+delay enumeration results obtained for general graphs (Chapter 8 in
+\cite{faulon2010}). While the complexity of molecular graph
+enumeration remains an open problem it has nonetheless been shown that
+molecules can be sampled efficiently \cite{goldberg1999}. Sampling
+procedures based on the Metropolis or the Genetic Algorithms have been
+developed and used to elucidate natural compounds from NMR data.
 
@@
 the current enumeration software products do not generally produce
-stereoisomers [3] or tautomers [4]; these require specific enumeration
+stereoisomers or tautomers; these require specific enumeration
 procedures, which are still the subject of investigations by the
@@
 (UCSF) are valuable resources to search for new drugs. For instance
-the well known ligand-based design approach uses these
-databases along with structure similarity tools to discover new
-compounds binding to biological targets (protein or nucleic
-acid). Compounds retrieved through ligand-based design, however, must
-already be present in databases, which is a limitation of the
-technique. To overcome that limitation, algorithms have recently been
-developed to enumerate all compounds in the chemical space matching
-circular [5] or path fingerprints [6], these fingerprints being
+the well known ligand-based design approach uses these databases along
+with structure similarity tools to discover new compounds binding to
+biological targets (protein or nucleic acid). Compounds retrieved
+through ligand-based design, however, must already be present in
+databases, which is a limitation of the technique. To overcome that
+limitation, algorithms have recently been developed to enumerate all
+compounds in the chemical space matching circular \cite{faulon2003} or
+path fingerprints \cite{fujiwara2008}, these fingerprints being
 popular tools to measure similarity. Enumerating molecules from
@@
 that require the enumeration of reaction networks. As reviewed in
-Chapter 11 in [1] several network enumeration techniques have been
-developed. However, these techniques generally suffer from a
+Chapter 11 in \cite{faulon2010} several network enumeration techniques
+have been developed. However, these techniques generally suffer from a
 combinatorial explosion of intermediate compounds being produced. One

### 2011-07-24 paper.tex
Added in Jannas edits

@@
 \begin{abstract}
-  One of the most prominent success stories in all of science over the last decade has
+b  One of the most prominent success stories in all the sciences over the last decade has
   been the advance of bioinformatics: interdisciplinary collaboration between
-  computer scientists and molecular biologists led to the sequencing the human genome
-  (among many other accomplishments), because researchers connected biological
+ computer scientists and molecular biologists led to the sequencing of the human genome
+  and many other accomplishments, since researchers connected biological
   concepts to the theory of string algorithms. However, despite this great
@@
   discipline---cheminformatics, the use of computers to represent
-  small molecules structures and analyze their properties based on aspects
-  of their structures. Cheminformatics has a wide applicability
-  from the discovery of novel drug candidates to optimizing
+ the structures of small molecules and analyze their properties. 
+  Cheminformatics has a wide applicability,
+  from the discovery of novel drug candidates to the optimization of
   certain physicochemical properties of small molecules. Until
-  recently, the data and techniques of chemoinformatics have been closely guarded
+ recently, the data and techniques of cheminformatics have been closely guarded
   secrets of companies whose financial success depended on being the first to produce
   the new ``miracle molecule.'' Only within the last decade---and largely because of
-  chemists volunteering their time for an Open Science ``movement''---do researchers
-  now have access to freely available software packages and databases of tens of
-  millions of chemicals. As a result, chemists now confront a variety of unsolved
+ chemists volunteering their time for an Open Science ``movement''---have researchers
+  gained access to freely available software packages and databases of tens of
+  millions of chemicals. Academic chemists now confront a variety of unsolved
   algorithmic problems that could not have been tackled a decade ago, but whose
@@
 increasing rate---public data stores such as at the European Bioinformatics
-Institute (EBI) contain on the order of 10 petabytes of biological information.
-However, while biological information has been available publicly for decades,
+Institute (EBI) contain in the order of 10 petabytes of biological information.
+However, while biological information has been publicly available for decades,
 the same has not been true for chemical information until very recently. It was
 not until 2004 that a large public small molecule structure repository (PubChem)
-became freely available. This was followed by other databases, as discuss later
+became freely available. This was followed by other databases, as discussed later
 in this article. In the same vein, open source software and the publication of
@@
 
-Making chemical information public is important because the drugs we
-take are the result of intensive and costly research, and the more
-information is shared (a process sometimes very hard to achieve in
-pharmaceutical companies) the easier it should be to develop novel
-treatments. How chemical information relates to computer science
-becomes clear when realizing the amount of data available
+Making chemical information public is important because the
+drugs we take are the result of intensive and costly research, 
+which is facilitated by the extent of information available. The
+more information is shared (a process very hard to achieve
+in pharmaceutical companies), the more information is available to each 
+researcher, facilitating the development of
+novel treatments. The relevance of computer
+science in this research becomes clear in considering the quantity of data available
 currently---74 million different molecules are currently stored in
-various public databases (such as PubChem) with a large amount of
-annotations, and the design of data structures for the representation
-and manipulation of chemical structure, as well as subsequent data
-mining of this information, are areas where expertise in computer
-science is dearly needed. (For a recent, detailed introduction to
+public databases (such as PubChem), together with an even larger number of annotations.
+The design of data structures for this chemical information, as well as subsequent
+data mining of this information, are areas where expertise in computer
+science is urgently needed. (For a recent, detailed introduction to
 cheminformatics, written for computer scientists, see
@@
 the bottom left hand corner of the figure, and the precise 3D shape is
-relevant in many situations (although not in others).  Nevertheless,
-the 3D information is lost when one only considers the graph of a
+relevant in many situations (although not in others).  Nevertheless, the
+3D information is lost when one only considers the graph of a
 molecule. To extend the concept of three-dimensionality further, in
@@
 promising avenue to explore, one needs to keep in mind that not all
-molecular properties are encoded in the graph and thus often other
-`descriptors' need to be used to capture the essentials of a chemical
-system \cite{bender2004}.
+molecular properties are encoded in the graph (for example, neither
+rotational states, nor different charge states are encoded) and thus
+often other `descriptors' need to be used to capture the essentials of
+a chemical system \cite{bender2004}.
 
-From the above it becomes clear that the nature of many problems in
-cheminformatics require that a complex mixture of combinatorial and
-continuous techniques.  If we oversimplify the nature of chemical
-structures and their properties, we can not describe the world around
-us in sufficient detail. On the other hand, if we try to encode every
-detail (\emph{e.g.}, all possible 3D structures and their properties,
-all metabolites of a structure in humans in different tissues), the
-complexity of collecting, storing, and mining data increases
+From the above it becomes clear that the nature of chemical problems
+is often a complex mixture of combinatorial and continuous
+problems. If we oversimplify the nature of chemical structures and
+their properties, we can not describe the world around us in
+sufficient detail. On the other hand, if we try to encode every detail
+(\emph{e.g.}, all possible 3D structures and their properties, all
+metabolites of a structure in humans in different tissues), the
+complexity of collecting, storing, and mining these data increases
 dramatically. For the field of cheminformatics to flourish, we need
@@
 and where the latter are able to devise ways of capturing, storing,
-and analyzing chemical data by bridging the gap between risk in
-generalizable encodings and complexity (space and time). The goals of
-cheminformatics are to support better chemical decision making by
-1. storing and integrating data in maintainable ways, 2.  providing
-enough open standards and tools so software engineering concepts allow
-applications and data to be used across heterogeneous platforms, and
-3. mining the many chemical property spaces in a time- and
-space-efficient way. We will discuss all three goals in the following
-sections.
-
-\subsection{Outline}
-\label{sec:outline}
-
-The remainder of the article discusses various topics in
-cheminformatics by placing them in the context of the relevant
-subfields of computer science, aiming to highlight the challenges that
-need to be addressed and possible entry points for computer science
-research. Section \ref{sec:databases} discusses issues related to the
-storage of chemical structures in relational databases and how
-ontologies allow us to query such collections beyond the standard
-graph based subgraph matching approaches. While databases usually
-contain explicit molecules, enumeration of sctructures based on
-various constraints is key to a variety of chemical problems ranging
-from mass spectrometry to reaction planning. Section
-\ref{sec:struct-enum} discusses challenges in this area. Section
-\ref{sec:activity-mining} highlights the problem of making predictions
-of biological activity from chemical structure and tieing it all
-together, Section \ref{sec:knowledge-management} discusses the
-challenges of managing chemical structure data along with associated
-biological data in production scenarios. Finally Section
-\ref{sec:role-open-source} discusses the public sources of chemical
-information and tools available for the handling of chemical structre
--- both of which are required for for the development of
-cheminformatics research programs.
+and analyzing chemical data which achieve optimal balances between the different risks in 
+generalizable encodings and complexity (space and time). The goals of cheminformatics are to
+support better chemical decision making by 1. storing and integrating
+data in maintainable ways, 2.  providing enough open standards and
+tools so software engineering concepts allow applications and data to
+be used across heterogeneous platforms, and 3. mining the many
+chemical property spaces in a time- and space-efficient way. We will
+discuss all three goals in the following sections.
 
@@
 is still limited. For example, it cannot differentiate between
-certain types of stereoisomers: Figure 2 illustrates two stereoisomers for which
+certain types of stereoisomers: Figure \ref{cistrans} illustrates two stereoisomers for which
 the generated InChI is the same.
 
-%Figure 2: The generated InChI is the same for the stereoisomers cisplatin and
+%Figure {cistrans}: The generated InChI is the same for the stereoisomers cisplatin and
 %transplatin.
 
-InChI or other identity-mapping algorithms allow for exact searching
-(within the limitations of the algorithms), but two other requirements
-for scientific discovery based on chemical databases are
-\emph{substructure searching} and \emph{similarity searching}. In
-substructure searching, the database is searched for a specified
+InChI or other identity-mapping algorithms allow for exact searching (within the
+limitations of the algorithms), but two other requirements for scientific
+discovery based on chemical databases are \emph{substructure searching} and \emph{similarity
+searching}. In substructure searching, the database is searched for a specified
 wholly contained part of the structures in the database; in similarity
-searching, the database is searched for structures similar to a
-provided search structure. Chemical search packages are often
-implemented and optimized for a given database technology, for
-example, the OrChem package is an open source chemical search package
-for the Oracle database \cite{rijnbeek2009}.
+searching, the database is searched for structures similar to a provided search
+structure. Chemical search packages are often implemented and optimized for a
+given database technology, for example, the OrChem package is an open source
+chemical search package for the Oracle database \cite{rijnbeek2009}.
 
-Graph substructure matching is a variant of \emph{graph isomorphism
-  detection},which is well known to be computationally intractable
-[3]. To execute a graph isomorphism search across a full chemical
-database of thousands or even millions of structures is completely
-infeasible. For that reason, complex heuristics must be used to
-drastically filter the candidate search space. Structure-based
-\emph{fingerprints} are commonly used for this purpose. Fingerprints
-encode characteristic features of a given chemical structure in a
-Boolean array, or bitmap. The fingerprint is created by an algorithm
-which generates bitmap patterns for each atom in the structure; then
-each atom and its nearest neighbors, including the bonds between them;
-then each group of atoms connected by paths of increasing lengths up
-to the maximum implemented for the algorithm (commonly around
-eight). The bitmap patterns are added together into the final
-fingerprint. If all bits in a query fingerprint are also present in
-the target fingerprint of a stored database structure, only then is
-the structure subjected to the computationally expensive subgraph
-matching algorithm. Bit operations are very fast and independent of
-the number of atoms in a structure due to the fixed length of the
-fingerprint.
+Graph substructure matching is a variant of \emph{graph isomorphism detection},which is
+well known to be computationally intractable [3]. To execute a graph isomorphism
+search across a full chemical database of thousands or even millions of
+structures is infeasible. For that reason, complex heuristics must be
+used to drastically filter the candidate search space. Structure-based
+\emph{fingerprints} are commonly used for this purpose. Fingerprints encode
+characteristic features of a given chemical structure in a Boolean array, or
+bitmap. The fingerprint is created by an algorithm which generates bitmap
+patterns for each atom in the structure; then each atom and its nearest
+neighbors, including the bonds between them; then each group of atoms connected
+by paths of increasing lengths up to the maximum implemented for the algorithm
+(commonly around eight). The bitmap patterns are added together into the final
+fingerprint. If all bits in a query fingerprint are also present in the target
+fingerprint of a stored database structure, only then is the structure subjected
+to the computationally expensive subgraph matching algorithm. Bit operations are
+very fast and independent of the number of atoms in a structure due to the fixed
+length of the fingerprint.
 
 Fingerprints also facilitate the rapid calculation of quantitative
-structure-based similarity measures. An example measure of similarity
-is the \emph{Tanimoto coefficient}, calculated as the ratio $T(a,b) =
-\frac{c}{(a+b-c)}$, where c is the count of bits on in the same
-position in both the two fingerprints, a is the count of bits on in
-the first structure, and b is the count of bits on in the second
-structure. The Tanimoto coefficient varies in the range 0.0 -- 1.0,
-with a score of 1.0 indicating that the two structures are very
-similar (i.e. their fingerprints are the same). Structure-based
-similarity is of limited use in grouping of chemicals based on
-non-structural features, such as shared bioactivity profiles. For this
-purpose, a large-scale and flexible classification structure is needed
-for chemicals and their linked biological information. One such answer
-to this need is provided in the form of \emph{ontologies}.  An
-ontology is a formal specification of entities and their relationships
-in a domain of interest. It is structured around an underlying logical
-representation such as Description Logic ~\cite{baaderdl2007}. The
-logic-based representation allows automated reasoning to derive new
-knowledge (inferences) from the knowledge encoded, making the
-representation both compact and powerful.  Ontologies are already in
-widespread use across the biomedical domain. A similarity measure
-based on the information encoded in ontologies is termed
-\emph{semantic similarity}, and such measures have already been
-applied to enhance chemical classification in several problem areas
-\cite{couto2010}. Many open questions still remain about the best way
-to combine the information encoded in chemical structures with
-chemical ontologies \cite{hastingsowled2010}.
+structure-based similarity measures. An example measure of similarity is the
+\emph{Tanimoto coefficient}, calculated as the ratio $T(a,b) = \frac{c}{(a+b-c)}$,
+where $c$ is the count of bits on in the same position in both the two
+fingerprints, $a$ is the count of bits on in the first structure, and $b$ is the
+count of bits on in the second structure. The Tanimoto coefficient varies in the
+range 0.0 -- 1.0, with a score of 1.0 indicating that the two structures are very
+similar (i.e. their fingerprints are the same). Structure-based similarity is of
+limited use in grouping of chemicals based on non-structural features, such as
+shared bioactivity profiles. For this purpose, a large-scale and flexible
+classification structure is needed for chemicals and their linked biological
+information. One such answer to this need is provided in the form of \emph{ontologies}.
+An ontology is a formal specification of entities and their relationships in a
+domain of interest. It is structured around an underlying logical representation
+such as Description Logics~\cite{baaderdl2007}. The logic-based representation allows
+automated reasoning to derive new knowledge (inferences) from the knowledge
+encoded, making the representation both compact and powerful.
+Ontologies are already in widespread use across the biomedical domain. A
+similarity measure based on the information encoded in ontologies is termed
+\emph{semantic similarity}, and such measures have already been applied to enhance
+chemical classification in several problem areas \cite{couto2010}. Many open
+questions still remain about the best way to combine the information encoded in
+chemical structures with chemical ontologies \cite{hastingsowled2010}.
 
@@
 
-With the advent of high-throughput screening technologies, large
-libraries of compounds can be screened against multiple targets in an
-efficient manner. Such panel assays provide a broad, systems-level
-view of small molecule activities.  Models developed on such data
-afford us the opportunity to identify targets, characterize off-target
-effects and so on. However, most approaches to this problem are rather
-simplistic since they develop multiple individual models.  Instead one
-could imagine an approach that takes into account the correlations in
-observed activities against multiple targets within a single
-model. Such an approach should lead to more robust predictions for
-panel assays. Finally, does this also better reflect clinical relevant
-compound profiles \cite{kuhn2010} and a 'personalized medicine'
-concept (e.g. drug-drug interaction profiles).
+With the advent of high-throughput screening technologies, large libraries of
+compounds can be screened against multiple targets in an efficient manner. Such
+panel assays provide a broad, systems-level view of small molecule activities.
+Models developed on such data afford us the opportunity to identify targets,
+characterize off-target effects and so on. However, most approaches to this
+problem are rather simplistic since they develop multiple individual models.
+Instead one could imagine an approach that takes into account the correlations
+in observed activities against multiple targets within a single model. Such an
+approach should lead to more robust predictions for panel assays. Finally, 
+this also better reflects clinically relevant compound profiles \cite{kuhn2010}
+and a `personalized medicine' concept (e.g. drug-drug interaction profiles) [The
+Decision Tree].
 

### 2011-08-01 paper.tex
Added Aarons latest edits

@@
 \begin{abstract}
-b  One of the most prominent success stories in all the sciences over the last decade has
+One of the most prominent success stories in all the sciences over the last decade has
   been the advance of bioinformatics: interdisciplinary collaboration between
@@
 Novel technologies in the life sciences churn out information at an ever
-increasing rate---public data stores such as at the European Bioinformatics
-Institute (EBI) contain in the order of 10 petabytes of biological information.
+increasing rate---public data stores such as the one at the European Bioinformatics
+Institute (EBI) contain on the order of 10 petabytes of biological information.
 However, while biological information has been publicly available for decades,
@@
 
-There is a considerable progress in open chemical and bioactivity databases like
+There is considerable progress in open chemical and bioactivity databases like
 ChEMBL and NCBI-PubChem, which are both freely available. Still, the integration
@@
 generation of a single canonical SMILES for a particular structure. However,
-with time several
-different SMILES implementations became available with no universal
-standard of generating a canonical SMILES. 
+over the course of time, several
+different SMILES implementations became available, with no universal
+standard for generating a canonical SMILES.
 %After years of research and various
@@
 Graph substructure matching is a variant of \emph{graph isomorphism detection},which is
-well known to be computationally intractable [3]. To execute a graph isomorphism
+widely believed to be computationally intractable [3]. To execute a graph isomorphism
 search across a full chemical database of thousands or even millions of
@@
 development of tools that are at the interface between
-chemoinformatics and bioinformatics.
+cheminformatics and bioinformatics.
 
@@
 black box methods can be less useful, though attempts have been made
-to interpret vectorial models built using neural networks, SVMs or
+to interpret vectorial models built using neural networks, Support Vector Machines or
 random forests. Related to this, there is a never-ending debate on the
@@
 ``Enterprise 2.0,'' required delivering on the promise to increase
-interconnectivity, serve as silo-breaker, and to improve collaboration
+interconnectivity, and to improve collaboration
 opportunities, and indeed it already performed these functions many years ago.
@@
 1960s, with the creation of DENDRAL, the first expert system of any
-kind.  Much of its development from the 1960s until the new millenium
+kind.  Much of the development of cheminformatics from the 1960s until the new millenium
 was performed in secret, because of competition in the pharmaceutical

### 2011-08-02 paper.tex
Added in Joergs edits

@@
 \begin{itemize}
-\item \textit{Tautomerism} - Tautomers are isomers (same molecular formula but different structural formula) of organic compounds that readily interconvert by a chemical reaction called tautomerization.
+\item \textit{Bioassay} - A system measuring the biological activity of a chemical compound on a particular biological assay (with all its parameters).
+\item \textit{Biodegradation} - is the chemical dissolution of materials by bacteria or other biological means.
+\item \textit{Conformation and conformational isomerism} - In chemistry, conformational isomerism is a form of stereoisomerism in which the isomers can be interconverted exclusively by rotations about formally single bonds.
+Single possible states of compounds are called conformers.
+\item \textit{Ligand (biochemistry)} - a substance that binds to a protein.
+\item \textit{Ligand (coordination chemistry)} - an atom, ion, or functional group that donates one or more of its electrons through a coordinate covalent bond to one or more central atoms or ions.
+\item \textit{Mass spectrometry} - Mass spectrometry (MS) is an analytical technique that measures the mass-to-charge ratio of charged particles.
+It is used for determining masses of particles, for determining the elemental composition of a sample or molecule, and for elucidating the chemical structures of molecules, such as peptides and other chemical compounds.
+\item \textit{Metabolites} - Metabolites are the intermediates and products of metabolism. The term metabolite is usually restricted to small molecules (chemical compounds).
+\item \textit{Proteochemometric modeling} - combining data mining for chemical compounds (ligands) and proteins within the same machine learning framework.
 \item \textit{Protonation and protonation state} - In chemistry, protonation is the addition of a proton (H$^+$) to an atom, molecule, or ion.
@@
 de-protonated, mono-protonated, di-protonated, or any variations of those states).
-\item \textit{Bioassay} - A system measuring the biological activity of a chemical compound on a particular biological assay (with all its parameters).
-\item \textit{Metabolites} - Metabolites are the intermediates and products of metabolism. The term metabolite is usually restricted to small molecules (chemical compounds).
-\item \textit{Mass spectrometry} - Mass spectrometry (MS) is an analytical technique that measures the mass-to-charge ratio of charged particles.
-It is used for determining masses of particles, for determining the elemental composition of a sample or molecule, and for elucidating the chemical structures of molecules, such as peptides and other chemical compounds.
-\item \textit{Ligand (coordination chemistry)} - an atom, ion, or functional group that donates one or more of its electrons through a coordinate covalent bond to one or more central atoms or ions.
-\item \textit{Ligand (biochemistry)} - a substance that binds to a protein.
+\item \textit{Reaction catalysis} - Catalysis is the change in rate of a chemical reaction due to the participation of a substance called a catalyst.
 \item \textit{Receptor (biochemistry)} - in biochemistry, a protein molecule that receives and responds to a neurotransmitter, or other substance.
-\item \textit{Proteochemometric modeling} - combining data mining for chemical compounds (ligands) and proteins within the same machine learning framework.
-\item \textit{Biodegradation} - is the chemical dissolution of materials by bacteria or other biological means.
-\item \textit{Reaction catalysis} - Catalysis is the change in rate of a chemical reaction due to the participation of a substance called a catalyst.
-\item \textit{Conformation and conformational isomerism} - In chemistry, conformational isomerism is a form of stereoisomerism in which the isomers can be interconverted exclusively by rotations about formally single bonds.
-Single possible states of compounds are called conformers.
+\item \textit{Regulation authorities} - In the US the Food and Drug Administration (FDA) is responsible for protecting and promoting public health through the regulation and supervision of e.g.
+dietary supplements, prescription and over-the-counter pharmaceutical drugs (medications), vaccines, biopharmaceuticals, medical devices, and many other health related items.
+The European Medicines Agency (EMA) is a European agency for the evaluation of medicinal products.
 \item \textit{SMARTS} - SMiles ARbitrary Target Specification \\(SMARTS) is a language for specifying substructure patterns in molecules
+\item \textit{Stereoisomer/stereoisomerism} - Stereoisomers are isomeric molecules that have the same molecular formula and sequence of bonded atoms (constitution), but that differ only in the three-dimensional orientations of their atoms in space.
+\item \textit{Tautomerism} - Tautomers are isomers (same molecular formula but different structural formula) of organic compounds that readily interconvert by a chemical reaction called tautomerization.
 \end{itemize}
@@
 
-Most cheminformatics applications rely on large databases of chemical structures
-and their properties. Organizing and maintaining such databases, searching based
+Most cheminformatics applications rely on large databases of chemical structures, 
+their properties, and relationships to e.g. biological targets.
+Organizing and maintaining such databases, searching based
 on features of chemical structures, and clustering similar structures together,
@@
 growth, usage, and collaboration scenarios between private and public parties.
-Only when the technical and scientific questions are solved can we ensure that
-legal questions about contribution metrics will get enough room to grow.
+In a business context we need to address ROI (return-of-investment) questions, not only for single
+institutions and companies, but also for collaboration networks of legal partners.
+So, only when the technical and scientific questions are solved can we ensure that
+legal questions about contribution metrics will get enough room to grow. Solving challenges is one thing (science and technical),
+ensuring that the contributing partners get rewarded by their contribution another (legal).
 

### 2011-11-08 supplement_history.tex
Added suppl history document. Updated refs

\documentclass{sig-alternate}

\begin{document}

\title{Supplement History of Chem(o)informatics as part of Cheminformatics: The Computer Science of Chemical Discovery}
\numberofauthors{9}
\author{
\alignauthor
Joerg Kurt Wegner\\
       \affaddr{Tibotec BVBA}\\
       \affaddr{Turnhoutseweg 30}\\
       \affaddr{2340 Beerse Turnhout, Belgium}\\
       \email{jwegner@its.jnj.com}
% 2nd. author
\alignauthor
Aaron Sterling\\
       \affaddr{Department of Computer Science}\\
       \affaddr{Iowa State University}\\
       \affaddr{Ames, Iowa, USA}\\
       \email{sterling@iastate.edu}
% 3rd author
\alignauthor
Rajarshi Guha\\
\affaddr{NIH Center for Translational Therapeutics}\\
\affaddr{9800 Medical Center Drive}\\
\affaddr{Rockville, MD 20850}\\
\email{guhar@mail.nih.gov}
}

\additionalauthors{Additional authors:
Andreas Bender (University of Cambridge, email: {\texttt{andreas.bender@cantab.net}}),
Jean-Loup Faulon (University of Evry, email: {\texttt{Jean-Loup.Faulon@issb.genopole.fr}}),
Janna Hastings (European Bioinformatics Institute, Cambridge, UK, email: {\texttt{janna.hastings@gmail.com}}),
Noel O'Boyle (University College Cork, Cork, Ireland, email: {\texttt{baoilleach@gmail.com}}),
John Overington (European Bioinformatics Institute, Cambridge, UK, email: {\texttt{jpo@ebi.ac.uk}}),
Herman Van Vlijmen (Tibotec, Beerse, Belgium, email: {\texttt{hvvlijme@its.jnj.com}}), and
Egon Willighagen (Karolinska Institutet, Stockholm, Sweden, email: {\texttt{egon.willighagen@ki.se}})
.}
\date{25 June 2011}
\maketitle

\section{History of Chem(o)informatics}
The aim of this brief historic review of cheminformatics is to put the
content of the main article into a broader perspective. Therefore it
might be useful to clarify some other terms and how cheminformatics
might relate to them.  Still, Chen \cite{Chen2006} and Brown
\cite{brown2009} cited various definitions of pioniers in the field
and Brown concluded rightfully: '\textit{The differences in
  definitions (of the term cheminformatics) are largely a result of
  the types of analyses that particular scientists practice and no
  single definition is intended to be all-encompassing}.'
\begin{itemize}
\item \textbf{Quantum chemistry} is something most chemists would
  relate with 'theoretical chemistry'. In short, it allows to describe
  chemical molecules on an electron density level, which allows also
  to describe physical properties, e.g infrared spectra of a chemical
  compound.  The results of quantum chemistry calculation are used in
  molecular modeling (energy force field parameters) and
  cheminformatics (atom partial charges, or full molecule properties,
  e.g molecule polarizabilities).
\item \textbf{Molecular modeling} or \textbf{Computational chemistry}
  considers typically energy calculations of single molecules or
  interactions with other molecules, e.g. proteins. 'Molecular
  docking' belongs into this category, since it places a chemical
  molecule in the active site of enzymes and calculates
  energy/interaction scores between a chemical molecule and the
  protein. Surely are there various cheminformatics concepts required
  to perform such a task. Still, strictly speaking is cheminformatics
  rather on the chemical molecule side than on the protein side,
  though are both sides affecting each other. This area is typically
  strongly related to structural biology and technologies like XRay
  crystallography and NMR spectra structure determination.
\item \textbf{Chemometrics} can be considered as applied
  cheminformatics to the field of chemical/physical analysis,
  e.g. infrared, mass and NMR spectra analysis.
\item '\textbf{Translational research} aims to improve the health and
  longevity of the world's populations and depends on developing
  broad-based teams of scientists and scholars who are able to focus
  their efforts to link basic scientific discoveries with the arena of
  clinical investigation, and translating the results of clinical
  trials into changes in clinical practice, informed by evidence from
  the social and political sciences.' (Source:Wikipedia).  Since
  clinical trials can be connected to drugs, chemical molecules, even
  all of translational research and biobanking could be considered as
  a field of cheminformatics, which would be on the other hand just a
  small contribution to a full knowledge management in this field
  \cite{Szalma2010}.
\end{itemize}
If we consider all information, analysis, and optimization of a
chemical molecule as 'cheminformatics', then the field is very large,
since a chemical molecule plays the central role in many related
disciplines. Still, for keeping it practical the aim of this review is
not being exhaustive, but rather to point the reader to more detailed
references and to highlight some historic milestones.

One of the most comprehensive cheminformatic series was published by
Prof. Gasteiger in 2003 \cite{Gasteiger2003}, one of the pioniers in
the field.  Compacter introductory books followed from Gasteiger/Engel
\cite{gasteigerengel2003} and Leach/Gillet \cite{leachgillet2007}.
The books of Bajorath \cite{Bajorath2004} and Oprea \cite{oprea2005}
focus on the applications of cheminformatics in drug design.  The
first books discussing chemical graph theory and algorithms were
published in 1989 by Zupan \cite{zupan1989}, 1991 by Bonchev/Rouvrey
\cite{bonchevrouvrey1991,bonchevrouvrey2003} and 1992 by Trinastic
\cite{Trinajstic1992}.  The most recent one was published by
Faulon/Bender \cite{faulon2010}.  Other books focus more on specific
fields/applications like mathematical challenges
\cite{mathchallenges1995}, molecular diversity
\cite{moleculardiversity1999}, factor analysis \cite{Malinowski2002},
evolutionary algorithms \cite{clark2000}, molecular descriptors
\cite{todeschini2000}.

\section{Historic milestones}
The first chemical graphs were drawn by the Scottish chemist Willam
Cullen in 1758 \cite{bonchevrouvrey1991}, which called them initially
affinity graphs.  After developing the concept of bonds between atoms
(Couper, 1858) the first chemical graphs occured in publications of
Brown in 1864 and Cayley in 1874 \cite{bonchevrouvrey1991,brown2009}.

The year 1946 may be regarded as the birth year of chemoinformatics
\cite{Chen2006}: '\textit{In 1946 King et al.\cite{kct1946} published
  an article illustrating the use of IBM's business accounting
  machines in carrying out the construction of the rotational spectra
  of asymmetric rotors by the evaluation of mathematical equations for
  line position and line intensity.}'

The first records of managing chemical information go back to chemical
literature being indexed since the year 1771.  In 1881 the first
edition of \textit{Beilstein's Handbuch der Organischen Chemie}
ecyclopedia was published \cite{polanski2009} registering 1500
chemical compounds. The short term 'chemoinformatics' was coined by
Brown in 1998\cite{brown1998}, while the \textit{Journal of Chemical
  Documentation} (founded in 1961) changed already its name in 1975 to
\textit{Journal of Chemical Information and Computer Sciences} to
reflect the tight connection between chemical information and computer
science. The Chemical Abstract Service (CAS) of the American Chemical
Society (ACS) formed in 1955 a research department and starting from
1965 they provided their Chemical Registry System, and in 1968 they
made the first computer-readable file of all abstracted documents
available \cite{Chen2006}.

Ray and Kirsch published in 1957 the first substructure searching
algorithm for allowing to find chemical records via computers
\cite{RayKirsch1957}.

One of the key algorithms for molecular graph canonicalization is the
Morgan algorithm \cite{Morgan1965}. It will search for the most
central bond in a graph and allows then to traverse other atoms,
resolve ties, and assigning new atom indices. Still, there are various
options how different software implementations are handling the 'ties'
for assigning new atom indices. So, by using the canonicalization of
multiple software implementations and chemical expert systems can be
even used to create randomized (canonical, only when sticking to one
software) molecular graphs, which can be used as stochastic input for
other algorithms \cite{cokfl06}. The central bond is also being used
to create 3D conformational search trees \cite{Schwab2001}.

In 1965 the DENDRAL expert system started with the aim to
automatically determine the structure of an unknown chemical compound
from the corresponding mass spectrum \cite{Gray1986}. The system is
also often being cited as one of the earliest artificial intelligence
and expert systems \cite{Chen2006}.

The first system supporting chemical synthesis planning was OCSS
(Organic Chemical Simulation of Syntheses) developed by Corey and
Wipke in 1969 \cite{CoreyWipke1969}.

In the early 1970s the collaboration between Lempel-Ziv-Welch has led
to the LZW algorithm, since the Cheminformatician Lynch was discussing
chemical fragment screening approaches with the computer scientists
Lempel and Ziv \cite{brown2009}.

In 1978 Gasteiger/Marsili published a fast algorithm to calculate
partial charges in organic molecules by a Partial Equalization of
Orbital Electronegativities (PEOE) \cite{gm78}, which is till today
one of the gold-standards to calculate partial charges of atoms.

Instead of converting molecular graphs to a vector representation for
applying machine learning methods, it is also possible to use direct
molecular graph mining methods \cite{okada2006}. Some of the first
published methods use neural networks \cite{kireev1995}, inductive
logic programming \cite{yh02a}, or graph kernels \cite{kti03}.

Molecular docking of chemical molecules into proteins is an important
application in molecular modeling \cite{Leach2001}. It is noteworthy
that in some cases molecular shape matching methods alone
(cheminformatics), can outperform energy-based molecular docking
algorithms (molecular modeling) \cite{hsn07}.

\bibliographystyle{abbrv}
\bibliography{paper}

\end{document}

### 2011-11-08 supplement_history.tex
updated with joergs latest version of the supplment

@@
 
-\title{Supplement History of Chem(o)informatics as part of Cheminformatics: The Computer Science of Chemical Discovery}
+\title{Supplement History of Chem(o)informaqtics as part of Cheminformatics: The Computer Science of Chemical Discovery}
 \numberofauthors{9}
@@
 \section{History of Chem(o)informatics}
-The aim of this brief historic review of cheminformatics is to put the
+The aim of this brief review of the history of cheminformatics is to put the
 content of the main article into a broader perspective. Therefore it
-might be useful to clarify some other terms and how cheminformatics
-might relate to them.  Still, Chen \cite{Chen2006} and Brown
-\cite{brown2009} cited various definitions of pioniers in the field
-and Brown concluded rightfully: '\textit{The differences in
-  definitions (of the term cheminformatics) are largely a result of
-  the types of analyses that particular scientists practice and no
-  single definition is intended to be all-encompassing}.'
+might be useful to provide some definitions of cheminformatics, which
+might also shed some light on the founding principles of the
+field. Chen \cite{Chen2006} and Brown \cite{brown2009} cited various
+definitions of pioniers in the field and Brown concluded rightfully:
+'\textit{The differences in definitions (of the term cheminformatics)
+  are largely a result of the types of analyses that particular
+  scientists practice and no single definition is intended to be
+  all-encompassing}.'
+\begin{itemize}
+\item '\textit{The mixing of information resources to transform data
+    into information, and information into knowledge, for the intended
+    purpose of making better decisions faster in the arena of drug
+    lead identification and optimization.}' Frank K. Brown, 1998.
+\item '[Chemoinformatics involves]... \textit{the computer
+    manipulation of two- or threedimensional chemical structures and
+    excludes textual information. This distinguishes the term from
+    chemical information, largely a discipline of chemical librarians
+    and does not include the development of computational methods.}'
+  Peter Willett, 2002.
+\item '\textit{...the application of informatics to solve chemical
+    problems.}' and '\textit{chemoinformatics makes the point that
+    you’re using one scientific discipline to understand another
+    scientific discipline.}' Johann Gasteiger, 2002
+\end{itemize}
+Furthermore is it useful to clarify some other terms and how
+cheminformatics might relate to them:
 \begin{itemize}
@@
   e.g molecule polarizabilities).
-\item \textbf{Molecular modeling} or \textbf{Computational chemistry}
-  considers typically energy calculations of single molecules or
-  interactions with other molecules, e.g. proteins. 'Molecular
-  docking' belongs into this category, since it places a chemical
-  molecule in the active site of enzymes and calculates
-  energy/interaction scores between a chemical molecule and the
-  protein. Surely are there various cheminformatics concepts required
-  to perform such a task. Still, strictly speaking is cheminformatics
-  rather on the chemical molecule side than on the protein side,
-  though are both sides affecting each other. This area is typically
-  strongly related to structural biology and technologies like XRay
-  crystallography and NMR spectra structure determination.
-\item \textbf{Chemometrics} can be considered as applied
-  cheminformatics to the field of chemical/physical analysis,
-  e.g. infrared, mass and NMR spectra analysis.
-\item '\textbf{Translational research} aims to improve the health and
-  longevity of the world's populations and depends on developing
-  broad-based teams of scientists and scholars who are able to focus
-  their efforts to link basic scientific discoveries with the arena of
-  clinical investigation, and translating the results of clinical
-  trials into changes in clinical practice, informed by evidence from
-  the social and political sciences.' (Source:Wikipedia).  Since
-  clinical trials can be connected to drugs, chemical molecules, even
-  all of translational research and biobanking could be considered as
-  a field of cheminformatics, which would be on the other hand just a
-  small contribution to a full knowledge management in this field
-  \cite{Szalma2010}.
+\item \textbf{Molecular modeling} or \textbf{Computational chemistry} considers typically energy calculations of single molecules or interactions
+with other molecules, e.g. proteins. 'Molecular docking' belongs into this category, since it places a chemical molecule in the active site of enzymes
+and calculates energy/interaction scores between a chemical molecule and the protein. Surely are there various cheminformatics concepts required to
+perform such a task. Still, strictly speaking is cheminformatics rather on the chemical molecule side than on the protein side, though are both sides
+affecting each other. This area is typically strongly related to structural biology and technologies like XRay crystallography and 
+NMR spectra structure determination. 
+\item \textbf{Chemometrics} can be considered as applied cheminformatics to the field of chemical/physical analysis, e.g. infrared, mass and NMR spectra analysis.
+\item '\textbf{Translational research/medicine} aims to improve the health and longevity of the world's populations and depends on developing broad-based 
+teams of scientists and scholars who are able to focus their efforts to link basic scientific discoveries with the arena of clinical investigation, 
+and translating the results of clinical trials into changes in clinical practice, informed by evidence from the social and political sciences.' (Source:Wikipedia).
+Since clinical trials can be connected to drugs, chemical molecules, even all of translational research and biobanking could be considered as a field
+of cheminformatics, which would be on the other hand just a small contribution to a full knowledge management in this field \cite{Szalma2010}.
 \end{itemize}
-If we consider all information, analysis, and optimization of a
-chemical molecule as 'cheminformatics', then the field is very large,
-since a chemical molecule plays the central role in many related
-disciplines. Still, for keeping it practical the aim of this review is
-not being exhaustive, but rather to point the reader to more detailed
-references and to highlight some historic milestones.
+If we consider all information, analysis, and optimization of a chemical molecule as 'cheminformatics', then the field is very large, since a chemical
+molecule plays the central role in many related disciplines. Still, for keeping it practical the aim of this review is not being exhaustive, but rather to 
+point the reader to more detailed references and to highlight some historic milestones.
 
-One of the most comprehensive cheminformatic series was published by
-Prof. Gasteiger in 2003 \cite{Gasteiger2003}, one of the pioniers in
-the field.  Compacter introductory books followed from Gasteiger/Engel
-\cite{gasteigerengel2003} and Leach/Gillet \cite{leachgillet2007}.
-The books of Bajorath \cite{Bajorath2004} and Oprea \cite{oprea2005}
-focus on the applications of cheminformatics in drug design.  The
-first books discussing chemical graph theory and algorithms were
-published in 1989 by Zupan \cite{zupan1989}, 1991 by Bonchev/Rouvrey
-\cite{bonchevrouvrey1991,bonchevrouvrey2003} and 1992 by Trinastic
-\cite{Trinajstic1992}.  The most recent one was published by
-Faulon/Bender \cite{faulon2010}.  Other books focus more on specific
-fields/applications like mathematical challenges
-\cite{mathchallenges1995}, molecular diversity
-\cite{moleculardiversity1999}, factor analysis \cite{Malinowski2002},
-evolutionary algorithms \cite{clark2000}, molecular descriptors
-\cite{todeschini2000}.
+One of the most comprehensive cheminformatic series was published by Prof. Gasteiger in 2003 \cite{Gasteiger2003}, one of the pioniers in the field.
+Compacter introductory books followed from Gasteiger/Engel \cite{gasteigerengel2003} and Leach/Gillet \cite{leachgillet2007}.
+The books of Bajorath \cite{Bajorath2004} and Oprea \cite{oprea2005} focus on the applications of cheminformatics in drug design.
+The first books discussing chemical graph theory and algorithms were published in 1989 by Zupan \cite{zupan1989}, 
+1991 by Bonchev/Rouvrey \cite{bonchevrouvrey1991,bonchevrouvrey2003} and 1992 by Trinastic \cite{Trinajstic1992}. 
+The most recent one was published by Faulon/Bender \cite{faulon2010}.
+Other books focus more on specific fields/applications like mathematical challenges \cite{mathchallenges1995}, 
+molecular diversity \cite{moleculardiversity1999}, factor analysis \cite{Malinowski2002}, evolutionary algorithms \cite{clark2000},
+molecular descriptors \cite{todeschini2000}.
 
 \section{Historic milestones}
-The first chemical graphs were drawn by the Scottish chemist Willam
-Cullen in 1758 \cite{bonchevrouvrey1991}, which called them initially
-affinity graphs.  After developing the concept of bonds between atoms
-(Couper, 1858) the first chemical graphs occured in publications of
-Brown in 1864 and Cayley in 1874 \cite{bonchevrouvrey1991,brown2009}.
+The first chemical graphs were drawn by the Scottish chemist Willam Cullen in 1758 \cite{bonchevrouvrey1991}, which called them initially affinity graphs.
+After developing the concept of bonds between atoms (Couper, 1858) the first chemical graphs occured in publications 
+of Brown in 1864 and Cayley in 1874 \cite{bonchevrouvrey1991,brown2009}.
 
-The year 1946 may be regarded as the birth year of chemoinformatics
-\cite{Chen2006}: '\textit{In 1946 King et al.\cite{kct1946} published
-  an article illustrating the use of IBM's business accounting
-  machines in carrying out the construction of the rotational spectra
-  of asymmetric rotors by the evaluation of mathematical equations for
-  line position and line intensity.}'
+The year 1946 may be regarded as the birth year of chemoinformatics \cite{Chen2006}: '\textit{In 1946 King et al.\cite{kct1946} published an article illustrating 
+the use of IBM's business accounting machines in carrying out the construction of the rotational spectra of asymmetric rotors
+by the evaluation of mathematical equations for line position and line intensity.}'
 
-The first records of managing chemical information go back to chemical
-literature being indexed since the year 1771.  In 1881 the first
-edition of \textit{Beilstein's Handbuch der Organischen Chemie}
-ecyclopedia was published \cite{polanski2009} registering 1500
-chemical compounds. The short term 'chemoinformatics' was coined by
-Brown in 1998\cite{brown1998}, while the \textit{Journal of Chemical
-  Documentation} (founded in 1961) changed already its name in 1975 to
-\textit{Journal of Chemical Information and Computer Sciences} to
-reflect the tight connection between chemical information and computer
-science. The Chemical Abstract Service (CAS) of the American Chemical
-Society (ACS) formed in 1955 a research department and starting from
-1965 they provided their Chemical Registry System, and in 1968 they
-made the first computer-readable file of all abstracted documents
-available \cite{Chen2006}.
+The first records of managing chemical information go back to chemical literature being indexed since the year 1771.
+In 1881 the first edition of \textit{Beilstein's Handbuch der Organischen Chemie} ecyclopedia was 
+published \cite{polanski2009} registering 1500 chemical compounds. The short term 'chemoinformatics' was coined by Brown 
+in 1998\cite{brown1998}, while the \textit{Journal of Chemical Documentation} (founded in 1961) changed already its name in 1975 to 
+\textit{Journal of Chemical Information and Computer Sciences} to reflect the tight connection between chemical information 
+and computer science. The Chemical Abstract Service (CAS) of the American Chemical Society (ACS) formed in 1955 a research department
+and starting from 1965 they provided their Chemical Registry System, and in 1968 they made the first computer-readable file of all
+abstracted documents available \cite{Chen2006}.
 
-Ray and Kirsch published in 1957 the first substructure searching
-algorithm for allowing to find chemical records via computers
-\cite{RayKirsch1957}.
+Ray and Kirsch published in 1957 the first substructure searching algorithm for allowing to find chemical records via computers \cite{RayKirsch1957}.
 
-One of the key algorithms for molecular graph canonicalization is the
-Morgan algorithm \cite{Morgan1965}. It will search for the most
-central bond in a graph and allows then to traverse other atoms,
-resolve ties, and assigning new atom indices. Still, there are various
-options how different software implementations are handling the 'ties'
-for assigning new atom indices. So, by using the canonicalization of
-multiple software implementations and chemical expert systems can be
-even used to create randomized (canonical, only when sticking to one
-software) molecular graphs, which can be used as stochastic input for
-other algorithms \cite{cokfl06}. The central bond is also being used
-to create 3D conformational search trees \cite{Schwab2001}.
+One of the key algorithms for molecular graph canonicalization is the Morgan algorithm \cite{Morgan1965}. It will search for the most central bond
+in a graph and allows then to traverse other atoms, resolve ties, and assigning new atom indices. Still, there are various options how different 
+software implementations are handling the 'ties' for assigning new atom indices. So, by using the canonicalization of multiple software implementations and chemical
+expert systems can be even used to create randomized (canonical, only when sticking to one software) molecular graphs, which can be used as stochastic input for 
+other algorithms \cite{cokfl06}. The central bond is also being used to create 3D conformational search trees \cite{Schwab2001}.
 
-In 1965 the DENDRAL expert system started with the aim to
-automatically determine the structure of an unknown chemical compound
-from the corresponding mass spectrum \cite{Gray1986}. The system is
-also often being cited as one of the earliest artificial intelligence
-and expert systems \cite{Chen2006}.
+In 1965 the DENDRAL expert system started with the aim to automatically determine the structure of an unknown chemical compound from the 
+corresponding mass spectrum \cite{Gray1986}. The system is also often being cited as one of the earliest artificial intelligence and expert systems  \cite{Chen2006}.
 
-The first system supporting chemical synthesis planning was OCSS
-(Organic Chemical Simulation of Syntheses) developed by Corey and
-Wipke in 1969 \cite{CoreyWipke1969}.
+The first system supporting chemical synthesis planning was OCSS (Organic Chemical Simulation of Syntheses) developed by Corey and Wipke in 1969 \cite{CoreyWipke1969}.
 
-In the early 1970s the collaboration between Lempel-Ziv-Welch has led
-to the LZW algorithm, since the Cheminformatician Lynch was discussing
-chemical fragment screening approaches with the computer scientists
-Lempel and Ziv \cite{brown2009}.
+In the early 1970s the collaboration between Lempel-Ziv-Welch has led to the LZW algorithm, since the Cheminformatician Lynch was
+discussing chemical fragment screening approaches with the computer scientists Lempel and Ziv \cite{brown2009}.
 
-In 1978 Gasteiger/Marsili published a fast algorithm to calculate
-partial charges in organic molecules by a Partial Equalization of
-Orbital Electronegativities (PEOE) \cite{gm78}, which is till today
-one of the gold-standards to calculate partial charges of atoms.
+In 1978 Gasteiger/Marsili published a fast algorithm to calculate partial charges in organic molecules by a Partial Equalization of Orbital 
+Electronegativities (PEOE) \cite{gm78}, which is till today one of the gold-standards to calculate partial charges of atoms.
 
-Instead of converting molecular graphs to a vector representation for
-applying machine learning methods, it is also possible to use direct
-molecular graph mining methods \cite{okada2006}. Some of the first
-published methods use neural networks \cite{kireev1995}, inductive
-logic programming \cite{yh02a}, or graph kernels \cite{kti03}.
+Instead of converting molecular graphs to a vector representation for applying machine learning methods, it is also possible to use direct 
+molecular graph mining methods \cite{okada2006}. Some of the first published methods use neural networks \cite{kireev1995}, 
+inductive logic programming \cite{yh02a}, or graph kernels \cite{kti03}.
 
-Molecular docking of chemical molecules into proteins is an important
-application in molecular modeling \cite{Leach2001}. It is noteworthy
-that in some cases molecular shape matching methods alone
-(cheminformatics), can outperform energy-based molecular docking
-algorithms (molecular modeling) \cite{hsn07}.
+Molecular docking of chemical molecules into proteins is an important application in molecular modeling \cite{Leach2001}. It is noteworthy that in some cases
+molecular shape matching methods alone (cheminformatics), can outperform energy-based molecular docking algorithms (molecular modeling) \cite{hsn07}.
 

### 2011-11-08 paper.tex
put in submitted version of the paper from Aaron

--- 2011-08-01 paper.tex
@@
 \documentclass{sig-alternate}
+
+\usepackage{array}
+\usepackage{pifont}
+\usepackage{url}
+\usepackage{graphicx}
+\usepackage{multirow}
+
+\newcommand{\none}{\ding{55}}
+\newcommand{\least}{\ding{51}}
+\newcommand{\little}{\ding{51}\ding{51}}
+\newcommand{\lots}{\ding{51}\ding{51}\ding{51}}
 
@@
 \additionalauthors{Additional authors:
-Andreas Bender (University of Cambridge, email: {\texttt{bender.andreas@gmail.com}}),
+Andreas Bender (University of Cambridge, email: {\texttt{andreas.bender@cantab.net}}),
 Jean-Loup Faulon (University of Evry, email: {\texttt{Jean-Loup.Faulon@issb.genopole.fr}}),
@@
 One of the most prominent success stories in all the sciences over the last decade has
-  been the advance of bioinformatics: interdisciplinary collaboration between
- computer scientists and molecular biologists led to the sequencing of the human genome
-  and many other accomplishments, since researchers connected biological
-  concepts to the theory of string algorithms. However, despite this great
+  been the advance of bioinformatics: The interdisciplinary collaboration between
+ computer scientists and molecular biologists led as one of its major accomplishments to the sequencing of the human
+ genome, where researchers connected biological concepts to the theory of string algorithms. However, despite this
+ great
   success, few computer scientists are familiar with a related (and older!)
-  discipline---cheminformatics, the use of computers to represent
+  discipline: cheminformatics, the use of computers to represent
  the structures of small molecules and analyze their properties.
-  Cheminformatics has a wide applicability,
-  from the discovery of novel drug candidates to the optimization of
+  Cheminformatics has wide applicability, from the discovery of novel drug candidates to the optimization of
   certain physicochemical properties of small molecules. Until
- recently, the data and techniques of cheminformatics have been closely guarded
+ recently, the data and techniques employed in the cheminformatics field have been closely guarded
   secrets of companies whose financial success depended on being the first to produce
-  the new ``miracle molecule.'' Only within the last decade---and largely because of
- chemists volunteering their time for an Open Science ``movement''---have researchers
-  gained access to freely available software packages and databases of tens of
+  the new ``miracle molecule.'' Only within the last decade---and as an effect of a change in mindsets,
+  government-mandated policy changes and last but not least because of  chemists volunteering their time for an Open
+  Science ``movement''---have researchers gained access to freely available software packages and databases of tens of
   millions of chemicals. Academic chemists now confront a variety of unsolved
@@
   solutions are critical to research ranging from determining the behavior of small
-  molecules in biological pathways, to finding a cure for malaria.
+  molecules in biological pathways, to finding a cure for malaria and other ``neglected'' diseases.
 \end{abstract}
 
-\category{H.4}{Information Systems Applications}{Miscellaneous}
-\terms{
-\begin{itemize}
-\item \textit{Tautomerism} - Tautomers are isomers (same molecular formula but different structural formula) of organic compounds that readily interconvert by a chemical reaction called tautomerization.
-\item \textit{Protonation and protonation state} - In chemistry, protonation is the addition of a proton (H$^+$) to an atom, molecule, or ion.
-The different states (different number of H$^+$ or on different atoms) are called protonation states for a molecular compound (e.g.
-de-protonated, mono-protonated, di-protonated, or any variations of those states).
-\item \textit{Bioassay} - A system measuring the biological activity of a chemical compound on a particular biological assay (with all its parameters).
-\item \textit{Metabolites} - Metabolites are the intermediates and products of metabolism. The term metabolite is usually restricted to small molecules (chemical compounds).
-\item \textit{Mass spectrometry} - Mass spectrometry (MS) is an analytical technique that measures the mass-to-charge ratio of charged particles.
-It is used for determining masses of particles, for determining the elemental composition of a sample or molecule, and for elucidating the chemical structures of molecules, such as peptides and other chemical compounds.
-\item \textit{Ligand (coordination chemistry)} - an atom, ion, or functional group that donates one or more of its electrons through a coordinate covalent bond to one or more central atoms or ions.
-\item \textit{Ligand (biochemistry)} - a substance that binds to a protein.
-\item \textit{Receptor (biochemistry)} - in biochemistry, a protein molecule that receives and responds to a neurotransmitter, or other substance.
-\item \textit{Proteochemometric modeling} - combining data mining for chemical compounds (ligands) and proteins within the same machine learning framework.
-\item \textit{Biodegradation} - is the chemical dissolution of materials by bacteria or other biological means.
-\item \textit{Reaction catalysis} - Catalysis is the change in rate of a chemical reaction due to the participation of a substance called a catalyst.
-\item \textit{Conformation and conformational isomerism} - In chemistry, conformational isomerism is a form of stereoisomerism in which the isomers can be interconverted exclusively by rotations about formally single bonds.
-Single possible states of compounds are called conformers.
-\item \textit{SMARTS} - SMiles ARbitrary Target Specification \\(SMARTS) is a language for specifying substructure patterns in molecules
-\end{itemize}
-}
-\keywords{cheminformatics, graphs, chemistry, databases, knowledge management, open source, open data}
+\category{J.2}{Computer Applications}{Physical Sciences and Engineering}[Chemistry]
+\terms{Algorithms, Design, Human Factors, Theory}
+\keywords{cheminformatics, chemoinformatics, graphs, chemistry, databases, knowledge management, open source, open
+data}
 
@@
 However, while biological information has been publicly available for decades,
-the same has not been true for chemical information until very recently. It was
+the same has not been true for chemical information, until recently. It was
 not until 2004 that a large public small molecule structure repository (PubChem)
@@
 drugs we take are the result of intensive and costly research,
-which is facilitated by the extent of information available. The
+which is facilitated by the amount of information available. The
 more information is shared (a process very hard to achieve
@@
 novel treatments. The relevance of computer
-science in this research becomes clear in considering the quantity of data available
-currently---74 million different molecules are currently stored in
+science to this research becomes clear when we consider the quantity of data available---
+currently 74 million different molecules are stored in
 public databases (such as PubChem), together with an even larger number of annotations.
@@
 bioactivity prediction, which might involve algorithms based on graphs).
-However, a major difference from bioinformatics lies in the type of
-information being analyzed: While bioinformatics often deals with sequences, the
+However, the type of information being analyzed is different from biological information: while bioinformatics often deals with sequences, the
 domain of cheminformatics is chemical structures. In the case of the former,
 information can very frequently be represented as one-dimensional strings, which
-are relatively easy to handle computationally. In the case of the latter, chemical
+are relatively easy to handle computationally. In the case of the latter however, chemical
 structures may possess rings, branches as well as multiple valid representations
-of the same structure, potentially giving rise to ambiguities. Hence,
-chemical structure is often reported to be more difficult to standardize.
+of the same structure (such as so-called ``tautomers'', where hydrogen atoms can be positioned in different places of the
+molecules), potentially giving rise to ambiguities. Hence,
+chemical structures are often reported to be more difficult to standardize.
 
 To illustrate the difficulty of standardizing chemical structures, different
-representations of the same chemical structure are shown in Figure 1. The top
-left hand corner shows a structure a chemist would draw it---every `corner'
+representations of the same chemical structure are shown in Figure~\ref{figure:chemical-structures}. The top
+left hand corner shows a structure a chemist would draw it---every ``corner''
 representing a carbon atom (which can also be represented as a C), every O
@@
 be collapsed into one node, and sets of atoms might have the same relevant
-properties assigned. Hence, the top right hand corner of Figure 1 shows an
+properties assigned. Hence, the top right hand corner of Figure~\ref{figure:chemical-structures} shows an
 annotated graph representation, taking into account that atoms, though being of
 different elemental atom type, can have similar properties---\emph{e.g.}, both oxygen
-and nitrogen carry an `On' label in this graph.
+and nitrogen carry an ``On'' label in this graph.
 
-% \begin{figure}
-% \centering
-% \psfig{file=placeholder.ps}
-% \caption{The many different facets of representing chemical structures. Shown
-% are a chemist’s usually representation of a molecule (top left) and a
-% graph-based representation amenable to data mining (top right). These
-% representations are supplemented by a 3D representation of the same molecule
-% (bottom left) that is closer to reality and a representation using surface
-% properties (bottom right), which is often more relevant for the properties the
-% molecule exhibits in an experimental setup (such as when measuring its
-% solubility, or bioactivity).}
-% \end{figure}
+\begin{figure}
+\centering
+\includegraphics[height=2in]{chemical-structures.png}
+\caption{The many different facets of representing chemical structures. Shown
+are a chemist's usual representation of a molecule (top left) and a
+graph-based representation amenable to data mining (top right). These
+representations are supplemented by a 3D representation of the same molecule
+(bottom left) that is closer to reality and a representation using surface
+properties (bottom right), which is often more relevant for the properties the
+molecule exhibits in an experimental setup (such as when measuring its
+solubility, or bioactivity).}
+\label{figure:chemical-structures}
+\end{figure}
 
@@
 rotational states, nor different charge states are encoded) and thus
-often other `descriptors' need to be used to capture the essentials of
-a chemical system \cite{bender2004}.
+often other ``descriptors'' need to be used to capture the essentials of
+a chemical system \cite{bender2004}. Hence, one of the difficulties of a cheminformatician is that the equivalence of
+particular atoms (or sets of atoms) is usually not known from the onset but it rather depends on the particular system
+considered---involving a large amount of trial-and-error in the choice of a ``proper'' description for molecules in a
+particular case.
 
@@
 
-\section{Bridging Cheminformatics \& Computer Science}
+\section{Bridging Cheminformatics and Computer Science}
 
@@
 
-While cheminformatics has a broad range of applicability, ranging from
+While cheminformatics has a broad applicability, ranging from
 agrochemical research and drug discovery to the design of novel
-materials, we have chosen to use drug discovery as the context of this
-article. The remainder of this article is structured around the
-concept of ``risk minimization'' in drug discovery---how does one
-minimize the chances of a small molecule failing as a drug candidate
-and how do cheminformatics methods play a key role in these efforts.
-
+materials (since everything around us consists of ``molecules''), we have chosen to use drug discovery as the context of
+this
+article due to its central relevance for human wellbeing. Given that the safety of drugs on the market is a key concern
+both for the general audience as well as researchers in pharmaceutical companies, the remainder of this article is
+structured around the
+concept of ``risk minimization'' in drug discovery---the question of how to minimize the chances of a small molecule
+failing in the various research and development stages as a drug candidate, and how to successfully employ
+cheminformatics methods in these efforts.
+Table~\ref{table:properties} (in the Appendix) summarizes properties a drug candidate must satisfy to be considered ``promising,'' and sketches the role of cheminformatics at each stage of drug development.
+%
 \subsection{Databases \& Ontologies}
 \label{sec:databases}
-
-Most cheminformatics applications rely on large databases of chemical structures
-and their properties. Organizing and maintaining such databases, searching based
-on features of chemical structures, and clustering similar structures together,
-are essential utilities to enable many scientific applications. Yet, each one of
+%
+Most cheminformatics applications rely on large databases of chemical structures,
+their properties, and relationships to, \textit{e.g.}, biological targets.
+Organizing and maintaining such databases, as well as searching and clustering similar structures together,
+are essential  to enable many scientific applications. Yet, each one of
 those areas poses its unique computer science challenges.
@@
 There is considerable progress in open chemical and bioactivity databases like
-ChEMBL and NCBI-PubChem, which are both freely available. Still, the integration
+ChEMBL and NCBI-PubChem, which are both freely available and contain in the order of millions (ChEMBL) and tens of
+millions (PubChem) of data points each. Still, the integration
 of chemical databases with each other remains a challenge, not only due to the
 normalization problems in the chemical and bioactivity landscape, but also due
-to sheer data volume. As mentioned earlier, we have to be able to
-search in complex data, for example multi-labelled molecular graphs, where
-graph labels can be different. Still, we need to be able to capture
-molecular graphs in a risk-bounded machine encoded way.
+to sheer data volume. One problem is the annotation with the ``protein targets'' a molecule binds to---here dozens of
+protein and gene identifiers exist, generally without the availability of a one-to-one mapping from each ontology, which makes merging databases often a cumbersome exercise.
 
-Chemical graphs can differ based on the manner in which the structures
-are drawn. Also chemical behaviors can result in multiple encodings of
-``the same'' molecule and its tautomeric states, different protonation
-states, or many variations of implicit chemical expert
-knowledge. Checking whether two graphs represent the same chemical is
+Given the requirement to ``make sense'' of as much chemical data at hand as possible, we have to be able to search in
+complex data, for example multi-labelled molecular graphs, where graph labels can be different. Hence, we need to be
+able to capture molecular graphs in a risk-bounded machine encoded way.
+
+One main challenge in the cheminformatics field is that not one unique way exists to represent every molecule in a
+machine-readable form. This results in multiple encodings of
+``the same'' molecule and its different tautomeric or protonation states (see the Glossary in the Appendix for an explanation of technical chemistry terms).
+Checking whether two graphs represent the same chemical structure is
 an expensive operation (the next section will discuss this in-depth);
 thus, a representation of the structure that is unique and invariant
-with respect to how the structure is drawn is needed for
+with respect to a particular graphical display of that chemical is needed for
 identification. An early effort in this direction was the Simplified
 Molecular Input Line Specification (SMILES), and specifically the
-generation of a single canonical SMILES for a particular structure. However,
+generation of a single, ``canonical'' SMILES for a particular structure. The idea was to devise rules for a unique
+ordering of the atoms in a molecular structure, hence resulting in the same representation of the same molecule.
+However,
 over the course of time, several
 different SMILES implementations became available, with no universal
-standard for generating a canonical SMILES.
-%After years of research and various
-%commercial products, there is today no single public standard for a
-%line-notation of molecular labeled graphs, which is available to a
-%broad audience (academic and
-%commercial). %The very same problem does also exist for
-%bioassay ontologies and the lack of options to normalize the property spaces of
-%chemical compounds, e.g. biological assays or clinical data outcome
-%(biobanking).
+standard for generating a canonical SMILES. Nonetheless, the SMILES format remains one of the major representations of
+molecules used in practice today.
 
@@
 automated exchange of data between different databases was hindered.
-As more and more data became available online,
-structure-based identification became a pressing need, and the IUPAC
-International Chemical Identifier (InChI) code was developed. The
+As more and more data became available online, the use of unique
+structure-based identifiers became a pressing need, resulting in the IUPAC
+International Chemical Identifier (InChI) to be developed. The
 InChI is a non-proprietary, structured textual identifier for chemical
-entities \cite{inchi}. InChI identifiers are not intended to be read and
+entities~\cite{inchi}. InChI identifiers are not intended to be read and
 understood by humans, but are useful for computational matching of
@@
 hashed key for the InChI, which has an invariant length of 14
-characters. Figure \ref{smiles} illustrates the SMILES, InChI and InChIKey for
+characters. (Unfortunately, after implementation and publication of the hashing scheme it became apparent that in
+particular in case of stereochemically related molecules clashes of the InChI key take place in some, albeit rare
+cases.) Figure~\ref{figure:smiles} illustrates the SMILES, InChI and InChIKey for
 lipoic acid.
 
-%Figure {smiles}: The standard chemical graph illustration together with the SMILES,
-%InChI and InChIKey codes for lipoic acid. Note that the SMILES is much more
-%human-readable than the others.
+\begin{figure}
+\centering
+\includegraphics[height=1in]{lipoicacid.png}
+\caption{The standard chemical graph illustration together with the SMILES, InChI and InChIKey codes for lipoic acid. Note that the SMILES is much more human-readable than the others.}
+\label{figure:smiles}
+\end{figure}
 
@@
 is still limited. For example, it cannot differentiate between
-certain types of stereoisomers: Figure \ref{cistrans} illustrates two stereoisomers for which
+certain types of stereoisomers: Figure~\ref{figure:cistrans} (in the Appendix) illustrates two stereoisomers for which
 the generated InChI is the same.
 
-%Figure {cistrans}: The generated InChI is the same for the stereoisomers cisplatin and
-%transplatin.
-
 InChI or other identity-mapping algorithms allow for exact searching (within the
-limitations of the algorithms), but two other requirements for scientific
+limitations of the algorithms). However, two other practically relevant algorithms for scientific
 discovery based on chemical databases are \emph{substructure searching} and \emph{similarity
-searching}. In substructure searching, the database is searched for a specified
-wholly contained part of the structures in the database; in similarity
-searching, the database is searched for structures similar to a provided search
+searching}, which are required to generalize from the data points contained in a particular database to other, related
+molecules. In substructure searching, the database is searched for a specified
+wholly contained part of the structures in the database, while in similarity
+searching, the database is searched for structures similar (in some property space) to a provided search, or query,
 structure. Chemical search packages are often implemented and optimized for a
-given database technology, for example, the OrChem package is an open source
+given database technology; for example, the OrChem package is an open source
 chemical search package for the Oracle database \cite{rijnbeek2009}.
 
-Graph substructure matching is a variant of \emph{graph isomorphism detection},which is
-widely believed to be computationally intractable [3]. To execute a graph isomorphism
-search across a full chemical database of thousands or even millions of
+Graph substructure matching is a variant of \emph{graph isomorphism detection}, which is widely believed to be computationally
+intractable~\cite{cordella2001}. To execute a graph isomorphism search
+across a full chemical database of thousands or even millions of
 structures is infeasible. For that reason, complex heuristics must be
 used to drastically filter the candidate search space. Structure-based
-\emph{fingerprints} are commonly used for this purpose. Fingerprints encode
-characteristic features of a given chemical structure in a Boolean array, or
-bitmap. The fingerprint is created by an algorithm which generates bitmap
-patterns for each atom in the structure; then each atom and its nearest
-neighbors, including the bonds between them; then each group of atoms connected
-by paths of increasing lengths up to the maximum implemented for the algorithm
-(commonly around eight). The bitmap patterns are added together into the final
-fingerprint. If all bits in a query fingerprint are also present in the target
-fingerprint of a stored database structure, only then is the structure subjected
-to the computationally expensive subgraph matching algorithm. Bit operations are
-very fast and independent of the number of atoms in a structure due to the fixed
-length of the fingerprint.
+\emph{fingerprints} are commonly used for this purpose. Fingerprints
+encode characteristic features of a given chemical structure in a
+Boolean array, or bitmap, and they can be implemented in a variety of
+ways. Common fingerprints are created by an algorithm which generates
+bitmap patterns for each atom in the structure; then each atom and its
+nearest neighbors, including the bonds between them; then each group
+of atoms connected by paths of increasing lengths up to the maximum
+implemented for the algorithm (commonly around eight). In this way,
+characteristic parts of molecule are captured and the bitmap patterns
+are added together into the final fingerprint. If all bits in a query
+fingerprint are also present in the target fingerprint of a stored
+database structure, only then is the structure subjected to the
+computationally expensive subgraph matching algorithm. Bit operations
+are very fast and independent of the number of atoms in a structure
+due to the fixed length of the fingerprint.
 
-Fingerprints also facilitate the rapid calculation of quantitative
-structure-based similarity measures. An example measure of similarity is the
-\emph{Tanimoto coefficient}, calculated as the ratio $T(a,b) = \frac{c}{(a+b-c)}$,
-where $c$ is the count of bits on in the same position in both the two
-fingerprints, $a$ is the count of bits on in the first structure, and $b$ is the
-count of bits on in the second structure. The Tanimoto coefficient varies in the
-range 0.0 -- 1.0, with a score of 1.0 indicating that the two structures are very
-similar (i.e. their fingerprints are the same). Structure-based similarity is of
+%Fingerprints also facilitate the rapid calculation of quantitative
+%structure-based similarity measures. An example measure of similarity is the
+%\emph{Tanimoto coefficient}, calculated as the ratio $T(a,b) = \frac{c}{(a+b-c)}$,
+%where $c$ is the count of bits on in the same position in both the two
+%fingerprints, $a$ is the count of bits on in the first structure, and $b$ is the
+%count of bits on in the second structure. Historically, this coefficient originated from the comparison of ecosystems
+%around lakes, where a measure was needed to describe how similar those ecosystems are with respect to the animals,
+%plants etc. they contain. However, it can be applied to every other system as well where features can be described in a
+%binary manner. The Tanimoto coefficient varies in the
+%range 0.0 -- 1.0, with a score of 1.0 indicating that the two structures are either very
+%similar or even identical (i.e. their fingerprints are the same).
+%
+Structure-based similarity is however of
 limited use in grouping of chemicals based on non-structural features, such as
@@
 encoded, making the representation both compact and powerful.
-Ontologies are already in widespread use across the biomedical domain. A
+Ontologies are already in widespread use across the biomedical domain, and in the cheminformatics field ontologies
+relating to the chemical structures themselves and the protein targets the molecules are active against are probably
+most common. A
 similarity measure based on the information encoded in ontologies is termed
@@
 Enumerating molecules is a combinatorial problem that has fascinated
-chemists, computer scientists and mathematicians alike for more than a
+chemists, computer scientists and mathematicians alike, for more than a
 century. It is one important risk-reduction element in still being
@@
 principles in graph theory and combinatorics were developed by Cayley,
-Polya and others. Even terms that are widely-used nowadays like graph
-and tree were originally coined in a chemistry context (Chapters 1 and
+Polya and others. Even terms that are widely-used nowadays, like ``graph''
+and ``tree'', were originally coined in a chemistry context (Chapters 1 and
 8 in \cite{faulon2010}). More than four decades ago, Joshua Lederberg,
@@
 intelligence textbooks as a pioneer project that produced the first
-expert system.
+expert system~\cite{DENDRAL}.
 
@@
 practical utility as well. The foremost application of enumeration is \emph{structure
-elucidation} of natural compounds such as metabolites.
+elucidation} of natural compounds such as metabolites. Metabolites are small molecules that can be found an organism
+(such as a human) and which in many cases indicate the state the organism is in---such as indicating a disease, or a
+fasted state, etc.---and with a sufficiently large database the idea is to be able to detect the chemical composition of
+body fluids and to arrive at conclusions regarding the particular state of the organism at this point in time.
 Ideally, the wishful bench chemist collects experimental data
@@
 solution, there are commercial software products such as MOLGEN
-(http://molgen.de) that can, for instance, list all structures matching a given
+(\url{http://molgen.de}) that can, for instance, list all structures matching a given
 molecular formula, or mass spectrometry (MS) data. Another important application
@@
 elucidation, molecular design has introduced some novel solutions to molecular
-enumeration. Finally, with the advent of combinatorial chemistry, molecular
+enumeration. Finally, with the advent of \emph{combinatorial chemistry} (the rapid synthesis of different but related molecules), molecular
 enumeration has taken a central role as it allows computational chemists to
@@
 
-The major difficulty with enumeration is that the \emph{in silico}
+The major difficulty with enumeration (apart from the size of the search space, which is thought to be in the order of
+$10^{60}$ molecules for 30 heavy atoms) is that the \emph{in silico}
 representation of molecules, the so-called molecular graphs, are
@@
 this problem is to consider orbits of labeled molecular graphs under
-the operation of the symmetric group. When enumerating molecular
-structures, one has thus to solve the so--called isomorphism problem.
+the operation of the symmetric group. However, when enumerating molecular
+structures, one has thus to solve the so--called isomorphism problem, where molecules that seem different are actually
+equivalent (i.e., their atoms can be mapped on top of each other by keeping the connectivity between atoms intact).
 The good news is that isomorphism can be solved in polynomial time for
@@
 same restriction prevents applying to a molecular graph the polynomial
-delay enumeration results obtained for general graphs (Chapter 8 in
-\cite{faulon2010}). While the complexity of molecular graph
+delay enumeration results obtained for general graphs (Chapter 8 in~\cite{faulon2010}). While the complexity of molecular graph
 enumeration remains an open problem it has nonetheless been shown that
-molecules can be sampled efficiently \cite{goldberg1999}. Sampling
+molecules can be sampled efficiently~\cite{goldberg1999}. Sampling
 procedures based on the Metropolis or the Genetic Algorithms have been
@@
 comprising almost a trillion chemical
-structures\footnote{http://www.dcb-server.unibe.ch/groups/reymond/gdb/home.html}. Yet,
+structures~\cite{GDB}. Yet,
 the current enumeration software products do not generally produce
@@
 procedures, which are still the subject of investigations by the
-cheminformatics community.
+cheminformatics community. In addition, GDB-13 only enumerates molecules up to 13 heavy atoms---given the possibly
+exponential increase in search space with the number of atoms, it seems unlikely that the practically relevant chemical
+space of about 30 heavy atoms will be exhaustively enumerated in the near future.
 
-GDB-13 along with other databases such as PubChem (NCBI/NIH) or ZINC
-(UCSF) are valuable resources to search for new drugs. For instance
-the well known ligand-based design approach uses these databases along
-with structure similarity tools to discover new compounds binding to
-biological targets (protein or nucleic acid). Compounds retrieved
-through ligand-based design, however, must already be present in
-databases, which is a limitation of the technique. To overcome that
-limitation, algorithms have recently been developed to enumerate all
-compounds in the chemical space matching circular \cite{faulon2003} or
-path fingerprints \cite{fujiwara2008}, these fingerprints being
-popular tools to measure similarity. Enumerating molecules from
-fingerprints, or more generally enumerating molecules from molecular
-descriptors, is still very much under development. At stake is the
-design of new chemicals or drugs that have not yet been synthesized.
-
-Another area where enumeration plays a role is in the \emph{generation
+%GDB-13 along with other databases such as PubChem (NCBI/NIH) or ZINC
+%(UCSF) are valuable resources to search for new drugs. For instance
+%the well known ligand-based design approach uses these databases along
+%with structure similarity tools to discover new compounds binding to
+%biological targets (protein or nucleic acid). Compounds retrieved
+%through ligand-based design, however, must already be present in
+%databases, which is a limitation of the technique. To overcome that
+%limitation, algorithms have recently been developed to enumerate all
+%compounds in the chemical space matching circular \cite{faulon2003} or
+%path fingerprints \cite{fujiwara2008}, these fingerprints being
+%%popular tools to measure similarity. Enumerating molecules from
+%fingerprints, or more generally enumerating molecules from molecular
+%descriptors, is still very much under development. At stake is the
+%design of new chemicals or drugs that have not yet been synthesized.
+%
+Apart from analytical chemistry and compound design, enumeration plays a role is in the \emph{generation
   of chemical reaction networks}. The problem here consists of
@@
 the lowest possible risk, and also to describe the likelihood of error
-(confidence) when using such models for decision making. In many
-cases, the activity of a small molecule is due to its interaction with
-a receptor. Traditional QSAR approaches ignore receptor interactions,
+(confidence) when using such models for decision making.
+Figure~\ref{figure:applicability-domain} (in the Appendix) illustrates the use of substructure matching to produce a space of chemical structures that are likely to share similar biological activity; this is called the generation of an \emph{applicability domain}.
+
+In many cases, the activity of a small molecule is due to its interaction with
+a receptor. Traditional ``QSAR'' approaches ignore receptor interactions,
 focusing only on small molecule features, and therefore loose valuable
@@
 The first step in the predictive modeling of biological activities is
-to compare compounds. Either we transform molecular graphs into a
+to describe compounds in a way amenable to computerized data mining, while at the same time capturing information
+relevant for the property one attempts to model. Here multiple approaches exist: Either we transform molecular graphs
+into a
 numerical vector representation (a.k.a. \emph{descriptors} or
@@
 processing time and space. Note that the public PubChem dataset alone
-contains over 30 million chemical compounds and over 70 million other
-substances.
+contains over 30 million chemical compounds and over 70 million substances.
 
-First, we must consider the goal of a predictive model in a
-cheminformatics setting. In some cases, pure predictive ability is
-desired---such as in a virtual screening setting. For such cases, the
+In order to judge which predictive modeling approach is applicable in every case, we firstly have to consider the goal
+of a the model in a cheminformatics setting. In some cases, pure predictive ability is
+desired---such as in a virtual screening setting. (Basically, only the reliability of the model matters, but other
+aspects such as interpretability are of minor importance.) For such cases, the
 complexity of the modeling approach is immaterial, as it leads to
@@
 \emph{in silico} compounds into real-world compounds. In such cases,
-black box methods can be less useful, though attempts have been made
+``black-box methods'' can be less useful, though attempts have been made
 to interpret vectorial models built using neural networks, Support Vector Machines or
-random forests. Related to this, there is a never-ending debate on the
-choice of descriptors, \emph{e.g.}, by applying feature selection.
+random forests as well.
 
-Second, the concept of model applicability---when is the prediction
-for a new object reliable (confidence)?---has grown in importance with
+Secondly, the concept of model applicability---the question of when the prediction of the model
+for a new object is reliable---has grown in importance with
 the increasing use of predictive models in regulatory settings. A
@@
 proteochemometric approach as first reported by Lapinsh
-\textit{et al.} \cite{lapinsh2001}.
+\textit{et al.}~\cite{lapinsh2001}.
 
@@
 approach should lead to more robust predictions for panel assays. Finally,
-this also better reflects clinically relevant compound profiles \cite{kuhn2010}
-and a `personalized medicine' concept (e.g. drug-drug interaction profiles) [The
-Decision Tree].
+this also better reflects clinically relevant compound profiles~\cite{kuhn2010}
+and a ``personalized medicine'' concept (\emph{e.g.}, drug-drug interaction profiles).
 
@@
 scientists already switched many years ago from paper lab-notebooks to
-online collaboration platforms, called ELNs (Electronic Lab Notebooks). So, why
-have chemists and companies already adopted ``Enterprise 2.0,'' a social
-online collaboration culture? Many other disciplines are still busy with solving
-ROI (return-of-investment), risk reduction, opportunity creation, change
-culture, and technical collaboration challenges for scientists and scientific
-data.
+online collaboration platforms, called ELNs (Electronic Lab Notebooks). So, what were the drivers behind chemists, and
+companies, adopting ``Enterprise 2.0,'' a social online collaboration culture?
 
@@
 lab-notebooks, multiple compound registration, and search tools. Compared to
-today there were many pitfalls and drawbacks with this. The overall architecture
+today there were many pitfalls and drawbacks associated with this. The overall architecture
 was too inflexible to adopt to fast changing data standards, scientists spent
 too much time with administrative work, chemical data quality was inconsistent,
-an alignment with other working groups was inefficient (e.g. for running
+an alignment with other working groups was inefficient (\emph{e.g.}, for running
 analytical experiments), a legal dispute required to find data in paper
@@
 delays, and local silos hampered an efficient global collaboration. Now,
-chemists prefer ELNs, because they can run searches in the data of colleagues. This does
+chemists prefer ELNs, because they can run searches in the data across the organization. This does
 not only allow to find experts faster, but also allows access to highly trusted
@@
 ELNs are typically much more accepted and used within the cheminformatics
-domain than in the bioinformatics domain, which is picking up speed lately. The
-question is why? In chemistry, every experiment requires a lot of expert
+domain than in the bioinformatics domain, which is picking up speed lately, the
+question being why this is the case. In chemistry, every experiment requires a lot of expert
 knowledge. Therefore, approaches for reducing risks by communicating
@@
 growth, usage, and collaboration scenarios between private and public parties.
-Only when the technical and scientific questions are solved can we ensure that
-legal questions about contribution metrics will get enough room to grow.
+In a business context we need to address ROI (return-of-investment) questions, not only for single
+institutions and companies, but also for collaboration networks of legal partners.
+So, only when the technical and scientific questions are solved can we ensure that
+legal questions about contribution metrics will get enough room to grow. Solving challenges is one thing (science and
+technical),
+ensuring that the contributing partners get rewarded by their contribution another (legal).
 
@@
 activity data types. On the other hand, there is a distinct lack of
-textual data (such as journal articles) that are Open in nature. While
+textual data (such as journal articles) that are ``Open'' in nature. While
 PubMed abstracts serve as a proxy for journal articles, text mining
-methods in cheminformatics are hindered by this state of affairs.
+methods in cheminformatics are hindered by not being able to mine the full text of many scientific publications.
 Interestingly, patent information is publicly accessible (cf. Google
@@
 contribution to this has been the Blue Obelisk group
-(http://blueobelisk.org) of chemists, who have worked together to
+(\url{http://blueobelisk.org}) of chemists, who have worked together to
 create interoperable open-source chemistry software (among other
@@
 including the Chemistry Development
-Kit (CDK) \cite{steinbeck2003}, Open Babel, RDKit and Indigo. These toolkits are written in
+Kit (CDK)~\cite{steinbeck2003}, Open Babel, RDKit and Indigo. These toolkits are written in
 Java or C++ (though toolkits written in the latter invariably have
@@
 for example, the molecular viewer Avogadro
-(http://avogadro.openmolecules.net) uses Open Babel, while the
+(\url{http://avogadro.openmolecules.net}) uses Open Babel, while the
 molecular workbench Bioclipse \cite{Bioclipse2} uses the CDK.
@@
 optimization and supports the interconversion of a huge number of
-chemical file formats.
+chemical file formats.  Table~\ref{tab:features} (in the Appendix) compares features offered by the open-source toolkits currently available.
 
@@
 on the mailing list. Anyone interested should email the respective
-developers' mailing list (either cdk-devel@lists.sf.net or
-openbabel-devel@lists.sf.net) with their interests and background, and
+developers' mailing list (either \texttt{cdk-devel@lists.sf.net} or
+\texttt{openbabel-devel@lists.sf.net}) with their interests and background, and
 a mentor will be assigned to help them join the project.
+%
 
+%
 \section{Conclusions}
@@
 detailed and subtle to admit as clean an abstraction as ``algorithms
-on a string,'' the way many bioinformatics algorithms can be abstracted into a CS
+on a string,'' the way many bioinformatics algorithms can be abstracted into a computer science
 framework.  Nevertheless, because of the changing landscape, there are
@@
 algorithmically, and more computer scientists think ``chemically,''
-%we may discover ``miracle drugs'' more consistently than ever before.
 we are much better positioned taking not only ``simple'' chemicals
@@
 but also their complex relationships, transformations, and
-combinatorial challenges. This will allow us to
-normalize, search, and use all chemical information, around us and
-within us.
+combinatorial challenges.
 
-\section*{Acknowledgements}
+In 2010, drug company Glaxo made public over 13,000 compounds that they have investigated as potential cures for malaria~\cite{glaxo-malaria}.
+(Malaria has traditionally been a ``neglected'' disease in drug research, because the populations who suffer from it tend not to have enough money to be a stable consumer base for a new pharmaceutical.)  Perhaps we will now be able to find cures for some of these neglected diseases, by using the open source, open data approach to the computer science of chemical discovery.
+
+\section{Acknowledgements}
 We would like to thank Danny Verbinnen for sharing his insights into
@@
 Department of Electrical Engineering and Computer Science,
-Northwestern University, supported in part by NSF grant CCF-1049899.
+Northwestern University, supported in part by NSF grant CCF-1049899. A. Bender thanks Unilever for funding.
 
@@
 
+\appendix
+
+\section{Glossary of Terms}
+\begin{itemize}
+\item \textit{Bioassay} - A system measuring the biological activity of a chemical compound on a particular
+    biological assay (with all its parameters).
+\item \textit{Biodegradation} - is the chemical dissolution of materials by bacteria or other biological means.
+\item \textit{Conformation and conformational isomerism} - In chemistry, conformational isomerism is a form of
+    stereoisomerism in which the isomers can be interconverted exclusively by rotations about formally single
+    bonds.
+    Single possible states of compounds are called conformers.
+\item \textit{Ligand (biochemistry)} - a substance that binds to a protein.
+\item \textit{Ligand (coordination chemistry)} - an atom, ion, or functional group that donates one or more of its
+    electrons through a coordinate covalent bond to one or more central atoms or ions.
+\item \textit{Mass spectrometry} - Mass spectrometry (MS) is an analytical technique that measures the
+    mass-to-charge ratio of charged particles. It is used for determining masses of particles, for determining the
+    elemental composition of a sample or molecule, and for elucidating the chemical structures of molecules, such
+    as
+    peptides and other chemical compounds.
+\item \textit{Metabolites} - Metabolites are the intermediates and products of metabolism. The term metabolite is
+    usually restricted to small molecules (chemical compounds).
+\item \textit{Proteochemometric modeling} - combining data mining for chemical compounds (ligands) and proteins
+    within the same machine learning framework.
+\item \textit{Protonation and protonation state} - In chemistry, protonation is the addition of a proton (H$^+$) to
+    an atom, molecule, or ion. The different states (different number of H$^+$ or on different atoms) are called
+    protonation states for a molecular compound (e.g. de-protonated, mono-protonated, di-protonated, or any
+    variations of those states).
+\item \textit{QSAR} - Quantitative Structure-Activity Relationships; methods to relate the structure and the
+    biological activity of a chemical structure in a quantitative manner.
+\item \textit{Reaction catalysis} - Catalysis is the change in rate of a chemical reaction due to the participation
+    of a substance called a catalyst.
+\item \textit{Receptor (biochemistry)} - in biochemistry, a protein molecule that receives and responds to a
+    neurotransmitter, or other substance.
+\item \textit{Regulation authorities} - In the US the Food and Drug Administration (FDA) is responsible for
+    protecting and promoting public health through the regulation and supervision of e.g. dietary supplements,
+    prescription and over-the-counter pharmaceutical drugs (medications), vaccines, biopharmaceuticals, medical
+    devices, and many other health related items. The European Medicines Agency (EMA) is a European agency for the
+    evaluation of medicinal products.
+\item \textit{SMARTS} - SMiles ARbitrary Target Specification \\(SMARTS) is a language for specifying substructure
+    patterns in molecules
+\item \textit{Stereoisomer/stereoisomerism} - Stereoisomers are isomeric molecules that have the same molecular
+    formula and sequence of bonded atoms (constitution), but that differ only in the three-dimensional orientations
+    of their atoms in space.
+\item \textit{Tautomerism} - Tautomers are isomers (same molecular formula but different structural formula) of
+    organic compounds that readily interconvert by a chemical reaction called tautomerization.
+\end{itemize}
+%
+\section{Additional Figure and Tables}
+%
+\begin{table*}
+\begin{tabular}{|l|l|l|} \hline
+\textbf{Property} & \textbf{Requirement} & \textbf{Potential of Cheminformatics Approaches} \\ \hline
+\multicolumn{3}{|l|}{\textbf{Research Stage}} \\ \hline
+\multirow{3}{*}{Bioactivity} & Compound is active & \textsc{High} -- computer models can often predict \\
+& in \emph{in vitro} (``test tube'')
+studies &  which compounds are bioactive, \\
+&& the so-called \emph{virtual screening} approaches \\ \hline
+\multirow{3}{*}{Solubility} & Compound is soluble  & \textsc{High} -- computer models can often predict \\
+& at effective concentration & which compounds are insoluble, \\
+&& decreasing the need for experimental testing \\ \hline
+\multirow{3}{*}{No Obvious Toxicity} & No known ``toxicophores''  & \textsc{High} -- approaches such as substructure
+searching \\
+& contained within molecule & are routinely applied to remove
+chemical structures \\
+&& with functional groups that
+are known to be toxic \\ \hline
+\multirow{3}{*}{Selectivity} & Compound does not modulate  & \textsc{Medium} -- \emph{in silico} approaches can often make
+guesses  \\
+& other than  & which proteins are modulated by a
+molecule, but the \\
+& the intended protein
+target & precise prediction of
+bioactivity profiles remains a challenge \\ \hline
+\multicolumn{3}{|l|}{\textbf{Development Stage (All of the above, plus...)}} \\ \hline
+\multirow{3}{*}{Efficacy} & Compound is not only
+active
+& \textsc{Low} -- bioinformatics approaches can partially
+be used  \\
+ & in the test tube, but also & to anticipate whether modulation of a
+particular protein  \\
+in Animal/Human Models & in the
+organism  & shows an effect in humans via pathway modeling\\ \hline
+\multirow{3}{*}{Safety in} & Compound exhibits
+tolerable & \textsc{Low} -- bioinformatics approaches can partially
+be used \\
+& side effects in
+animal
+models & to anticipate safety concerns in
+humans, \\
+Animal/Human Models & and/or humans & however experimental confirmation is
+required \\ \hline
+\multirow{4}{*}{Bioavailability} & Compound gets
+dissolved & \textsc{Medium} -- cheminformatics models for
+bioavailability  \\
+& in the
+gastrointestinal tract & can be generated, however due
+to the complexity  \\
+& and it is absorbed into
+the  & of the process they are in
+many cases  \\
+& bloodstream & not sufficiently reliable for practical
+use \\ \hline
+\multicolumn{3}{|l|}{\textbf{Marketed Drug (All of the above, plus...)}} \\ \hline
+\multirow{3}{*}{Cost of Synthesis} & The compound can be
+ & \textsc{Medium} -- some cheminformatics approaches
+exist \\
+& synthesized at a  & that can estimate the difficulty of \\
+& reasonable cost & synthesizing a particular chemical structure \\ \hline
+\multirow{2}{*}{}Better Efficacy & The compound has & \textsc{Medium} -- the profile of the final product will
+determine  \\
+than Competitor Products & a market advantage & the particular advantage of the drug
+in the market \\ \hline
+\multirow{4}{*}{} & Patients with a disease & \textsc{Medium} -- business decisions are quite relevant \\
+Existence of  & treated by the drug exist & when drugs are developed in private pharmaceutical \\
+Relevant Market Need   & and they possess the means  &  companies (but often to a lesser extent \\
+& to buy the drug &  in government-owned pharmaceutical research laboratories) \\  \hline
+\end{tabular}
+\caption{Some of the properties potential drug candidates need to have to be considered as ``promising,'' either in the different research and development stages, as well as a drug that is finally admitted to the market.
+Note that the particular requirements are also heavily project-dependent, and that also in cases where cheminformatics approaches can be employed in drug design an experimental confirmation of the prediction is required---the computer can hence decrease the experimental burden, but not replace it altogether.}
+\label{table:properties}
+\end{table*}
+%
+\begin{table*}
+  %\onehalfspacing
+  \centering
+  \begin{tabular}{lcccc}  \hline
+    \textbf{Feature} & \textbf{CDK} & \textbf{OpenBabel} &
+    \textbf{RDKit} & \textbf{Indigo}\textsuperscript{1} \tabularnewline \hline
+    License & LGPL & GPL & new BSD & GPL or commercial \tabularnewline
+    Language & Java & C++ & C++ & C++ / Python \tabularnewline
+    SLOC\textsuperscript{2} & 188,554 & 194,358 & 173,219 & ? \tabularnewline
+    Structure-based Fingerprints & \lots\textsuperscript{3} & \lots & \lots & yes \tabularnewline
+    %\ \ \ Substructure & \lots & \lots & \lots \tabularnewline
+    Support for multiple chemical file formats & \little & \lots & \least & \little \tabularnewline
+    %Aromaticity models & \least & \least & \least \tabularnewline
+    %Stereochemistry & \least & \little & \lots \tabularnewline
+    Canonical labeling of graphs & \lots & \lots & \lots & yes \tabularnewline
+    Calculation of molecular descriptors & \lots & \least & \lots & no \tabularnewline
+    %2D coordinate generation & \lots & \none & \lots & no \tabularnewline
+    3D coordinate generation & \least & \lots & \lots & no \tabularnewline
+    2D depictions & \lots & \little & \lots & \little \tabularnewline
+    Generation of conformational isomers & \none & \least & \least & no \tabularnewline
+    %Rigid alignment & \lots & \lots & \lots \tabularnewline
+    SMARTS searching/matching & \lots & \lots & \lots & yes \tabularnewline
+    %Pharmacophore searching & \little & \none & \lots \tabularnewline
+    Chemical reaction enumeration & no & no & no & yes \tabularnewline
+    \hline
+    \multicolumn{5}{l}{1. Indigo is a new toolkit (published late 2010), and we have not investigated its features in depth.} \\
+    \multicolumn{5}{l}{2. Source Lines of Code as measured by the tool \emph{sloccount} (\url{http://www.dwheeler.com/sloccount/}).} \\
+    \multicolumn{5}{l}{The count includes all source files, in any language, for the
+  project. In addition, only the trunk} \\
+    \multicolumn{5}{l}{for each project was considered.  We did not count Indigo's lines of code.} \\
+    \multicolumn{5}{l}{3. More checks indicates more complete implementation of a feature.} \\
+  \end{tabular}
+  \caption{Cheminformatics features provided by open-source toolkits.}
+  \label{tab:features}
+\end{table*}
+%
+\begin{figure}
+\centering
+\includegraphics[height=1.3in]{platins.png}
+\caption{The generated InChI is the same for the stereoisomers cisplatin and transplatin.}
+\label{figure:cistrans}
+\end{figure}
+%
+\begin{figure}
+\centering
+\includegraphics[height=2in]{CACM_ApplicabilityDomain.pdf}
+\caption{Illustration of the ``Applicability Domain'' concept.
+In order to estimate whether model predictions are reliable, it is crucial to define areas of ``chemical space'' where the model is applicable, and where it is not.
+In this case, the ``model domain'' includes the molecule at the top.
+The (relatively similar) molecules to the left are likely to be included in the Applicability Domain of the model, while the (more dissimilar) molecules to the right are likely located outside this domain, hence the generated model is probably not applicable.}
+\label{figure:applicability-domain}
+\end{figure}
+%
 \end{document}

### 2011-11-08 supplement_history.tex
more updates to the suppl

--- 2011-11-08 supplement_history.tex
@@
 
-\title{Supplement History of Chem(o)informatics as part of Cheminformatics: The Computer Science of Chemical Discovery}
+\title{Supplement History of Cheminformatics as part of Cheminformatics: The Computer Science of Chemical Discovery}
 \numberofauthors{9}
@@
 
-\section{History of Chem(o)informatics}
-The aim of this brief historic review of cheminformatics is to put the
-content of the main article into a broader perspective. Therefore it
-might be useful to clarify some other terms and how cheminformatics
-might relate to them.  Still, Chen \cite{Chen2006} and Brown
-\cite{brown2009} cited various definitions of pioniers in the field
-and Brown concluded rightfully: '\textit{The differences in
-  definitions (of the term cheminformatics) are largely a result of
-  the types of analyses that particular scientists practice and no
-  single definition is intended to be all-encompassing}.'
+\section{History of Cheminformatics}
+The aim of this brief review of the history of cheminformatics is to
+put the content of the main article into a broader perspective. In
+addition, we also discuss some terms as they relate to
+cheminformatics.  It is thus useful to first provide some definitions
+of cheminformatics itself, given that they also shed some light on the
+founding principles of the field. For example, we have the following
+definitions:
 \begin{itemize}
-\item \textbf{Quantum chemistry} is something most chemists would
-  relate with 'theoretical chemistry'. In short, it allows to describe
-  chemical molecules on an electron density level, which allows also
-  to describe physical properties, e.g infrared spectra of a chemical
-  compound.  The results of quantum chemistry calculation are used in
-  molecular modeling (energy force field parameters) and
+\item ``\textit{The mixing of information resources to transform data
+    into information, and information into knowledge, for the intended
+    purpose of making better decisions faster in the arena of drug
+    lead identification and optimization.}'' Frank K. Brown, 1998.
+\item ``\textit{[Chemoinformatics involves]... the computer
+    manipulation of two- or threedimensional chemical structures and
+    excludes textual information. This distinguishes the term from
+    chemical information, largely a discipline of chemical librarians
+    and does not include the development of computational methods.}''
+  Peter Willett, 2002.
+\item ``\textit{\ldots the application of informatics to solve chemical
+    problems.}'' and ``\textit{\ldots chemoinformatics makes the point that
+    you're using one scientific discipline to understand another
+    scientific discipline.}'' Johann Gasteiger, 2002
+\end{itemize}
+More generally, Chen\cite{Chen2006} and Brown \cite{brown2009} cited various
+definitions of pioneers in the field with Brown concluding that
+``'\textit{The differences in definitions (of the term
+  cheminformatics) are largely a result of the types of analyses that
+  particular scientists practice and no single definition is intended
+  to be all-encompassing}'', as is evident from the three definitions above.
+
+It is also useful to clarify some other terms and how
+cheminformatics might relate to them:
+\begin{itemize}
+\item \textbf{Quantum chemistry} is generally associated with
+  theoretical chemistry or chemical physics. Briefly, it focuses on
+  the description of chemical systems starting from irst principles
+  and is based on the Schr\"{o}dinger wave equation and frameworks
+built on top of it. The use of quantum mechanics allows one to
+describe molecules in terms of electron density, which thus forms the
+basis of physically accurate calculations of physical properties
+(e.g., infrred spectra). The results of quantum chemistry calculations
+are used in molecular modeling (energy force field parameters) and
 cheminformatics (atom partial charges, or full molecule properties,
 e.g molecule polarizabilities).
-\item \textbf{Molecular modeling} or \textbf{Computational chemistry}
-  considers typically energy calculations of single molecules or
-  interactions with other molecules, e.g. proteins. 'Molecular
-  docking' belongs into this category, since it places a chemical
-  molecule in the active site of enzymes and calculates
-  energy/interaction scores between a chemical molecule and the
-  protein. Surely are there various cheminformatics concepts required
-  to perform such a task. Still, strictly speaking is cheminformatics
-  rather on the chemical molecule side than on the protein side,
-  though are both sides affecting each other. This area is typically
-  strongly related to structural biology and technologies like XRay
-  crystallography and NMR spectra structure determination.
-\item \textbf{Chemometrics} can be considered as applied
-  cheminformatics to the field of chemical/physical analysis,
-  e.g. infrared, mass and NMR spectra analysis.
-\item '\textbf{Translational research} aims to improve the health and
-  longevity of the world's populations and depends on developing
-  broad-based teams of scientists and scholars who are able to focus
-  their efforts to link basic scientific discoveries with the arena of
-  clinical investigation, and translating the results of clinical
-  trials into changes in clinical practice, informed by evidence from
-  the social and political sciences.' (Source:Wikipedia).  Since
-  clinical trials can be connected to drugs, chemical molecules, even
-  all of translational research and biobanking could be considered as
-  a field of cheminformatics, which would be on the other hand just a
-  small contribution to a full knowledge management in this field
-  \cite{Szalma2010}.
+\item \textbf{Molecular modeling} or \textbf{Computational chemistry} considers typically energy calculations of single molecules or interactions
+with other molecules, e.g. proteins. 'Molecular docking' belongs into this category, since it places a chemical molecule in the active site of enzymes
+and calculates energy/interaction scores between a chemical molecule and the protein. Surely are there various cheminformatics concepts required to
+perform such a task. Still, strictly speaking is cheminformatics rather on the chemical molecule side than on the protein side, though are both sides
+affecting each other. This area is typically strongly related to structural biology and technologies like XRay crystallography and 
+NMR spectra structure determination. 
+\item \textbf{Chemometrics} can be considered as applied cheminformatics to the field of chemical/physical analysis, e.g. infrared, mass and NMR spectra analysis.
+\item '\textbf{Translational research/medicine} aims to improve the health and longevity of the world's populations and depends on developing broad-based 
+teams of scientists and scholars who are able to focus their efforts to link basic scientific discoveries with the arena of clinical investigation, 
+and translating the results of clinical trials into changes in clinical practice, informed by evidence from the social and political sciences.' (Source:Wikipedia).
+Since clinical trials can be connected to drugs, chemical molecules, even all of translational research and biobanking could be considered as a field
+of cheminformatics, which would be on the other hand just a small contribution to a full knowledge management in this field \cite{Szalma2010}.
 \end{itemize}
-If we consider all information, analysis, and optimization of a
-chemical molecule as 'cheminformatics', then the field is very large,
-since a chemical molecule plays the central role in many related
-disciplines. Still, for keeping it practical the aim of this review is
-not being exhaustive, but rather to point the reader to more detailed
-references and to highlight some historic milestones.
+If we consider all information, analysis, and optimization of a chemical molecule as 'cheminformatics', then the field is very large, since a chemical
+molecule plays the central role in many related disciplines. Still, for keeping it practical the aim of this review is not being exhaustive, but rather to 
+point the reader to more detailed references and to highlight some historic milestones.
 
-One of the most comprehensive cheminformatic series was published by
-Prof. Gasteiger in 2003 \cite{Gasteiger2003}, one of the pioniers in
-the field.  Compacter introductory books followed from Gasteiger/Engel
-\cite{gasteigerengel2003} and Leach/Gillet \cite{leachgillet2007}.
-The books of Bajorath \cite{Bajorath2004} and Oprea \cite{oprea2005}
-focus on the applications of cheminformatics in drug design.  The
-first books discussing chemical graph theory and algorithms were
-published in 1989 by Zupan \cite{zupan1989}, 1991 by Bonchev/Rouvrey
-\cite{bonchevrouvrey1991,bonchevrouvrey2003} and 1992 by Trinastic
-\cite{Trinajstic1992}.  The most recent one was published by
-Faulon/Bender \cite{faulon2010}.  Other books focus more on specific
-fields/applications like mathematical challenges
-\cite{mathchallenges1995}, molecular diversity
-\cite{moleculardiversity1999}, factor analysis \cite{Malinowski2002},
-evolutionary algorithms \cite{clark2000}, molecular descriptors
-\cite{todeschini2000}.
+One of the most comprehensive cheminformatic series was published by Prof. Gasteiger in 2003 \cite{Gasteiger2003}, one of the pioniers in the field.
+Compacter introductory books followed from Gasteiger/Engel \cite{gasteigerengel2003} and Leach/Gillet \cite{leachgillet2007}.
+The books of Bajorath \cite{Bajorath2004} and Oprea \cite{oprea2005} focus on the applications of cheminformatics in drug design.
+The first books discussing chemical graph theory and algorithms were published in 1989 by Zupan \cite{zupan1989}, 
+1991 by Bonchev/Rouvrey \cite{bonchevrouvrey1991,bonchevrouvrey2003} and 1992 by Trinastic \cite{Trinajstic1992}. 
+The most recent one was published by Faulon/Bender \cite{faulon2010}.
+Other books focus more on specific fields/applications like mathematical challenges \cite{mathchallenges1995}, 
+molecular diversity \cite{moleculardiversity1999}, factor analysis \cite{Malinowski2002}, evolutionary algorithms \cite{clark2000},
+molecular descriptors \cite{todeschini2000}.
 
 \section{Historic milestones}
-The first chemical graphs were drawn by the Scottish chemist Willam
-Cullen in 1758 \cite{bonchevrouvrey1991}, which called them initially
-affinity graphs.  After developing the concept of bonds between atoms
-(Couper, 1858) the first chemical graphs occured in publications of
-Brown in 1864 and Cayley in 1874 \cite{bonchevrouvrey1991,brown2009}.
+The first chemical graphs were drawn by the Scottish chemist Willam Cullen in 1758 \cite{bonchevrouvrey1991}, which called them initially affinity graphs.
+After developing the concept of bonds between atoms (Couper, 1858) the first chemical graphs occured in publications 
+of Brown in 1864 and Cayley in 1874 \cite{bonchevrouvrey1991,brown2009}.
 
-The year 1946 may be regarded as the birth year of chemoinformatics
-\cite{Chen2006}: '\textit{In 1946 King et al.\cite{kct1946} published
-  an article illustrating the use of IBM's business accounting
-  machines in carrying out the construction of the rotational spectra
-  of asymmetric rotors by the evaluation of mathematical equations for
-  line position and line intensity.}'
+The year 1946 may be regarded as the birth year of chemoinformatics \cite{Chen2006}: '\textit{In 1946 King et al.\cite{kct1946} published an article illustrating 
+the use of IBM's business accounting machines in carrying out the construction of the rotational spectra of asymmetric rotors
+by the evaluation of mathematical equations for line position and line intensity.}'
 
-The first records of managing chemical information go back to chemical
-literature being indexed since the year 1771.  In 1881 the first
-edition of \textit{Beilstein's Handbuch der Organischen Chemie}
-ecyclopedia was published \cite{polanski2009} registering 1500
-chemical compounds. The short term 'chemoinformatics' was coined by
-Brown in 1998\cite{brown1998}, while the \textit{Journal of Chemical
-  Documentation} (founded in 1961) changed already its name in 1975 to
-\textit{Journal of Chemical Information and Computer Sciences} to
-reflect the tight connection between chemical information and computer
-science. The Chemical Abstract Service (CAS) of the American Chemical
-Society (ACS) formed in 1955 a research department and starting from
-1965 they provided their Chemical Registry System, and in 1968 they
-made the first computer-readable file of all abstracted documents
-available \cite{Chen2006}.
+The first records of managing chemical information go back to chemical literature being indexed since the year 1771.
+In 1881 the first edition of \textit{Beilstein's Handbuch der Organischen Chemie} ecyclopedia was 
+published \cite{polanski2009} registering 1500 chemical compounds. The short term 'chemoinformatics' was coined by Brown 
+in 1998\cite{brown1998}, while the \textit{Journal of Chemical Documentation} (founded in 1961) changed already its name in 1975 to 
+\textit{Journal of Chemical Information and Computer Sciences} to reflect the tight connection between chemical information 
+and computer science. The Chemical Abstract Service (CAS) of the American Chemical Society (ACS) formed in 1955 a research department
+and starting from 1965 they provided their Chemical Registry System, and in 1968 they made the first computer-readable file of all
+abstracted documents available \cite{Chen2006}.
 
-Ray and Kirsch published in 1957 the first substructure searching
-algorithm for allowing to find chemical records via computers
-\cite{RayKirsch1957}.
+Ray and Kirsch published in 1957 the first substructure searching algorithm for allowing to find chemical records via computers \cite{RayKirsch1957}.
 
-One of the key algorithms for molecular graph canonicalization is the
-Morgan algorithm \cite{Morgan1965}. It will search for the most
-central bond in a graph and allows then to traverse other atoms,
-resolve ties, and assigning new atom indices. Still, there are various
-options how different software implementations are handling the 'ties'
-for assigning new atom indices. So, by using the canonicalization of
-multiple software implementations and chemical expert systems can be
-even used to create randomized (canonical, only when sticking to one
-software) molecular graphs, which can be used as stochastic input for
-other algorithms \cite{cokfl06}. The central bond is also being used
-to create 3D conformational search trees \cite{Schwab2001}.
+One of the key algorithms for molecular graph canonicalization is the Morgan algorithm \cite{Morgan1965}. It will search for the most central bond
+in a graph and allows then to traverse other atoms, resolve ties, and assigning new atom indices. Still, there are various options how different 
+software implementations are handling the 'ties' for assigning new atom indices. So, by using the canonicalization of multiple software implementations and chemical
+expert systems can be even used to create randomized (canonical, only when sticking to one software) molecular graphs, which can be used as stochastic input for 
+other algorithms \cite{cokfl06}. The central bond is also being used to create 3D conformational search trees \cite{Schwab2001}.
 
-In 1965 the DENDRAL expert system started with the aim to
-automatically determine the structure of an unknown chemical compound
-from the corresponding mass spectrum \cite{Gray1986}. The system is
-also often being cited as one of the earliest artificial intelligence
-and expert systems \cite{Chen2006}.
+In 1965 the DENDRAL expert system started with the aim to automatically determine the structure of an unknown chemical compound from the 
+corresponding mass spectrum \cite{Gray1986}. The system is also often being cited as one of the earliest artificial intelligence and expert systems  \cite{Chen2006}.
 
-The first system supporting chemical synthesis planning was OCSS
-(Organic Chemical Simulation of Syntheses) developed by Corey and
-Wipke in 1969 \cite{CoreyWipke1969}.
+The first system supporting chemical synthesis planning was OCSS (Organic Chemical Simulation of Syntheses) developed by Corey and Wipke in 1969 \cite{CoreyWipke1969}.
 
-In the early 1970s the collaboration between Lempel-Ziv-Welch has led
-to the LZW algorithm, since the Cheminformatician Lynch was discussing
-chemical fragment screening approaches with the computer scientists
-Lempel and Ziv \cite{brown2009}.
+In the early 1970s the collaboration between Lempel-Ziv-Welch has led to the LZW algorithm, since the Cheminformatician Lynch was
+discussing chemical fragment screening approaches with the computer scientists Lempel and Ziv \cite{brown2009}.
 
-In 1978 Gasteiger/Marsili published a fast algorithm to calculate
-partial charges in organic molecules by a Partial Equalization of
-Orbital Electronegativities (PEOE) \cite{gm78}, which is till today
-one of the gold-standards to calculate partial charges of atoms.
+In 1978 Gasteiger/Marsili published a fast algorithm to calculate partial charges in organic molecules by a Partial Equalization of Orbital 
+Electronegativities (PEOE) \cite{gm78}, which is till today one of the gold-standards to calculate partial charges of atoms.
 
-Instead of converting molecular graphs to a vector representation for
-applying machine learning methods, it is also possible to use direct
-molecular graph mining methods \cite{okada2006}. Some of the first
-published methods use neural networks \cite{kireev1995}, inductive
-logic programming \cite{yh02a}, or graph kernels \cite{kti03}.
+Instead of converting molecular graphs to a vector representation for applying machine learning methods, it is also possible to use direct 
+molecular graph mining methods \cite{okada2006}. Some of the first published methods use neural networks \cite{kireev1995}, 
+inductive logic programming \cite{yh02a}, or graph kernels \cite{kti03}.
 
-Molecular docking of chemical molecules into proteins is an important
-application in molecular modeling \cite{Leach2001}. It is noteworthy
-that in some cases molecular shape matching methods alone
-(cheminformatics), can outperform energy-based molecular docking
-algorithms (molecular modeling) \cite{hsn07}.
+Molecular docking of chemical molecules into proteins is an important application in molecular modeling \cite{Leach2001}. It is noteworthy that in some cases
+molecular shape matching methods alone (cheminformatics), can outperform energy-based molecular docking algorithms (molecular modeling) \cite{hsn07}.
 

### 2011-11-09 supplement_history.tex
more updates to suppl history

@@
 \begin{itemize}
-\item \textbf{Quantum chemistry} is generally associated with
+\item \textbf{Quantum chemistry} (QM) is generally associated with
   theoretical chemistry or chemical physics. Briefly, it focuses on
-  the description of chemical systems starting from irst principles
+  the description of chemical systems starting from first principles
   and is based on the Schr\"{o}dinger wave equation and frameworks
   built on top of it. The use of quantum mechanics allows one to
-describe molecules in terms of electron density, which thus forms the
-basis of physically accurate calculations of physical properties
-(e.g., infrred spectra). The results of quantum chemistry calculations
-are used in molecular modeling (energy force field parameters) and
+  describe molecules in terms of electron density, and hence calculate
+  molecular properties (spectra, energies et.) in a physically
+  accurate manner.  The results of quantum chemistry calculations are
+  used in molecular modeling (energy force field parameters) and
   cheminformatics (atom partial charges, or full molecule properties,
   e.g molecule polarizabilities).
-\item \textbf{Molecular modeling} or \textbf{Computational chemistry} considers typically energy calculations of single molecules or interactions
-with other molecules, e.g. proteins. 'Molecular docking' belongs into this category, since it places a chemical molecule in the active site of enzymes
-and calculates energy/interaction scores between a chemical molecule and the protein. Surely are there various cheminformatics concepts required to
-perform such a task. Still, strictly speaking is cheminformatics rather on the chemical molecule side than on the protein side, though are both sides
-affecting each other. This area is typically strongly related to structural biology and technologies like XRay crystallography and 
-NMR spectra structure determination. 
-\item \textbf{Chemometrics} can be considered as applied cheminformatics to the field of chemical/physical analysis, e.g. infrared, mass and NMR spectra analysis.
-\item '\textbf{Translational research/medicine} aims to improve the health and longevity of the world's populations and depends on developing broad-based 
-teams of scientists and scholars who are able to focus their efforts to link basic scientific discoveries with the arena of clinical investigation, 
-and translating the results of clinical trials into changes in clinical practice, informed by evidence from the social and political sciences.' (Source:Wikipedia).
-Since clinical trials can be connected to drugs, chemical molecules, even all of translational research and biobanking could be considered as a field
-of cheminformatics, which would be on the other hand just a small contribution to a full knowledge management in this field \cite{Szalma2010}.
+\item \textbf{Molecular modeling} or \textbf{Computational chemistry}
+  can be considered an approximation of quantum chemistry and aims to
+  evaluate many of the properties considered by QM methods. The key
+  difference is that these methods employ a variety of
+  parametrizations (usually, but not always, derived from QM studies),
+  allowing one to evaluate energies for small molecules very rapidly
+  as well as handle larger systems (such as proteins) that are too
+  time consuming for QM methods. A variety of techniques including
+  semi-empirical methods, molecular mechanics and docking belong to
+  this category. While it is true that many of these methods involve
+  cheminformatics concepts (choice of tautomer, partial charge
+  assignments, etc.), the methodologies are fundamentally physical in
+  nature. We note that computational chemistry has close ties to
+  topics such as x-ray crystallography and NMR structure elucidation.
+\item \textbf{Chemometrics} is a focused application of statistical
+  methods to problems in analytical chemistry, such as infrared, mass
+  and NMR spectroscopy.
+\item \textbf{Translational research/medicine} aims to make more
+  direct links between academic (or early stage) research and clinical
+  practice in an effort to speed up the results of research into
+  actual treatments. When applied to medical research this usually
+  implies efforts to speed up the process by which a compounds goes
+  from the lead stage ot the clinical trial stage. Naturally,
+  cheminformatics plays a role at various stages of this process (lead
+  identification, lead optimization, ADMET modeling) as well as more
+  broadly such as in drug repurposing efforts.
 \end{itemize}
-If we consider all information, analysis, and optimization of a chemical molecule as 'cheminformatics', then the field is very large, since a chemical
-molecule plays the central role in many related disciplines. Still, for keeping it practical the aim of this review is not being exhaustive, but rather to 
-point the reader to more detailed references and to highlight some historic milestones.
+If we consider all information, analysis, and optimization of a
+molecule as ``cheminformatics'', then the field is very large, since a
+molecule plays the central role in many related disciplines. Clearly,
+this article is not mean to be an exhaustive overview; rather, we
+point the reader to more detailed references and to highlight some
+historic milestones.
 
-One of the most comprehensive cheminformatic series was published by Prof. Gasteiger in 2003 \cite{Gasteiger2003}, one of the pioniers in the field.
-Compacter introductory books followed from Gasteiger/Engel \cite{gasteigerengel2003} and Leach/Gillet \cite{leachgillet2007}.
-The books of Bajorath \cite{Bajorath2004} and Oprea \cite{oprea2005} focus on the applications of cheminformatics in drug design.
-The first books discussing chemical graph theory and algorithms were published in 1989 by Zupan \cite{zupan1989}, 
-1991 by Bonchev/Rouvrey \cite{bonchevrouvrey1991,bonchevrouvrey2003} and 1992 by Trinastic \cite{Trinajstic1992}. 
-The most recent one was published by Faulon/Bender \cite{faulon2010}.
-Other books focus more on specific fields/applications like mathematical challenges \cite{mathchallenges1995}, 
-molecular diversity \cite{moleculardiversity1999}, factor analysis \cite{Malinowski2002}, evolutionary algorithms \cite{clark2000},
-molecular descriptors \cite{todeschini2000}.
+One of the most comprehensive cheminformatic series was published by
+Prof. Gasteiger in 2003 \cite{Gasteiger2003}, one of the pioniers in
+the field.  Compacter introductory books followed from Gasteiger/Engel
+\cite{gasteigerengel2003} and Leach/Gillet \cite{leachgillet2007}.
+The books of Bajorath \cite{Bajorath2004} and Oprea \cite{oprea2005}
+focus on the applications of cheminformatics in drug design.  The
+first books discussing chemical graph theory and algorithms were
+published in 1989 by Zupan \cite{zupan1989}, 1991 by Bonchev/Rouvrey
+\cite{bonchevrouvrey1991,bonchevrouvrey2003} and 1992 by Trinastic
+\cite{Trinajstic1992}.  The most recent one was published by
+Faulon/Bender \cite{faulon2010}.  Other books focus more on specific
+fields/applications like mathematical challenges
+\cite{mathchallenges1995}, molecular diversity
+\cite{moleculardiversity1999}, factor analysis \cite{Malinowski2002},
+evolutionary algorithms \cite{clark2000}, molecular descriptors
+\cite{todeschini2000}.
 

### 2011-11-09 supplement_history.tex
more updates

--- 2011-11-08 supplement_history.tex
@@
 
-\title{Supplement History of Chem(o)informaqtics as part of Cheminformatics: The Computer Science of Chemical Discovery}
+\title{Supplement History of Cheminformatics as part of Cheminformatics: The Computer Science of Chemical Discovery}
 \numberofauthors{9}
@@
 
-\section{History of Chem(o)informatics}
-The aim of this brief review of the history of cheminformatics is to put the
-content of the main article into a broader perspective. Therefore it
-might be useful to provide some definitions of cheminformatics, which
-might also shed some light on the founding principles of the
-field. Chen \cite{Chen2006} and Brown \cite{brown2009} cited various
-definitions of pioniers in the field and Brown concluded rightfully:
-'\textit{The differences in definitions (of the term cheminformatics)
-  are largely a result of the types of analyses that particular
-  scientists practice and no single definition is intended to be
-  all-encompassing}.'
+\section{Definitions and References}
+The aim of this brief review of the history of cheminformatics is to
+put the content of the main article into a broader perspective. If we
+consider all information, analysis, and \emph{in silico} optimization
+of a molecule as ``cheminformatics'', then the field is very large,
+since a molecule plays the central role in many related
+disciplines. Clearly, this article is not mean to be an exhaustive
+overview; rather, we point the reader to more detailed references and
+to highlight some historic milestones.  In addition, we also discuss
+some terms as they relate to cheminformatics.  It is thus useful to
+first provide some definitions of cheminformatics itself, given that
+they also shed some light on the founding principles of the field. For
+example, we have the following definitions:
 \begin{itemize}
-\item '\textit{The mixing of information resources to transform data
+\item ``\textit{The mixing of information resources to transform data
     into information, and information into knowledge, for the intended
     purpose of making better decisions faster in the arena of drug
-    lead identification and optimization.}' Frank K. Brown, 1998.
-\item '[Chemoinformatics involves]... \textit{the computer
+    lead identification and optimization.}'' Frank K. Brown, 1998.
+\item ``\textit{[Chemoinformatics involves]... the computer
     manipulation of two- or threedimensional chemical structures and
@@
     chemical information, largely a discipline of chemical librarians
-    and does not include the development of computational methods.}'
+    and does not include the development of computational methods.}''
   Peter Willett, 2002.
-\item '\textit{...the application of informatics to solve chemical
-    problems.}' and '\textit{chemoinformatics makes the point that
-    you’re using one scientific discipline to understand another
-    scientific discipline.}' Johann Gasteiger, 2002
+\item ``\textit{\ldots the application of informatics to solve chemical
+    problems.}'' and ``\textit{\ldots chemoinformatics makes the point that
+    you're using one scientific discipline to understand another
+    scientific discipline.}'' Johann Gasteiger, 2002
 \end{itemize}
-Furthermore is it useful to clarify some other terms and how
+More generally, Chen\cite{Chen2006} and Brown \cite{brown2009} cited various
+definitions of pioneers in the field with Brown concluding that
+``'\textit{The differences in definitions (of the term
+  cheminformatics) are largely a result of the types of analyses that
+  particular scientists practice and no single definition is intended
+  to be all-encompassing}'', as is evident from the three definitions above. 
+It is also useful to clarify some other terms and how
 cheminformatics might relate to them:
 \begin{itemize}
-\item \textbf{Quantum chemistry} is something most chemists would
-  relate with 'theoretical chemistry'. In short, it allows to describe
-  chemical molecules on an electron density level, which allows also
-  to describe physical properties, e.g infrared spectra of a chemical
-  compound.  The results of quantum chemistry calculation are used in
-  molecular modeling (energy force field parameters) and
+\item \textbf{Quantum chemistry} (QM) is generally associated with
+  theoretical chemistry or chemical physics. Briefly, it focuses on
+  the description of chemical systems starting from first principles
+  and is based on the Schr\"{o}dinger wave equation and frameworks
+  built on top of it. The use of quantum mechanics allows one to
+  describe molecules in terms of electron density, and hence calculate
+  molecular properties (spectra, energies et.) in a physically
+  accurate manner.  The results of quantum chemistry calculations are
+  used in molecular modeling (energy force field parameters) and
   cheminformatics (atom partial charges, or full molecule properties,
   e.g molecule polarizabilities).
-\item \textbf{Molecular modeling} or \textbf{Computational chemistry} considers typically energy calculations of single molecules or interactions
-with other molecules, e.g. proteins. 'Molecular docking' belongs into this category, since it places a chemical molecule in the active site of enzymes
-and calculates energy/interaction scores between a chemical molecule and the protein. Surely are there various cheminformatics concepts required to
-perform such a task. Still, strictly speaking is cheminformatics rather on the chemical molecule side than on the protein side, though are both sides
-affecting each other. This area is typically strongly related to structural biology and technologies like XRay crystallography and 
-NMR spectra structure determination. 
-\item \textbf{Chemometrics} can be considered as applied cheminformatics to the field of chemical/physical analysis, e.g. infrared, mass and NMR spectra analysis.
-\item '\textbf{Translational research/medicine} aims to improve the health and longevity of the world's populations and depends on developing broad-based 
-teams of scientists and scholars who are able to focus their efforts to link basic scientific discoveries with the arena of clinical investigation, 
-and translating the results of clinical trials into changes in clinical practice, informed by evidence from the social and political sciences.' (Source:Wikipedia).
-Since clinical trials can be connected to drugs, chemical molecules, even all of translational research and biobanking could be considered as a field
-of cheminformatics, which would be on the other hand just a small contribution to a full knowledge management in this field \cite{Szalma2010}.
+\item \textbf{Molecular modeling} or \textbf{Computational chemistry}
+  can be considered an approximation of quantum chemistry and aims to
+  evaluate many of the properties considered by QM methods. The key
+  difference is that these methods employ a variety of
+  parametrizations (usually, but not always, derived from QM studies),
+  allowing one to evaluate energies for small molecules very rapidly
+  as well as handle larger systems (such as proteins) that are too
+  time consuming for QM methods. A variety of techniques including
+  semi-empirical methods, molecular mechanics and docking belong to
+  this category. While it is true that many of these methods involve
+  cheminformatics concepts (choice of tautomer, partial charge
+  assignments, etc.), the methodologies are fundamentally physical in
+  nature. We note that computational chemistry has close ties to
+  topics such as x-ray crystallography and NMR structure elucidation.
+\item \textbf{Chemometrics} is a focused application of statistical
+  methods to problems in analytical chemistry, such as infrared, mass
+  and NMR spectroscopy.
+\item \textbf{Translational research/medicine} aims to make more
+  direct links between academic (or early stage) research and clinical
+  practice in an effort to speed up the results of research into
+  actual treatments. When applied to medical research this usually
+  implies efforts to speed up the process by which a compounds goes
+  from the lead stage ot the clinical trial stage. Naturally,
+  cheminformatics plays a role at various stages of this process (lead
+  identification, lead optimization, ADMET modeling) as well as more
+  broadly such as in drug repurposing efforts.
 \end{itemize}
-If we consider all information, analysis, and optimization of a chemical molecule as 'cheminformatics', then the field is very large, since a chemical
-molecule plays the central role in many related disciplines. Still, for keeping it practical the aim of this review is not being exhaustive, but rather to 
-point the reader to more detailed references and to highlight some historic milestones.
+Obviously, we cannot cover all domains that involve or refer to
+cheminformatics concepts. However, there are a number of publications
+that cover cheminformatics topics in different ways.  Probably the
+most comprehensive cheminformatic treatise was published by Gasteiger
+\cite{Gasteiger2003} in 2003.  More concise introductory texts
+followed from Gasteiger \& Engel \cite{gasteigerengel2003} and Leach
+\& Gillet \cite{leachgillet2007}.  The books of Bajorath
+\cite{Bajorath2004} and Oprea \cite{oprea2005} focus on the
+applications of cheminformatics in drug design and span a number of
+topics. The first books discussing chemical graph theory and
+algorithms were published in 1989 by Zupan \cite{zupan1989}, 1991 by
+Bonchev/Rouvrey \cite{bonchevrouvrey1991,bonchevrouvrey2003} and 1992
+by Trinastic \cite{Trinajstic1992}.  More recently, Faulon \& Bender
+\cite{faulon2010} have edited a collection focusing on algorithms in
+cheminformatics, including various applications of graph theory.
+Other books focus more on specific topcs such as mathematical
+challenges \cite{mathchallenges1995}, molecular diversity
+\cite{moleculardiversity1999}, factor analysis \cite{Malinowski2002},
+evolutionary algorithms \cite{clark2000} and molecular descriptors
+\cite{todeschini2000}.
 
-One of the most comprehensive cheminformatic series was published by Prof. Gasteiger in 2003 \cite{Gasteiger2003}, one of the pioniers in the field.
-Compacter introductory books followed from Gasteiger/Engel \cite{gasteigerengel2003} and Leach/Gillet \cite{leachgillet2007}.
-The books of Bajorath \cite{Bajorath2004} and Oprea \cite{oprea2005} focus on the applications of cheminformatics in drug design.
-The first books discussing chemical graph theory and algorithms were published in 1989 by Zupan \cite{zupan1989}, 
-1991 by Bonchev/Rouvrey \cite{bonchevrouvrey1991,bonchevrouvrey2003} and 1992 by Trinastic \cite{Trinajstic1992}. 
-The most recent one was published by Faulon/Bender \cite{faulon2010}.
-Other books focus more on specific fields/applications like mathematical challenges \cite{mathchallenges1995}, 
-molecular diversity \cite{moleculardiversity1999}, factor analysis \cite{Malinowski2002}, evolutionary algorithms \cite{clark2000},
-molecular descriptors \cite{todeschini2000}.
+\section{Historical Milestones}
+The first chemical graphs were drawn by the Scottish chemist Willam
+Cullen in 1758 \cite{bonchevrouvrey1991}, who initially called them
+affinity graphs.  After developing the concept of bonds between atoms
+(Couper, 1858) the first chemical graphs occured in publications of
+Brown in 1864 and Cayley in 1874 \cite{bonchevrouvrey1991,brown2009}.
 
-\section{Historic milestones}
-The first chemical graphs were drawn by the Scottish chemist Willam Cullen in 1758 \cite{bonchevrouvrey1991}, which called them initially affinity graphs.
-After developing the concept of bonds between atoms (Couper, 1858) the first chemical graphs occured in publications 
-of Brown in 1864 and Cayley in 1874 \cite{bonchevrouvrey1991,brown2009}.
+The year 1946 may be regarded as the birth year of chemoinformatics
+\cite{Chen2006}: \textit{``In 1946 King et al.\cite{kct1946} published
+  an article illustrating the use of IBM's business accounting
+  machines in carrying out the construction of the rotational spectra
+  of asymmetric rotors by the evaluation of mathematical equations for
+  line position and line intensity''}. One could argue that this is
+more of a computational chemistry application. The shorter term
+'chemoinformatics' was coined by Brown in 1998\cite{brown1998}, while
+the \textit{Journal of Chemical Documentation} (founded in 1961)
+changed its name in 1975 to \textit{Journal of Chemical Information
+  and Computer Sciences} to reflect the tight connection between
+chemical information and computer science. The Chemical Abstract
+Service (CAS) of the American Chemical Society (ACS) formed a research
+department in 1955 and starting from 1965 they provided their Chemical
+Registry System, and in 1968 they made the first computer-readable
+file of all abstracted documents available \cite{Chen2006}.
 
-The year 1946 may be regarded as the birth year of chemoinformatics \cite{Chen2006}: '\textit{In 1946 King et al.\cite{kct1946} published an article illustrating 
-the use of IBM's business accounting machines in carrying out the construction of the rotational spectra of asymmetric rotors
-by the evaluation of mathematical equations for line position and line intensity.}'
+From an informatics perspective, the first record of managing chemical
+information goes back to the chemical literature being indexed since
+the year 1771.  In 1881 the first edition of the \textit{Beilstein's
+  Handbuch der Organischen Chemie} encyclopedia was published
+\cite{polanski2009} registering 1500 chemical compounds.  Many key
+cheminformatics algorithms appeared in the early to mid 20th century
+(though in some cases, the underlying mathematics was known much
+earlier). Ray and Kirsch published in 1957 the first substructure
+searching algorithm to support retrieval of computerized structure
+records \cite{RayKirsch1957}. Subsequent developments improved
+substructure searches, notably the use of pre-screening via fragments
+\cite{Adamson:1973fk,Feldman:1975uq} in the 1970's. One of the key
+algorithms for molecular graph canonicalization is the Morgan
+algorithm \cite{Morgan1965}, described in 1965.
 
-The first records of managing chemical information go back to chemical literature being indexed since the year 1771.
-In 1881 the first edition of \textit{Beilstein's Handbuch der Organischen Chemie} ecyclopedia was 
-published \cite{polanski2009} registering 1500 chemical compounds. The short term 'chemoinformatics' was coined by Brown 
-in 1998\cite{brown1998}, while the \textit{Journal of Chemical Documentation} (founded in 1961) changed already its name in 1975 to 
-\textit{Journal of Chemical Information and Computer Sciences} to reflect the tight connection between chemical information 
-and computer science. The Chemical Abstract Service (CAS) of the American Chemical Society (ACS) formed in 1955 a research department
-and starting from 1965 they provided their Chemical Registry System, and in 1968 they made the first computer-readable file of all
-abstracted documents available \cite{Chen2006}.
+In 1965 the DENDRAL expert system started with the aim to
+automatically determine the structure of an unknown chemical compound
+from the corresponding mass spectrum \cite{Gray1986}. The system is
+also often being cited as one of the earliest artificial intelligence
+and expert systems \cite{Chen2006}.
 
-Ray and Kirsch published in 1957 the first substructure searching algorithm for allowing to find chemical records via computers \cite{RayKirsch1957}.
+The first system supporting chemical synthesis planning was OCSS
+(Organic Chemical Simulation of Syntheses) developed by Corey and
+Wipke in 1969 \cite{CoreyWipke1969} which in turn was an
+implementation of theories described by Vleduts in 1963
+\cite{Vleduts:1963kx}.
 
-One of the key algorithms for molecular graph canonicalization is the Morgan algorithm \cite{Morgan1965}. It will search for the most central bond
-in a graph and allows then to traverse other atoms, resolve ties, and assigning new atom indices. Still, there are various options how different 
-software implementations are handling the 'ties' for assigning new atom indices. So, by using the canonicalization of multiple software implementations and chemical
-expert systems can be even used to create randomized (canonical, only when sticking to one software) molecular graphs, which can be used as stochastic input for 
-other algorithms \cite{cokfl06}. The central bond is also being used to create 3D conformational search trees \cite{Schwab2001}.
+In 1978 Gasteiger and Marsili published a fast algorithm to calculate
+partial charges in organic molecules by a Partial Equalization of
+Orbital Electronegativities (PEOE) \cite{gm78}, which is till today
+one of the gold-standards to calculate partial charges of atoms.
 
-In 1965 the DENDRAL expert system started with the aim to automatically determine the structure of an unknown chemical compound from the 
-corresponding mass spectrum \cite{Gray1986}. The system is also often being cited as one of the earliest artificial intelligence and expert systems  \cite{Chen2006}.
+Instead of converting molecular graphs to a vector representation for
+applying machine learning methods, it is also possible to use direct
+molecular graph mining methods \cite{okada2006}. Some of the first
+published methods use neural networks \cite{kireev1995}, inductive
+logic programming \cite{yh02a}, or graph kernels \cite{kti03}.
 
-The first system supporting chemical synthesis planning was OCSS (Organic Chemical Simulation of Syntheses) developed by Corey and Wipke in 1969 \cite{CoreyWipke1969}.
-
-In the early 1970s the collaboration between Lempel-Ziv-Welch has led to the LZW algorithm, since the Cheminformatician Lynch was
-discussing chemical fragment screening approaches with the computer scientists Lempel and Ziv \cite{brown2009}.
-
-In 1978 Gasteiger/Marsili published a fast algorithm to calculate partial charges in organic molecules by a Partial Equalization of Orbital 
-Electronegativities (PEOE) \cite{gm78}, which is till today one of the gold-standards to calculate partial charges of atoms.
-
-Instead of converting molecular graphs to a vector representation for applying machine learning methods, it is also possible to use direct 
-molecular graph mining methods \cite{okada2006}. Some of the first published methods use neural networks \cite{kireev1995}, 
-inductive logic programming \cite{yh02a}, or graph kernels \cite{kti03}.
-
-Molecular docking of chemical molecules into proteins is an important application in molecular modeling \cite{Leach2001}. It is noteworthy that in some cases
-molecular shape matching methods alone (cheminformatics), can outperform energy-based molecular docking algorithms (molecular modeling) \cite{hsn07}.
-
-\bibliographystyle{abbrv}
-\bibliography{paper}
+\bibliographystyle{abbrv} \bibliography{paper}
 

### 2011-11-09 supplement_history.tex
more updates

@@
 \section{Historical Milestones}
-The first chemical graphs were drawn by the Scottish chemist Willam
-Cullen in 1758 \cite{bonchevrouvrey1991}, who initially called them
-affinity graphs.  After developing the concept of bonds between atoms
-(Couper, 1858) the first chemical graphs occured in publications of
-Brown in 1864 and Cayley in 1874 \cite{bonchevrouvrey1991,brown2009}.
+In this section we highlight the development of a variety of
+cheminformatics concepts and techniques over time. The first chemical
+graphs were drawn by the Scottish chemist Willam Cullen in 1758
+\cite{bonchevrouvrey1991}, who initially called them affinity graphs.
+After developing the concept of bonds between atoms (Couper, 1858) the
+first chemical graphs occured in publications of Brown in 1864 and
+Cayley in 1874 \cite{bonchevrouvrey1991,brown2009}.
 
@@
 
+QSAR is one of the most well known applications of predictive
+cheminformatics and originates with the work of Hansch
+\cite{Hansch:1962vn} in 1962. While Hansch's approach focused on
+property-activity relationships, true structure-activity relationships
+originated with the work of Free and Wilson in 1964
+\cite{Free:1964ys}. Subsequent work in this area led to 3D-QSAR
+techniques including CoMFA \cite{Cramer:1988zr} and
+CoMSIA\cite{Klebe:1994ly}.
+
 In 1978 Gasteiger and Marsili published a fast algorithm to calculate

### 2011-11-09 supplement_history.tex
fixed some typos in refs. minor changes to the text

@@
 algorithms were published in 1989 by Zupan \cite{zupan1989}, 1991 by
-Bonchev/Rouvrey \cite{bonchevrouvrey1991,bonchevrouvrey2003} and 1992
-by Trinastic \cite{Trinajstic1992}.  More recently, Faulon \& Bender
-\cite{faulon2010} have edited a collection focusing on algorithms in
-cheminformatics, including various applications of graph theory.
-Other books focus more on specific topcs such as mathematical
+Bonchev and Rouvrey \cite{bonchevrouvrey1991,bonchevrouvrey2003} and
+1992 by Trinastic \cite{Trinajstic1992}.  More recently, Faulon \&
+Bender \cite{faulon2010} have edited a collection focusing on
+algorithms in cheminformatics, including various applications of graph
+theory.  Other books focus more on specific topcs such as mathematical
 challenges \cite{mathchallenges1995}, molecular diversity

### 2011-11-09 rebuttal.tex
Added document for rebuttal. Updated some refs and added them to the suppl history document

\documentclass[11pt,a4paper]{article}
\usepackage{times}

\setlength{\parindent}{0cm}

\begin{document}
\newcommand{\comment}[1]{\textit{``\ldots #1''}\par\vspace{0.5em}}
\newcommand{\response}[1]{#1\vspace{1em}}

\fbox{\textbf{Reviewer 1}}
\
\comment{when the authors state 'together with an even larger number
  of annotations', they shold explain what are annotations and how
  much is the large number;} 
\response{Foo}

\comment{For an overview I would expect a more comprehensive set of
  references.}
\response{We agree with the reviewer that a more comprehensive list of
references would be appropriate and useful. However, due to format
limitations of the article, we have been restricted to the use of XXX
references. To alleviate this problem, we have included a supplement
that provides a more extensive list of references, specifically
including a number of texts covering cheminformatics broadly.}

\fbox{\textbf{Reviewer 2}}

\fbox{\textbf{Reviewer 3}}

\end{document}

### 2011-11-09 paper.tex
Added document for rebuttal. Updated some refs and added them to the suppl history document

@@
 \begin{abstract}
-One of the most prominent success stories in all the sciences over the last decade has
-  been the advance of bioinformatics: The interdisciplinary collaboration between
- computer scientists and molecular biologists led as one of its major accomplishments to the sequencing of the human
- genome, where researchers connected biological concepts to the theory of string algorithms. However, despite this
- great
-  success, few computer scientists are familiar with a related (and older!)
-  discipline: cheminformatics, the use of computers to represent
- the structures of small molecules and analyze their properties.
-  Cheminformatics has wide applicability, from the discovery of novel drug candidates to the optimization of
-  certain physicochemical properties of small molecules. Until
- recently, the data and techniques employed in the cheminformatics field have been closely guarded
-  secrets of companies whose financial success depended on being the first to produce
-  the new ``miracle molecule.'' Only within the last decade---and as an effect of a change in mindsets,
-  government-mandated policy changes and last but not least because of  chemists volunteering their time for an Open
-  Science ``movement''---have researchers gained access to freely available software packages and databases of tens of
-  millions of chemicals. Academic chemists now confront a variety of unsolved
-  algorithmic problems that could not have been tackled a decade ago, but whose
-  solutions are critical to research ranging from determining the behavior of small
-  molecules in biological pathways, to finding a cure for malaria and other ``neglected'' diseases.
+  One of the most prominent success stories in all the sciences over
+  the last decade has been the advance of bioinformatics: The
+  interdisciplinary collaboration between computer scientists and
+  molecular biologists led as one of its major accomplishments to the
+  sequencing of the human genome, where researchers connected
+  biological concepts to the theory of string algorithms. However,
+  despite this great success, few computer scientists are familiar
+  with a related (and older!)  discipline: cheminformatics, the use of
+  computers to represent the structures of small molecules and analyze
+  their properties.  Cheminformatics has wide applicability, from the
+  discovery of novel drug candidates to the optimization of certain
+  physicochemical properties of small molecules. Until recently, the
+  data and techniques employed in the cheminformatics field have been
+  closely guarded secrets of companies whose financial success
+  depended on being the first to produce the new ``miracle molecule.''
+  Only within the last decade---and as an effect of a change in
+  mindsets, government-mandated policy changes and last but not least
+  because of chemists volunteering their time for an Open Science
+  ``movement''---have researchers gained access to freely available
+  software packages and databases of tens of millions of
+  chemicals. Academic chemists now confront a variety of unsolved
+  algorithmic problems that could not have been tackled a decade ago,
+  but whose solutions are critical to research ranging from
+  determining the behavior of small molecules in biological pathways,
+  to finding a cure for malaria and other ``neglected'' diseases.
 \end{abstract}
@@
 
-Novel technologies in the life sciences churn out information at an ever
-increasing rate---public data stores such as the one at the European Bioinformatics
-Institute (EBI) contain on the order of 10 petabytes of biological information.
-However, while biological information has been publicly available for decades,
-the same has not been true for chemical information, until recently. It was
-not until 2004 that a large public small molecule structure repository (PubChem)
-became freely available. This was followed by other databases, as discussed later
-in this article. In the same vein, open source software and the publication of
-algorithms have only recently become a main focus of the field
-\cite{faulon2010}.
+Novel technologies in the life sciences churn out information at an
+ever increasing rate---public data stores such as the one at the
+European Bioinformatics Institute (EBI) contain on the order of 10
+petabytes of biological information.  While biological
+information has been publicly available for decades, the same has not
+been true for chemical information, until recently. It was not until
+2004 that a large public small molecule structure repository (PubChem)
+became freely available. This was followed by other databases, as
+discussed later in this article. In the same vein, open source
+software and the publication of algorithms have only recently become a
+main focus of the field \cite{faulon2010}.
 
@@
 
-Making chemical information public is important because the
-drugs we take are the result of intensive and costly research,
-which is facilitated by the amount of information available. The
-more information is shared (a process very hard to achieve
-in pharmaceutical companies), the more information is available to each
-researcher, facilitating the development of
-novel treatments. The relevance of computer
-science to this research becomes clear when we consider the quantity of data available---
-currently 74 million different molecules are stored in
-public databases (such as PubChem), together with an even larger number of annotations.
-The design of data structures for this chemical information, as well as subsequent
+Making chemical information public is important because the drugs we
+take are the result of intensive and costly research, which is
+facilitated by the amount of information available. The more
+information is shared (a process very hard to achieve in
+pharmaceutical companies), the more information is available to each
+researcher, facilitating the development of novel treatments. The
+relevance of computer science to this research becomes clear when we
+consider the quantity of data available--- currently 74 million
+different molecules are stored in public databases (such as PubChem),
+together with an even larger number of annotations.  The design of
+data structures for this chemical information, as well as subsequent
 data mining of this information, are areas where expertise in computer

### 2011-11-09 paper.tex
Somre more edits to the suppl history

@@
 \begin{document}
+\pagenumbering{arabic}
 
@@
 relevance of computer science to this research becomes clear when we
-consider the quantity of data available--- currently 74 million
-different molecules are stored in public databases (such as PubChem),
-together with an even larger number of annotations.  The design of
-data structures for this chemical information, as well as subsequent
-data mining of this information, are areas where expertise in computer
-science is urgently needed. (For a recent, detailed introduction to
-cheminformatics, written for computer scientists, see
-\cite{brown2009}).
+consider the quantity and types of data available--- currently a
+single data base such as PubChem contains more than 34 million
+chemical structure records, along with an even larger number of
+annotations such as synonyms, known targets, mode of action,
+regulatory approval status and so on.  Simply considering the synonym
+list, PubChem stores 50.1 million synonym for 19.6 million compounds.
+
+The design of data structures for this chemical information, as well
+as subsequent data mining of this information, are areas where
+expertise in computer science is urgently needed. (For a recent,
+detailed introduction to cheminformatics, written for computer
+scientists, see \cite{brown2009}).
 

### 2011-11-09 paper.tex
Remove statement that 3d info is completely lost. Replace by other statements on the topic.

--- 2011-11-08 paper.tex
@@
 the bottom left hand corner of the figure, and the precise 3D shape is
-relevant in many situations (although not in others).  Nevertheless, the
-3D information is lost when one only considers the graph of a
-molecule. To extend the concept of three-dimensionality further, in
+relevant in many situations (although not in others). Although the
+graph implicitly encodes the 3D structure of a molecule (when combined
+with knowledge about favored bond angles and distances), many
+different low energy 3D structures, or conformers, may be consistent
+with the same graph structure. In addition, the 3D structure may have
+a geometrical arrangement of features (such as the presence of a
+right-handed helix) which cannnot be encoded in the graph.
+
+To extend the concept of three-dimensionality further, in
 many biological systems the \emph{surface properties} of the molecule

### 2011-11-09 supplement_history.tex
Added document for rebuttal. Updated some refs and added them to the suppl history document

@@
   implies efforts to speed up the process by which a compounds goes
-  from the lead stage ot the clinical trial stage. Naturally,
+  from the lead stage to the clinical trial stage. Naturally,
   cheminformatics plays a role at various stages of this process (lead
   identification, lead optimization, ADMET modeling) as well as more
-  broadly such as in drug repurposing efforts.
+  broadly such as in drug repurposing
+  efforts \cite{Dudley:2011fk,Swamidass:2011uq}.
 \end{itemize}
@@
 most comprehensive cheminformatic treatise was published by Gasteiger
-\cite{Gasteiger2003} in 2003.  More concise introductory texts
+\cite{Gasteiger2003} in 2003.  More concise, introductory texts
 followed from Gasteiger \& Engel \cite{gasteigerengel2003} and Leach
@@
 In this section we highlight the development of a variety of
-cheminformatics concepts and techniques over time. The first chemical
-graphs were drawn by the Scottish chemist Willam Cullen in 1758
+cheminformatics concepts and techniques over time. The shorter term
+'chemoinformatics' was coined by Brown in
+1998\cite{brown1998}. However, the main journal for this field
+(\textit{Journal of Chemical Documentation} was founded in
+1961. Interestingly, the journal was renamed to \textit{Journal of
+  Chemical Information and Computer Sciences} in 1975 to reflect the
+tight connection between chemical information and computer
+science. The Chemical Abstract Service (CAS) of the American Chemical
+Society (ACS) formed a research department in 1955 and starting from
+1965 they provided their Chemical Registry System, and in 1968 they
+made the first computer-readable file of all abstracted documents
+available \cite{Chen2006}.
+
+From a more technical viewpoint, the first chemical graphs were drawn
+by the Scottish chemist Willam Cullen in 1758
 \cite{bonchevrouvrey1991}, who initially called them affinity graphs.
@@
   line position and line intensity''}. One could argue that this is
-more of a computational chemistry application. The shorter term
-'chemoinformatics' was coined by Brown in 1998\cite{brown1998}, while
-the \textit{Journal of Chemical Documentation} (founded in 1961)
-changed its name in 1975 to \textit{Journal of Chemical Information
-  and Computer Sciences} to reflect the tight connection between
-chemical information and computer science. The Chemical Abstract
-Service (CAS) of the American Chemical Society (ACS) formed a research
-department in 1955 and starting from 1965 they provided their Chemical
-Registry System, and in 1968 they made the first computer-readable
-file of all abstracted documents available \cite{Chen2006}.
+more of a computational chemistry application.
 

### 2011-11-09 rebuttal.tex
Somre more edits to the suppl history
pulled from noel. upodated rebuttal

@@
   much is the large number;} 
-\response{Foo}
+\response{We have updated the text on page 1 to be more explicit in
+  terms of what annotations are and provide a concete example of the
+  sizes involved.}
 
@@
 
+\comment{The authors mention limitations of SMILES implementations,
+  but do not mention canonicalization in this context, often using the
+  Morgan algorithm published in 1965 in J. Chem. Doc.}
+\response{We have updated the text on page 3 to note the Morgan
+  algorithm in the context of chemical structure representation}
+
+\comment{The authors state that 3D information is lost when only
+  considering the molecular graph. However, they do not make it clear
+  that the 3D space is implicitly encoded in the molecular graph –
+  this is essential to cover properly.}
+\response{We have update the text on page 2 to note that 3D
+  information is implicit in 2D representations}
+
 \fbox{\textbf{Reviewer 3}}

### 2011-11-09 rebuttal.tex
Updated rebuttal with some points

@@
 \fbox{\textbf{Reviewer 1}}
-\
+
 \comment{when the authors state 'together with an even larger number
@@
 
+\comment{If this manuscript were to be published it would be
+  completely rewriting the history of cheminformatics}
+\response{While we agree with the reviewer that some key milestones in
+  the history of cheminformatics were not included in the original
+  article, we believe that the addition of the supplemental history
+  addresses this problem. Furthermore, our original aim was \emph{not}
+  to provide a complete historical overview of the field. Others have
+  already provided such overviews. Our aim was to present a brief
+  overview of \emph{topics} in cheminformatics, selected because we
+  believed that they represent on going challenges for the field and
+  thus amenable to coss-disciplinary efforts.}
+
 \fbox{\textbf{Reviewer 3}}

### 2011-11-09 rebuttal.tex
more rebuttal

@@
 
+\comment{page 1, column 1, line 24: the authors need to provide a
+  reference for the assertion that cheminformatics is an older field,
+  particularly given the comments later in this
+  manuscript.}
+\response{We have removed the assertion that cheminformatics is older
+  than bioinformatics. Primarily, because there is no obvious
+  reference that claims this. But alse because, depending on what one
+  considers bioinformatics and cheminformatics, one field can be
+  identified as being older or younger than the other. We feel that
+  claims on age do not strengthen the paper and have thus removed it}
+
+\comment{p1, c1, l29-37: I would dispute that until recently the
+  cheminformatics techniques have been closely guarded
+  secrets. J. Chem. Doc. was founded in 1961 and other techniques even
+  older, such as the Wiener index (1947), Wiswesser Line Notation
+  (1949), etc. Not to mention the actual foundations of what we call
+  cheminformatics back to the atomistic theory of the 19th century.}
+
 \end{document}

### 2011-11-09 paper.tex
more rebuttal

@@
   despite this great success, few computer scientists are familiar
-  with a related (and older!)  discipline: cheminformatics, the use of
-  computers to represent the structures of small molecules and analyze
-  their properties.  Cheminformatics has wide applicability, from the
+  with a related discipline: cheminformatics, the use of computers to
+  represent the structures of small molecules and analyze their
+  properties.  Cheminformatics has wide applicability, from the
   discovery of novel drug candidates to the optimization of certain
@@
 
-Cheminformatics comprises different areas, which broadly speaking can be divided
-into three fields: \emph{capturing data} (using lab notebooks or potentially using
-formats such as the Chemical Markup Language for publications); \emph{storing data}
-(designing database schemas, devising ontologies) and \emph{mining data} (such as for
-bioactivity prediction, which might involve algorithms based on graphs).
-However, the type of information being analyzed is different from biological information: while bioinformatics often deals with sequences, the
-domain of cheminformatics is chemical structures. In the case of the former,
-information can very frequently be represented as one-dimensional strings, which
-are relatively easy to handle computationally. In the case of the latter however, chemical
-structures may possess rings, branches as well as multiple valid representations
-of the same structure (such as so-called ``tautomers'', where hydrogen atoms can be positioned in different places of the
-molecules), potentially giving rise to ambiguities. Hence,
-chemical structures are often reported to be more difficult to standardize.
+Cheminformatics comprises different areas, which broadly speaking can
+be divided into three fields: \emph{capturing data} (using lab
+notebooks or potentially using formats such as the Chemical Markup
+Language for publications); \emph{storing data} (designing database
+schemas, devising ontologies) and \emph{mining data} (such as for
+bioactivity prediction, which might involve algorithms based on
+graphs).  However, the type of information being analyzed is different
+from biological information: while bioinformatics often deals with
+sequences, the domain of cheminformatics is chemical structures. In
+the case of the former, information can very frequently be represented
+as one-dimensional strings, which are relatively easy to handle
+computationally. In the case of the latter however, chemical
+structures may possess rings, branches as well as multiple valid
+representations of the same structure (such as so-called
+``tautomers'', where hydrogen atoms can be positioned in different
+places of the molecules), potentially giving rise to
+ambiguities. Hence, chemical structures are often reported to be more
+difficult to standardize.
 
-To illustrate the difficulty of standardizing chemical structures, different
-representations of the same chemical structure are shown in Figure~\ref{figure:chemical-structures}. The top
-left hand corner shows a structure a chemist would draw it---every ``corner''
-representing a carbon atom (which can also be represented as a C), every O
-denoting oxygen, every N nitrogen, every H hydrogen, and so on. This is the
-standard format chemists use when exchanging information with their peers.
-However, this format is not directly amenable to data mining---different atoms might
-be collapsed into one node, and sets of atoms might have the same relevant
-properties assigned. Hence, the top right hand corner of Figure~\ref{figure:chemical-structures} shows an
-annotated graph representation, taking into account that atoms, though being of
-different elemental atom type, can have similar properties---\emph{e.g.}, both oxygen
-and nitrogen carry an ``On'' label in this graph.
+To illustrate the difficulty of standardizing chemical structures,
+different representations of the same chemical structure are shown in
+Figure~\ref{figure:chemical-structures}. The top left hand corner
+shows a structure a chemist would draw it---every ``corner''
+representing a carbon atom (which can also be represented as a C),
+every O denoting oxygen, every N nitrogen, every H hydrogen, and so
+on. This is the standard format chemists use when exchanging
+information with their peers.  However, this format is not directly
+amenable to data mining---different atoms might be collapsed into one
+node, and sets of atoms might have the same relevant properties
+assigned. Hence, the top right hand corner of
+Figure~\ref{figure:chemical-structures} shows an annotated graph
+representation, taking into account that atoms, though being of
+different elemental atom type, can have similar
+properties---\emph{e.g.}, both oxygen and nitrogen carry an ``On''
+label in this graph.
 

### 2011-11-09 rebuttal.tex
more updates to main text and rebuttal

@@
   cheminformatics back to the atomistic theory of the 19th century.}
-
+\response{We have the updated the text on page 1 to attenuate this
+  statement, stressing on the fact that much of the data in
+  cheminformatics is proprietary, whereas only some of the techniques (such
+  as SMILES canonicalization) are. Regarding the statement that atomistic
+  theory laid the foundations of cheminformatics - this is certainly
+  true in a very broad sense. For that matter the atomistic theory
+  laid the ground for computational chemistry in general. In the
+  supplemental history document, we make a distinction between
+  computational chemistry and cheminformatics}
 \end{document}

### 2011-11-09 paper.tex
more updates to main text and rebuttal

@@
   discovery of novel drug candidates to the optimization of certain
-  physicochemical properties of small molecules. Until recently, the
-  data and techniques employed in the cheminformatics field have been
-  closely guarded secrets of companies whose financial success
-  depended on being the first to produce the new ``miracle molecule.''
-  Only within the last decade---and as an effect of a change in
-  mindsets, government-mandated policy changes and last but not least
-  because of chemists volunteering their time for an Open Science
-  ``movement''---have researchers gained access to freely available
-  software packages and databases of tens of millions of
-  chemicals. Academic chemists now confront a variety of unsolved
+  physicochemical properties of small molecules. Until recently, much
+  of the the data and some of the techniques employed in the
+  cheminformatics field have been closely guarded secrets of companies
+  whose financial success depended on being the first to produce the
+  new ``miracle molecule.''  Only within the last decade---and as an
+  effect of a change in mindsets, government-mandated policy changes
+  and last but not least because of chemists volunteering their time
+  for an Open Science ``movement''---have researchers gained access to
+  freely available software packages and databases of tens of millions
+  of chemicals. Academic chemists now confront a variety of unsolved
   algorithmic problems that could not have been tackled a decade ago,

### 2011-11-09 rebuttal.tex
Addressed more reviewer points

@@
   computational chemistry and cheminformatics}
+
+\comment{p1, c1, l32: I think the phrase miracle molecule could be
+  more usefully replaced with new drugs or new small molecule
+  therapeutics.}
+\response{The text has been updated to use the term ``therapeutic
+  molecule''}
+
+\comment{P4, c2, l41: the GDB-13 database contains 970 million
+  molecules, which is nearly a billion, not a trillion.}
+\response{The text has been updated to use the correct number}
+
+\comment{P5, c1, l29: the normal phrase used to describe this concept
+  is the similar property principle. The authors should also provide a
+  reference.}
+\response{This portion of the text has been restructured to refer to
+  the similarity property principle and also include the relevant reference}
+
 \end{document}

### 2011-11-09 paper.tex
Addressed more reviewer points

--- 2011-11-09 paper.tex
@@
   discovery of novel drug candidates to the optimization of certain
-  physicochemical properties of small molecules. Until recently, the
-  data and techniques employed in the cheminformatics field have been
-  closely guarded secrets of companies whose financial success
-  depended on being the first to produce the new ``miracle molecule.''
-  Only within the last decade---and as an effect of a change in
-  mindsets, government-mandated policy changes and last but not least
-  because of chemists volunteering their time for an Open Science
-  ``movement''---have researchers gained access to freely available
-  software packages and databases of tens of millions of
-  chemicals. Academic chemists now confront a variety of unsolved
+  physicochemical properties of small molecules. Until recently, much
+  of the the data and some of the techniques employed in the
+  cheminformatics field have been closely guarded secrets of companies
+  whose financial success depended on being the first to produce the
+  new therapeutic molecules.  Only within the last decade---and as an
+  effect of a change in mindsets, government-mandated policy changes
+  and last but not least because of chemists volunteering their time
+  for an Open Science ``movement''---have researchers gained access to
+  freely available software packages and databases of tens of millions
+  of chemicals. Academic chemists now confront a variety of unsolved
   algorithmic problems that could not have been tackled a decade ago,
@@
 enumeration tools to generate large chemical libraries such as GDB-13,
-comprising almost a trillion chemical
-structures~\cite{GDB}. Yet,
-the current enumeration software products do not generally produce
+comprising almost a billion chemical structures~\cite{GDB}. Yet, the
+current enumeration software products do not generally produce
 stereoisomers or tautomers; these require specific enumeration
 procedures, which are still the subject of investigations by the
-cheminformatics community. In addition, GDB-13 only enumerates molecules up to 13 heavy atoms---given the possibly
-exponential increase in search space with the number of atoms, it seems unlikely that the practically relevant chemical
-space of about 30 heavy atoms will be exhaustively enumerated in the near future.
+cheminformatics community. In addition, GDB-13 only enumerates
+molecules up to 13 heavy atoms---given the possibly exponential
+increase in search space with the number of atoms, it seems unlikely
+that the practically relevant chemical space of about 30 heavy atoms
+will be exhaustively enumerated in the near future.
 
@@
 
-The basis of predictive modeling in cheminformatics is that
-\emph{chemical structure implies biological activity}. Thus, the goal
-of any modeling approach is to capture and characterize the
-correlations between structural features and the observed biological
-activities. The goal is to describe cause-effect relationships with
-the lowest possible risk, and also to describe the likelihood of error
-(confidence) when using such models for decision making.
-Figure~\ref{figure:applicability-domain} (in the Appendix) illustrates the use of substructure matching to produce a space of chemical structures that are likely to share similar biological activity; this is called the generation of an \emph{applicability domain}.
+The basis of predictive modeling in cheminformatics is that the
+biological activity of a molecule is a function of its chemical
+structure. Together with the \emph{similarity property principle}
+\cite{Johnson:1990ys} that states that structurally similar molecules
+will exhibit similar properties, the goal of any modeling approach is
+to capture and characterize the correlations between structural
+features and the observed biological activities. Simultaneously, such
+approaches must also describe the likelihood of error (confidence)
+when using such models for decision making.  A variety of approaches
+can be employed to assess the error in (or conversely, the reliability
+of) a prediction, ranging from statistical approaches (XXX) to more
+empirical approaches such as defining an \emph{applicability domain} -
+the region of the input (i.e., chemical structures) that can be
+reliably predicted, usually deliniated using some form of similarity
+to the training set.
 
@@
 a receptor. Traditional ``QSAR'' approaches ignore receptor interactions,
-focusing only on small molecule features, and therefore loose valuable
+focusing only on small molecule features, and therefore lose valuable
 information. As a result, techniques such as proteochemometric

### 2011-11-09 rebuttal.tex
more response to reviews

@@
 \usepackage{times}
-
+\usepackage{fullpage}
 \setlength{\parindent}{0cm}
@@
 \response{This portion of the text has been restructured to refer to
-  the similarity property principle and also include the relevant reference}
+  the similarity property principle and also include the relevant
+  reference}
+
+\comment{P5, c1, l40: QSAR should be referenced to Hansch et al. Also
+  perhaps some discussion on what the authors mean by referring to
+  these approaches as traditional.}
+\response{We have updated text to include references to Hansch and
+  Free \& Wilson. Regarding the use of ``traditional'', we believe the
+  text explicitly explains why - the fact that QSAR as original
+  defined only considers ligand features. However, we have rephrased it to use
+  ``traditionally'' since one can argue that methods such as docking
+  and pharmacophores are also QSAR methods, but consider both ligand
+  and receptor.}
+
+\comment{P5, c2, l42: formal citation for one of Hopfinger?s papers is
+  required here.}
+\response{The relevant reference has been added}
+
+\comment{P5, c2, l58: this assertion is made with no evidence to back
+  it up. Why should multi-target models be more reliable? It is not
+  clear to me that they should.}\marginpar{\textbf{Needs more work}}
+\response{We have rephrased this statement to be less dogmatic}
 

### 2011-11-09 paper.tex
more response to reviews

@@
 structure. Together with the \emph{similarity property principle}
-\cite{Johnson:1990ys} that states that structurally similar molecules
-will exhibit similar properties, the goal of any modeling approach is
-to capture and characterize the correlations between structural
-features and the observed biological activities. Simultaneously, such
-approaches must also describe the likelihood of error (confidence)
-when using such models for decision making.  A variety of approaches
-can be employed to assess the error in (or conversely, the reliability
-of) a prediction, ranging from statistical approaches (XXX) to more
+\cite{Johnson:1990ys} (structurally similar molecules will exhibit
+similar properties), the goal of any modeling approach is to capture
+and characterize the correlations between structural features and the
+observed biological activities. Simultaneously, such approaches must
+also describe the likelihood of error when using such
+models for decision making.  A variety of approaches can be employed
+to assess the error in (or conversely, the reliability or confidence of) a
+prediction, ranging from statistical approaches (XXX) to more
 empirical approaches such as defining an \emph{applicability domain} -
@@
 
-In many cases, the activity of a small molecule is due to its interaction with
-a receptor. Traditional ``QSAR'' approaches ignore receptor interactions,
-focusing only on small molecule features, and therefore lose valuable
-information. As a result, techniques such as proteochemometric
-modeling have been designed to take into account both ligand and
-receptor structures.
+In many cases, the activity of a small molecule is due to its
+interaction with a receptor. Traditionally,  QSAR' \cite{Hansch:1962vn,
+  Free:1964ys} approaches ignore receptor interactions, focusing only
+on small molecule features, and therefore lose valuable
+information. As a result, techniques such as docking, pharmacophore
+modeling and proteochemometric methods have been designed to take into
+account both ligand and receptor structures.
 
@@
 simultaneously. Multi-conformer modeling has been addressed in the
-4D-QSAR methodology described by Hopfinger and co-workers.  Computer
-science techniques such as \emph{multiple-instance learning} could
-also be applied to the multi-conformer problem. More recently,
-simultaneous modeling of receptor and ligand has been addressed by the
-proteochemometric approach as first reported by Lapinsh
-\textit{et al.}~\cite{lapinsh2001}.
-
-With the advent of high-throughput screening technologies, large libraries of
-compounds can be screened against multiple targets in an efficient manner. Such
-panel assays provide a broad, systems-level view of small molecule activities.
-Models developed on such data afford us the opportunity to identify targets,
-characterize off-target effects and so on. However, most approaches to this
-problem are rather simplistic since they develop multiple individual models.
-Instead one could imagine an approach that takes into account the correlations
-in observed activities against multiple targets within a single model. Such an
-approach should lead to more robust predictions for panel assays. Finally,
-this also better reflects clinically relevant compound profiles~\cite{kuhn2010}
-and a ``personalized medicine'' concept (\emph{e.g.}, drug-drug interaction profiles).
+4D-QSAR methodology described by Hopfinger and co-workers
+\cite{Albuquerque:1998ys}.  Computer science techniques such as
+\emph{multiple-instance learning} could also be applied to the
+multi-conformer problem. More recently, simultaneous modeling of
+receptor and ligand has been addressed by the proteochemometric
+approach as first reported by Lapinsh \textit{et
+  al.}~\cite{lapinsh2001}.
+
+With the advent of high-throughput screening technologies, large
+libraries of compounds can be screened against multiple targets in an
+efficient manner. Such panel assays provide a broad, systems-level
+view of small molecule activities.  Models developed on such data
+afford us the opportunity to identify targets, characterize off-target
+effects and so on. However, most approaches to this problem are rather
+simplistic since they develop multiple individual models
+\cite{Chen:2010zr}, leading to multiple, independent predictions for a
+given input molecule.  Instead, one could imagine an approach that
+takes into account the covariance structure of the multiple observed
+activities and the structural descriptors within a single model. Such
+an approach could lead to more robust predictions for panel
+assays. Finally, this approach also better reflects clinically
+relevant compound profiles~\cite{kuhn2010} and a ``personalized
+medicine'' concept (\emph{e.g.}, drug-drug interaction profiles).
 

### 2011-11-09 paper.tex
more updates

@@
 open-source cheminformatics projects.  If more chemists think
-algorithmically, and more computer scientists think ``chemically,''
+algorithmically, and more computer scientists think chemically
 we are much better positioned taking not only ``simple'' chemicals
@@
 We would like to thank Danny Verbinnen for sharing his insights into
-ELN's.  This article was written while A. Sterling was visiting the
+ELNs.  This article was written while A. Sterling was visiting the
 Department of Electrical Engineering and Computer Science,
-Northwestern University, supported in part by NSF grant CCF-1049899. A. Bender thanks Unilever for funding.
+Northwestern University, supported in part by NSF grant
+CCF-1049899. A. Bender thanks Unilever for funding.
 

### 2011-11-09 rebuttal.tex
more updates
some minor changes. added review comments doc to repo

@@
   Free \& Wilson. Regarding the use of ``traditional'', we believe the
-  text explicitly explains why - the fact that QSAR as original
+  text explicitly explains why - the fact that QSAR as originaly
   defined only considers ligand features. However, we have rephrased it to use
@@
 
+\comment{P7, c2, l51: why is chemically in inverted commas and not
+  algorithmically? I would prefer inverted commas for neither.}
+\response{Quotes have been removed}
+
+\comment{P8, c1, l10: ELN's should be ELNs.}
+\response{Corrected}
 \end{document}

### 2011-11-09 paper.tex
some minor changes. added review comments doc to repo

@@
 
-\begin{figure}
-\centering
-\includegraphics[height=2in]{chemical-structures.png}
-\caption{The many different facets of representing chemical structures. Shown
-are a chemist's usual representation of a molecule (top left) and a
-graph-based representation amenable to data mining (top right). These
-representations are supplemented by a 3D representation of the same molecule
-(bottom left) that is closer to reality and a representation using surface
-properties (bottom right), which is often more relevant for the properties the
-molecule exhibits in an experimental setup (such as when measuring its
-solubility, or bioactivity).}
-\label{figure:chemical-structures}
-\end{figure}
+% \begin{figure}
+% \centering
+% \includegraphics[height=2in]{chemical-structures.png}
+% \caption{The many different facets of representing chemical structures. Shown
+% are a chemist's usual representation of a molecule (top left) and a
+% graph-based representation amenable to data mining (top right). These
+% representations are supplemented by a 3D representation of the same molecule
+% (bottom left) that is closer to reality and a representation using surface
+% properties (bottom right), which is often more relevant for the properties the
+% molecule exhibits in an experimental setup (such as when measuring its
+% solubility, or bioactivity).}
+% \label{figure:chemical-structures}
+% \end{figure}
 
@@
 
-\begin{figure}
-\centering
-\includegraphics[height=1in]{lipoicacid.png}
-\caption{The standard chemical graph illustration together with the SMILES, InChI and InChIKey codes for lipoic acid. Note that the SMILES is much more human-readable than the others.}
-\label{figure:smiles}
-\end{figure}
+% \begin{figure}
+% \centering
+% \includegraphics[height=1in]{lipoicacid.png}
+% \caption{The standard chemical graph illustration together with the SMILES, InChI and InChIKey codes for lipoic acid. Note that the SMILES is much more human-readable than the others.}
+% \label{figure:smiles}
+% \end{figure}
 
@@
 
-In 2010, drug company Glaxo made public over 13,000 compounds that they have investigated as potential cures for malaria~\cite{glaxo-malaria}.
-(Malaria has traditionally been a ``neglected'' disease in drug research, because the populations who suffer from it tend not to have enough money to be a stable consumer base for a new pharmaceutical.)  Perhaps we will now be able to find cures for some of these neglected diseases, by using the open source, open data approach to the computer science of chemical discovery.
+In 2010, drug company Glaxo made public over 13,000 compounds that
+they have investigated as potential cures for
+malaria~\cite{glaxo-malaria}.  (Malaria has traditionally been a
+neglected disease in drug research, because the populations who
+suffer from it tend not to have enough money to be a stable consumer
+base for a new pharmaceutical.)  Perhaps we will now be able to find
+cures for some of these neglected diseases, by using the open source,
+open data approach to the computer science of chemical discovery.
 
@@
 %
-\begin{figure}
-\centering
-\includegraphics[height=1.3in]{platins.png}
-\caption{The generated InChI is the same for the stereoisomers cisplatin and transplatin.}
-\label{figure:cistrans}
-\end{figure}
-%
-\begin{figure}
-\centering
-\includegraphics[height=2in]{CACM_ApplicabilityDomain.pdf}
-\caption{Illustration of the ``Applicability Domain'' concept.
+% \begin{figure}
+% \centering
+% \includegraphics[height=1.3in]{platins.png}
+% \caption{The generated InChI is the same for the stereoisomers cisplatin and transplatin.}
+% \label{figure:cistrans}
+% \end{figure}
+% %
+% \begin{figure}
+% \centering
+% \includegraphics[height=2in]{CACM_ApplicabilityDomain.pdf}
+% \caption{Illustration of the ``Applicability Domain'' concept.
 In order to estimate whether model predictions are reliable, it is crucial to define areas of ``chemical space'' where the model is applicable, and where it is not.
@@
 The (relatively similar) molecules to the left are likely to be included in the Applicability Domain of the model, while the (more dissimilar) molecules to the right are likely located outside this domain, hence the generated model is probably not applicable.}
-\label{figure:applicability-domain}
-\end{figure}
-%
+% \label{figure:applicability-domain}
+% \end{figure}
+% %
 \end{document}

### 2011-11-10 paper.tex
addresed another reviewer comment and added a ref

@@
 
-InChI or other identity-mapping algorithms allow for exact searching (within the
-limitations of the algorithms). However, two other practically relevant algorithms for scientific
-discovery based on chemical databases are \emph{substructure searching} and \emph{similarity
-searching}, which are required to generalize from the data points contained in a particular database to other, related
-molecules. In substructure searching, the database is searched for a specified
-wholly contained part of the structures in the database, while in similarity
-searching, the database is searched for structures similar (in some property space) to a provided search, or query,
-structure. Chemical search packages are often implemented and optimized for a
-given database technology; for example, the OrChem package is an open source
-chemical search package for the Oracle database \cite{rijnbeek2009}.
-
-Graph substructure matching is a variant of \emph{graph isomorphism detection}, which is widely believed to be computationally
+InChI or other identity-mapping algorithms allow for exact searching
+(within the limitations of the algorithms). However, two other
+practically relevant algorithms for scientific discovery based on
+chemical databases are \emph{substructure searching} and
+\emph{similarity searching}, which are required to generalize from the
+data points contained in a particular database to other, related
+molecules. In substructure searching, the database is searched for a
+specified wholly contained part of the structures in the database,
+while in similarity searching, the database is searched for structures
+similar (in some property space) to a provided search, or query,
+structure. Chemical search packages are often implemented and
+optimized for a given database technology; for example, the OrChem
+package is an open source chemical search package for the Oracle
+database \cite{rijnbeek2009}.
+
+Graph substructure matching is a variant of \emph{graph isomorphism
+  detection}, which is widely believed to be computationally
 intractable~\cite{cordella2001}. To execute a graph isomorphism search
@@
 
-The major difficulty with enumeration (apart from the size of the search space, which is thought to be in the order of
-$10^{60}$ molecules for 30 heavy atoms) is that the \emph{in silico}
-representation of molecules, the so-called molecular graphs, are
-labeled objects (i.e. labeled graphs), while the atoms of a molecule
-are of course not uniquely labeled. The mathematical concept to tackle
-this problem is to consider orbits of labeled molecular graphs under
-the operation of the symmetric group. However, when enumerating molecular
-structures, one has thus to solve the so--called isomorphism problem, where molecules that seem different are actually
-equivalent (i.e., their atoms can be mapped on top of each other by keeping the connectivity between atoms intact).
-The good news is that isomorphism can be solved in polynomial time for
+The major difficulty with enumeration (apart from the size of the
+search space, which is thought to be in the order of $10^{60}$
+molecules for 30 heavy atoms \cite{Bohacek:1996ve}) is that the
+\emph{in silico} representation of molecules, the so-called molecular
+graphs, are labeled objects (i.e. labeled graphs), while the atoms of
+a molecule are of course not uniquely labeled. The mathematical
+concept to tackle this problem is to consider orbits of labeled
+molecular graphs under the operation of the symmetric group. However,
+when enumerating molecular structures, one has thus to solve the
+so--called isomorphism problem, where molecules that seem different
+are actually equivalent (i.e., their atoms can be mapped on top of
+each other by keeping the connectivity between atoms intact).  The
+good news is that isomorphism can be solved in polynomial time for
 molecular graphs due to the fact that molecules belong to a restricted
@@
 same restriction prevents applying to a molecular graph the polynomial
-delay enumeration results obtained for general graphs (Chapter 8 in~\cite{faulon2010}). While the complexity of molecular graph
+delay enumeration results obtained for general graphs (Chapter 8
+in~\cite{faulon2010}). While the complexity of molecular graph
 enumeration remains an open problem it has nonetheless been shown that

### 2011-11-10 rebuttal.tex
addresed another reviewer comment and added a ref
fixed some missing stuff

@@
 
+\comment{P4, c2, l17: citation needed on the size of chemistry space.}
+\response{The appropriate reference has been added}
+
 \comment{P4, c2, l41: the GDB-13 database contains 970 million
@@
 
+\comment{P5, c1, l41: I would dispute the assertion that these methods
+  "ignore reception interactions" since the aim is to identify a
+  correlation of biological response (e.g. pIC50) with chemical
+  structure. The biological response is an explicit measurement of
+  receptor interactions on the protein.}  
+\response{We have updated
+  the text to explicitly note that QSAR models do not usually take
+  into account \emph{receptor features}. However we note that while
+  receptor information is implicit in the $IC_{50}$, the value also
+  includes other non-receptor related features such as
+  permeability. Furthermore, lack of receptor information in QSAR
+  models has been noted as the origin of activity cliffs (Guha and Van
+  Drie, \textit{J.~Chem.~Inf.~Model.}, \textbf{2008}, \textit{48},
+  1716--1728), thus supporting the statement that traditional QSAR
+  models do indeed miss important information on receptor-ligand
+  interactions.}
+
 \comment{P5, c2, l42: formal citation for one of Hopfinger?s papers is
@@
 \response{Corrected}
+
 \end{document}

### 2011-11-10 paper.tex
fixed some missing stuff

@@
 
-To extend the concept of three-dimensionality further, in
-many biological systems the \emph{surface properties} of the molecule
-are what appear to be relevant, as shown in the bottom right hand
-corner.  Hence, while graph mining on chemicals is certainly a
-promising avenue to explore, one needs to keep in mind that not all
-molecular properties are encoded in the graph (for example, neither
-rotational states, nor different charge states are encoded) and thus
-often other ``descriptors'' need to be used to capture the essentials of
-a chemical system \cite{bender2004}. Hence, one of the difficulties of a cheminformatician is that the equivalence of
-particular atoms (or sets of atoms) is usually not known from the onset but it rather depends on the particular system
-considered---involving a large amount of trial-and-error in the choice of a ``proper'' description for molecules in a
-particular case.
+To extend the concept of three-dimensionality further, in many
+biological systems the \emph{surface properties} of the molecule are
+what appear to be relevant, as shown in the bottom right hand corner.
+Hence, while graph mining on chemicals is certainly a promising avenue
+to explore, one needs to keep in mind that not all molecular
+properties are encoded in the graph (for example, neither rotational
+states, nor different charge states are encoded) and thus often other
+``descriptors'' need to be used to capture the essentials of a
+chemical system \cite{bender2004}. Hence, one of the difficulties of a
+cheminformatician is that the equivalence of particular atoms (or sets
+of atoms) is usually not known from the onset but it rather depends on
+the particular system considered---involving a large amount of
+trial-and-error in the choice of a ``proper'' description for
+molecules in a particular case.
 
@@
 
-Given the requirement to ``make sense'' of as much chemical data at hand as possible, we have to be able to search in
-complex data, for example multi-labelled molecular graphs, where graph labels can be different. Hence, we need to be
-able to capture molecular graphs in a risk-bounded machine encoded way.
-
-One main challenge in the cheminformatics field is that not one unique way exists to represent every molecule in a
-machine-readable form. This results in multiple encodings of
-``the same'' molecule and its different tautomeric or protonation states (see the Glossary in the Appendix for an explanation of technical chemistry terms).
-Checking whether two graphs represent the same chemical structure is
-an expensive operation (the next section will discuss this in-depth);
-thus, a representation of the structure that is unique and invariant
-with respect to a particular graphical display of that chemical is needed for
-identification. An early effort in this direction was the Simplified
-Molecular Input Line Specification (SMILES), and specifically the
-generation of a single, ``canonical'' SMILES for a particular structure. The idea was to devise rules for a unique
-ordering of the atoms in a molecular structure, hence resulting in the same representation of the same molecule.
-However,
-over the course of time, several
-different SMILES implementations became available, with no universal
-standard for generating a canonical SMILES. Nonetheless, the SMILES format remains one of the major representations of
-molecules used in practice today.
+Given the requirement to ``make sense'' of as much chemical data at
+hand as possible, we have to be able to search in complex data, for
+example multi-labelled molecular graphs, where graph labels can be
+different. Hence, we need to be able to capture molecular graphs in a
+risk-bounded machine encoded way.
+
+A fundamental challenge in cheminformatics is the unique
+representation of chemical structures in a machine readable format. A
+(chemical) graph can be encoded in multiple ways, depending on how one
+orders the nodes, resulting in multiple representations of the same
+molecule. When one considers the need to include other aspects such as
+protonation and tautomer states, the challenge increases. The need for
+a unique representation that is invariant with respect to the atom
+ordering arises due to the fact that graph isomorphism (i.e., checking
+whether two molecules are the same) is an expensive operation. The
+first such canonicalization algorithm was described by Morgan
+\cite{Morgan1965}, allowing one to generate unique string
+representations of chemical graphs letting one compare structures via
+string comparisons. The SMILES (Simplified Molecular Input Line
+Specification) format \cite{Weininger:1988kx} is an example of a
+representation that can be canonicalized. Since the original
+canonicalization algorithm was proprietary, multiple implementations
+of the format have become available, each one employing a different
+canonicalization algorithm, usually based on the Morgan algorithm. See
+Warr (Warr \cite{Warr:2011vn}) for an extensive dicussion on chemical
+structure representations.
 
@@
 As more and more data became available online, the use of unique
-structure-based identifiers became a pressing need, resulting in the IUPAC
-International Chemical Identifier (InChI) to be developed. The
+structure-based identifiers became a pressing need, resulting in the
+IUPAC International Chemical Identifier (InChI) to be developed. The
 InChI is a non-proprietary, structured textual identifier for chemical
-entities~\cite{inchi}. InChI identifiers are not intended to be read and
-understood by humans, but are useful for computational matching of
+entities~\cite{inchi}. InChI identifiers are not intended to be read
+and understood by humans, but are useful for computational matching of
 chemical entities. For rapid database lookups, the InChIKey is a
 hashed key for the InChI, which has an invariant length of 14
-characters. (Unfortunately, after implementation and publication of the hashing scheme it became apparent that in
-particular in case of stereochemically related molecules clashes of the InChI key take place in some, albeit rare
-cases.) Figure~\ref{figure:smiles} illustrates the SMILES, InChI and InChIKey for
-lipoic acid.
+characters. (Unfortunately, after implementation and publication of
+the hashing scheme it became apparent that in particular in case of
+stereochemically related molecules clashes of the InChI key take place
+in some, albeit rare cases.) Figure~\ref{figure:smiles} illustrates
+the SMILES, InChI and InChIKey for lipoic acid.
 
@@
 
-The InChI is now widely used for matching identical chemical structures, but it
-is still limited. For example, it cannot differentiate between
-certain types of stereoisomers: Figure~\ref{figure:cistrans} (in the Appendix) illustrates two stereoisomers for which
-the generated InChI is the same.
+The InChI is now widely used for matching identical chemical
+structures, but it is still limited. For example, it cannot
+differentiate between certain types of stereoisomers:
+Figure~\ref{figure:cistrans} (in the Appendix) illustrates two
+stereoisomers for which the generated InChI is the same.
 
@@
 %
-Structure-based similarity is however of
-limited use in grouping of chemicals based on non-structural features, such as
-shared bioactivity profiles. For this purpose, a large-scale and flexible
-classification structure is needed for chemicals and their linked biological
-information. One such answer to this need is provided in the form of \emph{ontologies}.
-An ontology is a formal specification of entities and their relationships in a
-domain of interest. It is structured around an underlying logical representation
-such as Description Logics~\cite{baaderdl2007}. The logic-based representation allows
-automated reasoning to derive new knowledge (inferences) from the knowledge
-encoded, making the representation both compact and powerful.
-Ontologies are already in widespread use across the biomedical domain, and in the cheminformatics field ontologies
-relating to the chemical structures themselves and the protein targets the molecules are active against are probably
-most common. A
-similarity measure based on the information encoded in ontologies is termed
-\emph{semantic similarity}, and such measures have already been applied to enhance
-chemical classification in several problem areas \cite{couto2010}. Many open
-questions still remain about the best way to combine the information encoded in
-chemical structures with chemical ontologies \cite{hastingsowled2010}.
+Structure-based similarity is however of limited use in grouping of
+chemicals based on non-structural features, such as shared bioactivity
+profiles. For this purpose, a large-scale and flexible classification
+structure is needed for chemicals and their linked biological
+information. One such answer to this need is provided in the form of
+\emph{ontologies}.  An ontology is a formal specification of entities
+and their relationships in a domain of interest. It is structured
+around an underlying logical representation such as Description
+Logics~\cite{baaderdl2007}. The logic-based representation allows
+automated reasoning to derive new knowledge (inferences) from the
+knowledge encoded, making the representation both compact and
+powerful.  Ontologies are already in widespread use across the
+biomedical domain, and in the cheminformatics field ontologies
+relating to the chemical structures themselves and the protein targets
+the molecules are active against are probably most common. A
+similarity measure based on the information encoded in ontologies is
+termed \emph{semantic similarity}, and such measures have already been
+applied to enhance chemical classification in several problem areas
+\cite{couto2010}. Many open questions still remain about the best way
+to combine the information encoded in chemical structures with
+chemical ontologies \cite{hastingsowled2010}.
 
@@
 
-Enumerating molecules is not only an interesting academic exercise but has
-practical utility as well. The foremost application of enumeration is \emph{structure
-elucidation} of natural compounds such as metabolites. Metabolites are small molecules that can be found an organism
-(such as a human) and which in many cases indicate the state the organism is in---such as indicating a disease, or a
-fasted state, etc.---and with a sufficiently large database the idea is to be able to detect the chemical composition of
-body fluids and to arrive at conclusions regarding the particular state of the organism at this point in time.
-Ideally, the wishful bench chemist collects experimental data
-from an unknown compound, the data is fed into a code, and the resulting
-structure is returned. Although that streamlined process may not yield a unique
-solution, there are commercial software products such as MOLGEN
-(\url{http://molgen.de}) that can, for instance, list all structures matching a given
-molecular formula, or mass spectrometry (MS) data. Another important application
-of enumeration is \emph{molecular design}. Here the problem is to design compounds
-(drugs, for instance) that optimize some physical, chemical, or biological
-property or activity. Although less prolific and more recent than structure
-elucidation, molecular design has introduced some novel solutions to molecular
-enumeration. Finally, with the advent of \emph{combinatorial chemistry} (the rapid synthesis of different but related molecules), molecular
-enumeration has taken a central role as it allows computational chemists to
-construct virtual libraries and test hypotheses, and it provides guidance in the
-design of optimal combinatorial experiments.
+Enumerating molecules is not only an interesting academic exercise but
+has practical utility as well. The foremost application of enumeration
+is \emph{structure elucidation} of natural compounds such as
+metabolites. Metabolites are small molecules that can be found an
+organism (such as a human) and which in many cases indicate the state
+the organism is in---such as indicating a disease, or a fasted state,
+etc.---and with a sufficiently large database the idea is to be able
+to detect the chemical composition of body fluids and to arrive at
+conclusions regarding the particular state of the organism at this
+point in time.  Ideally, the wishful bench chemist collects
+experimental data from an unknown compound, the data is fed into a
+code, and the resulting structure is returned. Although that
+streamlined process may not yield a unique solution, there are
+commercial software products such as MOLGEN (\url{http://molgen.de})
+that can, for instance, list all structures matching a given molecular
+formula, or mass spectrometry (MS) data. Another important application
+of enumeration is \emph{molecular design}. Here the problem is to
+design compounds (drugs, for instance) that optimize some physical,
+chemical, or biological property or activity. Although less prolific
+and more recent than structure elucidation, molecular design has
+introduced some novel solutions to molecular enumeration. Finally,
+with the advent of \emph{combinatorial chemistry} (the rapid synthesis
+of different but related molecules), molecular enumeration has taken a
+central role as it allows computational chemists to construct virtual
+libraries and test hypotheses, and it provides guidance in the design
+of optimal combinatorial experiments.
 
@@
 
-%GDB-13 along with other databases such as PubChem (NCBI/NIH) or ZINC
-%(UCSF) are valuable resources to search for new drugs. For instance
-%the well known ligand-based design approach uses these databases along
-%with structure similarity tools to discover new compounds binding to
-%biological targets (protein or nucleic acid). Compounds retrieved
-%through ligand-based design, however, must already be present in
-%databases, which is a limitation of the technique. To overcome that
-%limitation, algorithms have recently been developed to enumerate all
-%compounds in the chemical space matching circular \cite{faulon2003} or
-%path fingerprints \cite{fujiwara2008}, these fingerprints being
-%%popular tools to measure similarity. Enumerating molecules from
-%fingerprints, or more generally enumerating molecules from molecular
-%descriptors, is still very much under development. At stake is the
-%design of new chemicals or drugs that have not yet been synthesized.
-%
-Apart from analytical chemistry and compound design, enumeration plays a role is in the \emph{generation
-  of chemical reaction networks}. The problem here consists of
-enumerating all the possible compounds that can be produced by
-applying reaction rules to a set of initial molecules. By reversing
-the reaction rules one can also find sets of starting compounds
-necessary for producing a given target. This latter process is known
-in chemistry as \emph{retrosynthesis}, ever since the works of
-E.J. Corey, who received the Nobel Prize in Chemistry in
+Apart from analytical chemistry and compound design, enumeration plays
+a role is in the \emph{generation of chemical reaction networks}. The
+problem here consists of enumerating all the possible compounds that
+can be produced by applying reaction rules to a set of initial
+molecules. By reversing the reaction rules one can also find sets of
+starting compounds necessary for producing a given target. This latter
+process is known in chemistry as \emph{retrosynthesis}, ever since the
+works of E.J. Corey, who received the Nobel Prize in Chemistry in
 1990. Designing new drugs or chemicals, understanding the kinetics of
@@
 requires encompassing both chemical and sequence information and the
-development of tools that are at the interface between
-cheminformatics and bioinformatics.
+development of tools that are at the interface between cheminformatics
+and bioinformatics.
 
@@
 structure. Together with the \emph{similarity property principle}
-\cite{Johnson:1990ys} (structurally similar molecules will exhibit
+\cite{Johnson:1990qf} (structurally similar molecules will exhibit  
 similar properties), the goal of any modeling approach is to capture
@@
 interaction with a receptor. Traditionally, QSAR' \cite{Hansch:1962vn,
-  Free:1964ys} approaches ignore receptor interactions, focusing only
-on small molecule features, and therefore lose valuable
-information. As a result, techniques such as docking, pharmacophore
-modeling and proteochemometric methods have been designed to take into
-account both ligand and receptor structures.
+  Free:1964ys} approaches do not take into account receptor features,
+focusing only on small molecule features, and therefore lose valuable
+information on ligand-receptor interactions. As a result, techniques
+such as docking, pharmacophore modeling and proteochemometric methods
+have been designed to take into account both ligand and receptor
+structures.
 
 The first step in the predictive modeling of biological activities is
-to describe compounds in a way amenable to computerized data mining, while at the same time capturing information
-relevant for the property one attempts to model. Here multiple approaches exist: Either we transform molecular graphs
-into a
-numerical vector representation (a.k.a. \emph{descriptors} or
-\emph{features}) of chemical structures, or we use algorithms for
-comparing molecular graphs directly (\emph{molecule kernel}). In the
-first case thousands of descriptors (many of which are equivalent)
-have been described \cite{todeschini2000},
-and a variety of tools are available for their calculation. Given a
-set of molecules, observed activities and a descriptor vector, the
-problem can be considered a traditional statistical modeling
-problem. In the second case no explicit vector representation is
-required, which prevents the challenge of transforming the molecular
-graph into a vector without losing information. On the other hand, the
-implicit kernel encoding imposes the challenge that we have to compare all compounds against
-all compounds, which imposes large-scale mining challenge, which risks overrunning available
-processing time and space. Note that the public PubChem dataset alone
-contains over 30 million chemical compounds and over 70 million substances.
-
-In order to judge which predictive modeling approach is applicable in every case, we firstly have to consider the goal
-of a the model in a cheminformatics setting. In some cases, pure predictive ability is
-desired---such as in a virtual screening setting. (Basically, only the reliability of the model matters, but other
-aspects such as interpretability are of minor importance.) For such cases, the
+to describe compounds in a way amenable to computerized data mining,
+while at the same time capturing information relevant for the property
+one attempts to model. Here multiple approaches exist: Either we
+transform molecular graphs into a numerical vector representation
+(a.k.a. \emph{descriptors} or \emph{features}) of chemical structures,
+or we use algorithms for comparing molecular graphs directly
+(\emph{molecule kernel}). In the first case thousands of descriptors
+(many of which are equivalent) have been described
+\cite{todeschini2000}, and a variety of tools are available for their
+calculation. Given a set of molecules, observed activities and a
+descriptor vector, the problem can be considered a traditional
+statistical modeling problem. In the second case no explicit vector
+representation is required, which prevents the challenge of
+transforming the molecular graph into a vector without losing
+information. On the other hand, the implicit kernel encoding imposes
+the challenge that we have to compare all compounds against all
+compounds, which imposes large-scale mining challenge, which risks
+overrunning available processing time and space. Note that the public
+PubChem dataset alone contains over 30 million chemical compounds and
+over 70 million substances.
+
+In order to judge which predictive modeling approach is applicable in
+every case, we firstly have to consider the goal of a the model in a
+cheminformatics setting. In some cases, pure predictive ability is
+desired -- such as in a virtual screening setting. (Basically, only the
+reliability of the model matters, but other aspects such as
+interpretability are of minor importance.) For such cases, the
 complexity of the modeling approach is immaterial, as it leads to
@@
 \emph{in silico} compounds into real-world compounds. In such cases,
-``black-box methods'' can be less useful, though attempts have been made
-to interpret vectorial models built using neural networks, Support Vector Machines or
-random forests as well.
-
-Secondly, the concept of model applicability---the question of when the prediction of the model
-for a new object is reliable---has grown in importance with
-the increasing use of predictive models in regulatory settings. A
-variety of methods have been developed that attempt to characterize
-the domain of a model. These methods not only determine whether models
-are applicable, but can also be used for deciding if additional
-biological experiments are required for reducing the risk on certain
-compound classes.
+``black-box methods'' can be less useful, though attempts have been
+made to interpret vectorial models built using neural networks,
+Support Vector Machines or random forests as well.
+
+Secondly, the concept of model applicability -- the question of when
+the prediction of the model for a new object is reliable -- has grown
+in importance with the increasing use of predictive models in
+regulatory settings. A variety of methods have been developed that
+attempt to characterize the domain of a model. These methods not only
+determine whether models are applicable, but can also be used for
+deciding if additional biological experiments are required for
+reducing the prediction error on certain compound classes.
 
@@
 
-The relevance of scientific knowledge management is increasing not only for
-reducing the risk in current research, but also for enabling new
-collaboration/innovation opportunities with internal partners and a rapidly growing
-number of external/public partners. In fact, in cheminformatics and chemistry,
-scientists already switched many years ago from paper lab-notebooks to
-online collaboration platforms, called ELNs (Electronic Lab Notebooks). So, what were the drivers behind chemists, and
-companies, adopting ``Enterprise 2.0,'' a social online collaboration culture?
+The relevance of scientific knowledge management is increasing not
+only for reducing the risk in current research, but also for enabling
+new collaboration/innovation opportunities with internal partners and
+a rapidly growing number of external/public partners. In fact, in
+cheminformatics and chemistry, scientists already switched many years
+ago from paper lab-notebooks to online collaboration platforms, called
+ELNs (Electronic Lab Notebooks). So, what were the drivers behind
+chemists, and companies, adopting ``Enterprise 2.0,'' a social online
+collaboration culture?
 
 Before the year 2000, many chemists were still using paper
-lab-notebooks, multiple compound registration, and search tools. Compared to
-today there were many pitfalls and drawbacks associated with this. The overall architecture
-was too inflexible to adopt to fast changing data standards, scientists spent
-too much time with administrative work, chemical data quality was inconsistent,
-an alignment with other working groups was inefficient (\emph{e.g.}, for running
-analytical experiments), a legal dispute required to find data in paper
-lab-notebooks, data synchronization between systems caused lag times and access
-delays, and local silos hampered an efficient global collaboration. Now,
-chemists prefer ELNs, because they can run searches in the data across the organization. This does
-not only allow to find experts faster, but also allows access to highly trusted
-data with approved quality workflows. Since this also gives access to
-analytical and legal data directly every new ELN release is aiming at increasing
-the integration of scientific and business workflows for reducing the risk due
-to irrational or bounded-rational decision making and increasing decision making
-efficiency.
-
-ELNs are typically much more accepted and used within the cheminformatics
-domain than in the bioinformatics domain, which is picking up speed lately, the
-question being why this is the case. In chemistry, every experiment requires a lot of expert
+lab-notebooks, multiple compound registration, and search
+tools. Compared to today there were many pitfalls and drawbacks
+associated with this. The overall architecture was too inflexible to
+adopt to fast changing data standards, scientists spent too much time
+with administrative work, chemical data quality was inconsistent, an
+alignment with other working groups was inefficient (\emph{e.g.}, for
+running analytical experiments), a legal dispute required to find data
+in paper lab-notebooks, data synchronization between systems caused
+lag times and access delays, and local silos hampered an efficient
+global collaboration. Now, chemists prefer ELNs, because they can run
+searches in the data across the organization. This does not only allow
+to find experts faster, but also allows access to highly trusted data
+with approved quality workflows. Since this also gives access to
+analytical and legal data directly every new ELN release is aiming at
+increasing the integration of scientific and business workflows for
+reducing the risk due to irrational or bounded-rational decision
+making and increasing decision making efficiency.
+
+ELNs are typically much more accepted and used within the
+cheminformatics domain than in the bioinformatics domain, which is
+picking up speed lately, the question being why this is the case. In
+chemistry, every experiment requires a lot of expert
 knowledge. Therefore, approaches for reducing risks by communicating
-with fellow chemists about similar reactions or compounds is critical. This
-requires capturing the reaction details, compound details, and time consuming
-analysis details which ensure quality data, such that these data can be searched by others.
+with fellow chemists about similar reactions or compounds is
+critical. This requires capturing the reaction details, compound
+details, and time consuming analysis details which ensure quality
+data, such that these data can be searched by others.
 
@@
 
-However, the future still leaves many challenges ahead. Using external/public databases
-with chemical and bioactivity data remains a challenge due to differences in
-ontologies, synchronization and maintenance, and efficient
-substructure and similarity search within complex data types
-(compounds). A
-collaboration with external parties, e.g. contract research, poses other
-problems with duplicate checking for compounds (e.g. tautomeric or salt
-structures) and efficient data synchronization, and most often a double-registration is
-required for fulfilling the data quality processes on both sides. If external
+However, the future still leaves many challenges ahead. Using
+external/public databases with chemical and bioactivity data remains a
+challenge due to differences in ontologies, synchronization and
+maintenance, and efficient substructure and similarity search within
+complex data types (compounds). A collaboration with external parties,
+e.g. contract research, poses other problems with duplicate checking
+for compounds (e.g. tautomeric or salt structures) and efficient data
+synchronization, and most often a double-registration is required for
+fulfilling the data quality processes on both sides. If external
 partners are small they often do not even have sufficient IT resources
-themselves and thus rely on external services. Cloud services could help
-not only to provide a service infrastructure for all parties involved,
-but also to provide the required private/public access toolbox. Furthermore,
-there are still many data sources not being indexed properly, since chemistry
-patents are often cryptic and image/text mining remains a puzzle. One could
-argue that the image/text mining in scientific journals is less challenging; in
-fact, it remains a huge challenge and many chemists do not trust automated
-non-expert validated mining sources. The closed CAS (Chemical Abstract Service)
-is a highly trusted source, and public chemistry and bioactivity space has to
-improve quality and interconnectivity to compare with it. Efforts like ChEMBL and PubChem-BioAssay
-are on the right track. Still, improving data quality and standards between
-public and closed sources will be absolutely critical for ensuring constant
-growth, usage, and collaboration scenarios between private and public parties.
-In a business context we need to address ROI (return-of-investment) questions, not only for single
-institutions and companies, but also for collaboration networks of legal partners.
-So, only when the technical and scientific questions are solved can we ensure that
-legal questions about contribution metrics will get enough room to grow. Solving challenges is one thing (science and
-technical),
-ensuring that the contributing partners get rewarded by their contribution another (legal).
+themselves and thus rely on external services. Cloud services could
+help not only to provide a service infrastructure for all parties
+involved, but also to provide the required private/public access
+toolbox. Furthermore, there are still many data sources not being
+indexed properly, since chemistry patents are often cryptic and
+image/text mining remains a puzzle. One could argue that the
+image/text mining in scientific journals is less challenging; in fact,
+it remains a huge challenge and many chemists do not trust automated
+non-expert validated mining sources. The closed CAS (Chemical Abstract
+Service) is a highly trusted source, and public chemistry and
+bioactivity space has to improve quality and interconnectivity to
+compare with it. Efforts like ChEMBL and PubChem-BioAssay are on the
+right track. Still, improving data quality and standards between
+public and closed sources will be absolutely critical for ensuring
+constant growth, usage, and collaboration scenarios between private
+and public parties.  In a business context we need to address ROI
+(return-of-investment) questions, not only for single institutions and
+companies, but also for collaboration networks of legal partners.  So,
+only when the technical and scientific questions are solved can we
+ensure that legal questions about contribution metrics will get enough
+room to grow. Solving challenges is one thing (science and technical),
+ensuring that the contributing partners get rewarded by their
+contribution another (legal).
 
@@
 available. As a result, this enables much more robust validation of
-cheminformatics methodologies as well as large scale \emph{benchmarking of
-algorithms}. The latter is especially relevant for the plethora of
-data mining techniques that are employed in cheminformatics. At this
-point, the nature of Open Data focuses primarily on structure and
-activity data types. On the other hand, there is a distinct lack of
-textual data (such as journal articles) that are ``Open'' in nature. While
-PubMed abstracts serve as a proxy for journal articles, text mining
-methods in cheminformatics are hindered by not being able to mine the full text of many scientific publications.
-Interestingly, patent information is publicly accessible (cf. Google
-Patents) and there have been efforts to mine these for chemical
-information using Open tools.
+cheminformatics methodologies as well as large scale
+\emph{benchmarking of algorithms}. The latter is especially relevant
+for the plethora of data mining techniques that are employed in
+cheminformatics. At this point, the nature of Open Data focuses
+primarily on structure and activity data types. On the other hand,
+there is a distinct lack of textual data (such as journal articles)
+that are ``Open'' in nature. While PubMed abstracts serve as a proxy
+for journal articles, text mining methods in cheminformatics are
+hindered by not being able to mine the full text of many scientific
+publications.  Interestingly, patent information is publicly
+accessible (cf. Google Patents) and there have been efforts to mine
+these for chemical information using Open tools.
 
@@
 developers' mailing list (either \texttt{cdk-devel@lists.sf.net} or
-\texttt{openbabel-devel@lists.sf.net}) with their interests and background, and
-a mentor will be assigned to help them join the project.
+\texttt{openbabel-devel@lists.sf.net}) with their interests and
+background, and a mentor will be assigned to help them join the
+project.
 %
@@
 1960s, with the creation of DENDRAL, the first expert system of any
-kind.  Much of the development of cheminformatics from the 1960s until the new millenium
-was performed in secret, because of competition in the pharmaceutical
-industry.  While chemists were among the first scientists to embrace
-web-based tools of knowledge management, research into fundamental
-algorithmic questions of structure enumeration and database management
-(and a host of other topics) have suffered, because of limited
-academic access to either tools or data.  As we enter the second
-decade of the 21st century, however, this landscape is rapidly
-shifting.  There now exist free-of-charge open-source cheminformatics
-toolkits, and open databases with tens of millions of compounds.  We
-believe the prospects for academic and industrial collaboration
-between chemists and computer scientists are bright, and we
-would like to encourage more computer scientists to participate in
-cheminformatics research.  There is an admittedly steep learning
-curve, as the structural issues faced by chemists are often too
-detailed and subtle to admit as clean an abstraction as ``algorithms
-on a string,'' the way many bioinformatics algorithms can be abstracted into a computer science
-framework.  Nevertheless, because of the changing landscape, there are
-many research questions we are only now in position to try to answer:
-for a theorist, graph-theoretic questions of three-dimensional
-enumeration; for a database designer, more effective search algorithms
-and ontologies; for a practical programmer, expansion of the many
-open-source cheminformatics projects.  If more chemists think
-algorithmically, and more computer scientists think chemically
-we are much better positioned taking not only ``simple'' chemicals
-into account (only one identifier and isomer possible per molecule),
-but also their complex relationships, transformations, and
-combinatorial challenges.
+kind.  Much of the development of cheminformatics from the 1960s until
+the new millenium was performed in secret, because of competition in
+the pharmaceutical industry.  While chemists were among the first
+scientists to embrace web-based tools of knowledge management,
+research into fundamental algorithmic questions of structure
+enumeration and database management (and a host of other topics) have
+suffered, because of limited academic access to either tools or data.
+As we enter the second decade of the 21st century, however, this
+landscape is rapidly shifting.  There now exist free-of-charge
+open-source cheminformatics toolkits, and open databases with tens of
+millions of compounds.  We believe the prospects for academic and
+industrial collaboration between chemists and computer scientists are
+bright, and we would like to encourage more computer scientists to
+participate in cheminformatics research.  There is an admittedly steep
+learning curve, as the structural issues faced by chemists are often
+too detailed and subtle to admit as clean an abstraction as
+``algorithms on a string,'' the way many bioinformatics algorithms can
+be abstracted into a computer science framework.  Nevertheless,
+because of the changing landscape, there are many research questions
+we are only now in position to try to answer: for a theorist,
+graph-theoretic questions of three-dimensional enumeration; for a
+database designer, more effective search algorithms and ontologies;
+for a practical programmer, expansion of the many open-source
+cheminformatics projects.  If more chemists think algorithmically, and
+more computer scientists think chemically we are much better
+positioned taking not only ``simple'' chemicals into account (only one
+identifier and isomer possible per molecule), but also their complex
+relationships, transformations, and combinatorial challenges.
 

### 2011-11-10 paper.tex
Added more review responses

@@
 cheminformatics setting. In some cases, pure predictive ability is
-desired -- such as in a virtual screening setting. (Basically, only the
-reliability of the model matters, but other aspects such as
+desired -- such as in a virtual screening setting. (Basically, only
+the reliability of the model matters, but other aspects such as
 interpretability are of minor importance.) For such cases, the
@@
 made to interpret vectorial models built using neural networks,
-Support Vector Machines or random forests as well.
+na\"{i}ve Bayes, Support Vector Machines or random forests as well.
 

### 2011-11-10 rebuttal.tex
Added more review responses
more review responses

@@
 
+\comment{P4, c2, l23: the isomorphism problem has already been
+  mentioned previously in this manuscript.}
+\response{The text has been updated and simplified}
+
 \comment{P4, c2, l41: the GDB-13 database contains 970 million
@@
 
+\comment{P5, c2, l22: I would mention naive Bayesian classifiers as well as
+this is perhaps the most widely applied method in the field.}
+\response{We have included the Na\"{i}ve Bayes as well}
+
 \comment{P5, c2, l42: formal citation for one of Hopfinger?s papers is

### 2011-11-10 paper.tex
more review responses

@@
 concept to tackle this problem is to consider orbits of labeled
-molecular graphs under the operation of the symmetric group. However,
-when enumerating molecular structures, one has thus to solve the
-so--called isomorphism problem, where molecules that seem different
-are actually equivalent (i.e., their atoms can be mapped on top of
-each other by keeping the connectivity between atoms intact).  The
-good news is that isomorphism can be solved in polynomial time for
-molecular graphs due to the fact that molecules belong to a restricted
-class of graphs known as \emph{bounded valence graphs}; however, that
-same restriction prevents applying to a molecular graph the polynomial
-delay enumeration results obtained for general graphs (Chapter 8
-in~\cite{faulon2010}). While the complexity of molecular graph
-enumeration remains an open problem it has nonetheless been shown that
-molecules can be sampled efficiently~\cite{goldberg1999}. Sampling
-procedures based on the Metropolis or the Genetic Algorithms have been
-developed and used to elucidate natural compounds from NMR data.
+molecular graphs under the operation of the symmetric group. While the
+complexity of molecular graph enumeration remains an open problem it
+has nonetheless been shown that molecules can be sampled
+efficiently \cite{goldberg1999}. Sampling procedures based on the
+Metropolis or the Genetic Algorithms have been developed and used to
+elucidate natural compounds from NMR data.
 

### 2011-11-10 rebuttal.tex
more reviewer comments addressed

@@
 
+\comment{P3, c1, l41: the authors need to cover the Morgan algorithm
+  (published in 1965) here and explain the canonicalization
+  process. The issues mentioned in different canonicalization
+  implementations providing different SMILES strings is also a
+  challenge with InChI codes with different softwares giving different
+  representation. Therefore, this is still not a solved problem as
+  suggested here. Could the authors clarify this in the text?}
+\response{We have updated the text to reference the Morgan
+  algorithm. Regarding the issue of InChI codes, we believe that this
+  is not the case. Currently, there is only one implementation of the
+  InChI algorithm and even if alternative implementations were to be
+  developed, the InChI specification is publically available. As a
+  result, there should be no differing InChI representations for a
+  given input structure.}
+
 \comment{P4, c2, l17: citation needed on the size of chemistry space.}

### 2011-11-10 paper.tex
Added more updates

@@
 agrochemical research and drug discovery to the design of novel
-materials (since everything around us consists of ``molecules''), we have chosen to use drug discovery as the context of
-this
-article due to its central relevance for human wellbeing. Given that the safety of drugs on the market is a key concern
-both for the general audience as well as researchers in pharmaceutical companies, the remainder of this article is
-structured around the
-concept of ``risk minimization'' in drug discovery---the question of how to minimize the chances of a small molecule
-failing in the various research and development stages as a drug candidate, and how to successfully employ
-cheminformatics methods in these efforts.
-Table~\ref{table:properties} (in the Appendix) summarizes properties a drug candidate must satisfy to be considered ``promising,'' and sketches the role of cheminformatics at each stage of drug development.
+materials, we have chosen to use drug discovery as the context of this
+article due to its central relevance for human wellbeing. Given that
+the safety of drugs on the market is a key concern both for the
+general audience as well as researchers in pharmaceutical companies,
+the remainder of this article is structured around the concept of
+``risk minimization'' in drug discovery -- the question of how to
+minimize the chances of a small molecule failing in the various
+research and development stages as a drug candidate, and how to
+successfully employ cheminformatics methods in these efforts.
+Table~\ref{table:properties} (in the Appendix) summarizes properties a
+drug candidate must satisfy to be considered therapeutically useful
+(such as being appropriately soluble, exhibiting minimal toxicity and
+being selective for the intended target) and sketches the role of
+cheminformatics at each stage of drug development.
 %

### 2011-11-10 paper.tex
Added more updates

@@
 ``risk minimization'' in drug discovery -- the question of how to
-minimize the chances of a small molecule failing in the various
-research and development stages as a drug candidate, and how to
-successfully employ cheminformatics methods in these efforts.
-Table~\ref{table:properties} (in the Appendix) summarizes properties a
-drug candidate must satisfy to be considered therapeutically useful
-(such as being appropriately soluble, exhibiting minimal toxicity and
-being selective for the intended target) and sketches the role of
-cheminformatics at each stage of drug development.
+minimize the chances of a small molecule failing, due to poor
+physical, chemical or biological properties, in the various research
+and development stages as a drug candidate, and how to successfully
+employ cheminformatics methods in these efforts. As an example, a
+molecule must be soluble and show a certain degree of bioavailability
+to be considered as drug candidate. In cases where these properties
+are poor, a cheminformatics approach can suggest replacement of
+certain functional groups, that maintain potency but improve the
+solubility and bioavailability.  Table~\ref{table:properties} (in the
+Appendix) summarizes a number of properties a drug candidate must
+satisfy to be considered therapeutically useful (such as being
+appropriately soluble, exhibiting minimal toxicity and being selective
+for the intended target) and sketches the role of cheminformatics at
+each stage of drug development.
 %

### 2011-11-10 paper.tex
Rewrite/rearrange/remove the entire 'Open Source' section as 'Support for the Development of Novel Algorithms'

--- 2011-11-10 paper.tex
@@
   despite this great success, few computer scientists are familiar
-  with a related discipline: cheminformatics, the use of computers to
-  represent the structures of small molecules and analyze their
-  properties.  Cheminformatics has wide applicability, from the
+  with a related (and older!)  discipline: cheminformatics, the use of
+  computers to represent the structures of small molecules and analyze
+  their properties.  Cheminformatics has wide applicability, from the
   discovery of novel drug candidates to the optimization of certain
-  physicochemical properties of small molecules. Until recently, much
-  of the the data and some of the techniques employed in the
-  cheminformatics field have been closely guarded secrets of companies
-  whose financial success depended on being the first to produce the
-  new therapeutic molecules.  Only within the last decade---and as an
-  effect of a change in mindsets, government-mandated policy changes
-  and last but not least because of chemists volunteering their time
-  for an Open Science ``movement''---have researchers gained access to
-  freely available software packages and databases of tens of millions
-  of chemicals. Academic chemists now confront a variety of unsolved
+  physicochemical properties of small molecules. Until recently, the
+  data and techniques employed in the cheminformatics field have been
+  closely guarded secrets of companies whose financial success
+  depended on being the first to produce the new ``miracle molecule.''
+  Only within the last decade---and as an effect of a change in
+  mindsets, government-mandated policy changes and last but not least
+  because of chemists volunteering their time for an Open Science
+  ``movement''---have researchers gained access to freely available
+  software packages and databases of tens of millions of
+  chemicals. Academic chemists now confront a variety of unsolved
   algorithmic problems that could not have been tackled a decade ago,
@@
 
-Cheminformatics comprises different areas, which broadly speaking can
-be divided into three fields: \emph{capturing data} (using lab
-notebooks or potentially using formats such as the Chemical Markup
-Language for publications); \emph{storing data} (designing database
-schemas, devising ontologies) and \emph{mining data} (such as for
-bioactivity prediction, which might involve algorithms based on
-graphs).  However, the type of information being analyzed is different
-from biological information: while bioinformatics often deals with
-sequences, the domain of cheminformatics is chemical structures. In
-the case of the former, information can very frequently be represented
-as one-dimensional strings, which are relatively easy to handle
-computationally. In the case of the latter however, chemical
-structures may possess rings, branches as well as multiple valid
-representations of the same structure (such as so-called
-``tautomers'', where hydrogen atoms can be positioned in different
-places of the molecules), potentially giving rise to
-ambiguities. Hence, chemical structures are often reported to be more
-difficult to standardize.
+Cheminformatics comprises different areas, which broadly speaking can be divided
+into three fields: \emph{capturing data} (using lab notebooks or potentially using
+formats such as the Chemical Markup Language for publications); \emph{storing data}
+(designing database schemas, devising ontologies) and \emph{mining data} (such as for
+bioactivity prediction, which might involve algorithms based on graphs).
+However, the type of information being analyzed is different from biological information: while bioinformatics often deals with sequences, the
+domain of cheminformatics is chemical structures. In the case of the former,
+information can very frequently be represented as one-dimensional strings, which
+are relatively easy to handle computationally. In the case of the latter however, chemical
+structures may possess rings, branches as well as multiple valid representations
+of the same structure (such as so-called ``tautomers'', where hydrogen atoms can be positioned in different places of the
+molecules), potentially giving rise to ambiguities. Hence,
+chemical structures are often reported to be more difficult to standardize.
 
-To illustrate the difficulty of standardizing chemical structures,
-different representations of the same chemical structure are shown in
-Figure~\ref{figure:chemical-structures}. The top left hand corner
-shows a structure a chemist would draw it---every ``corner''
-representing a carbon atom (which can also be represented as a C),
-every O denoting oxygen, every N nitrogen, every H hydrogen, and so
-on. This is the standard format chemists use when exchanging
-information with their peers.  However, this format is not directly
-amenable to data mining---different atoms might be collapsed into one
-node, and sets of atoms might have the same relevant properties
-assigned. Hence, the top right hand corner of
-Figure~\ref{figure:chemical-structures} shows an annotated graph
-representation, taking into account that atoms, though being of
-different elemental atom type, can have similar
-properties---\emph{e.g.}, both oxygen and nitrogen carry an ``On''
-label in this graph.
+To illustrate the difficulty of standardizing chemical structures, different
+representations of the same chemical structure are shown in Figure~\ref{figure:chemical-structures}. The top
+left hand corner shows a structure a chemist would draw it---every ``corner''
+representing a carbon atom (which can also be represented as a C), every O
+denoting oxygen, every N nitrogen, every H hydrogen, and so on. This is the
+standard format chemists use when exchanging information with their peers.
+However, this format is not directly amenable to data mining---different atoms might
+be collapsed into one node, and sets of atoms might have the same relevant
+properties assigned. Hence, the top right hand corner of Figure~\ref{figure:chemical-structures} shows an
+annotated graph representation, taking into account that atoms, though being of
+different elemental atom type, can have similar properties---\emph{e.g.}, both oxygen
+and nitrogen carry an ``On'' label in this graph.
 
-% \begin{figure}
-% \centering
-% \includegraphics[height=2in]{chemical-structures.png}
-% \caption{The many different facets of representing chemical structures. Shown
-% are a chemist's usual representation of a molecule (top left) and a
-% graph-based representation amenable to data mining (top right). These
-% representations are supplemented by a 3D representation of the same molecule
-% (bottom left) that is closer to reality and a representation using surface
-% properties (bottom right), which is often more relevant for the properties the
-% molecule exhibits in an experimental setup (such as when measuring its
-% solubility, or bioactivity).}
-% \label{figure:chemical-structures}
-% \end{figure}
+\begin{figure}
+\centering
+\includegraphics[height=2in]{chemical-structures.png}
+\caption{The many different facets of representing chemical structures. Shown
+are a chemist's usual representation of a molecule (top left) and a
+graph-based representation amenable to data mining (top right). These
+representations are supplemented by a 3D representation of the same molecule
+(bottom left) that is closer to reality and a representation using surface
+properties (bottom right), which is often more relevant for the properties the
+molecule exhibits in an experimental setup (such as when measuring its
+solubility, or bioactivity).}
+\label{figure:chemical-structures}
+\end{figure}
 
@@
 
-To extend the concept of three-dimensionality further, in many
-biological systems the \emph{surface properties} of the molecule are
-what appear to be relevant, as shown in the bottom right hand corner.
-Hence, while graph mining on chemicals is certainly a promising avenue
-to explore, one needs to keep in mind that not all molecular
-properties are encoded in the graph (for example, neither rotational
-states, nor different charge states are encoded) and thus often other
-``descriptors'' need to be used to capture the essentials of a
-chemical system \cite{bender2004}. Hence, one of the difficulties of a
-cheminformatician is that the equivalence of particular atoms (or sets
-of atoms) is usually not known from the onset but it rather depends on
-the particular system considered---involving a large amount of
-trial-and-error in the choice of a ``proper'' description for
-molecules in a particular case.
+To extend the concept of three-dimensionality further, in
+many biological systems the \emph{surface properties} of the molecule
+are what appear to be relevant, as shown in the bottom right hand
+corner.  Hence, while graph mining on chemicals is certainly a
+promising avenue to explore, one needs to keep in mind that not all
+molecular properties are encoded in the graph (for example, neither
+rotational states, nor different charge states are encoded) and thus
+often other ``descriptors'' need to be used to capture the essentials of
+a chemical system \cite{bender2004}. Hence, one of the difficulties of a cheminformatician is that the equivalence of
+particular atoms (or sets of atoms) is usually not known from the onset but it rather depends on the particular system
+considered---involving a large amount of trial-and-error in the choice of a ``proper'' description for molecules in a
+particular case.
 
@@
 
-Given the requirement to ``make sense'' of as much chemical data at
-hand as possible, we have to be able to search in complex data, for
-example multi-labelled molecular graphs, where graph labels can be
-different. Hence, we need to be able to capture molecular graphs in a
-risk-bounded machine encoded way.
+Given the requirement to ``make sense'' of as much chemical data at hand as possible, we have to be able to search in
+complex data, for example multi-labelled molecular graphs, where graph labels can be different. Hence, we need to be
+able to capture molecular graphs in a risk-bounded machine encoded way.
 
-A fundamental challenge in cheminformatics is the unique
-representation of chemical structures in a machine readable format. A
-(chemical) graph can be encoded in multiple ways, depending on how one
-orders the nodes, resulting in multiple representations of the same
-molecule. When one considers the need to include other aspects such as
-protonation and tautomer states, the challenge increases. The need for
-a unique representation that is invariant with respect to the atom
-ordering arises due to the fact that graph isomorphism (i.e., checking
-whether two molecules are the same) is an expensive operation. The
-first such canonicalization algorithm was described by Morgan
-\cite{Morgan1965}, allowing one to generate unique string
-representations of chemical graphs letting one compare structures via
-string comparisons. The SMILES (Simplified Molecular Input Line
-Specification) format \cite{Weininger:1988kx} is an example of a
-representation that can be canonicalized. Since the original
-canonicalization algorithm was proprietary, multiple implementations
-of the format have become available, each one employing a different
-canonicalization algorithm, usually based on the Morgan algorithm. See
-Warr (Warr \cite{Warr:2011vn}) for an extensive dicussion on chemical
-structure representations.
+One main challenge in the cheminformatics field is that not one unique way exists to represent every molecule in a
+machine-readable form. This results in multiple encodings of
+``the same'' molecule and its different tautomeric or protonation states (see the Glossary in the Appendix for an explanation of technical chemistry terms).
+Checking whether two graphs represent the same chemical structure is
+an expensive operation (the next section will discuss this in-depth);
+thus, a representation of the structure that is unique and invariant
+with respect to a particular graphical display of that chemical is needed for
+identification. An early effort in this direction was the Simplified
+Molecular Input Line Specification (SMILES), and specifically the
+generation of a single, ``canonical'' SMILES for a particular structure. The idea was to devise rules for a unique
+ordering of the atoms in a molecular structure, hence resulting in the same representation of the same molecule.
+However,
+over the course of time, several
+different SMILES implementations became available, with no universal
+standard for generating a canonical SMILES. Nonetheless, the SMILES format remains one of the major representations of
+molecules used in practice today.
 
@@
 As more and more data became available online, the use of unique
-structure-based identifiers became a pressing need, resulting in the
-IUPAC International Chemical Identifier (InChI) to be developed. The
+structure-based identifiers became a pressing need, resulting in the IUPAC
+International Chemical Identifier (InChI) to be developed. The
 InChI is a non-proprietary, structured textual identifier for chemical
-entities~\cite{inchi}. InChI identifiers are not intended to be read
-and understood by humans, but are useful for computational matching of
+entities~\cite{inchi}. InChI identifiers are not intended to be read and
+understood by humans, but are useful for computational matching of
 chemical entities. For rapid database lookups, the InChIKey is a
 hashed key for the InChI, which has an invariant length of 14
-characters. (Unfortunately, after implementation and publication of
-the hashing scheme it became apparent that in particular in case of
-stereochemically related molecules clashes of the InChI key take place
-in some, albeit rare cases.) Figure~\ref{figure:smiles} illustrates
-the SMILES, InChI and InChIKey for lipoic acid.
+characters. (Unfortunately, after implementation and publication of the hashing scheme it became apparent that in
+particular in case of stereochemically related molecules clashes of the InChI key take place in some, albeit rare
+cases.) Figure~\ref{figure:smiles} illustrates the SMILES, InChI and InChIKey for
+lipoic acid.
 
-% \begin{figure}
-% \centering
-% \includegraphics[height=1in]{lipoicacid.png}
-% \caption{The standard chemical graph illustration together with the SMILES, InChI and InChIKey codes for lipoic acid. Note that the SMILES is much more human-readable than the others.}
-% \label{figure:smiles}
-% \end{figure}
+\begin{figure}
+\centering
+\includegraphics[height=1in]{lipoicacid.png}
+\caption{The standard chemical graph illustration together with the SMILES, InChI and InChIKey codes for lipoic acid. Note that the SMILES is much more human-readable than the others.}
+\label{figure:smiles}
+\end{figure}
 
-The InChI is now widely used for matching identical chemical
-structures, but it is still limited. For example, it cannot
-differentiate between certain types of stereoisomers:
-Figure~\ref{figure:cistrans} (in the Appendix) illustrates two
-stereoisomers for which the generated InChI is the same.
+The InChI is now widely used for matching identical chemical structures, but it
+is still limited. For example, it cannot differentiate between
+certain types of stereoisomers: Figure~\ref{figure:cistrans} (in the Appendix) illustrates two stereoisomers for which
+the generated InChI is the same.
 
-InChI or other identity-mapping algorithms allow for exact searching
-(within the limitations of the algorithms). However, two other
-practically relevant algorithms for scientific discovery based on
-chemical databases are \emph{substructure searching} and
-\emph{similarity searching}, which are required to generalize from the
-data points contained in a particular database to other, related
-molecules. In substructure searching, the database is searched for a
-specified wholly contained part of the structures in the database,
-while in similarity searching, the database is searched for structures
-similar (in some property space) to a provided search, or query,
-structure. Chemical search packages are often implemented and
-optimized for a given database technology; for example, the OrChem
-package is an open source chemical search package for the Oracle
-database \cite{rijnbeek2009}.
+InChI or other identity-mapping algorithms allow for exact searching (within the
+limitations of the algorithms). However, two other practically relevant algorithms for scientific
+discovery based on chemical databases are \emph{substructure searching} and \emph{similarity
+searching}, which are required to generalize from the data points contained in a particular database to other, related
+molecules. In substructure searching, the database is searched for a specified
+wholly contained part of the structures in the database, while in similarity
+searching, the database is searched for structures similar (in some property space) to a provided search, or query,
+structure. Chemical search packages are often implemented and optimized for a
+given database technology; for example, the OrChem package is an open source
+chemical search package for the Oracle database \cite{rijnbeek2009}.
 
-Graph substructure matching is a variant of \emph{graph isomorphism
-  detection}, which is widely believed to be computationally
+Graph substructure matching is a variant of \emph{graph isomorphism detection}, which is widely believed to be computationally
 intractable~\cite{cordella2001}. To execute a graph isomorphism search
@@
 %
-Structure-based similarity is however of limited use in grouping of
-chemicals based on non-structural features, such as shared bioactivity
-profiles. For this purpose, a large-scale and flexible classification
-structure is needed for chemicals and their linked biological
-information. One such answer to this need is provided in the form of
-\emph{ontologies}.  An ontology is a formal specification of entities
-and their relationships in a domain of interest. It is structured
-around an underlying logical representation such as Description
-Logics~\cite{baaderdl2007}. The logic-based representation allows
-automated reasoning to derive new knowledge (inferences) from the
-knowledge encoded, making the representation both compact and
-powerful.  Ontologies are already in widespread use across the
-biomedical domain, and in the cheminformatics field ontologies
-relating to the chemical structures themselves and the protein targets
-the molecules are active against are probably most common. A
-similarity measure based on the information encoded in ontologies is
-termed \emph{semantic similarity}, and such measures have already been
-applied to enhance chemical classification in several problem areas
-\cite{couto2010}. Many open questions still remain about the best way
-to combine the information encoded in chemical structures with
-chemical ontologies \cite{hastingsowled2010}.
+Structure-based similarity is however of
+limited use in grouping of chemicals based on non-structural features, such as
+shared bioactivity profiles. For this purpose, a large-scale and flexible
+classification structure is needed for chemicals and their linked biological
+information. One such answer to this need is provided in the form of \emph{ontologies}.
+An ontology is a formal specification of entities and their relationships in a
+domain of interest. It is structured around an underlying logical representation
+such as Description Logics~\cite{baaderdl2007}. The logic-based representation allows
+automated reasoning to derive new knowledge (inferences) from the knowledge
+encoded, making the representation both compact and powerful.
+Ontologies are already in widespread use across the biomedical domain, and in the cheminformatics field ontologies
+relating to the chemical structures themselves and the protein targets the molecules are active against are probably
+most common. A
+similarity measure based on the information encoded in ontologies is termed
+\emph{semantic similarity}, and such measures have already been applied to enhance
+chemical classification in several problem areas \cite{couto2010}. Many open
+questions still remain about the best way to combine the information encoded in
+chemical structures with chemical ontologies \cite{hastingsowled2010}.
 
@@
 
-Enumerating molecules is not only an interesting academic exercise but
-has practical utility as well. The foremost application of enumeration
-is \emph{structure elucidation} of natural compounds such as
-metabolites. Metabolites are small molecules that can be found an
-organism (such as a human) and which in many cases indicate the state
-the organism is in---such as indicating a disease, or a fasted state,
-etc.---and with a sufficiently large database the idea is to be able
-to detect the chemical composition of body fluids and to arrive at
-conclusions regarding the particular state of the organism at this
-point in time.  Ideally, the wishful bench chemist collects
-experimental data from an unknown compound, the data is fed into a
-code, and the resulting structure is returned. Although that
-streamlined process may not yield a unique solution, there are
-commercial software products such as MOLGEN (\url{http://molgen.de})
-that can, for instance, list all structures matching a given molecular
-formula, or mass spectrometry (MS) data. Another important application
-of enumeration is \emph{molecular design}. Here the problem is to
-design compounds (drugs, for instance) that optimize some physical,
-chemical, or biological property or activity. Although less prolific
-and more recent than structure elucidation, molecular design has
-introduced some novel solutions to molecular enumeration. Finally,
-with the advent of \emph{combinatorial chemistry} (the rapid synthesis
-of different but related molecules), molecular enumeration has taken a
-central role as it allows computational chemists to construct virtual
-libraries and test hypotheses, and it provides guidance in the design
-of optimal combinatorial experiments.
+Enumerating molecules is not only an interesting academic exercise but has
+practical utility as well. The foremost application of enumeration is \emph{structure
+elucidation} of natural compounds such as metabolites. Metabolites are small molecules that can be found an organism
+(such as a human) and which in many cases indicate the state the organism is in---such as indicating a disease, or a
+fasted state, etc.---and with a sufficiently large database the idea is to be able to detect the chemical composition of
+body fluids and to arrive at conclusions regarding the particular state of the organism at this point in time.
+Ideally, the wishful bench chemist collects experimental data
+from an unknown compound, the data is fed into a code, and the resulting
+structure is returned. Although that streamlined process may not yield a unique
+solution, there are commercial software products such as MOLGEN
+(\url{http://molgen.de}) that can, for instance, list all structures matching a given
+molecular formula, or mass spectrometry (MS) data. Another important application
+of enumeration is \emph{molecular design}. Here the problem is to design compounds
+(drugs, for instance) that optimize some physical, chemical, or biological
+property or activity. Although less prolific and more recent than structure
+elucidation, molecular design has introduced some novel solutions to molecular
+enumeration. Finally, with the advent of \emph{combinatorial chemistry} (the rapid synthesis of different but related molecules), molecular
+enumeration has taken a central role as it allows computational chemists to
+construct virtual libraries and test hypotheses, and it provides guidance in the
+design of optimal combinatorial experiments.
 
-The major difficulty with enumeration (apart from the size of the
-search space, which is thought to be in the order of $10^{60}$
-molecules for 30 heavy atoms \cite{Bohacek:1996ve}) is that the
-\emph{in silico} representation of molecules, the so-called molecular
-graphs, are labeled objects (i.e. labeled graphs), while the atoms of
-a molecule are of course not uniquely labeled. The mathematical
-concept to tackle this problem is to consider orbits of labeled
-molecular graphs under the operation of the symmetric group. While the
-complexity of molecular graph enumeration remains an open problem it
-has nonetheless been shown that molecules can be sampled
-efficiently \cite{goldberg1999}. Sampling procedures based on the
-Metropolis or the Genetic Algorithms have been developed and used to
-elucidate natural compounds from NMR data.
+The major difficulty with enumeration (apart from the size of the search space, which is thought to be in the order of
+$10^{60}$ molecules for 30 heavy atoms) is that the \emph{in silico}
+representation of molecules, the so-called molecular graphs, are
+labeled objects (i.e. labeled graphs), while the atoms of a molecule
+are of course not uniquely labeled. The mathematical concept to tackle
+this problem is to consider orbits of labeled molecular graphs under
+the operation of the symmetric group. However, when enumerating molecular
+structures, one has thus to solve the so--called isomorphism problem, where molecules that seem different are actually
+equivalent (i.e., their atoms can be mapped on top of each other by keeping the connectivity between atoms intact).
+The good news is that isomorphism can be solved in polynomial time for
+molecular graphs due to the fact that molecules belong to a restricted
+class of graphs known as \emph{bounded valence graphs}; however, that
+same restriction prevents applying to a molecular graph the polynomial
+delay enumeration results obtained for general graphs (Chapter 8 in~\cite{faulon2010}). While the complexity of molecular graph
+enumeration remains an open problem it has nonetheless been shown that
+molecules can be sampled efficiently~\cite{goldberg1999}. Sampling
+procedures based on the Metropolis or the Genetic Algorithms have been
+developed and used to elucidate natural compounds from NMR data.
 
@@
 enumeration tools to generate large chemical libraries such as GDB-13,
-comprising almost a billion chemical structures~\cite{GDB}. Yet, the
-current enumeration software products do not generally produce
+comprising almost a trillion chemical
+structures~\cite{GDB}. Yet,
+the current enumeration software products do not generally produce
 stereoisomers or tautomers; these require specific enumeration
 procedures, which are still the subject of investigations by the
-cheminformatics community. In addition, GDB-13 only enumerates
-molecules up to 13 heavy atoms---given the possibly exponential
-increase in search space with the number of atoms, it seems unlikely
-that the practically relevant chemical space of about 30 heavy atoms
-will be exhaustively enumerated in the near future.
+cheminformatics community. In addition, GDB-13 only enumerates molecules up to 13 heavy atoms---given the possibly
+exponential increase in search space with the number of atoms, it seems unlikely that the practically relevant chemical
+space of about 30 heavy atoms will be exhaustively enumerated in the near future.
 
-Apart from analytical chemistry and compound design, enumeration plays
-a role is in the \emph{generation of chemical reaction networks}. The
-problem here consists of enumerating all the possible compounds that
-can be produced by applying reaction rules to a set of initial
-molecules. By reversing the reaction rules one can also find sets of
-starting compounds necessary for producing a given target. This latter
-process is known in chemistry as \emph{retrosynthesis}, ever since the
-works of E.J. Corey, who received the Nobel Prize in Chemistry in
+%GDB-13 along with other databases such as PubChem (NCBI/NIH) or ZINC
+%(UCSF) are valuable resources to search for new drugs. For instance
+%the well known ligand-based design approach uses these databases along
+%with structure similarity tools to discover new compounds binding to
+%biological targets (protein or nucleic acid). Compounds retrieved
+%through ligand-based design, however, must already be present in
+%databases, which is a limitation of the technique. To overcome that
+%limitation, algorithms have recently been developed to enumerate all
+%compounds in the chemical space matching circular \cite{faulon2003} or
+%path fingerprints \cite{fujiwara2008}, these fingerprints being
+%%popular tools to measure similarity. Enumerating molecules from
+%fingerprints, or more generally enumerating molecules from molecular
+%descriptors, is still very much under development. At stake is the
+%design of new chemicals or drugs that have not yet been synthesized.
+%
+Apart from analytical chemistry and compound design, enumeration plays a role is in the \emph{generation
+  of chemical reaction networks}. The problem here consists of
+enumerating all the possible compounds that can be produced by
+applying reaction rules to a set of initial molecules. By reversing
+the reaction rules one can also find sets of starting compounds
+necessary for producing a given target. This latter process is known
+in chemistry as \emph{retrosynthesis}, ever since the works of
+E.J. Corey, who received the Nobel Prize in Chemistry in
 1990. Designing new drugs or chemicals, understanding the kinetics of
@@
 requires encompassing both chemical and sequence information and the
-development of tools that are at the interface between cheminformatics
-and bioinformatics.
+development of tools that are at the interface between
+cheminformatics and bioinformatics.
 
@@
 
-The basis of predictive modeling in cheminformatics is that the
-biological activity of a molecule is a function of its chemical
-structure. Together with the \emph{similarity property principle}
-\cite{Johnson:1990qf} (structurally similar molecules will exhibit  
-similar properties), the goal of any modeling approach is to capture
-and characterize the correlations between structural features and the
-observed biological activities. Simultaneously, such approaches must
-also describe the likelihood of error when using such
-models for decision making.  A variety of approaches can be employed
-to assess the error in (or conversely, the reliability or confidence of) a
-prediction, ranging from statistical approaches (XXX) to more
-empirical approaches such as defining an \emph{applicability domain} -
-the region of the input (i.e., chemical structures) that can be
-reliably predicted, usually deliniated using some form of similarity
-to the training set.
+The basis of predictive modeling in cheminformatics is that
+\emph{chemical structure implies biological activity}. Thus, the goal
+of any modeling approach is to capture and characterize the
+correlations between structural features and the observed biological
+activities. The goal is to describe cause-effect relationships with
+the lowest possible risk, and also to describe the likelihood of error
+(confidence) when using such models for decision making.
+Figure~\ref{figure:applicability-domain} (in the Appendix) illustrates the use of substructure matching to produce a space of chemical structures that are likely to share similar biological activity; this is called the generation of an \emph{applicability domain}.
 
-In many cases, the activity of a small molecule is due to its
-interaction with a receptor. Traditionally, QSAR' \cite{Hansch:1962vn,
-  Free:1964ys} approaches do not take into account receptor features,
-focusing only on small molecule features, and therefore lose valuable
-information on ligand-receptor interactions. As a result, techniques
-such as docking, pharmacophore modeling and proteochemometric methods
-have been designed to take into account both ligand and receptor
-structures.
+In many cases, the activity of a small molecule is due to its interaction with
+a receptor. Traditional ``QSAR'' approaches ignore receptor interactions,
+focusing only on small molecule features, and therefore loose valuable
+information. As a result, techniques such as proteochemometric
+modeling have been designed to take into account both ligand and
+receptor structures.
 
 The first step in the predictive modeling of biological activities is
-to describe compounds in a way amenable to computerized data mining,
-while at the same time capturing information relevant for the property
-one attempts to model. Here multiple approaches exist: Either we
-transform molecular graphs into a numerical vector representation
-(a.k.a. \emph{descriptors} or \emph{features}) of chemical structures,
-or we use algorithms for comparing molecular graphs directly
-(\emph{molecule kernel}). In the first case thousands of descriptors
-(many of which are equivalent) have been described
-\cite{todeschini2000}, and a variety of tools are available for their
-calculation. Given a set of molecules, observed activities and a
-descriptor vector, the problem can be considered a traditional
-statistical modeling problem. In the second case no explicit vector
-representation is required, which prevents the challenge of
-transforming the molecular graph into a vector without losing
-information. On the other hand, the implicit kernel encoding imposes
-the challenge that we have to compare all compounds against all
-compounds, which imposes large-scale mining challenge, which risks
-overrunning available processing time and space. Note that the public
-PubChem dataset alone contains over 30 million chemical compounds and
-over 70 million substances.
+to describe compounds in a way amenable to computerized data mining, while at the same time capturing information
+relevant for the property one attempts to model. Here multiple approaches exist: Either we transform molecular graphs
+into a
+numerical vector representation (a.k.a. \emph{descriptors} or
+\emph{features}) of chemical structures, or we use algorithms for
+comparing molecular graphs directly (\emph{molecule kernel}). In the
+first case thousands of descriptors (many of which are equivalent)
+have been described \cite{todeschini2000},
+and a variety of tools are available for their calculation. Given a
+set of molecules, observed activities and a descriptor vector, the
+problem can be considered a traditional statistical modeling
+problem. In the second case no explicit vector representation is
+required, which prevents the challenge of transforming the molecular
+graph into a vector without losing information. On the other hand, the
+implicit kernel encoding imposes the challenge that we have to compare all compounds against
+all compounds, which imposes large-scale mining challenge, which risks overrunning available
+processing time and space. Note that the public PubChem dataset alone
+contains over 30 million chemical compounds and over 70 million substances.
 
-In order to judge which predictive modeling approach is applicable in
-every case, we firstly have to consider the goal of a the model in a
-cheminformatics setting. In some cases, pure predictive ability is
-desired -- such as in a virtual screening setting. (Basically, only
-the reliability of the model matters, but other aspects such as
-interpretability are of minor importance.) For such cases, the
+In order to judge which predictive modeling approach is applicable in every case, we firstly have to consider the goal
+of a the model in a cheminformatics setting. In some cases, pure predictive ability is
+desired---such as in a virtual screening setting. (Basically, only the reliability of the model matters, but other
+aspects such as interpretability are of minor importance.) For such cases, the
 complexity of the modeling approach is immaterial, as it leads to
@@
 \emph{in silico} compounds into real-world compounds. In such cases,
-``black-box methods'' can be less useful, though attempts have been
-made to interpret vectorial models built using neural networks,
-na\"{i}ve Bayes, Support Vector Machines or random forests as well.
+``black-box methods'' can be less useful, though attempts have been made
+to interpret vectorial models built using neural networks, Support Vector Machines or
+random forests as well.
 
-Secondly, the concept of model applicability -- the question of when
-the prediction of the model for a new object is reliable -- has grown
-in importance with the increasing use of predictive models in
-regulatory settings. A variety of methods have been developed that
-attempt to characterize the domain of a model. These methods not only
-determine whether models are applicable, but can also be used for
-deciding if additional biological experiments are required for
-reducing the prediction error on certain compound classes.
+Secondly, the concept of model applicability---the question of when the prediction of the model
+for a new object is reliable---has grown in importance with
+the increasing use of predictive models in regulatory settings. A
+variety of methods have been developed that attempt to characterize
+the domain of a model. These methods not only determine whether models
+are applicable, but can also be used for deciding if additional
+biological experiments are required for reducing the risk on certain
+compound classes.
 
@@
 simultaneously. Multi-conformer modeling has been addressed in the
-4D-QSAR methodology described by Hopfinger and co-workers
-\cite{Albuquerque:1998ys}.  Computer science techniques such as
-\emph{multiple-instance learning} could also be applied to the
-multi-conformer problem. More recently, simultaneous modeling of
-receptor and ligand has been addressed by the proteochemometric
-approach as first reported by Lapinsh \textit{et
-  al.}~\cite{lapinsh2001}.
+4D-QSAR methodology described by Hopfinger and co-workers.  Computer
+science techniques such as \emph{multiple-instance learning} could
+also be applied to the multi-conformer problem. More recently,
+simultaneous modeling of receptor and ligand has been addressed by the
+proteochemometric approach as first reported by Lapinsh
+\textit{et al.}~\cite{lapinsh2001}.
 
-With the advent of high-throughput screening technologies, large
-libraries of compounds can be screened against multiple targets in an
-efficient manner. Such panel assays provide a broad, systems-level
-view of small molecule activities.  Models developed on such data
-afford us the opportunity to identify targets, characterize off-target
-effects and so on. However, most approaches to this problem are rather
-simplistic since they develop multiple individual models
-\cite{Chen:2010zr}, leading to multiple, independent predictions for a
-given input molecule.  Instead, one could imagine an approach that
-takes into account the covariance structure of the multiple observed
-activities and the structural descriptors within a single model. Such
-an approach could lead to more robust predictions for panel
-assays. Finally, this approach also better reflects clinically
-relevant compound profiles~\cite{kuhn2010} and a ``personalized
-medicine'' concept (\emph{e.g.}, drug-drug interaction profiles).
+With the advent of high-throughput screening technologies, large libraries of
+compounds can be screened against multiple targets in an efficient manner. Such
+panel assays provide a broad, systems-level view of small molecule activities.
+Models developed on such data afford us the opportunity to identify targets,
+characterize off-target effects and so on. However, most approaches to this
+problem are rather simplistic since they develop multiple individual models.
+Instead one could imagine an approach that takes into account the correlations
+in observed activities against multiple targets within a single model. Such an
+approach should lead to more robust predictions for panel assays. Finally,
+this also better reflects clinically relevant compound profiles~\cite{kuhn2010}
+and a ``personalized medicine'' concept (\emph{e.g.}, drug-drug interaction profiles).
 
@@
 
-The relevance of scientific knowledge management is increasing not
-only for reducing the risk in current research, but also for enabling
-new collaboration/innovation opportunities with internal partners and
-a rapidly growing number of external/public partners. In fact, in
-cheminformatics and chemistry, scientists already switched many years
-ago from paper lab-notebooks to online collaboration platforms, called
-ELNs (Electronic Lab Notebooks). So, what were the drivers behind
-chemists, and companies, adopting ``Enterprise 2.0,'' a social online
-collaboration culture?
+The relevance of scientific knowledge management is increasing not only for
+reducing the risk in current research, but also for enabling new
+collaboration/innovation opportunities with internal partners and a rapidly growing
+number of external/public partners. In fact, in cheminformatics and chemistry,
+scientists already switched many years ago from paper lab-notebooks to
+online collaboration platforms, called ELNs (Electronic Lab Notebooks). So, what were the drivers behind chemists, and
+companies, adopting ``Enterprise 2.0,'' a social online collaboration culture?
 
 Before the year 2000, many chemists were still using paper
-lab-notebooks, multiple compound registration, and search
-tools. Compared to today there were many pitfalls and drawbacks
-associated with this. The overall architecture was too inflexible to
-adopt to fast changing data standards, scientists spent too much time
-with administrative work, chemical data quality was inconsistent, an
-alignment with other working groups was inefficient (\emph{e.g.}, for
-running analytical experiments), a legal dispute required to find data
-in paper lab-notebooks, data synchronization between systems caused
-lag times and access delays, and local silos hampered an efficient
-global collaboration. Now, chemists prefer ELNs, because they can run
-searches in the data across the organization. This does not only allow
-to find experts faster, but also allows access to highly trusted data
-with approved quality workflows. Since this also gives access to
-analytical and legal data directly every new ELN release is aiming at
-increasing the integration of scientific and business workflows for
-reducing the risk due to irrational or bounded-rational decision
-making and increasing decision making efficiency.
+lab-notebooks, multiple compound registration, and search tools. Compared to
+today there were many pitfalls and drawbacks associated with this. The overall architecture
+was too inflexible to adopt to fast changing data standards, scientists spent
+too much time with administrative work, chemical data quality was inconsistent,
+an alignment with other working groups was inefficient (\emph{e.g.}, for running
+analytical experiments), a legal dispute required to find data in paper
+lab-notebooks, data synchronization between systems caused lag times and access
+delays, and local silos hampered an efficient global collaboration. Now,
+chemists prefer ELNs, because they can run searches in the data across the organization. This does
+not only allow to find experts faster, but also allows access to highly trusted
+data with approved quality workflows. Since this also gives access to
+analytical and legal data directly every new ELN release is aiming at increasing
+the integration of scientific and business workflows for reducing the risk due
+to irrational or bounded-rational decision making and increasing decision making
+efficiency.
 
-ELNs are typically much more accepted and used within the
-cheminformatics domain than in the bioinformatics domain, which is
-picking up speed lately, the question being why this is the case. In
-chemistry, every experiment requires a lot of expert
+ELNs are typically much more accepted and used within the cheminformatics
+domain than in the bioinformatics domain, which is picking up speed lately, the
+question being why this is the case. In chemistry, every experiment requires a lot of expert
 knowledge. Therefore, approaches for reducing risks by communicating
-with fellow chemists about similar reactions or compounds is
-critical. This requires capturing the reaction details, compound
-details, and time consuming analysis details which ensure quality
-data, such that these data can be searched by others.
+with fellow chemists about similar reactions or compounds is critical. This
+requires capturing the reaction details, compound details, and time consuming
+analysis details which ensure quality data, such that these data can be searched by others.
 
@@
 
-However, the future still leaves many challenges ahead. Using
-external/public databases with chemical and bioactivity data remains a
-challenge due to differences in ontologies, synchronization and
-maintenance, and efficient substructure and similarity search within
-complex data types (compounds). A collaboration with external parties,
-e.g. contract research, poses other problems with duplicate checking
-for compounds (e.g. tautomeric or salt structures) and efficient data
-synchronization, and most often a double-registration is required for
-fulfilling the data quality processes on both sides. If external
+However, the future still leaves many challenges ahead. Using external/public databases
+with chemical and bioactivity data remains a challenge due to differences in
+ontologies, synchronization and maintenance, and efficient
+substructure and similarity search within complex data types
+(compounds). A
+collaboration with external parties, e.g. contract research, poses other
+problems with duplicate checking for compounds (e.g. tautomeric or salt
+structures) and efficient data synchronization, and most often a double-registration is
+required for fulfilling the data quality processes on both sides. If external
 partners are small they often do not even have sufficient IT resources
-themselves and thus rely on external services. Cloud services could
-help not only to provide a service infrastructure for all parties
-involved, but also to provide the required private/public access
-toolbox. Furthermore, there are still many data sources not being
-indexed properly, since chemistry patents are often cryptic and
-image/text mining remains a puzzle. One could argue that the
-image/text mining in scientific journals is less challenging; in fact,
-it remains a huge challenge and many chemists do not trust automated
-non-expert validated mining sources. The closed CAS (Chemical Abstract
-Service) is a highly trusted source, and public chemistry and
-bioactivity space has to improve quality and interconnectivity to
-compare with it. Efforts like ChEMBL and PubChem-BioAssay are on the
-right track. Still, improving data quality and standards between
-public and closed sources will be absolutely critical for ensuring
-constant growth, usage, and collaboration scenarios between private
-and public parties.  In a business context we need to address ROI
-(return-of-investment) questions, not only for single institutions and
-companies, but also for collaboration networks of legal partners.  So,
-only when the technical and scientific questions are solved can we
-ensure that legal questions about contribution metrics will get enough
-room to grow. Solving challenges is one thing (science and technical),
-ensuring that the contributing partners get rewarded by their
-contribution another (legal).
+themselves and thus rely on external services. Cloud services could help
+not only to provide a service infrastructure for all parties involved,
+but also to provide the required private/public access toolbox. Furthermore,
+there are still many data sources not being indexed properly, since chemistry
+patents are often cryptic and image/text mining remains a puzzle. One could
+argue that the image/text mining in scientific journals is less challenging; in
+fact, it remains a huge challenge and many chemists do not trust automated
+non-expert validated mining sources. The closed CAS (Chemical Abstract Service)
+is a highly trusted source, and public chemistry and bioactivity space has to
+improve quality and interconnectivity to compare with it. Efforts like ChEMBL and PubChem-BioAssay
+are on the right track. Still, improving data quality and standards between
+public and closed sources will be absolutely critical for ensuring constant
+growth, usage, and collaboration scenarios between private and public parties.
+In a business context we need to address ROI (return-of-investment) questions, not only for single
+institutions and companies, but also for collaboration networks of legal partners.
+So, only when the technical and scientific questions are solved can we ensure that
+legal questions about contribution metrics will get enough room to grow. Solving challenges is one thing (science and
+technical),
+ensuring that the contributing partners get rewarded by their contribution another (legal).
 
-\section{The Role of Open Source \& Open Data}
-\label{sec:role-open-source}
+\section{Support for the Development of Novel Algorithms}
+\label{sec:development-support}
 
-A key aspect of cheminformatics is that much of its development
-requires access to collections of molecular structures (and, usually,
-associated bioactivity data). As has been noted, it is only recently
-that large structure and bioactivity data collections have become
-available. As a result, this enables much more robust validation of
-cheminformatics methodologies as well as large scale
-\emph{benchmarking of algorithms}. The latter is especially relevant
-for the plethora of data mining techniques that are employed in
-cheminformatics. At this point, the nature of Open Data focuses
-primarily on structure and activity data types. On the other hand,
-there is a distinct lack of textual data (such as journal articles)
-that are ``Open'' in nature. While PubMed abstracts serve as a proxy
-for journal articles, text mining methods in cheminformatics are
-hindered by not being able to mine the full text of many scientific
-publications.  Interestingly, patent information is publicly
-accessible (cf. Google Patents) and there have been efforts to mine
-these for chemical information using Open tools.
+For researchers interested in the development of new algorithms in the
+field, a software library that handles chemical structures and related
+data is essential. A wide variety of such chemical toolkits are
+available, both commercial (which may offer free academic
+licenses) and Open Source.
 
-Obviously freely accessible data is important for research; but
-equally important are the tools to work with chemical data---more
-specifically, chemical structures. Chemical toolkits are the workhorses
-of cheminformatics research and a variety of toolkits are available,
-both commercial (which, in many cases, provide free academic
-licenses) and Open Source. In this article we do not debate the
-arguments for and against Open Source; rather, we focus on Open Source
-tools, since this ensures that a reader will be able to explore
+Due to the importance of cheminformatics to the pharmaceutical
+industry in particular, numerous commercial vendors provide libraries,
+applications and database cartridges to handle a wide variety of
+needs including Accelrys, BioSolveIT, ChemAxon, Chemical Computing Group,
+Daylight, Open Eye, Schrodinger, Tripos, and Xemistry. More recently
+there has been a growth in the development of Open
+Source cheminformatics software, and this presents opportunities for
+the rapid development and implementation of novel algorithms which can
+build on the existing Open Source ecosystem. In this article we do not debate the
+arguments for and against Open Source; we focus on Open Source
+tools simply because this ensures that a reader will be able to explore
 cheminformatics problems with a minimum of encumbrances.
-
-\subsection{Open Source Chemical Toolkits}
-\label{sec:toolkits}
 
@@
 rise in the amount of open-source cheminformatics software. One
-contribution to this has been the Blue Obelisk group
+contribution to this has been the Blue Obelisk
+group~\cite{BlueObelisk2011}
 (\url{http://blueobelisk.org}) of chemists, who have worked together to
 create interoperable open-source chemistry software (among other
-activities). Now there are a number of open-source toolkits available
+activities). Now there are a number of Open-Source chemical toolkits available
 including the Chemistry Development
-Kit (CDK)~\cite{steinbeck2003}, Open Babel, RDKit and Indigo. These toolkits are written in
+Kit (CDK)~\cite{steinbeck2003}, Open Babel~\cite{openbabel2011}, RDKit and Indigo. These toolkits are written in
 Java or C++ (though toolkits written in the latter invariably have
@@
 (\url{http://avogadro.openmolecules.net}) uses Open Babel, while the
-molecular workbench Bioclipse \cite{Bioclipse2} uses the CDK.
+molecular workbench Bioclipse~\cite{Bioclipse2} uses the CDK.
 
@@
 chemical file formats. Table~\ref{tab:features} (in the Appendix) compares features offered by the open-source toolkits currently available.
+Given that there is ample scope for software engineering (algorithm
+implementation, code quality and analysis, build systems etc), both
+the CDK and Open Babel are open to new contributions. Both are driven
+via public, active mailing lists and public issue tracking systems.
 
-While algorithmic challenges are certainly faced by toolkits, software
-engineering challenges probably affect the use and development of the
-toolkits on a daily basis. Examples include verifying code coverage,
-thread safety, documentation quality and so on. Some projects such as
-the CDK employ a variety of tools to address some of these issues.
-Examples include the use of PMD to analyze code quality and JUnit to
-perform unit testing. At the level of implementation, there are still
-inefficiencies in some toolkits. For example the CDK employs a JJTree
-parser to implement the SMARTS specification, but employs a hand built
-parser to handle SMILES. The latter is essentially a complex tangle of
-\emph{if \ldots then} statements and is not easily maintainable.
-
-It is also important to note that certain challenges faced by nearly
+It is important to note that certain challenges faced by nearly
 all cheminformatics toolkits stem from the graph representation of
@@
 merely derives from the chemical graph, but typically also from the 3D
-geometry of the chemical. However, more challenging is that chemical
-representation of a molecule really needs to capture non-static
-graphs, in order to take into account delocalisation and the tautomeric
-phenomena molecules can undergo in practice. It is important to
+geometry of the chemical. It is important to
 realize that although chemical representations have increased in
 sophistication over the years, many chiral molecules cannot be
-satisfactorily handled by current representation systems.
+satisfactorily handled by current representation systems. Perhaps a more challenging aspect is that
+the chemical representation of a molecule really needs to capture non-static
+graphs, in order to take into account delocalisation and the tautomeric
+phenomena molecules can undergo in practice.
 
-Given that there is ample scope for software engineering (algorithm
-implementation, code quality and analysis, build systems etc), both
-the CDK and Open Babel are open to new contributions. Both are driven
-via public, active mailing lists and public issue tracking systems.
-Development questions are actively and openly discussed to arrive at a
-consensus and contributors are encouraged to get involved with any of
-the following areas: development of new features, improvement of
-existing features, performance optimization, fixing bugs,
-documentation, website development, reporting bugs, and helping users
-on the mailing list. Anyone interested should email the respective
-developers' mailing list (either \texttt{cdk-devel@lists.sf.net} or
-\texttt{openbabel-devel@lists.sf.net}) with their interests and
-background, and a mentor will be assigned to help them join the
-project.
-%
+Of equal importance to the development of new algorithms is freely
+available data on which to train and test new methods. As has been noted, recently
+large structure and bioactivity data collections have become
+available. As a result, this enables much more robust validation of
+cheminformatics methodologies as well as large scale \emph{benchmarking of
+algorithms}. The latter is especially relevant for the plethora of
+data mining techniques that are employed in cheminformatics. At this
+point, the nature of Open Data focuses primarily on structure and
+activity data types. On the other hand, there is a distinct lack of
+textual data (such as journal articles) that are ``Open'' in nature. While
+PubMed abstracts serve as a proxy for journal articles, text mining
+methods in cheminformatics are hindered by not being able to mine the full text of many scientific publications.
+Interestingly, patent information is publicly accessible (cf. Google
+Patents) and there have been efforts to mine these for chemical
+information using Open tools.
 
@@
 1960s, with the creation of DENDRAL, the first expert system of any
-kind.  Much of the development of cheminformatics from the 1960s until
-the new millenium was performed in secret, because of competition in
-the pharmaceutical industry.  While chemists were among the first
-scientists to embrace web-based tools of knowledge management,
-research into fundamental algorithmic questions of structure
-enumeration and database management (and a host of other topics) have
-suffered, because of limited academic access to either tools or data.
-As we enter the second decade of the 21st century, however, this
-landscape is rapidly shifting.  There now exist free-of-charge
-open-source cheminformatics toolkits, and open databases with tens of
-millions of compounds.  We believe the prospects for academic and
-industrial collaboration between chemists and computer scientists are
-bright, and we would like to encourage more computer scientists to
-participate in cheminformatics research.  There is an admittedly steep
-learning curve, as the structural issues faced by chemists are often
-too detailed and subtle to admit as clean an abstraction as
-``algorithms on a string,'' the way many bioinformatics algorithms can
-be abstracted into a computer science framework.  Nevertheless,
-because of the changing landscape, there are many research questions
-we are only now in position to try to answer: for a theorist,
-graph-theoretic questions of three-dimensional enumeration; for a
-database designer, more effective search algorithms and ontologies;
-for a practical programmer, expansion of the many open-source
-cheminformatics projects.  If more chemists think algorithmically, and
-more computer scientists think chemically we are much better
-positioned taking not only ``simple'' chemicals into account (only one
-identifier and isomer possible per molecule), but also their complex
-relationships, transformations, and combinatorial challenges.
+kind.  Much of the development of cheminformatics from the 1960s until the new millenium
+was performed in secret, because of competition in the pharmaceutical
+industry.  While chemists were among the first scientists to embrace
+web-based tools of knowledge management, research into fundamental
+algorithmic questions of structure enumeration and database management
+(and a host of other topics) have suffered, because of limited
+academic access to either tools or data.  As we enter the second
+decade of the 21st century, however, this landscape is rapidly
+shifting.  There now exist free-of-charge open-source cheminformatics
+toolkits, and open databases with tens of millions of compounds.  We
+believe the prospects for academic and industrial collaboration
+between chemists and computer scientists are bright, and we
+would like to encourage more computer scientists to participate in
+cheminformatics research.  There is an admittedly steep learning
+curve, as the structural issues faced by chemists are often too
+detailed and subtle to admit as clean an abstraction as ``algorithms
+on a string,'' the way many bioinformatics algorithms can be abstracted into a computer science
+framework.  Nevertheless, because of the changing landscape, there are
+many research questions we are only now in position to try to answer:
+for a theorist, graph-theoretic questions of three-dimensional
+enumeration; for a database designer, more effective search algorithms
+and ontologies; for a practical programmer, expansion of the many
+open-source cheminformatics projects.  If more chemists think
+algorithmically, and more computer scientists think ``chemically,''
+we are much better positioned taking not only ``simple'' chemicals
+into account (only one identifier and isomer possible per molecule),
+but also their complex relationships, transformations, and
+combinatorial challenges.
 
-In 2010, drug company Glaxo made public over 13,000 compounds that
-they have investigated as potential cures for
-malaria~\cite{glaxo-malaria}.  (Malaria has traditionally been a
-neglected disease in drug research, because the populations who
-suffer from it tend not to have enough money to be a stable consumer
-base for a new pharmaceutical.)  Perhaps we will now be able to find
-cures for some of these neglected diseases, by using the open source,
-open data approach to the computer science of chemical discovery.
+In 2010, drug company Glaxo made public over 13,000 compounds that they have investigated as potential cures for malaria~\cite{glaxo-malaria}.
+(Malaria has traditionally been a ``neglected'' disease in drug research, because the populations who suffer from it tend not to have enough money to be a stable consumer base for a new pharmaceutical.)  Perhaps we will now be able to find cures for some of these neglected diseases, by using the open source, open data approach to the computer science of chemical discovery.
 
@@
 We would like to thank Danny Verbinnen for sharing his insights into
-ELNs.  This article was written while A. Sterling was visiting the
+ELN's.  This article was written while A. Sterling was visiting the
 Department of Electrical Engineering and Computer Science,
-Northwestern University, supported in part by NSF grant
-CCF-1049899. A. Bender thanks Unilever for funding.
+Northwestern University, supported in part by NSF grant CCF-1049899. A. Bender thanks Unilever for funding.
 
@@
 %
-% \begin{figure}
-% \centering
-% \includegraphics[height=1.3in]{platins.png}
-% \caption{The generated InChI is the same for the stereoisomers cisplatin and transplatin.}
-% \label{figure:cistrans}
-% \end{figure}
-% %
-% \begin{figure}
-% \centering
-% \includegraphics[height=2in]{CACM_ApplicabilityDomain.pdf}
-% \caption{Illustration of the ``Applicability Domain'' concept.
+\begin{figure}
+\centering
+\includegraphics[height=1.3in]{platins.png}
+\caption{The generated InChI is the same for the stereoisomers cisplatin and transplatin.}
+\label{figure:cistrans}
+\end{figure}
+%
+\begin{figure}
+\centering
+\includegraphics[height=2in]{CACM_ApplicabilityDomain.pdf}
+\caption{Illustration of the ``Applicability Domain'' concept.
 In order to estimate whether model predictions are reliable, it is crucial to define areas of ``chemical space'' where the model is applicable, and where it is not.
@@
 The (relatively similar) molecules to the left are likely to be included in the Applicability Domain of the model, while the (more dissimilar) molecules to the right are likely located outside this domain, hence the generated model is probably not applicable.}
-% \label{figure:applicability-domain}
-% \end{figure}
-% %
+\label{figure:applicability-domain}
+\end{figure}
+%
 \end{document}

### 2011-11-10 rebuttal.tex

@@
 
+\comment{p2, c2, l57: it might be worthwhile here clarifying in the text
+precisely the properties that require satisfaction to deliver a small
+molecule therapeutic, not ?promising?. A drug must be safe and
+efficacious ultimately; perhaps this should be mentioned first
+followed by the typical pitfalls and how they are assessed?}
+\response{The text has been updated to provide some examples of
+  properties that would characterize a therapeutically useful small molecule}
+
 \comment{P3, c1, l41: the authors need to cover the Morgan algorithm

### 2011-11-10 paper.tex

--- 2011-11-10 paper.tex
@@
 
-\section{The Role of Open Source \& Open Data}
-\label{sec:role-open-source}
-
-A key aspect of cheminformatics is that much of its development
-requires access to collections of molecular structures (and, usually,
-associated bioactivity data). As has been noted, it is only recently
-that large structure and bioactivity data collections have become
-available. As a result, this enables much more robust validation of
-cheminformatics methodologies as well as large scale
-\emph{benchmarking of algorithms}. The latter is especially relevant
-for the plethora of data mining techniques that are employed in
-cheminformatics. At this point, the nature of Open Data focuses
-primarily on structure and activity data types. On the other hand,
-there is a distinct lack of textual data (such as journal articles)
-that are ``Open'' in nature. While PubMed abstracts serve as a proxy
-for journal articles, text mining methods in cheminformatics are
-hindered by not being able to mine the full text of many scientific
-publications.  Interestingly, patent information is publicly
-accessible (cf. Google Patents) and there have been efforts to mine
-these for chemical information using Open tools.
-
-Obviously freely accessible data is important for research; but
-equally important are the tools to work with chemical data---more
-specifically, chemical structures. Chemical toolkits are the workhorses
-of cheminformatics research and a variety of toolkits are available,
-both commercial (which, in many cases, provide free academic
-licenses) and Open Source. In this article we do not debate the
-arguments for and against Open Source; rather, we focus on Open Source
-tools, since this ensures that a reader will be able to explore
+\section{Support for the Development of Novel Algorithms}
+\label{sec:development-support}
+
+For researchers interested in the development of new algorithms in the
+field, a software library that handles chemical structures and related
+data is essential. A wide variety of such chemical toolkits are
+available, both commercial (which may offer free academic
+licenses) and Open Source.
+
+Due to the importance of cheminformatics to the pharmaceutical
+industry in particular, numerous commercial vendors provide libraries,
+applications and database cartridges to handle a wide variety of
+needs including Accelrys, BioSolveIT, ChemAxon, Chemical Computing Group,
+Daylight, Open Eye, Schrodinger, Tripos, and Xemistry. More recently
+there has been a growth in the development of Open
+Source cheminformatics software, and this presents opportunities for
+the rapid development and implementation of novel algorithms which can
+build on the existing Open Source ecosystem. In this article we do not debate the
+arguments for and against Open Source; we focus on Open Source
+tools simply because this ensures that a reader will be able to explore
 cheminformatics problems with a minimum of encumbrances.
-
-\subsection{Open Source Chemical Toolkits}
-\label{sec:toolkits}
 
@@
 chemical file formats. Table~\ref{tab:features} (in the Appendix) compares features offered by the open-source toolkits currently available.
-
-While algorithmic challenges are certainly faced by toolkits, software
-engineering challenges probably affect the use and development of the
-toolkits on a daily basis. Examples include verifying code coverage,
-thread safety, documentation quality and so on. Some projects such as
-the CDK employ a variety of tools to address some of these issues.
-Examples include the use of PMD to analyze code quality and JUnit to
-perform unit testing. At the level of implementation, there are still
-inefficiencies in some toolkits. For example the CDK employs a JJTree
-parser to implement the SMARTS specification, but employs a hand built
-parser to handle SMILES. The latter is essentially a complex tangle of
-\emph{if \ldots then} statements and is not easily maintainable.
-
-It is also important to note that certain challenges faced by nearly
+Given that there is ample scope for software engineering (algorithm
+implementation, code quality and analysis, build systems etc), both
+the CDK and Open Babel are open to new contributions. Both are driven
+via public, active mailing lists and public issue tracking systems.
+
+It is important to note that certain challenges faced by nearly
 all cheminformatics toolkits stem from the graph representation of
@@
 merely derives from the chemical graph, but typically also from the 3D
-geometry of the chemical. However, more challenging is that chemical
-representation of a molecule really needs to capture non-static
-graphs, in order to take into account delocalisation and the tautomeric
-phenomena molecules can undergo in practice. It is important to
+geometry of the chemical. It is important to
 realize that although chemical representations have increased in
 sophistication over the years, many chiral molecules cannot be
-satisfactorily handled by current representation systems.
-
-Given that there is ample scope for software engineering (algorithm
-implementation, code quality and analysis, build systems etc), both
-the CDK and Open Babel are open to new contributions. Both are driven
-via public, active mailing lists and public issue tracking systems.
-Development questions are actively and openly discussed to arrive at a
-consensus and contributors are encouraged to get involved with any of
-the following areas: development of new features, improvement of
-existing features, performance optimization, fixing bugs,
-documentation, website development, reporting bugs, and helping users
-on the mailing list. Anyone interested should email the respective
-developers' mailing list (either \texttt{cdk-devel@lists.sf.net} or
-\texttt{openbabel-devel@lists.sf.net}) with their interests and
-background, and a mentor will be assigned to help them join the
-project.
-%
+satisfactorily handled by current representation systems. Perhaps a more challenging aspect is that
+the chemical representation of a molecule really needs to capture non-static
+graphs, in order to take into account delocalisation and the tautomeric
+phenomena molecules can undergo in practice.
+
+Of equal importance to the development of new algorithms is freely
+available data on which to train and test new methods. As has been noted, recently
+large structure and bioactivity data collections have become
+available. As a result, this enables much more robust validation of
+cheminformatics methodologies as well as large scale \emph{benchmarking of
+algorithms}. The latter is especially relevant for the plethora of
+data mining techniques that are employed in cheminformatics. At this
+point, the nature of Open Data focuses primarily on structure and
+activity data types. On the other hand, there is a distinct lack of
+textual data (such as journal articles) that are ``Open'' in nature. While
+PubMed abstracts serve as a proxy for journal articles, text mining
+methods in cheminformatics are hindered by not being able to mine the full text of many scientific publications.
+Interestingly, patent information is publicly accessible (cf. Google
+Patents) and there have been efforts to mine these for chemical
+information using Open tools.
 

### 2011-11-10 rebuttal.tex
Update rebuttal. Added ref and updated history

@@
 
+\comment{P7, c2, l25: the authors should go back even further in the
+  history of cheminformatics here before DENDRAL. While this is an
+  excellent example of crossover, it is not by any means the founding
+  of our field. Aspects of mathematical chemistry should also be
+  mentioned in this article, which date back even further, such as
+  1894 with the publication of ?The Principles of Mathematical
+  Chemistry? published by Helm.}
+\response{As noted previously, we have included a supplemental
+  document that lists some of the milestones in the history of
+  cheminformatics. However, we note that the suggested reference by
+  Helm is not specifically related to cheminformatics. Rather, it is a
+  mathematical treatment of physical chemistry. Given that DENDRAL is
+  referenced in the context of structure enumeration, we feel that
+  inclusion of this reference would be somewhat irrelevant.}
+
 \comment{P7, c2, l51: why is chemically in inverted commas and not

### 2011-11-10 rebuttal.tex
some more updates

@@
   referenced in the context of structure enumeration, we feel that
-  inclusion of this reference would be somewhat irrelevant.}
+  inclusion of this reference would be somewhat irrelevant. We have
+  included this reference in the supplemental history}
 

### 2011-11-10 rebuttal.tex
adressed another reviewer point

@@
 
+\comment{P8, c1, l5: the users emphasise open source, open data but
+  this should be an article on all of cheminformatics not just a
+  particular subset.}  
+\response{We have restructured the section on
+  open source and open data. It is true that we still focus on open
+  source/open data, because this ensures that a reader will be able to
+  explore cheminformatics problems with a minimum of encumbrances}
+
 \end{document}

### 2011-11-10 paper.tex
Update rebuttal. Added ref and updated history
some reformatting

@@
 Enumerating molecules is a combinatorial problem that has fascinated
-chemists, computer scientists and mathematicians alike, for more than a
-century. It is one important risk-reduction element in still being
+chemists, computer scientists and mathematicians alike, for more than
+a century. It is one important risk-reduction element in still being
 able to encode closely-related molecules and for keeping the space
@@
 principles in graph theory and combinatorics were developed by Cayley,
-Polya and others. Even terms that are widely-used nowadays, like ``graph''
-and ``tree'', were originally coined in a chemistry context (Chapters 1 and
-8 in \cite{faulon2010}). More than four decades ago, Joshua Lederberg,
-Carl Djerassi, Edward Feigenbaum and others from Stanford University
-started developing codes to explicitly enumerate molecules. Their
-project, named DENDRAL, was funded by NASA with the goal of developing
-an automated system to elucidate compounds from spectral data. While
-studying this challenging problem, important concepts in computer
-science were developed. DENDRAL is generally quoted in artificial
-intelligence textbooks as a pioneer project that produced the first
-expert system~\cite{DENDRAL}.
+Polya and others. Even terms that are widely-used nowadays, like
+``graph'' and ``tree'', were originally coined in a chemistry context
+(Chapters 1 and 8 in \cite{faulon2010}). More than four decades ago,
+Joshua Lederberg, Carl Djerassi, Edward Feigenbaum and others from
+Stanford University started developing codes to explicitly enumerate
+molecules. DENDRAL is generally quoted in artificial intelligence
+textbooks as a pioneer project that produced the first expert
+system~\cite{DENDRAL}. However, even before that
 
@@
 The extent of open-source software in cheminformatics has somewhat
-lagged behind related fields such as bioinformatics, partly
-due to the existence of a lucrative pharmaceutical market for
-proprietary software. However, in recent years there has been a
-rise in the amount of open-source cheminformatics software. One
-contribution to this has been the Blue Obelisk
-group~\cite{BlueObelisk2011}
-(\url{http://blueobelisk.org}) of chemists, who have worked together to
-create interoperable open-source chemistry software (among other
-activities). Now there are a number of Open-Source chemical toolkits available
-including the Chemistry Development
-Kit (CDK)~\cite{steinbeck2003}, Open Babel~\cite{openbabel2011}, RDKit and Indigo. These toolkits are written in
-Java or C++ (though toolkits written in the latter invariably have
-bindings to a variety of languages). Such toolkits are used, for
-example, to read/write chemical file formats, manipulate chemical
-structures, measure similarity of molecules, search for substructures,
-and generate 2D diagrams. The underlying algorithms used include graph
-algorithms (e.g. maximal common substructure, canonical labelling of
-graphs), geometrical methods (e.g. Kabsch alignment) and vector
-manipulation (e.g. converting coordinates in various systems, 3D
-structure generation). As well as being of direct use to practising
+lagged behind related fields such as bioinformatics, partly due to the
+existence of a lucrative pharmaceutical market for proprietary
+software. However, in recent years there has been a rise in the amount
+of open-source cheminformatics software. One contribution to this has
+been the Blue Obelisk group~\cite{BlueObelisk2011}
+(\url{http://blueobelisk.org}) of chemists, who have worked together
+to create interoperable open-source chemistry software (among other
+activities). Now there are a number of Open-Source chemical toolkits
+available including the Chemistry Development Kit
+(CDK)~\cite{steinbeck2003}, Open Babel~\cite{openbabel2011}, RDKit and
+Indigo. These toolkits are written in Java or C++ (though toolkits
+written in the latter invariably have bindings to a variety of
+languages). Such toolkits are used, for example, to read/write
+chemical file formats, manipulate chemical structures, measure
+similarity of molecules, search for substructures, and generate 2D
+diagrams. The underlying algorithms used include graph algorithms
+(e.g. maximal common substructure, canonical labelling of graphs),
+geometrical methods (e.g. Kabsch alignment) and vector manipulation
+(e.g. converting coordinates in various systems, 3D structure
+generation). As well as being of direct use to practising
 cheminformaticians in their day-to-day work, many applications have
@@
 have certain distinguishing features. For example, the CDK implements
-a large collection of molecular descriptors. On the other hand,
-Open Babel has robust support for 3D structure generation and
-optimization and supports the interconversion of a huge number of
-chemical file formats. Table~\ref{tab:features} (in the Appendix) compares features offered by the open-source toolkits currently available.
-Given that there is ample scope for software engineering (algorithm
+a large collection of molecular descriptors. On the other hand, Open
+Babel has robust support for 3D structure generation and optimization
+and supports the interconversion of a huge number of chemical file
+formats. Table~\ref{tab:features} (in the Appendix) compares features
+offered by the open-source toolkits currently available.  Given that
+there is ample scope for software engineering (algorithm
 implementation, code quality and analysis, build systems etc), both
@@
 
-It is important to note that certain challenges faced by nearly
-all cheminformatics toolkits stem from the graph representation of
+It is important to note that certain challenges faced by nearly all
+cheminformatics toolkits stem from the graph representation of
 chemistry. This representation, while succinct and amenable to
@@
 merely derives from the chemical graph, but typically also from the 3D
-geometry of the chemical. It is important to
-realize that although chemical representations have increased in
-sophistication over the years, many chiral molecules cannot be
-satisfactorily handled by current representation systems. Perhaps a more challenging aspect is that
-the chemical representation of a molecule really needs to capture non-static
-graphs, in order to take into account delocalisation and the tautomeric
-phenomena molecules can undergo in practice.
+geometry of the chemical. It is important to realize that although
+chemical representations have increased in sophistication over the
+years, many chiral molecules cannot be satisfactorily handled by
+current representation systems. Perhaps a more challenging aspect is
+that the chemical representation of a molecule really needs to capture
+non-static graphs, in order to take into account delocalisation and
+the tautomeric phenomena molecules can undergo in practice.
 
 Of equal importance to the development of new algorithms is freely
-available data on which to train and test new methods. As has been noted, recently
-large structure and bioactivity data collections have become
-available. As a result, this enables much more robust validation of
-cheminformatics methodologies as well as large scale \emph{benchmarking of
-algorithms}. The latter is especially relevant for the plethora of
-data mining techniques that are employed in cheminformatics. At this
-point, the nature of Open Data focuses primarily on structure and
-activity data types. On the other hand, there is a distinct lack of
-textual data (such as journal articles) that are ``Open'' in nature. While
-PubMed abstracts serve as a proxy for journal articles, text mining
-methods in cheminformatics are hindered by not being able to mine the full text of many scientific publications.
-Interestingly, patent information is publicly accessible (cf. Google
-Patents) and there have been efforts to mine these for chemical
-information using Open tools.
+available data on which to train and test new methods. As has been
+noted, recently large structure and bioactivity data collections have
+become available. As a result, this enables much more robust
+validation of cheminformatics methodologies as well as large scale
+\emph{benchmarking of algorithms}. The latter is especially relevant
+for the plethora of data mining techniques that are employed in
+cheminformatics. At this point, the nature of Open Data focuses
+primarily on structure and activity data types. On the other hand,
+there is a distinct lack of textual data (such as journal articles)
+that are ``Open'' in nature. While PubMed abstracts serve as a proxy
+for journal articles, text mining methods in cheminformatics are
+hindered by not being able to mine the full text of many scientific
+publications.  Interestingly, patent information is publicly
+accessible (cf. Google Patents) and there have been efforts to mine
+these for chemical information using Open tools.
 
@@
 malaria~\cite{glaxo-malaria}.  (Malaria has traditionally been a
-neglected disease in drug research, because the populations who
-suffer from it tend not to have enough money to be a stable consumer
-base for a new pharmaceutical.)  Perhaps we will now be able to find
-cures for some of these neglected diseases, by using the open source,
-open data approach to the computer science of chemical discovery.
+neglected disease in drug research, because the populations who suffer
+from it tend not to have enough money to be a stable consumer base for
+a new pharmaceutical.)  Perhaps we will now be able to find cures for
+some of these neglected diseases, by using the open source, open data
+approach to the computer science of chemical discovery.
 

### 2011-11-10 rebuttal.tex
more updates

@@
 
+\comment{P3, c2, l42: is this true given the lower complexity of
+  molecular graphs compared with other much larger and denser graphs?
+  I think a reference to a known paper that states the problem clearly
+  would be of benefit here.}
+\response{While it is true that isomorphism on smaller graphs that are
+  characteristic of small molecule is in an absolute sense, not very
+  slow, there are a number of cases (polycyclic hydrocarbons, steroids
+  etc.) that can take significantly longer. More generally, the fact
+  that subgraph isomorphism is NP-complete, means that no time guarantees
+  can be provided. Thus when performing this on \emph{millions} of
+  molecules as in a database search, we may not be able to complete
+  the operation. In practice this is usually not the case, but still,
+  isomorphism algorithms do take longer than fingerprint
+  screening. Hence in practice performing isomorphsim tests for large
+  databases is infeasible. We have included a reference that
+  explicitly talks about this problem.}
+
 \comment{P4, c2, l17: citation needed on the size of chemistry space.}
@@
 \response{The text has been updated and simplified}
+
+\comment{p4, c1, l30: a citation is required regarding these
+  assertions of graph theory and combinatorics improvements.}
+\response{\textbf{What is he referring to?}}
 

### 2011-11-10 paper.tex
more updates

@@
 across a full chemical database of thousands or even millions of
-structures is infeasible. For that reason, complex heuristics must be
-used to drastically filter the candidate search space. Structure-based
-\emph{fingerprints} are commonly used for this purpose. Fingerprints
-encode characteristic features of a given chemical structure in a
-Boolean array, or bitmap, and they can be implemented in a variety of
-ways. Common fingerprints are created by an algorithm which generates
-bitmap patterns for each atom in the structure; then each atom and its
-nearest neighbors, including the bonds between them; then each group
-of atoms connected by paths of increasing lengths up to the maximum
-implemented for the algorithm (commonly around eight). In this way,
-characteristic parts of molecule are captured and the bitmap patterns
-are added together into the final fingerprint. If all bits in a query
-fingerprint are also present in the target fingerprint of a stored
-database structure, only then is the structure subjected to the
-computationally expensive subgraph matching algorithm. Bit operations
-are very fast and independent of the number of atoms in a structure
-due to the fixed length of the fingerprint.
+structures is infeasible \cite{Weininger:2011ly}. For that reason,
+heuristics must be used to drastically filter the candidate search
+space. Structure-based \emph{fingerprints} are commonly used for this
+purpose. Fingerprints encode characteristic features of a given
+chemical structure in a Boolean array, or bitmap, and they can be
+implemented in a variety of ways. Common fingerprints are created by
+an algorithm which generates bitmap patterns for each atom in the
+structure; then each atom and its nearest neighbors, including the
+bonds between them; then each group of atoms connected by paths of
+increasing lengths up to the maximum implemented for the algorithm
+(commonly around eight). In this way, characteristic parts of molecule
+are captured and the bitmap patterns are added together into the final
+fingerprint. If all bits in a query fingerprint are also present in
+the target fingerprint of a stored database structure, only then is
+the structure subjected to the computationally expensive subgraph
+matching algorithm. Bit operations are very fast and independent of
+the number of atoms in a structure due to the fixed length of the
+fingerprint.
 

### 2011-11-10 rebuttal.tex
Added some more text on fingeprints

@@
   given input structure.}
+
+\comment{P3, c2: structure-based fingerprints referred to here sound
+  like Daylight-style fingerprints to this reviewer. Is this the case?
+  I would emphasise the distinction of structure-key and hash-key
+  fingerprints. My understanding is that Daylight-style fingerprints
+  enumerate paths of 7 in length, not 8, could the authors provide a
+  reference for this?}
+\response{The section on fingerprints has been rewritten to be more
+  clearer as well as distinguish between the two braod classes of
+  fingerpritns. We have also mentioned the use of fingerprints in
+  similarity searchs and the Tanimoto coeffcient}
 

### 2011-11-10 paper.tex
Added some more text on fingeprints

@@
 
-InChI or other identity-mapping algorithms allow for exact searching
-(within the limitations of the algorithms). However, two other
+InChI or other identity-mapping algorithms allow for exact searching. However, two other
 practically relevant algorithms for scientific discovery based on
@@
 
-Graph substructure matching is a variant of \emph{graph isomorphism
-  detection}, which is widely believed to be computationally
+Graph substructure matching is a variant of \emph{graph isomorphism},
+which is widely believed to be computationally
 intractable~\cite{cordella2001}. To execute a graph isomorphism search
 across a full chemical database of thousands or even millions of
-structures is infeasible \cite{Weininger:2011ly}. For that reason,
-heuristics must be used to drastically filter the candidate search
-space. Structure-based \emph{fingerprints} are commonly used for this
-purpose. Fingerprints encode characteristic features of a given
-chemical structure in a Boolean array, or bitmap, and they can be
-implemented in a variety of ways. Common fingerprints are created by
-an algorithm which generates bitmap patterns for each atom in the
-structure; then each atom and its nearest neighbors, including the
-bonds between them; then each group of atoms connected by paths of
-increasing lengths up to the maximum implemented for the algorithm
-(commonly around eight). In this way, characteristic parts of molecule
-are captured and the bitmap patterns are added together into the final
-fingerprint. If all bits in a query fingerprint are also present in
-the target fingerprint of a stored database structure, only then is
-the structure subjected to the computationally expensive subgraph
-matching algorithm. Bit operations are very fast and independent of
-the number of atoms in a structure due to the fixed length of the
-fingerprint.
+structures is infeasible \cite{Weininger:2011ly}. Speedups can be
+obtained via the use of heuristics such as structural
+\emph{fingerprints}. Fingerprints encode characteristic features of a
+given chemical structure, usually in a fixed-length
+bitmap. Fingerprints fall broadly into two categories: structure keys
+and hashed keys. In the former, each bit position corresponds to a
+distinct substructure such as an alcohol or keto group. Examples
+include MACCS and PubChem keys. The latter class are designed such
+that substructural patterns are represented as strings and then hashed
+to a random bit position. As a result, a given position can encode
+multiple substructures. The advantage of such fingerprints is that
+they can cover an arbitrarily large collection of substructures such
+as paths of length $N$ or circular environments and so on. Examples
+include ECFPs and Daylight fingerprints.
+
+Given a binary fingerprint, we can first ``pre-screen'' a database, to
+ignore molecules that cannot possibly match the query. This is
+achieved by requiring that all bits in a query fingerprint must also
+be present in the target fingerprint. Since the target fingerprints
+are pre-computed, comparing binary strings can be performed extremely
+rapidly on modern hardware. As a result, we apply the actual
+isomoprhism test only on those molecules that pass the
+screen. Fingerprints can also be used to rapidly search databases for
+similar molecules, in which case a similarity metric such as the
+Tanimoto coefficient is used to compare the query and target
+fingerprints. While fast, heuristics have been developed
+\cite{Swamidass:2007ve} to further
+speed up similarity searches.
 
@@
 textbooks as a pioneer project that produced the first expert
-system~\cite{DENDRAL}. However, even before that
+system~\cite{DENDRAL}.
 

### 2011-11-10 paper.tex
added figs back to main doc

@@
 
-% \begin{figure}
-% \centering
-% \includegraphics[height=2in]{chemical-structures.png}
-% \caption{The many different facets of representing chemical structures. Shown
-% are a chemist's usual representation of a molecule (top left) and a
-% graph-based representation amenable to data mining (top right). These
-% representations are supplemented by a 3D representation of the same molecule
-% (bottom left) that is closer to reality and a representation using surface
-% properties (bottom right), which is often more relevant for the properties the
-% molecule exhibits in an experimental setup (such as when measuring its
-% solubility, or bioactivity).}
-% \label{figure:chemical-structures}
-% \end{figure}
+\begin{figure}
+\centering
+\includegraphics[height=2in]{chemical-structures.png}
+\caption{The many different facets of representing chemical structures. Shown
+are a chemist's usual representation of a molecule (top left) and a
+graph-based representation amenable to data mining (top right). These
+representations are supplemented by a 3D representation of the same molecule
+(bottom left) that is closer to reality and a representation using surface
+properties (bottom right), which is often more relevant for the properties the
+molecule exhibits in an experimental setup (such as when measuring its
+solubility, or bioactivity).}
+\label{figure:chemical-structures}
+\end{figure}
 
@@
 
-% \begin{figure}
-% \centering
-% \includegraphics[height=1in]{lipoicacid.png}
-% \caption{The standard chemical graph illustration together with the SMILES, InChI and InChIKey codes for lipoic acid. Note that the SMILES is much more human-readable than the others.}
-% \label{figure:smiles}
-% \end{figure}
+\begin{figure}
+\centering
+\includegraphics[height=1in]{lipoicacid.png}
+\caption{The standard chemical graph illustration together with the SMILES, InChI and InChIKey codes for lipoic acid. Note that the SMILES is much more human-readable than the others.}
+\label{figure:smiles}
+\end{figure}
 
@@
 \end{table*}
+
+\begin{figure}
+\centering
+\includegraphics[height=1.3in]{platins.png}
+\caption{The generated InChI is the same for the stereoisomers cisplatin and transplatin.}
+\label{figure:cistrans}
+\end{figure}
 %
-% \begin{figure}
-% \centering
-% \includegraphics[height=1.3in]{platins.png}
-% \caption{The generated InChI is the same for the stereoisomers cisplatin and transplatin.}
-% \label{figure:cistrans}
-% \end{figure}
-% %
-% \begin{figure}
-% \centering
-% \includegraphics[height=2in]{CACM_ApplicabilityDomain.pdf}
-% \caption{Illustration of the ``Applicability Domain'' concept.
+\begin{figure}
+\centering
+\includegraphics[height=2in]{CACM_ApplicabilityDomain.pdf}
+\caption{Illustration of the ``Applicability Domain'' concept.
 In order to estimate whether model predictions are reliable, it is crucial to define areas of ``chemical space'' where the model is applicable, and where it is not.

### 2011-11-11 rebuttal.tex
Added another update

@@
 \fbox{\textbf{Reviewer 3}}
+
+\comments{p1, c2, l31: it is not clear to this reviewer what software
+  and algorithms have only recently been made available. Certainly,
+  the referenced book to which they refer contains a large number of
+  algorithms that have already been published in the primary
+  literature and frequently many decades ago. If the open source
+  community has provided true novelty then please articulate this,
+  otherwise cite the primary references that report these
+  discoveries.}
+\response{The text has been updated to note that most of the
+  algorithms in cheminformatics have been known since the mid 20th
+  century, but that freely accessible toolkits implementing them are
+  relatively recent}
 

### 2011-11-11 paper.tex
Added another update

@@
 Novel technologies in the life sciences churn out information at an
-ever increasing rate---public data stores such as the one at the
+ever increasing rate -- public data stores such as the one at the
 European Bioinformatics Institute (EBI) contain on the order of 10
-petabytes of biological information.  While biological
-information has been publicly available for decades, the same has not
-been true for chemical information, until recently. It was not until
-2004 that a large public small molecule structure repository (PubChem)
-became freely available. This was followed by other databases, as
-discussed later in this article. In the same vein, open source
-software and the publication of algorithms have only recently become a
-main focus of the field \cite{faulon2010}.
+petabytes of biological information.  While biological information has
+been publicly available for decades, the same has not been true for
+chemical information, until recently. It was not until 2004 that a
+large public small molecule structure repository (PubChem) became
+freely available. This was followed by other databases, as discussed
+later in this article. While much of the foundational algorithms of
+cheminformatics have been described since the 1950s, publically
+available software implementing many of these algorithms have only
+been accessible in the last ten to fifteen years \cite{faulon2010}.
 
 So why should we actually care about chemical information being made
-public---and how does this relate to the field of computer science?
+public -- and how does this relate to the field of computer science?
 

### 2011-11-11 rebuttal.tex
responded to more comments

@@
 
-\comments{p1, c2, l31: it is not clear to this reviewer what software
+The article over-simplifies concepts, such as molecular representations, for an audience which is perhaps even more well versed in graph theory than many cheminformaticians.
+The casual reader should be left with an understanding of specific challenges to which they can bring to bear methods from their field.
+Vague challenges are mentioned, but not in any clarity enough to go away and tackle them.
+
+The manuscript provides a very specific overview on the aspects of the open source community in cheminformatics.
+Therefore, I would recommend that this be clarified in the title to avoid confusion or the manuscript broadened to
+cover the actual extent of the field, which is far vaster than is covered here.
+
+The authors make no (or very little) mention of the founding principles of cheminformatics. Very few references cover the algorithms that have been developed and published.
+
+The manuscript is poor in relevant citations with none of the classical references provided, such as Morgan 1965, Weininger 1988, Johnson and Maggiora 1990, Barnard 1993, Willett et al. 1998, etc.
+
+Molecular docking is also not mentioned, but is regarded as one of the key methods applied in cheminformatics.
+Docking should be included, along with other molecular modelling methods such as pharmacophores, shape searching,
+molecular dynamics, as a distinct section.
+
+Proteochemometric methods are mentioned a couple of times but these are largely un-noticed by the community currently.
+I do not think that it is necessarily appropriate to include these as tried-and-tested methods in an over-arching review such as this.
+
+\comment{RDKit is not mentioned at all in this article and is widely used as a
+cheminformatics toolkit in industry and academia and free and
+open-source (although from commercial beginnings).}
+\response{We agree that this toolkit should have been included. We
+  have updated the text to take note of RDKit, though we still
+  primarily focus on the CDK and OpenBabel. However, Table 2
+  includes RDKit in the feature comparison}
+
+\comment{p1, c2, l31: it is not clear to this reviewer what software
   and algorithms have only recently been made available. Certainly,

### 2011-11-11 rebuttal.tex
Added more responses

@@
 
-Molecular docking is also not mentioned, but is regarded as one of the key methods applied in cheminformatics.
-Docking should be included, along with other molecular modelling methods such as pharmacophores, shape searching,
-molecular dynamics, as a distinct section.
+\comment{Molecular docking is also not mentioned, but is regarded as
+  one of the key methods applied in cheminformatics. Docking should be
+  included, along with other molecular modelling methods such as
+  pharmacophores, shape searching, molecular dynamics, as a distinct
+  section.}  
+\response{These methods are mentioned on page 5 (column
+  1). However, we have chosen not to place a more in depth discussion
+  of these methods in their own section in the main document for a few
+  reasons. First, due to space limitations this would require us to
+  compress or drop other sections which we believe to be more
+  relevant. Second, we are of the view that dokcing, for example, is
+  not really a cheminformatics technique, though it makes extensive
+  use of cheminformatic concepts. Instead we place it under molecular
+  modeling. We realize that this is a somewhat subjective
+  ``classification'' of methods. However we do make explicit note of
+  these methods in the supplemental document where we make a
+  distinction between cheminformatics and related subject areas such
+  as quantum mechanics and molecular modeling.}
 
-Proteochemometric methods are mentioned a couple of times but these are largely un-noticed by the community currently.
-I do not think that it is necessarily appropriate to include these as tried-and-tested methods in an over-arching review such as this.
+\comment{Proteochemometric methods are mentioned a couple of times but
+  these are largely un-noticed by the community currently. I do not
+  think that it is necessarily appropriate to include these as
+  tried-and-tested methods in an over-arching review such as this.}
+\response{While it is true that proteochemometric methodologies are
+  not as well known as say docking or pharmacophore modeling, we
+  choose to include them since they are an example of a (statistical)
+  QSAR technique that does explicitly include ligand and receptor features}
 
-\comment{RDKit is not mentioned at all in this article and is widely used as a
-cheminformatics toolkit in industry and academia and free and
-open-source (although from commercial beginnings).}
+\comment{RDKit is not mentioned at all in this article and is widely
+  used as a cheminformatics toolkit in industry and academia and free
+  and open-source (although from commercial beginnings).}
 \response{We agree that this toolkit should have been included. We
-  have updated the text to take note of RDKit, though we still
-  primarily focus on the CDK and OpenBabel. However, Table 2
+  have updated the text on page 7 to take note of RDKit, though we
+  still primarily focus on the CDK and OpenBabel. However, Table 2
   includes RDKit in the feature comparison}

### 2011-11-11 paper.tex
Added more responses

@@
 have been designed to take into account both ligand and receptor
-structures.
+structures. The last method is an example of an extension of
+statistical QSAR methods to simultaneous model the f receptor and
+ligand as first reported by Lapinsh \textit{et  al.}~\cite{lapinsh2001}.
 
@@
 \emph{multiple-instance learning} could also be applied to the
-multi-conformer problem. More recently, simultaneous modeling of
-receptor and ligand has been addressed by the proteochemometric
-approach as first reported by Lapinsh \textit{et
-  al.}~\cite{lapinsh2001}.
+multi-conformer problem.
 

### 2011-11-11 rebuttal.tex
Added more responses

@@
 
-The manuscript is poor in relevant citations with none of the classical references provided, such as Morgan 1965, Weininger 1988, Johnson and Maggiora 1990, Barnard 1993, Willett et al. 1998, etc.
+\comment{The manuscript is poor in relevant citations with none of the
+  classical references provided, such as Morgan 1965, Weininger 1988,
+  Johnson and Maggiora 1990, Barnard 1993, Willett et al. 1998, etc.}
+\response{The text has been updated to include the ``classic''
+  references. In addition the supplemental history document lists a
+  number of milestones in cheminformatics and their relevant references.}
 

### 2011-11-11 rebuttal.tex
Added more responses

@@
 
-The article over-simplifies concepts, such as molecular representations, for an audience which is perhaps even more well versed in graph theory than many cheminformaticians.
-The casual reader should be left with an understanding of specific challenges to which they can bring to bear methods from their field.
-Vague challenges are mentioned, but not in any clarity enough to go away and tackle them.
-
-The manuscript provides a very specific overview on the aspects of the open source community in cheminformatics.
-Therefore, I would recommend that this be clarified in the title to avoid confusion or the manuscript broadened to
-cover the actual extent of the field, which is far vaster than is covered here.
-
-The authors make no (or very little) mention of the founding principles of cheminformatics. Very few references cover the algorithms that have been developed and published.
+\comment{The manuscript provides a very specific overview on the
+  aspects of the open source community in cheminformatics. }
+\response{As we have noted below, the current article is not a
+  comprehensive historical review of the field. We have included a
+  supplement  that does provide a very brief overview of historic
+  milestones. Furthermore, we have updated the text to note commercial
+  vendors and solutions; but we still stress on Open Source and Open
+  Access solutions to allow reasders to explore the problems and
+  topics we have discussed with a minimum of encumbrances}
 

### 2011-11-11 rebuttal.tex
Added more responses

@@
 \newcommand{\response}[1]{#1\vspace{1em}}
+
+We would like to thank the anonymous reviewers for their constructive
+comments. We have updated the main text in several places and have
+also included a supplemental document that provides a brief
+description of terminology and a list of some historic milestones in
+the field. The following sections provide a detailed response to
+individual reviewer comments.
 

### 2011-11-11 rebuttal.tex
Added more responses

@@
   includes RDKit in the feature comparison}
+
+\comment{Figure 1: what is meant by 'closer to reality' This needs to
+  be clarified. This figure and its caption requires a little more
+  consideration to appropriately convey the message, which this
+  reviewer takes to be that there are many useful ways to represent
+  molecular structures.}
+\response{We have updated the caption to remove the phrase ``closer to
+reality'' since this is not the main point. Instead we have noted that
+the representations contain differing amounts and types of information
+that are equally valid, but suited for different purposes.}
 

### 2011-11-11 paper.tex
Added more responses

@@
   whose financial success depended on being the first to produce the
-  new therapeutic molecules.  Only within the last decade---and as an
+  new therapeutic molecules.  Only within the last decade -- and as an
   effect of a change in mindsets, government-mandated policy changes
   and last but not least because of chemists volunteering their time
-  for an Open Science ``movement''---have researchers gained access to
+  for an Open Science ``movement'' -- have researchers gained access to
   freely available software packages and databases of tens of millions
@@
   determining the behavior of small molecules in biological pathways,
-  to finding a cure for malaria and other ``neglected'' diseases.
+  to finding therapies for rare and neglected diseases.
 \end{abstract}
@@
 relevance of computer science to this research becomes clear when we
-consider the quantity and types of data available--- currently a
+consider the quantity and types of data available -- currently a
 single data base such as PubChem contains more than 34 million
@@
 Figure~\ref{figure:chemical-structures}. The top left hand corner
-shows a structure a chemist would draw it---every ``corner''
+shows a structure a chemist would draw it -- every ``corner''
 representing a carbon atom (which can also be represented as a C),
@@
 information with their peers.  However, this format is not directly
-amenable to data mining---different atoms might be collapsed into one
+amenable to data mining -- different atoms might be collapsed into one
 node, and sets of atoms might have the same relevant properties
@@
 different elemental atom type, can have similar
-properties---\emph{e.g.}, both oxygen and nitrogen carry an ``On''
+properties -- \emph{e.g.}, both oxygen and nitrogen carry an ``On''
 label in this graph.
@@
 \centering
-\includegraphics[height=2in]{chemical-structures.png}
-\caption{The many different facets of representing chemical structures. Shown
-are a chemist's usual representation of a molecule (top left) and a
-graph-based representation amenable to data mining (top right). These
-representations are supplemented by a 3D representation of the same molecule
-(bottom left) that is closer to reality and a representation using surface
-properties (bottom right), which is often more relevant for the properties the
-molecule exhibits in an experimental setup (such as when measuring its
-solubility, or bioactivity).}
+\includegraphics[height=2in]{chemstructs.png}
+\caption{A hierarchy of chemical structure representations, increasing
+  in information content from left to right and top to bottom. A
+  chemists traditional 2D depiction (A), a labeled graph (B), a 3D
+  conformation (C) and a surface represention (D). All show differing
+  (and valid) aspects of the same molecule and are suited for
+  differing purposes. The choice of representation is guided by the
+  intended use of the structure data.}
 \label{figure:chemical-structures}
@@
 of atoms) is usually not known from the onset but it rather depends on
-the particular system considered---involving a large amount of
+the particular system considered -- involving a large amount of
 trial-and-error in the choice of a ``proper'' description for
@@
 closer collaboration between chemists (or, more generally, life
-scientists) and computer scientists---where the former need to be able
+scientists) and computer scientists -- where the former need to be able
 to pose their problems in a way relevant for practical applications;
@@
 normalization problems in the chemical and bioactivity landscape, but also due
-to sheer data volume. One problem is the annotation with the ``protein targets'' a molecule binds to---here dozens of
+to sheer data volume. One problem is the annotation with the ``protein targets'' a molecule binds to -- here dozens of
 protein and gene identifiers exist, generally without the availability of a one-to-one mapping from each ontology, which makes merging databases often a cumbersome exercise.
@@
 organism (such as a human) and which in many cases indicate the state
-the organism is in---such as indicating a disease, or a fasted state,
-etc.---and with a sufficiently large database the idea is to be able
+the organism is in -- such as indicating a disease, or a fasted state,
+etc. -- and with a sufficiently large database the idea is to be able
 to detect the chemical composition of body fluids and to arrive at
@@
 cheminformatics community. In addition, GDB-13 only enumerates
-molecules up to 13 heavy atoms---given the possibly exponential
+molecules up to 13 heavy atoms -- given the possibly exponential
 increase in search space with the number of atoms, it seems unlikely
@@
 chemistry. This representation, while succinct and amenable to
-computation, is only an approximation of reality---and edge cases
+computation, is only an approximation of reality -- and edge cases
 abound. Existing areas of interest are enumeration of colored graphs,
@@
 \caption{Some of the properties potential drug candidates need to have to be considered as ``promising,'' either in the different research and development stages, as well as a drug that is finally admitted to the market.
-Note that the particular requirements are also heavily project-dependent, and that also in cases where cheminformatics approaches can be employed in drug design an experimental confirmation of the prediction is required---the computer can hence decrease the experimental burden, but not replace it altogether.}
+Note that the particular requirements are also heavily project-dependent, and that also in cases where cheminformatics approaches can be employed in drug design an experimental confirmation of the prediction is required -- the computer can hence decrease the experimental burden, but not replace it altogether.}
 \label{table:properties}

### 2011-11-11 paper.tex
Added more responses
more updates

@@
 different representations of the same chemical structure are shown in
-Figure~\ref{figure:chemical-structures}. The top left hand corner
-shows a structure a chemist would draw it -- every ``corner''
-representing a carbon atom (which can also be represented as a C),
-every O denoting oxygen, every N nitrogen, every H hydrogen, and so
-on. This is the standard format chemists use when exchanging
-information with their peers.  However, this format is not directly
-amenable to data mining -- different atoms might be collapsed into one
-node, and sets of atoms might have the same relevant properties
-assigned. Hence, the top right hand corner of
-Figure~\ref{figure:chemical-structures} shows an annotated graph
-representation, taking into account that atoms, though being of
-different elemental atom type, can have similar
-properties -- \emph{e.g.}, both oxygen and nitrogen carry an ``On''
-label in this graph.
+Figure~\ref{figure:chemical-structures}. The representations range
+from a graph (Fig.~\ref{figure:chemical-structures}A) that contains
+much implicit information (such as unlabeled nodes represent carbons
+with valencies satisified by hydrogens) to a fully labeled graph,
+where all information is explicit
+(Fig.~\ref{figure:chemical-structures}B). The former is what is
+usually exchanged informally between chemists, but to actually mine
+chemical structures computationally, something like the latter is
+required.  For example, graph mining algorithms can be applied to
+labeled graph representations, to discover patterns in large-scale
+datasets \cite{wegner2006,horst2009}.
 
@@
 
-Graph mining algorithms can be applied to this type of annotated graph
-(employing large databases of thousands of molecules), to discover
-patterns in large-scale datasets.  Many successful applications of
-this concept have been published \cite{wegner2006,horst2009}. Still,
-this ``annotated graph'' representation is far from being the end of
-the story. Molecules are three-dimensional entities, as represented in
-the bottom left hand corner of the figure, and the precise 3D shape is
-relevant in many situations (although not in others). Although the
-graph implicitly encodes the 3D structure of a molecule (when combined
-with knowledge about favored bond angles and distances), many
-different low energy 3D structures, or conformers, may be consistent
-with the same graph structure. In addition, the 3D structure may have
-a geometrical arrangement of features (such as the presence of a
-right-handed helix) which cannnot be encoded in the graph.
-
-To extend the concept of three-dimensionality further, in many
-biological systems the \emph{surface properties} of the molecule are
-what appear to be relevant, as shown in the bottom right hand corner.
-Hence, while graph mining on chemicals is certainly a promising avenue
-to explore, one needs to keep in mind that not all molecular
-properties are encoded in the graph (for example, neither rotational
-states, nor different charge states are encoded) and thus often other
-``descriptors'' need to be used to capture the essentials of a
-chemical system \cite{bender2004}. Hence, one of the difficulties of a
-cheminformatician is that the equivalence of particular atoms (or sets
-of atoms) is usually not known from the onset but it rather depends on
-the particular system considered -- involving a large amount of
-trial-and-error in the choice of a ``proper'' description for
-molecules in a particular case.
+Although graphs implicitly encodes the 3D structure of a molecule
+(when combined with knowledge about favored bond angles and
+distances), many different low energy 3D structures, or conformers,
+may be consistent with the same graph structure. In addition, the 3D
+structure may have a ``secondary'' geometrical arrangement of features
+(such as the presence of a right-handed helix) which cannnot be
+encoded in the graph. Thus we have 3D representations
+(Fig.~\ref{figure:chemical-structures}C) that make the 3D arrangement
+of atoms and bonds explicit. Fig.~\ref{figure:chemical-structures}D
+goes further and represents a molecular surface, encoding some
+property (such as lipophilicity) that varies over the surface.
+
+Even though some structure representations contain more explicit
+information than others, they are all equally valid and are suited for
+different problems. Indeed, Fig.~\ref{figure:chemical-structures}
+represents a hierarchy of representations. Thus, when searching for
+substructures, a 2D representation is sufficient. When trying to
+rationalize why a molecule binds to a protein, a 3D structure is
+vital.
+
+A key challenge in chemical data mining is the choice of
+representation and how that can affect subsequent analysis
+efforts. But even after one has been chosen, one faces the challenge
+of balancing computational costs with the utility of the
+representation. For example, a full 3D description of a molecule,
+taking into account all possible conformers would let us accurately
+predict many properties. But the size of this representation and the
+time required to evaluate it would be prohibitive. Instead, can we
+obtain accurate predictions from a subset of conformers? Or, can we
+obtain comparable accuracy by using a 2D representation? And is so,
+what type of labels are required? Currently, many of these questions
+are answered using trial and error.
 
@@
 closer collaboration between chemists (or, more generally, life
-scientists) and computer scientists -- where the former need to be able
-to pose their problems in a way relevant for practical applications;
-and where the latter are able to devise ways of capturing, storing,
-and analyzing chemical data which achieve optimal balances between the different risks in
-generalizable encodings and complexity (space and time). The goals of cheminformatics are to
+scientists) and computer scientists -- where the former need to be
+able to pose their problems in a way relevant for practical
+applications; and where the latter are able to devise ways of
+capturing, storing, and analyzing chemical data which achieve optimal
+balances between the different risks in generalizable encodings and
+complexity (space and time). The goals of cheminformatics are to
 support better chemical decision making by 1. storing and integrating

### 2011-11-11 rebuttal.tex
more updates

--- 2011-11-11 rebuttal.tex
@@
   includes RDKit in the feature comparison}
+
+\comment{The authors could more appropriately cover the graph theory
+  aspects of molecular structures by correctly applying the
+  terminology to which computer scientists and cheminformaticians will
+  most certainly be aware}
+\response{The text has been updated}
+
+\comment{p2, c1: I think the authors could benefit from a re-write of
+  the section detailing molecular structure representations. \ldots
+  Figure 1: what is meant by 'closer to reality' This needs to be
+  clarified. This figure and its caption requires a little more
+  consideration to appropriately convey the message, }
+\response{We have updated the caption to remove the phrase ``closer to
+  reality'' since this is not the main point. Instead we have noted
+  that the representations contain differing amounts and types of
+  information that are equally valid, but suited for different
+  purposes. The text on page 2 has been updated to better represent
+  this view}
 

### 2011-11-11 paper.tex
more updates

@@
 
-A key challenge in chemical data mining is the choice of
-representation and how that can affect subsequent analysis
-efforts. But even after one has been chosen, one faces the challenge
+Given that a fundamental principle of the field is that \emph{similar
+  molecules exhibit similar properties} \cite{Johnson:1990qf}, the
+choice of representation is key in determining how such similarities
+are evalaued, and thus the effectiveness of subsequent analyses. But
+even after one has selected a representation one faces the challenge
 of balancing computational costs with the utility of the

### 2011-11-12 rebuttal.tex
more updates
Added in Aarons edits

@@
 individual reviewer comments.
+\vspace{2em}
 
@@
 
+\comment{P6, c1, l7: I think it is important for the authors to cover ELNs by using actual recorded data.
+Do we know that many scientists changed to ELNs ?many years ago?
+as stated?
+Is there a reference for this or data to back it up?}
+\response{We have expanded our original wording, and we now cite two references about the use and growth of ELNs in chemistry and elsewhere.}
+
+\comment{P6, c1, l7: the authors do not mention the leading players in ELNs (e.g.
+CambridgeSoft, Accelrys, etc.).
+Perhaps this is because open-source ELNs do not exist, or are in their infancy?
+I think the key software houses that develop and supply these ELN systems should be mentioned.
+Furthermore, systems such as Reaxys should be mentioned to ensure that it is understood that these systems do exist to provide these results to these searches and are used daily by chemists.}
+\response{We now mention CambridgeSoft, Accelrys and Reaxys in the article, and provide a citation to a review of 35 ELNs that was published in November 2011.}
+
 \comment{P7, c2, l25: the authors should go back even further in the

### 2011-11-12 paper.tex
Added in Aarons edits

@@
 \centering
-\includegraphics[height=2in]{chemstructs.png}
+\includegraphics[height=2in]{chemical-structures.png}
 \caption{A hierarchy of chemical structure representations, increasing
@@
 new collaboration/innovation opportunities with internal partners and
-a rapidly growing number of external/public partners. In fact, in
+a rapidly growing number of external/public partners. In
 cheminformatics and chemistry, scientists already switched many years
@@
 
-ELNs are typically much more accepted and used within the
-cheminformatics domain than in the bioinformatics domain, which is
-picking up speed lately, the question being why this is the case. In
-chemistry, every experiment requires a lot of expert
+While ELNs technically existed as long ago as the 1990s, chemists
+adopted their use in large numbers in 2004 and 2005, and the market is
+still growing~\cite{ELNstatus}. As one report notes, ``The initial
+drive for the development of ELNs came from the field of chemistry,
+perhaps driven by the early adoption by chemists of computer
+technologies for drawing chemical structures and for storing and
+searching them graphically within databases.... The pharmaceutical
+industry shifted from a position of `if we get an ELN' to `when we get
+an ELN'.''~\cite{ELNstatus} ELNs are typically much more accepted and
+used within the cheminformatics domain than in the bioinformatics
+domain, which is picking up speed lately, the question being why this
+is the case. In chemistry, every experiment requires a lot of expert
 knowledge. Therefore, approaches for reducing risks by communicating
@@
 
-It is worth mentioning that chemistry is traditionally a field where many people
-have been used to huge bookshelves of reactions and compounds. Chemists changed
-to online ELN collaboration not only to increase efficiency, but also
-because it allowed collaboration with trusted colleagues. Third party paper
-catalogs and academic publications proved to be
-inefficient, and even more critically, they did not take compounds of trusted
-colleagues into account. Overall the change of management to ELNs, similar to
+At this point, ``electronic lab notebooks have replaced the
+traditional paper lab book across the pharmaceutical industry,'' but
+are not yet common in academia, because of their high
+cost~\cite{WavingGoodbye}.  It is worth mentioning that chemistry is
+traditionally a field where many people have been used to huge
+bookshelves of reactions and compounds. Chemists changed to online ELN
+collaboration not only to increase efficiency, but also because it
+allowed collaboration with trusted colleagues. Third party paper
+catalogs and academic publications proved to be inefficient, and even
+more critically, they did not take compounds of trusted colleagues
+into account. Overall the change of management to ELNs, similar to
 ``Enterprise 2.0,'' required delivering on the promise to increase
-interconnectivity, and to improve collaboration
-opportunities, and indeed it already performed these functions many years ago.
+interconnectivity, and to improve collaboration opportunities.  The
+leading players in this area include CambridgeSoft's E-Notebook and
+Accelrys' Symyx Notebook, but there are at least 35 different
+companies producing ELNs as of February 2011~\cite{ELNreview}, for use
+in chemistry, biology, quality control, and other research.  Another
+knowledge management tool specifically for chemists is the web-based
+service Reaxys, founded in 2009, whose company goal is to provide ``a
+single, fully integrated chemical workflow solution.''
 

### 2011-11-12 paper.tex
Updated ontoogy section

@@
 %
-\subsection{Databases \& Ontologies}
+\subsection{Representing and Searching Structures}
 \label{sec:databases}
@@
 
-There is considerable progress in open chemical and bioactivity databases like
-ChEMBL and NCBI-PubChem, which are both freely available and contain in the order of millions (ChEMBL) and tens of
-millions (PubChem) of data points each. Still, the integration
-of chemical databases with each other remains a challenge, not only due to the
-normalization problems in the chemical and bioactivity landscape, but also due
-to sheer data volume. One problem is the annotation with the ``protein targets'' a molecule binds to -- here dozens of
-protein and gene identifiers exist, generally without the availability of a one-to-one mapping from each ontology, which makes merging databases often a cumbersome exercise.
+There is considerable progress in open chemical and bioactivity
+databases like ChEMBL and NCBI-PubChem, which are both freely
+available and contain in the order of millions (ChEMBL) and tens of
+millions (PubChem) of data points each. Still, the integration of
+chemical databases with each other remains a challenge, not only due
+to the normalization problems in the chemical and bioactivity
+landscape, but also due to sheer data volume. One problem is the
+annotation with the ``protein targets'' a molecule binds to -- here
+dozens of protein and gene identifiers exist, generally without the
+availability of a one-to-one mapping from each ontology, which makes
+merging databases often a cumbersome exercise.
 
@@
 
-InChI or other identity-mapping algorithms allow for exact searching. However, two other
-practically relevant algorithms for scientific discovery based on
-chemical databases are \emph{substructure searching} and
-\emph{similarity searching}, which are required to generalize from the
-data points contained in a particular database to other, related
-molecules. In substructure searching, the database is searched for a
-specified wholly contained part of the structures in the database,
-while in similarity searching, the database is searched for structures
-similar (in some property space) to a provided search, or query,
-structure. Chemical search packages are often implemented and
-optimized for a given database technology; for example, the OrChem
-package is an open source chemical search package for the Oracle
-database \cite{rijnbeek2009}.
+InChI or other identity-mapping algorithms allow for exact
+searching. However, two other practically relevant algorithms for
+scientific discovery based on chemical databases are
+\emph{substructure searching} and \emph{similarity searching}, which
+are required to generalize from the data points contained in a
+particular database to other, related molecules. In substructure
+searching, the database is searched for a specified wholly contained
+part of the structures in the database, while in similarity searching,
+the database is searched for structures similar (in some property
+space) to a provided search, or query, structure. Chemical search
+packages are often implemented and optimized for a given database
+technology; for example, the OrChem package is an open source chemical
+search package for the Oracle database \cite{rijnbeek2009}.
 
@@
 
-%Fingerprints also facilitate the rapid calculation of quantitative
-%structure-based similarity measures. An example measure of similarity is the
-%\emph{Tanimoto coefficient}, calculated as the ratio $T(a,b) = \frac{c}{(a+b-c)}$,
-%where $c$ is the count of bits on in the same position in both the two
-%fingerprints, $a$ is the count of bits on in the first structure, and $b$ is the
-%count of bits on in the second structure. Historically, this coefficient originated from the comparison of ecosystems
-%around lakes, where a measure was needed to describe how similar those ecosystems are with respect to the animals,
-%plants etc. they contain. However, it can be applied to every other system as well where features can be described in a
-%binary manner. The Tanimoto coefficient varies in the
-%range 0.0 -- 1.0, with a score of 1.0 indicating that the two structures are either very
-%similar or even identical (i.e. their fingerprints are the same).
+\subsection{Linking Molecules to Other Data}
+\label{sec:prof-ident}
+
+The preceding section focuses on the choice of representation of
+chemical structures. With the profusion of databases that contain
+details of small molecules, an emerging problem is the integration of
+such databases. Such integration forms the basis of ``systems'' level
+investigations of biological systems. A current challege in
+cheminformatics is how to identify ``equivalent'' objects across data
+sources. At one level, equivalent compounds can be identified via
+canonical structure representations (InChI). But there are other
+equivalences beyond structure. For example, compounds may bind to the
+same or similar targets; they may be have similar biological function
+(say, anti-neoplastic) and so on. To assess such higher level
+characteristics, ontologies and semantic technologies have been
+employed, to first, support annotations of chemical systems and
+second, enable semantic analyses of chemical systems in conjunction
+with other biological systems (such as pathways). A number of
+challenges remain in this area - the use of ontologies allows one to
+build a semntic network of concepts that links molecules to other
+molecules as well as other objects such as targets, genes and
+pathways. How does one evaluate similarities in such a heterogenous
+network \cite{couto2010}? How does one integrate chemical structure
+information with ontologies \cite{hastingsowled2010}? Integrating
+small molecule data with other biological data types will be key to
+acheiving a systems level understanding of small
+molecules \cite{Oprea:2007fk}.
+
+% Fingerprints also facilitate the rapid calculation of quantitative
+% structure-based similarity measures. An example measure of
+% similarity is the \emph{Tanimoto coefficient}, calculated as the
+% ratio $T(a,b) = \frac{c}{(a+b-c)}$, where $c$ is the count of bits
+% on in the same position in both the two fingerprints, $a$ is the
+% count of bits on in the first structure, and $b$ is the count of
+% bits on in the second structure. Historically, this coefficient
+% originated from the comparison of ecosystems around lakes, where a
+% measure was needed to describe how similar those ecosystems are with
+% respect to the animals, plants etc. they contain. However, it can be
+% applied to every other system as well where features can be
+% described in a binary manner. The Tanimoto coefficient varies in the
+% range 0.0 -- 1.0, with a score of 1.0 indicating that the two
+% structures are either very similar or even identical (i.e. their
+% fingerprints are the same).
 %
-Structure-based similarity is however of limited use in grouping of
-chemicals based on non-structural features, such as shared bioactivity
-profiles. For this purpose, a large-scale and flexible classification
-structure is needed for chemicals and their linked biological
-information. One such answer to this need is provided in the form of
-\emph{ontologies}.  An ontology is a formal specification of entities
-and their relationships in a domain of interest. It is structured
-around an underlying logical representation such as Description
-Logics~\cite{baaderdl2007}. The logic-based representation allows
-automated reasoning to derive new knowledge (inferences) from the
-knowledge encoded, making the representation both compact and
-powerful.  Ontologies are already in widespread use across the
-biomedical domain, and in the cheminformatics field ontologies
-relating to the chemical structures themselves and the protein targets
-the molecules are active against are probably most common. A
-similarity measure based on the information encoded in ontologies is
-termed \emph{semantic similarity}, and such measures have already been
-applied to enhance chemical classification in several problem areas
-\cite{couto2010}. Many open questions still remain about the best way
-to combine the information encoded in chemical structures with
-chemical ontologies \cite{hastingsowled2010}.
+
+% Structure-based similarity is however of limited use in grouping of
+% chemicals based on non-structural features, such as shared bioactivity
+% profiles. For this purpose, a large-scale and flexible classification
+% structure is needed for chemicals and their linked biological
+% information. One such answer to this need is provided in the form of
+% \emph{ontologies}.  An ontology is a formal specification of entities
+% and their relationships in a domain of interest. It is structured
+% around an underlying logical representation such as Description
+% Logics~\cite{baaderdl2007}. The logic-based representation allows
+% automated reasoning to derive new knowledge (inferences) from the
+% knowledge encoded, making the representation both compact and
+% powerful.  Ontologies are already in widespread use across the
+% biomedical domain, and in the cheminformatics field ontologies
+% relating to the chemical structures themselves and the protein targets
+% the molecules are active against are probably most common. A
+% similarity measure based on the information encoded in ontologies is
+% termed \emph{semantic similarity}, and such measures have already been
+% applied to enhance chemical classification in several problem areas
+% \cite{couto2010}. Many open questions still remain about the best way
+% to combine the information encoded in chemical structures with
+% chemical ontologies \cite{hastingsowled2010}.
 

### 2011-11-12 rebuttal.tex
Updated text to complete response to reviewer 1

@@
   references.} 
-\response{We agree with the reviewer that a more comprehensive list of
-references would be appropriate and useful. However, due to format
-limitations of the article, we have been restricted to the use of XXX
-references. To alleviate this problem, we have included a supplement
-that provides a more extensive list of references, specifically
-including a number of texts covering cheminformatics broadly.}
+ \response{We agree with the reviewer that a more
+  comprehensive list of references would be appropriate and
+  useful. However, due to format limitations of the article, we have
+  been restricted to the use of 40 references. To alleviate this
+  problem, we have included a supplement that provides a more
+  extensive list of references, specifically including a number of
+  texts covering cheminformatics broadly.}
+
+\comment{1) The connection between sections could be improved,
+  i.e. the authors should try to enhance the correlation that exists
+  between the topics described in each section and map them to a
+  global flow of cheminfomatics changes. 2) The authors state in the
+  beginning that the paper will be around the concept of risk
+  minimization in drug discovery, but I found that this topic is not
+  sufficiently explored. In many sections is not straightforward teh
+  connection between the challenges described and their impact on risk
+  minimization in drug discovery; 3) The paper misses some important
+  efforts and more detail on translational medicine and semantic web }
+\response{The reviewer is correct in noting the disjointed character
+  of the original article. We have updated the text throughout to flow
+  smoothly. As part of this we have made more explicit how each topic
+  that is discussed plays a role in derisking various stages of the
+  drug disocvery process. Finally, we have updated the text on page 4
+  to address the issue of linked data and the role that semantic
+  technologies can play in this area. We have, however, chosen to not
+  address the issue of translational medicine. While there have been
+  some cases where cheminformatics have played a role in translational
+  medicine, the bulk of cheminformatics lies upstream of this
+  topic. Hence, to keep the article focused we have chosen to not
+  include this topic.}
 

### 2011-11-12 paper.tex
Updated text to complete response to reviewer 1

@@
 network \cite{couto2010}? How does one integrate chemical structure
-information with ontologies \cite{hastingsowled2010}? Integrating
-small molecule data with other biological data types will be key to
-acheiving a systems level understanding of small
-molecules \cite{Oprea:2007fk}.
+information with ontologies \cite{hastingsowled2010}? And more
+generally, how can we achieve automated inferencing in such
+heterogenous linked systems, allowing us to acheive a systems level
+understanding of small molecules \cite{Oprea:2007fk}.
 

### 2011-11-12 rebuttal.tex
more updates

@@
 \response{The text has been updated}
+
+\comment{The inclusion of ontologies is not really useful here as the
+  conclusions drawn are vague and do not really inform the current
+  state-of-the-art and usefulness of ontologies in cheminformatics.}
+\response{The tetx has been updated on page 4 to make the discussion
+  of ontologies more focused, specifically identfying challenges in
+  linking chemical information to other data types and how ontologies
+  and semantic methods can be used ot address this.}
 

### 2011-11-13 paper.tex
smoothened section on ELNs

@@
 %
-Most cheminformatics applications rely on large databases of chemical structures,
-their properties, and relationships to, \textit{e.g.}, biological targets.
-Organizing and maintaining such databases, as well as searching and clustering similar structures together,
-are essential  to enable many scientific applications. Yet, each one of
+Most cheminformatics applications rely on large databases of chemical
+structures, their properties, and relationships to, \textit{e.g.},
+biological targets.  Organizing and maintaining such databases, as
+well as searching and clustering similar structures together, are
+essential to enable many scientific applications. Yet, each one of
 those areas poses its unique computer science challenges.
@@
 \includegraphics[height=1in]{lipoicacid.png}
-\caption{The standard chemical graph illustration together with the SMILES, InChI and InChIKey codes for lipoic acid. Note that the SMILES is much more human-readable than the others.}
+\caption{The standard chemical graph illustration together with the
+  SMILES, InChI and InChIKey codes for lipoic acid. Note that the
+  SMILES is much more human-readable than the others.}
 \label{figure:smiles}
@@
 perhaps driven by the early adoption by chemists of computer
-technologies for drawing chemical structures and for storing and
-searching them graphically within databases.... The pharmaceutical
-industry shifted from a position of `if we get an ELN' to `when we get
-an ELN'.''~\cite{ELNstatus} ELNs are typically much more accepted and
-used within the cheminformatics domain than in the bioinformatics
-domain, which is picking up speed lately, the question being why this
-is the case. In chemistry, every experiment requires a lot of expert
-knowledge. Therefore, approaches for reducing risks by communicating
-with fellow chemists about similar reactions or compounds is
-critical. This requires capturing the reaction details, compound
-details, and time consuming analysis details which ensure quality
-data, such that these data can be searched by others.
+technologies \ldots The pharmaceutical industry shifted from a
+position of `if we get an ELN' to `when we get an
+ELN'.''~\cite{ELNstatus}. The growth of the use of ELNs has been
+driven by many factors including ease of use and regulatory
+compliance. In chemistry, every experiment requires significant domain
+expertise and provenance information (especially in industrial
+settings where an experiment might be a key component of a patent
+filing). ELNs play an important role in reducing the risks associated
+with information loss by allowing easy sharing and retrieval of domain
+knowledge, securing sensitive data, ensuring data is time stamped and
+in general maintaining an audit trail of all experiemnts and data
+involved in a project.
 
@@
 traditionally a field where many people have been used to huge
-bookshelves of reactions and compounds. Chemists changed to online ELN
-collaboration not only to increase efficiency, but also because it
-allowed collaboration with trusted colleagues. Third party paper
-catalogs and academic publications proved to be inefficient, and even
-more critically, they did not take compounds of trusted colleagues
-into account. Overall the change of management to ELNs, similar to
+bookshelves of reactions and compounds. Chemists changed to online
+ELNs not only to increase efficiency, but also because it allowed easy
+collaboration with trusted colleagues. Third party paper catalogs and
+academic publications proved to be inefficient, and even more
+critically, they did not take compounds of trusted colleagues into
+account. Overall the change of management to ELNs, similar to
 ``Enterprise 2.0,'' required delivering on the promise to increase

### 2011-11-13 paper.tex
modified the eln section to be more concise and focus more on challenges. Still needs some more work

@@
 lab-notebooks, multiple compound registration, and search
-tools. Compared to today there were many pitfalls and drawbacks
-associated with this. The overall architecture was too inflexible to
+tools. The overall architecture was too inflexible to
 adopt to fast changing data standards, scientists spent too much time
@@
 alignment with other working groups was inefficient (\emph{e.g.}, for
-running analytical experiments), a legal dispute required to find data
-in paper lab-notebooks, data synchronization between systems caused
-lag times and access delays, and local silos hampered an efficient
-global collaboration. Now, chemists prefer ELNs, because they can run
-searches in the data across the organization. This does not only allow
-to find experts faster, but also allows access to highly trusted data
-with approved quality workflows. Since this also gives access to
-analytical and legal data directly every new ELN release is aiming at
-increasing the integration of scientific and business workflows for
-reducing the risk due to irrational or bounded-rational decision
-making and increasing decision making efficiency.
+running analytical experiments). Legal disputes would require manual
+searches through many paper lab notebooks and data synchornization was
+painful, hampering large scale collaborations.
+
+%  Now, chemists prefer ELNs, because they can run
+% searches in the data across the organization. This does not only allow
+% to find experts faster, but also allows access to highly trusted data
+% with approved quality workflows. Since this also gives access to
+% analytical and legal data directly every new ELN release is aiming at
+% increasing the integration of scientific and business workflows for
+% reducing the risk due to irrational or bounded-rational decision
+% making and increasing decision making efficiency.
 
@@
 ELN'.''~\cite{ELNstatus}. The growth of the use of ELNs has been
-driven by many factors including ease of use and regulatory
-compliance. In chemistry, every experiment requires significant domain
-expertise and provenance information (especially in industrial
-settings where an experiment might be a key component of a patent
-filing). ELNs play an important role in reducing the risks associated
-with information loss by allowing easy sharing and retrieval of domain
-knowledge, securing sensitive data, ensuring data is time stamped and
-in general maintaining an audit trail of all experiemnts and data
-involved in a project.
+driven by many factors including ease of use (rapid searches across
+organization-wide data sources, easy sharing of domain knowledge) and
+regulatory compliance (timestamped experimental data, experiment and
+data approvals and QC information). Given that any experiment could
+become the core of a patent filing, ELNs allow organizations to define
+and implement audit trails in a standardized and efficient fashion.
 
@@
 external/public databases with chemical and bioactivity data remains a
-challenge due to differences in ontologies, synchronization and
-maintenance, and efficient substructure and similarity search within
-complex data types (compounds). A collaboration with external parties,
-e.g. contract research, poses other problems with duplicate checking
-for compounds (e.g. tautomeric or salt structures) and efficient data
-synchronization, and most often a double-registration is required for
-fulfilling the data quality processes on both sides. If external
-partners are small they often do not even have sufficient IT resources
-themselves and thus rely on external services. Cloud services could
-help not only to provide a service infrastructure for all parties
-involved, but also to provide the required private/public access
-toolbox. Furthermore, there are still many data sources not being
-indexed properly, since chemistry patents are often cryptic and
-image/text mining remains a puzzle. One could argue that the
-image/text mining in scientific journals is less challenging; in fact,
-it remains a huge challenge and many chemists do not trust automated
-non-expert validated mining sources. The closed CAS (Chemical Abstract
+challenge due to differences in identifiers, synchronization, curation
+\& error correctin mechanisms and efficient substructure and
+similarity search within complex data types (compounds). A
+collaboration with external parties, e.g. contract research, poses
+other problems including compound deduplication and efficient data
+synchronization. If external partners are small they often do not even
+have sufficient IT resources themselves and thus rely on external
+services. Cloud services could help not only to provide a service
+infrastructure for all parties involved, but also to provide the
+required private/public access toolbox. Furthermore, there are still
+many data sources not being indexed properly: chemistry patents are
+often cryptic and chemical image and text mining remains a challenge
+(albeit, currently being addressed in academia and industry
+\cite{Jessop:2011fk,Sayle:2009uq}). The closed CAS (Chemical Abstract
 Service) is a highly trusted source, and public chemistry and
@@
 constant growth, usage, and collaboration scenarios between private
-and public parties.  In a business context we need to address ROI
-(return-of-investment) questions, not only for single institutions and
-companies, but also for collaboration networks of legal partners.  So,
-only when the technical and scientific questions are solved can we
-ensure that legal questions about contribution metrics will get enough
-room to grow. Solving challenges is one thing (science and technical),
-ensuring that the contributing partners get rewarded by their
-contribution another (legal).
+and public parties.
 

### 2011-11-13 rebuttal.tex
updated rebuttal

@@
 
-\comment{P6, c1, l7: I think it is important for the authors to cover ELNs by using actual recorded data.
-Do we know that many scientists changed to ELNs ?many years ago?
-as stated?
-Is there a reference for this or data to back it up?}
-\response{We have expanded our original wording, and we now cite two references about the use and growth of ELNs in chemistry and elsewhere.}
+\comment{P6, c1, l7: I think it is important for the authors to cover
+  ELNs by using actual recorded data. Do we know that many scientists
+  changed to ELNs ?many years ago? as stated? Is there a reference for
+  this or data to back it up?}
+\response{We have expanded our original wording, and we now cite two
+  references about the use and growth of ELNs in chemistry and
+  elsewhere.}
 
-\comment{P6, c1, l7: the authors do not mention the leading players in ELNs (e.g.
-CambridgeSoft, Accelrys, etc.).
-Perhaps this is because open-source ELNs do not exist, or are in their infancy?
-I think the key software houses that develop and supply these ELN systems should be mentioned.
-Furthermore, systems such as Reaxys should be mentioned to ensure that it is understood that these systems do exist to provide these results to these searches and are used daily by chemists.}
-\response{We now mention CambridgeSoft, Accelrys and Reaxys in the article, and provide a citation to a review of 35 ELNs that was published in November 2011.}
+\comment{P6, c1, l7: the authors do not mention the leading players in
+  ELNs (e.g. CambridgeSoft, Accelrys, etc.) \ldots I think the
+  key software houses that develop and supply these ELN systems should
+  be mentioned. Furthermore, systems such as Reaxys should be
+  mentioned \ldots} 
+ \response{We now mention CambridgeSoft, Accelrys and
+  Reaxys in the article, and provide a citation to a review of 35 ELNs
+  that was published in November 2011.}
 

### 2011-11-13 rebuttal.tex
redid the structure enumeration section

@@
   and semantic methods can be used ot address this.}
+
+\comment{Section 2.2. Structure Enumeration: it is not clear what this
+  section is trying to convey to the reader \ldots}
+\response{We have rewritten this section to simplify the language as
+  well as be more explicit in applications of and challenges in this
+  technique. We have also provide relevant references to key topics in
+  this area}
 

### 2011-11-13 paper.tex
redid the structure enumeration section

@@
 labeled graph representations, to discover patterns in large-scale
-datasets \cite{wegner2006,horst2009}.
+datasets \cite{horst2009}.
 
@@
 
-% Fingerprints also facilitate the rapid calculation of quantitative
-% structure-based similarity measures. An example measure of
-% similarity is the \emph{Tanimoto coefficient}, calculated as the
-% ratio $T(a,b) = \frac{c}{(a+b-c)}$, where $c$ is the count of bits
-% on in the same position in both the two fingerprints, $a$ is the
-% count of bits on in the first structure, and $b$ is the count of
-% bits on in the second structure. Historically, this coefficient
-% originated from the comparison of ecosystems around lakes, where a
-% measure was needed to describe how similar those ecosystems are with
-% respect to the animals, plants etc. they contain. However, it can be
-% applied to every other system as well where features can be
-% described in a binary manner. The Tanimoto coefficient varies in the
-% range 0.0 -- 1.0, with a score of 1.0 indicating that the two
-% structures are either very similar or even identical (i.e. their
-% fingerprints are the same).
-%
-
-% Structure-based similarity is however of limited use in grouping of
-% chemicals based on non-structural features, such as shared bioactivity
-% profiles. For this purpose, a large-scale and flexible classification
-% structure is needed for chemicals and their linked biological
-% information. One such answer to this need is provided in the form of
-% \emph{ontologies}.  An ontology is a formal specification of entities
-% and their relationships in a domain of interest. It is structured
-% around an underlying logical representation such as Description
-% Logics~\cite{baaderdl2007}. The logic-based representation allows
-% automated reasoning to derive new knowledge (inferences) from the
-% knowledge encoded, making the representation both compact and
-% powerful.  Ontologies are already in widespread use across the
-% biomedical domain, and in the cheminformatics field ontologies
-% relating to the chemical structures themselves and the protein targets
-% the molecules are active against are probably most common. A
-% similarity measure based on the information encoded in ontologies is
-% termed \emph{semantic similarity}, and such measures have already been
-% applied to enhance chemical classification in several problem areas
-% \cite{couto2010}. Many open questions still remain about the best way
-% to combine the information encoded in chemical structures with
-% chemical ontologies \cite{hastingsowled2010}.
-
-\subsection{Structure Enumeration}
+\subsection{Expanding Chemical Space}
 \label{sec:struct-enum} 
-
 Enumerating molecules is a combinatorial problem that has fascinated
 chemists, computer scientists and mathematicians alike, for more than
-a century. It is one important risk-reduction element in still being
-able to encode closely-related molecules and for keeping the space
-complexity of databases under (algorithmic) control. Indeed, while
-trying to solve the problem of counting the isomers of paraffin
-structures or counting substituted aromatic compounds, fundamental
-principles in graph theory and combinatorics were developed by Cayley,
-Polya and others. Even terms that are widely-used nowadays, like
-``graph'' and ``tree'', were originally coined in a chemistry context
-(Chapters 1 and 8 in \cite{faulon2010}). More than four decades ago,
-Joshua Lederberg, Carl Djerassi, Edward Feigenbaum and others from
-Stanford University started developing codes to explicitly enumerate
-molecules. DENDRAL is generally quoted in artificial intelligence
-textbooks as a pioneer project that produced the first expert
-system~\cite{DENDRAL}.
-
-Enumerating molecules is not only an interesting academic exercise but
-has practical utility as well. The foremost application of enumeration
-is \emph{structure elucidation} of natural compounds such as
-metabolites. Metabolites are small molecules that can be found an
-organism (such as a human) and which in many cases indicate the state
-the organism is in -- such as indicating a disease, or a fasted state,
-etc. -- and with a sufficiently large database the idea is to be able
-to detect the chemical composition of body fluids and to arrive at
-conclusions regarding the particular state of the organism at this
-point in time.  Ideally, the wishful bench chemist collects
-experimental data from an unknown compound, the data is fed into a
-code, and the resulting structure is returned. Although that
-streamlined process may not yield a unique solution, there are
-commercial software products such as MOLGEN (\url{http://molgen.de})
-that can, for instance, list all structures matching a given molecular
-formula, or mass spectrometry (MS) data. Another important application
-of enumeration is \emph{molecular design}. Here the problem is to
-design compounds (drugs, for instance) that optimize some physical,
-chemical, or biological property or activity. Although less prolific
-and more recent than structure elucidation, molecular design has
-introduced some novel solutions to molecular enumeration. Finally,
-with the advent of \emph{combinatorial chemistry} (the rapid synthesis
-of different but related molecules), molecular enumeration has taken a
-central role as it allows computational chemists to construct virtual
-libraries and test hypotheses, and it provides guidance in the design
-of optimal combinatorial experiments.
-
-The major difficulty with enumeration (apart from the size of the
-search space, which is thought to be in the order of $10^{60}$
-molecules for 30 heavy atoms \cite{Bohacek:1996ve}) is that the
-\emph{in silico} representation of molecules, the so-called molecular
-graphs, are labeled objects (i.e. labeled graphs), while the atoms of
-a molecule are of course not uniquely labeled. The mathematical
-concept to tackle this problem is to consider orbits of labeled
-molecular graphs under the operation of the symmetric group. While the
-complexity of molecular graph enumeration remains an open problem it
-has nonetheless been shown that molecules can be sampled
-efficiently \cite{goldberg1999}. Sampling procedures based on the
-Metropolis or the Genetic Algorithms have been developed and used to
-elucidate natural compounds from NMR data.
-
-The underlying computational complexity and potential intractability
-of molecular enumeration may not be a bottleneck after all, as
+a century. Indeed, many fundamental principles of graph therory and
+cominatorics were developed in the context of counting isomers of
+paraffin by Cayley, Polya and others. In the 1960's, Lederberg
+Dejerassi and others developed algorithms to enumerate structures
+based on spectral data, leading to the development of DENDRAL, widely
+considered as the first expert system \cite{DENDRAL}.
+
+From a risk reduction point of view efficient methods to enumerate
+structures allows us to explore new regions of chemical space (for
+example, allowing one to bypass patents), generate structures that
+exhibit desired properties, identify molecules matching experimental
+data and so on. It is important to realize that chemical space (the
+abstract, multi-dimensional space occupied by all possible chemical
+structures) is, in principle, infinite. Even considering molecules for
+just 30 heavy atoms, the size of this space is on the order of
+$10^{60}$ \cite{Bohacek:1996ve}. Any method that purports to enumerate
+molecules will inevitably face a combinatorial explosion if
+implemented na\"{i}vely.
+
+A key application of structure enumeration is the elucidation of
+structures based on spectral data \cite{Kind:2010zr}. This is
+especially relevant for the identifying metabolites, small molecules
+that are the byproducts of metabolic processes and which provide
+insight into the biological state (diseased, fasting, etc.) of an
+organism. Ideally, a chemist would gather spectral (mass, LC/MS, etc.)
+and the algorithm would provide a list of structures that can give
+rise to the observed spectra. Some commercial products such as MOLGEN
+(\url{http://molgen.de}) are able to perform this task in a rapid
+manner.
+
+Another application of structure enumeration is in the area of Markush
+structures and searches \cite{Barnard:1991vn}, where the user
+specifies a pattern such as ``\emph{\ldots an aromatic ring with two
+  alkyl groups attached to it}.'' Such a pattern is very general: an
+alkyl group can be methyl, ethyl or any chain of $n$ carbons and there
+are three possible positions for them on the ring. Even in this simple
+example there are $3n^2$ possible structures. For more complex
+Markushes, there can be billions of possible structures. Clearly,
+explicit enumeration is not feasible. Thus one challenge is to devise
+methods to search (based on structural or property similarity) through
+the implicit enumerated space. A number of commerical vendors such as
+Digitial Chemistry and ChemAxon provide toolkits to handle these
+problems.
+
+Structure enumeration plays a fundamental role in \emph{molecular
+  design} -- the design of compounds (drugs, for instance) that
+optimize some physical, chemical, or biological property or activity
+\cite{Schneider:2005uq}. A key challenge in this area is to combine
+enumeration algorithms with efficient property prediction and is
+closely related to methods in predictive modeling of chemical
+properties (Section \ref{sec:activity-mining}). This problem is also
+termed \emph{inverse QSAR}, where one must devise a set of features
+that describe a molecule with a specified property, and subsequently
+reconstruct the molecule from those features. As a side note, the
+reconstruction of molecules from feature sets is can be considered a
+cryptography problem. It is relevant to the pharmaceutical industry,
+as it can be necessary to share data on molecules without sharing
+explicit structures. Feature sets that allow easy reconstruction of
+molecular structures are thus undesirable. A number of workers have
+investigated the problem of identify such one-way molecular
+descriptors as well as methods to reconstruct molecules from
+descriptors \cite{Masek:2008kx} and new methods that generate such
+one-way molecular features but still correlate with physicochemical
+properties would be invaluable.
+
+While the complexity of molecular graph enumeration remains an open
+problem, an alternative to enumerating the space is to sample it
+\cite{goldberg1999}. Sampling procedures based on the Metropolis or
+the Genetic Algorithms have been developed and used to elucidate
+natural compounds from NMR data. It should be noted that even in the
+face of the computational complexity of molecular enumeration,
 computational chemists have developed and successfully used
 enumeration tools to generate large chemical libraries such as GDB-13,
-comprising almost a billion chemical structures~\cite{GDB}. Yet, the
-current enumeration software products do not generally produce
-stereoisomers or tautomers; these require specific enumeration
-procedures, which are still the subject of investigations by the
-cheminformatics community. In addition, GDB-13 only enumerates
-molecules up to 13 heavy atoms -- given the possibly exponential
-increase in search space with the number of atoms, it seems unlikely
-that the practically relevant chemical space of about 30 heavy atoms
-will be exhaustively enumerated in the near future.
-
-Apart from analytical chemistry and compound design, enumeration plays
-a role is in the \emph{generation of chemical reaction networks}. The
-problem here consists of enumerating all the possible compounds that
-can be produced by applying reaction rules to a set of initial
-molecules. By reversing the reaction rules one can also find sets of
-starting compounds necessary for producing a given target. This latter
-process is known in chemistry as \emph{retrosynthesis}, ever since the
-works of E.J. Corey, who received the Nobel Prize in Chemistry in
-1990. Designing new drugs or chemicals, understanding the kinetics of
-combustion and petroleum refining, studying the dynamics of metabolic
-networks, applying metabolic engineering and synthetic biology to
-produce heterologous compounds in microorganisms, all are applications
-that require the enumeration of reaction networks. As reviewed in
-Chapter 11 in \cite{faulon2010} several network enumeration techniques
-have been developed. However, these techniques generally suffer from a
-combinatorial explosion of intermediate compounds being produced. One
-way to limit the number of compounds generated is to simulate the
-dynamics of the network while it is being constructed and remove
-compounds of low concentration. Following that idea, methods have been
-developed based on the Gillespie Stochastic Simulation Algorithm (SSA)
-to compute on-the-fly species concentrations. Chemical reaction
-network enumeration and sampling is an active field of research,
-particularly in the context of metabolism, either to study
-biodegradation, or to propose metabolic engineering strategies to
-biosynthesize compounds of commercial interest. The difficulty with
-metabolic network design is that in addition to network generation
-based on reactions, one also needs to verify that there are enzymatic
-events possible for enabling reactions catalyses. That additional task
-requires encompassing both chemical and sequence information and the
+comprising almost a billion chemical structures~\cite{GDB} (but only
+considering molecules upto 13 heavy atoms and using only C, N, O, S
+and Cl). Yet, the current enumeration software products do not
+generally produce stereoisomers or tautomers; these require specific
+enumeration procedures, which are still the subject of investigations
+by the cheminformatics community.
+
+Another relevant application of enumeration methods is in the
+\emph{generation of chemical reaction networks}. The problem here
+consists of enumerating all the possible compounds that can be
+produced by applying reaction rules to a set of initial molecules. By
+reversing the reaction rules one can also find sets of starting
+compounds necessary for producing a given target, known as
+\emph{retrosynthesis}. Designing new drugs or chemicals, understanding
+the kinetics of combustion and petroleum refining, studying the
+dynamics of metabolic networks, applying metabolic engineering and
+synthetic biology to produce heterologous compounds in microorganisms,
+all involve enumeration of reaction networks. As reviewed in Chapter
+11 in \cite{faulon2010} several network enumeration techniques have
+been developed. However, these techniques generally suffer from a
+combinatorial explosion of product compounds. One way to limit the
+number of compounds generated is to simulate the dynamics of the
+network while it is being constructed and remove compounds of low
+concentration. Following that idea, methods have been developed based
+on the Gillespie Stochastic Simulation Algorithm (SSA) to compute
+on-the-fly species concentrations. Chemical reaction network
+enumeration and sampling is an active field of research, particularly
+in the context of metabolism, either to study biodegradation, or to
+propose metabolic engineering strategies to biosynthesize compounds of
+commercial interest. The difficulty with metabolic network design is
+that in addition to network generation based on reactions, one also
+needs to verify that there are enzymatic events possible for enabling
+reactions catalyses. That additional task requires encompassing both
+chemical structure and protein sequence information and the
 development of tools that are at the interface between cheminformatics
@@
 biological activity of a molecule is a function of its chemical
-structure. Together with the \emph{similarity property principle}
+structure. Together with the \emph{similar property principle}
 \cite{Johnson:1990qf} (structurally similar molecules will exhibit

### 2011-11-13 paper.tex
Updated glossary

@@
     peptides and other chemical compounds.
-\item \textit{Metabolites} - Metabolites are the intermediates and products of metabolism. The term metabolite is
-    usually restricted to small molecules (chemical compounds).
-\item \textit{Proteochemometric modeling} - combining data mining for chemical compounds (ligands) and proteins
-    within the same machine learning framework.
 \item \textit{Protonation and protonation state} - In chemistry, protonation is the addition of a proton (H$^+$) to
@@
     neurotransmitter, or other substance.
-\item \textit{Regulation authorities} - In the US the Food and Drug Administration (FDA) is responsible for
+\item \textit{Regulatory authorities} - In the US the Food and Drug Administration (FDA) is responsible for
     protecting and promoting public health through the regulation and supervision of e.g. dietary supplements,

### 2011-11-13 rebuttal.tex
restructured conclusions. responded to reviewer 2

@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \fbox{\textbf{Reviewer 2}}
@@
 
+\comment{ I entirely dispute that cheminformatics methods have been
+  'closely guarded secrets of companies...' since many methods have been
+  published in the public domain. The authors suggest that only in the
+  last decade has the field gained access to freely available
+  software, etc.}
+\response{We agree that secret was probably not the best choice of
+  words. However we do believe that much cheminformatics software and
+  data was proprietary till recently. We have updated the text on
+  pages 7 and 8 to expand on this and provide a more detailed
+  discussion}
+
 \comment{If this manuscript were to be published it would be
@@
   thus amenable to coss-disciplinary efforts.}
+
+%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+\comment{P7, c1, l9: the authors make an interesting point here
+  regarding why cheminformatics software has historically been from
+  commercial suppliers. Given the scope of this manuscript it would be
+  interesting to expand on this discussion.} 
+\response{We have
+  restructured the Conclusions section to expand on the role of
+  cheminformatics in industry vis a vis bioinformatics, and why much
+  of cheminformatics tools and data have been commercial, in contrast
+  to freely available tools and data in bioinformatics}
+
 \end{document}

### 2011-11-13 paper.tex
restructured conclusions. responded to reviewer 2

@@
 Cheminformatics, the computer science of chemical discovery, is an
-industrial and academic discipline with roots dating back to the
-1960s, with the creation of DENDRAL, the first expert system of any
-kind.  Much of the development of cheminformatics from the 1960s until
-the new millenium was performed in secret, because of competition in
-the pharmaceutical industry.  While chemists were among the first
-scientists to embrace web-based tools of knowledge management,
-research into fundamental algorithmic questions of structure
-enumeration and database management (and a host of other topics) have
-suffered, because of limited academic access to either tools or data.
-As we enter the second decade of the 21st century, however, this
+industrial and academic discipline with roots dating back to the 19th
+century and flowering in the 1960s with the advent of modern computing
+technologies. While many key cheminformatics techniques have been
+available in the literature since the 1960's, the bulk of
+cheminformatics tools and implementations have been proprietary. In
+addition, most data sources of chemical information have been
+proprietary (usually with restrictive licensing policies) until
+recently. There are many reasons for this, primarily revolving around
+issues of profits, competitive intelligence and intellectual property,
+which are relevant to the pharmaceutical as well as other chemical
+industries. This is in contrast to much of bioinformatics, where the
+data and tools have been freely available since the inception of the
+field. One could argue that the outcomes from cheminformatics methods
+and software have a more direct impact on profits, in terms of
+identifying lead like compounds and improving properties of drug
+candidates. In contrast, much bioinformatics software is employed in
+upstream areas such as target identification.  While certainly
+important, this type of research is not as directly related to
+possible profits as an actual small molecule.
+
+Furthermore, chemical data (structure, activity information etc.) is
+harder to aquire and when done on large scales can be extremely
+intensive in terms of time and effort. Thus much of it has been
+proprietary. Whereas a lot of bioinformatics data (sequences for
+example) are much easier to obtain. While both fields have theoretical
+components, it is possible that the free availability of large amounts
+of data in bioinformatics drove the development of publically
+available tools to handle it. In contrast, the proprietary nature of
+chemical data would imply that tools to process and analyse that data
+were of interest primarily to the owners. As a result, the incentive
+to making cheminformatics software public has been low and since the
+users of it were primarily industrial, commercial software was the
+norm.
+
+Of course, with the advent of systems biology and its large amounts of
+heterogenous data and wide array of (possibly profitable) commercial
+applications and, this may change. Indeed, in this field, the
+requriement to consider small molecules in the context of larger
+biosystems implies that cheminformatics and bioinformatics will both
+be required to address the problems that will be tackled.
+
+As we enter the second decade of the 21st century, the cheminformatics
 landscape is rapidly shifting.  There now exist free-of-charge
-open-source cheminformatics toolkits, and open databases with tens of
-millions of compounds.  We believe the prospects for academic and
-industrial collaboration between chemists and computer scientists are
-bright, and we would like to encourage more computer scientists to
-participate in cheminformatics research.  There is an admittedly steep
-learning curve, as the structural issues faced by chemists are often
-too detailed and subtle to admit as clean an abstraction as
-``algorithms on a string,'' the way many bioinformatics algorithms can
-be abstracted into a computer science framework.  Nevertheless,
-because of the changing landscape, there are many research questions
-we are only now in position to try to answer: for a theorist,
+open-source cheminformatics toolkits, applications and open databases
+with tens of millions of compounds and containing valuable
+experimental bioactivity data. Modern experimental techniques such as
+high throughput sequencing have made the generation of large amounts
+of structure-activity data much more accessible.  We believe the
+prospects for academic and industrial collaboration between chemists
+and computer scientists are bright, and we would like to encourage
+more computer scientists to participate in cheminformatics research.
+
+There is an admittedly steep learning curve, as significant domain
+knowledge is required to attack problems of significance. Indeed, many
+issues faced by chemists are often detailed and subtle and do not
+admit as clean an abstraction as ``algorithms on a string'', the way
+many bioinformatics algorithms can be abstracted into a computer
+science framework.  As a result, while one can certainly make
+significant contributions to cheminformatics by just considering
+structures as graphs, one is limited to somewhat abstract
+problems. Keeping in mind that cheminformatics is fundamentally a
+practical field that serves to help and advance experimental
+chemistry, the key challenges require an understanding of the
+underlying chemical systems. One could certainly study for a degree in
+chemistry. Alternatively one could collaborate with a chemist or
+cheminformatics practitioner.
+
+Nevertheless, because of the increasing availability of tools and
+data, the barrier to entry for non-chemists and non-cheminformaticians
+is significantly lower. Many questions that benefit from computer
+scientists can now be comprehensively taclked: for a theorist,
 graph-theoretic questions of three-dimensional enumeration; for a
@@
 relationships, transformations, and combinatorial challenges.
-
-In 2010, drug company Glaxo made public over 13,000 compounds that
-they have investigated as potential cures for
-malaria~\cite{glaxo-malaria}.  (Malaria has traditionally been a
-neglected disease in drug research, because the populations who suffer
-from it tend not to have enough money to be a stable consumer base for
-a new pharmaceutical.)  Perhaps we will now be able to find cures for
-some of these neglected diseases, by using the open source, open data
-approach to the computer science of chemical discovery.
 

### 2011-11-13 paper.tex
some fixes to the bibliogrpahy. rephrased conclusions

@@
 There is an admittedly steep learning curve, as significant domain
-knowledge is required to attack problems of significance. Indeed, many
+knowledge is required to attack many problems in cheminformatics due
+to the underlying chemistry of the systems being studies. Indeed, many
 issues faced by chemists are often detailed and subtle and do not

### 2011-11-13 paper.tex
some more updates

@@
 In many cases, the activity of a small molecule is due to its
-interaction with a receptor. Traditionally, QSAR' \cite{Hansch:1962vn,
+interaction with a receptor. Traditionally, QSAR \cite{Hansch:1962vn,
   Free:1964ys} approaches do not take into account receptor features,
@@
 structures. The last method is an example of an extension of
-statistical QSAR methods to simultaneous model the f receptor and
+statistical QSAR methods to simultaneous model the receptor and
 ligand as first reported by Lapinsh \textit{et  al.}~\cite{lapinsh2001}.
@@
 cheminformatics setting. In some cases, pure predictive ability is
-desired -- such as in a virtual screening setting. (Basically, only
-the reliability of the model matters, but other aspects such as
-interpretability are of minor importance.) For such cases, the
+desired -- such as in a virtual screening setting. For such cases, the
 complexity of the modeling approach is immaterial, as it leads to
@@
 biologists, the interpretability of a model and risk reduction
-arguments are paramount, since we have to translate better-predicted
-\emph{in silico} compounds into real-world compounds. In such cases,
-``black-box methods'' can be less useful, though attempts have been
-made to interpret vectorial models built using neural networks,
-na\"{i}ve Bayes, Support Vector Machines or random forests as well.
+arguments are paramount, since we have to translate the predicted
+outputs into real-world compounds. In such cases, ``black-box
+methods'' can be less useful, though attempts have been made to
+interpret vectorial models built using neural networks, na\"{i}ve
+Bayes, Support Vector Machines or random forests.
 
@@
 afford us the opportunity to identify targets, characterize off-target
-effects and so on. However, most approaches to this problem are rather
-simplistic since they develop multiple individual models
-\cite{Chen:2010zr}, leading to multiple, independent predictions for a
-given input molecule.  Instead, one could imagine an approach that
-takes into account the covariance structure of the multiple observed
-activities and the structural descriptors within a single model. Such
-an approach could lead to more robust predictions for panel
-assays. Finally, this approach also better reflects clinically
-relevant compound profiles~\cite{kuhn2010} and a ``personalized
-medicine'' concept (\emph{e.g.}, drug-drug interaction profiles).
+effects and so on. However, most approaches to this problem temd tp
+develop multiple individual models \cite{Chen:2010zr}, leading to
+multiple, independent predictions for a given input molecule.
+Instead, one could imagine an approach that takes into account the
+covariance structure of the multiple observed activities and the
+structural descriptors within a single model. Such an approach could
+lead to more robust predictions for panel assays. Finally, this
+approach also better reflects clinically relevant compound
+profiles~\cite{kuhn2010} and the ``personalized medicine'' concept
+(\emph{e.g.}, drug-drug interaction profiles).
 
@@
 publications.  Interestingly, patent information is publicly
-accessible (cf. Google Patents) and there have been efforts to mine
-these for chemical information using Open tools.
+accessible (cf. Google Patents) and represents an excellent resource
+to support these types of efforts.
 

### 2011-11-13 rebuttal.tex
more updats to the text

@@
 
+\comment{The paragraph in the second column discussing descriptors and
+  the proper description requires rewriting too since it provides a
+  very vague overview that is not particularly informative.}
+\response{We have updated the discussion of descriptors to be more
+  focused}
+
 \comment{page 1, column 1, line 24: the authors need to provide a
   reference for the assertion that cheminformatics is an older field,
-  particularly given the comments later in this
-  manuscript.}
+  particularly given the comments later in this manuscript.}
 \response{We have removed the assertion that cheminformatics is older
@@
   it up. Why should multi-target models be more reliable? It is not
-  clear to me that they should.}\marginpar{\textbf{Needs more work}}
-\response{We have rephrased this statement to be less dogmatic}
+  clear to me that they should.}
+\response{We have rephrased this statement to be less dogmatic. While
+  it is true that for the case of linear models, a multivariate
+  multiple regression is in general, equivalent to multiple individual
+regression models, one could argue that taking into account the
+correlation structure between the multiple y's could lead to }
 

### 2011-11-13 paper.tex
more updats to the text

@@
 The first step in the predictive modeling of biological activities is
-to describe compounds in a way amenable to computerized data mining,
-while at the same time capturing information relevant for the property
-one attempts to model. Here multiple approaches exist: Either we
-transform molecular graphs into a numerical vector representation
-(a.k.a. \emph{descriptors} or \emph{features}) of chemical structures,
-or we use algorithms for comparing molecular graphs directly
-(\emph{molecule kernel}). In the first case thousands of descriptors
-(many of which are equivalent) have been described
-\cite{todeschini2000}, and a variety of tools are available for their
-calculation. Given a set of molecules, observed activities and a
-descriptor vector, the problem can be considered a traditional
-statistical modeling problem. In the second case no explicit vector
-representation is required, which prevents the challenge of
-transforming the molecular graph into a vector without losing
-information. On the other hand, the implicit kernel encoding imposes
-the challenge that we have to compare all compounds against all
-compounds, which imposes large-scale mining challenge, which risks
-overrunning available processing time and space. Note that the public
-PubChem dataset alone contains over 30 million chemical compounds and
-over 70 million substances.
-
-In order to judge which predictive modeling approach is applicable in
-every case, we firstly have to consider the goal of a the model in a
-cheminformatics setting. In some cases, pure predictive ability is
-desired -- such as in a virtual screening setting. For such cases, the
-complexity of the modeling approach is immaterial, as it leads to
-accurate and generalizable results. Here a large-scale mining capacity
-is crucial. On the other hand, when working with chemists and
-biologists, the interpretability of a model and risk reduction
-arguments are paramount, since we have to translate the predicted
-outputs into real-world compounds. In such cases, ``black-box
-methods'' can be less useful, though attempts have been made to
-interpret vectorial models built using neural networks, na\"{i}ve
-Bayes, Support Vector Machines or random forests.
-
-Secondly, the concept of model applicability -- the question of when
-the prediction of the model for a new object is reliable -- has grown
-in importance with the increasing use of predictive models in
-regulatory settings. A variety of methods have been developed that
-attempt to characterize the domain of a model. These methods not only
-determine whether models are applicable, but can also be used for
-deciding if additional biological experiments are required for
-reducing the prediction error on certain compound classes.
+to generate molecular descriptors (a.k.a, features) that are numerical
+representations of different structural features. Here multiple
+approaches exist. We can transform molecular graphs into a numerical
+vector representation (a.k.a. \emph{descriptors} or \emph{features})
+of chemical structures \cite{todeschini2000}, ranging from counts of
+elements to eigenvalues of the Laplacian matrix.  Alternatively, we
+can compare molecular graphs directly, via kernel methods. In these
+cases, rather than compute vectorial representations, we directly
+operate on the graph representations.
+
+Both methods have advantages and disadvantages. The vectorial approach
+requires one to identify a subset of \emph{relevant} (to the property
+being modeled) descriptors. This is the feature selection problem and
+is well discussed in the data mining literature. A kernel approach
+does not require feature selection; but now, one faces the problem of
+evaluating a data set in a pairwise fashion. Furthermore, one must now
+identify an appropriate kernel. This is an important challenge as the
+kernel should be selected to satisfy Mercer's condition, and this is
+not always possible with traditional cheminformatics based kernels
+(such as a multiple common substructure based kernel). These
+challenges can make kernel based methods prohibitive on very large
+datasets.
+
+Having settled on a numerical representation and a possible class of
+model types, one must address the goal of the model. Are we
+looking for pure predictive power, with no interest in any explanatory
+features? Or are we more interested in decent predictive power, but
+some of explanation as to why a molecule is predicted as toxic or
+non-toxic, for example. The former scenario is common in virtual
+screening settings, where we require high degree of accuracy and fast
+predictions, but do not really care as to why one molecule is active
+and another inactive. These types of models can be black boxes (neural
+networks) and algorithmic models \cite{Breiman:2001fk} (random
+forests). The latter scenario is more common in exploratory and
+optimization settings, where we hope that the output of the model will
+guide us in chemical modifications to improve the property being
+modeled.  In this case, we require to understand the effect of a
+certain structural feature on the observed potency and expect that the
+model will provide insight. These types of models are generally
+distributional in nature (such as linear regression and PLS) though
+some algorithmic approaches such as random forests can provide some
+degree of insight.
+
+Having chosen a modeling approach, a vital question is that of model
+reliability which is closely tied to the concept of model
+applicability -- the question of when the prediction of the model for
+a new object is reliable. This issue has grown in importance with the
+increasing use of predictive models in regulatory settings. A
+misprediction (due to the model not bein applicable to a certain class
+of inputs) can have significant financial repurcussions. A variety of
+methods have been developed that attempt to characterize the domain of
+a model. These methods not only determine whether models are
+applicable, but can also be used for deciding if additional biological
+experiments are required for reducing the prediction error on certain
+compound classes.
 
@@
 4D-QSAR methodology described by Hopfinger and co-workers
-\cite{Albuquerque:1998ys}.  Computer science techniques such as
+\cite{Albuquerque:1998ys}.  Recent techniques such as
 \emph{multiple-instance learning} could also be applied to the

### 2011-11-15 paper.tex
added in Aarons latest edits

@@
   new therapeutic molecules.  Only within the last decade -- and as an
-  effect of a change in mindsets, government-mandated policy changes
-  and last but not least because of chemists volunteering their time
+  effect of a change in mindsets, government-mandated policy changes,
+  and, not least, because of chemists volunteering their time
   for an Open Science ``movement'' -- have researchers gained access to
@@
 chemical structure records, along with an even larger number of
-annotations such as synonyms, known targets, mode of action,
+\emph{annotations} (comments on the structure that are database searchable), such as synonyms (different names for the same molecule), known targets (drugs with which the molecule is known to interact), mode of action (how the molecule reacts with others),
 regulatory approval status and so on.  Simply considering the synonym
@@
 structures may possess rings, branches as well as multiple valid
-representations of the same structure (such as so-called
-``tautomers'', where hydrogen atoms can be positioned in different
+representations of the same structure (such as \emph{tautomers}, where hydrogen atoms can be positioned in different
 places of the molecules), potentially giving rise to
@@
 
-Although graphs implicitly encodes the 3D structure of a molecule
+Although graphs implicitly encode the 3D structure of a molecule
 (when combined with knowledge about favored bond angles and
@@
 goes further and represents a molecular surface, encoding some
-property (such as lipophilicity) that varies over the surface.
+property (such as lipophilicity, the ability to dissolve in fats) that varies over the surface.
 
@@
 millions (PubChem) of data points each. Still, the integration of
-chemical databases with each other remains a challenge, not only due
+chemical databases with one other remains a challenge, not only due
 to the normalization problems in the chemical and bioactivity
@@
 hand as possible, we have to be able to search in complex data, for
-example multi-labelled molecular graphs, where graph labels can be
-different. Hence, we need to be able to capture molecular graphs in a
+example multi-labelled molecular graphs, where graph labels can change from database to database. Hence, we need to be able to capture molecular graphs in a
 risk-bounded machine encoded way.
@@
 molecule. When one considers the need to include other aspects such as
-protonation and tautomer states, the challenge increases. The need for
+protonation states (how the molecule changes when one proton is added to it) and tautomer states (how the molecule can change when a proton migrates from one part of the molecule to another), the challenge increases.
+The need for
 a unique representation that is invariant with respect to the atom
@@
 whether two molecules are the same) is an expensive operation. The
-first such canonicalization algorithm was described by Morgan
+first such ``unique representation algorithm,'' or \emph{canonicalization} algorithm, was described by Morgan
 \cite{Morgan1965}, allowing one to generate unique string
@@
 canonicalization algorithm, usually based on the Morgan algorithm. See
-Warr (Warr \cite{Warr:2011vn}) for an extensive dicussion on chemical
+Warr~\cite{Warr:2011vn} for an extensive discussion on chemical
 structure representations.
@@
 hashed key for the InChI, which has an invariant length of 14
-characters. (Unfortunately, after implementation and publication of
-the hashing scheme it became apparent that in particular in case of
-stereochemically related molecules clashes of the InChI key take place
-in some, albeit rare cases.) Figure~\ref{figure:smiles} illustrates
+characters. Figure~\ref{figure:smiles} illustrates
 the SMILES, InChI and InChIKey for lipoic acid.
@@
 structures, but it is still limited. For example, it cannot
-differentiate between certain types of stereoisomers:
+differentiate between certain types of stereoisomers (informally, molecules that are ``3D mirror images'' of each other):
 Figure~\ref{figure:cistrans} (in the Appendix) illustrates two
@@
 and hashed keys. In the former, each bit position corresponds to a
-distinct substructure such as an alcohol or keto group. Examples
+distinct substructure such as a \emph{functional group} (a connected set of atoms that affects the characteristics of the chemical reactions of the molecule). Examples
 include MACCS and PubChem keys. The latter class are designed such
@@
 they can cover an arbitrarily large collection of substructures such
-as paths of length $N$ or circular environments and so on. Examples
-include ECFPs and Daylight fingerprints.
+as ``paths of length $N$,'' or circular environments, and so on. Examples
+include ECFPs (Extended Connectivity Fingerprints, which use local topological information) and Daylight fingerprints (which are ``folded'' to optimize information density and screening speed).
 
@@
 rapidly on modern hardware. As a result, we apply the actual
-isomoprhism test only on those molecules that pass the
+isomorphism test only on those molecules that pass the
 screen. Fingerprints can also be used to rapidly search databases for
@@
 such databases. Such integration forms the basis of ``systems'' level
-investigations of biological systems. A current challege in
+investigations of biological systems. A current challenge in
 cheminformatics is how to identify ``equivalent'' objects across data
@@
 same or similar targets; they may be have similar biological function
-(say, anti-neoplastic) and so on. To assess such higher level
+(say, anti-neoplastic, that is to say, it prevents the development of tumors) and so on. To assess such higher level
 characteristics, ontologies and semantic technologies have been
@@
 challenges remain in this area - the use of ontologies allows one to
-build a semntic network of concepts that links molecules to other
+build a semantic network of concepts that links molecules to other
 molecules as well as other objects such as targets, genes and
-pathways. How does one evaluate similarities in such a heterogenous
+pathways. Open questions include the following.  How does one evaluate similarities in such a heterogenous
 network \cite{couto2010}? How does one integrate chemical structure
@@
 comprising almost a billion chemical structures~\cite{GDB} (but only
-considering molecules upto 13 heavy atoms and using only C, N, O, S
-and Cl). Yet, the current enumeration software products do not
+considering molecules up to 13 heavy atoms and using only carbon, nitrogen, oxygen, sulfur
+and chlorine). Yet, the current enumeration software products do not
 generally produce stereoisomers or tautomers; these require specific
@@
 dynamics of metabolic networks, applying metabolic engineering and
-synthetic biology to produce heterologous compounds in microorganisms,
+synthetic biology to produce heterologous compounds (compounds from different species) in microorganisms,
 all involve enumeration of reaction networks. As reviewed in Chapter
@@
 needs to verify that there are enzymatic events possible for enabling
-reactions catalyses. That additional task requires encompassing both
+reactions catalyses (in other words, enzymes must be present to reduce the energy required for some reactions to take place). That additional task requires encompassing both
 chemical structure and protein sequence information and the
@@
 to assess the error in (or conversely, the reliability or confidence of) a
-prediction, ranging from statistical approaches (XXX) to more
+prediction, ranging from statistical approaches to more
 empirical approaches such as defining an \emph{applicability domain} -
@@
 information on ligand-receptor interactions. As a result, techniques
-such as docking, pharmacophore modeling and proteochemometric methods
+such as docking (which predicts how molecules fit together), pharmacophore modeling (which predicts the surface features of a molecule) and proteochemometric methods (which predict the interaction between small molecules and proteins)
 have been designed to take into account both ligand and receptor
@@
 The first step in the predictive modeling of biological activities is
-to generate molecular descriptors (a.k.a, features) that are numerical
+to generate \emph{molecular descriptors} (a.k.a. \emph{features}) that are numerical
 representations of different structural features. Here multiple
 approaches exist. We can transform molecular graphs into a numerical
-vector representation (a.k.a. \emph{descriptors} or \emph{features})
-of chemical structures \cite{todeschini2000}, ranging from counts of
+vector representation of chemical structures \cite{todeschini2000}, ranging from counts of
 elements to eigenvalues of the Laplacian matrix.  Alternatively, we
-can compare molecular graphs directly, via kernel methods. In these
+can compare molecular graphs directly, via \emph{kernel methods}, where a kernel on graphs $G$ and $G'$ provides a measure of how similar $G$ is to $G'$, or a kernel on a single graph compares measures similarities between the graph's nodes.
+In these
 cases, rather than compute vectorial representations, we directly
@@
 identify an appropriate kernel. This is an important challenge as the
-kernel should be selected to satisfy Mercer's condition, and this is
+kernel should be selected to satisfy \emph{Mercer's condition} (a well-known mathematical property in machine learning that makes a set of observations easier to make predictions about), and this is
 not always possible with traditional cheminformatics based kernels
@@
 features? Or are we more interested in decent predictive power, but
-some of explanation as to why a molecule is predicted as toxic or
+also some explanation as to why a molecule is predicted as toxic or
 non-toxic, for example. The former scenario is common in virtual
-screening settings, where we require high degree of accuracy and fast
-predictions, but do not really care as to why one molecule is active
+screening settings, where we require a high degree of accuracy and fast
+predictions, but do not really care why one molecule is active
 and another inactive. These types of models can be black boxes (neural
@@
 guide us in chemical modifications to improve the property being
-modeled.  In this case, we require to understand the effect of a
+modeled.  In this case, we need to understand the effect of a
 certain structural feature on the observed potency and expect that the
 model will provide insight. These types of models are generally
-distributional in nature (such as linear regression and PLS) though
-some algorithmic approaches such as random forests can provide some
-degree of insight.
+distributional in nature (such as linear regression and Partial Least Squares) though
+some algorithmic approaches such as random forests can provide insight.
 
@@
 increasing use of predictive models in regulatory settings. A
-misprediction (due to the model not bein applicable to a certain class
-of inputs) can have significant financial repurcussions. A variety of
+misprediction (due to the model not being applicable to a certain class
+of inputs) can have significant financial repercussions. A variety of
 methods have been developed that attempt to characterize the domain of
@@
 conformations. As a result, enhancing the accuracy of predictions
-ideally will require that such aspects be taking into account and, as
+ideally will require that the 3D geometries of that molecule be taking into account and, as
 far as possible, take into account the receptor
@@
 afford us the opportunity to identify targets, characterize off-target
-effects and so on. However, most approaches to this problem temd tp
+effects and so on. However, most approaches to this problem tend to
 develop multiple individual models \cite{Chen:2010zr}, leading to
@@
 structural descriptors within a single model. Such an approach could
-lead to more robust predictions for panel assays. Finally, this
+lead to more robust predictions for panel assays (that is, which battery of tests would be most useful to perform). Finally, this
 approach also better reflects clinically relevant compound
@@
 a rapidly growing number of external/public partners. In
-cheminformatics and chemistry, scientists already switched many years
+cheminformatics and chemistry, scientists already switched years
 ago from paper lab-notebooks to online collaboration platforms, called
@@
 regulatory compliance (timestamped experimental data, experiment and
-data approvals and QC information). Given that any experiment could
+data approvals and quality control information). Given that any experiment could
 become the core of a patent filing, ELNs allow organizations to define
@@
 collaboration with external parties, e.g. contract research, poses
-other problems including compound deduplication and efficient data
+other problems including compound duplication and efficient data
 synchronization. If external partners are small they often do not even
@@
 candidates. In contrast, much bioinformatics software is employed in
-upstream areas such as target identification.  While certainly
+upstream areas such as target identification (finding molecules that react well with the molecule under consideration).  While certainly
 important, this type of research is not as directly related to
@@
 Furthermore, chemical data (structure, activity information etc.) is
-harder to aquire and when done on large scales can be extremely
+harder to acquire and when done on large scales can be extremely
 intensive in terms of time and effort. Thus much of it has been
@@
 components, it is possible that the free availability of large amounts
-of data in bioinformatics drove the development of publically
+of data in bioinformatics drove the development of publicly
 available tools to handle it. In contrast, the proprietary nature of
-chemical data would imply that tools to process and analyse that data
+chemical data would imply that tools to process and analyze that data
 were of interest primarily to the owners. As a result, the incentive
-to making cheminformatics software public has been low and since the
+to make cheminformatics software public has been low and since the
 users of it were primarily industrial, commercial software was the
@@
 heterogenous data and wide array of (possibly profitable) commercial
-applications and, this may change. Indeed, in this field, the
-requriement to consider small molecules in the context of larger
+applications, this may change. Indeed, in biology, the
+requirement to consider small molecules in the context of larger
 biosystems implies that cheminformatics and bioinformatics will both
-be required to address the problems that will be tackled.
+be required to tackle future problems.
 
@@
 chemistry, the key challenges require an understanding of the
-underlying chemical systems. One could certainly study for a degree in
-chemistry. Alternatively one could collaborate with a chemist or
-cheminformatics practitioner.
+underlying chemical systems. 
+%One could certainly study for a degree in chemistry. Alternatively one could collaborate with a chemist or cheminformatics practitioner.
 
@@
 data, the barrier to entry for non-chemists and non-cheminformaticians
-is significantly lower. Many questions that benefit from computer
-scientists can now be comprehensively taclked: for a theorist,
+is significantly lower than it was a decade ago. Many questions that benefit from computer
+science can now be comprehensively tackled: for a theorist,
 graph-theoretic questions of three-dimensional enumeration; for a

### 2011-11-16 paper.tex
added more typo fixes from Aaron

@@
 choice of representation is key in determining how such similarities
-are evalaued, and thus the effectiveness of subsequent analyses. But
+are evaluated, and thus the effectiveness of subsequent analyses. But
 even after one has selected a representation one faces the challenge
@@
 obtain accurate predictions from a subset of conformers? Or, can we
-obtain comparable accuracy by using a 2D representation? And is so,
+obtain comparable accuracy by using a 2D representation? And so,
 what type of labels are required? Currently, many of these questions
@@
 generally, how can we achieve automated inferencing in such
-heterogenous linked systems, allowing us to acheive a systems level
+heterogenous linked systems, allowing us to achieve a systems level
 understanding of small molecules \cite{Oprea:2007fk}.
@@
 chemists, computer scientists and mathematicians alike, for more than
-a century. Indeed, many fundamental principles of graph therory and
-cominatorics were developed in the context of counting isomers of
+a century. Indeed, many fundamental principles of graph theory and
+combinatorics were developed in the context of counting isomers of
 paraffin by Cayley, Polya and others. In the 1960's, Lederberg
@@
 structures based on spectral data \cite{Kind:2010zr}. This is
-especially relevant for the identifying metabolites, small molecules
+especially relevant for identifying metabolites, small molecules
 that are the byproducts of metabolic processes and which provide
 insight into the biological state (diseased, fasting, etc.) of an
-organism. Ideally, a chemist would gather spectral (mass, LC/MS, etc.)
+organism. Ideally, a chemist would gather spectral data (mass, LC/MS, etc.),
 and the algorithm would provide a list of structures that can give
@@
 challenge due to differences in identifiers, synchronization, curation
-\& error correctin mechanisms and efficient substructure and
+\& error correcting mechanisms and efficient substructure and
 similarity search within complex data types (compounds). A
@@
 \bibliographystyle{abbrv}
-\bibliography{paper}
+\bibliography{resubmission}
 

### 2011-11-16 paper.tex
fixed a typo
more edits

@@
 large public small molecule structure repository (PubChem) became
-freely available. This was followed by other databases, as discussed
-later in this article. While much of the foundational algorithms of
-cheminformatics have been described since the 1950s, publically
-available software implementing many of these algorithms have only
-been accessible in the last ten to fifteen years \cite{faulon2010}.
-
-So why should we actually care about chemical information being made
-public -- and how does this relate to the field of computer science?
-
-Making chemical information public is important because the drugs we
-take are the result of intensive and costly research, which is
-facilitated by the amount of information available. The more
-information is shared (a process very hard to achieve in
-pharmaceutical companies), the more information is available to each
-researcher, facilitating the development of novel treatments. The
+freely available. This was followed by other databases. While much of
+the foundational algorithms of cheminformatics have been described
+since the 1950s, publicly available software implementing many of
+these algorithms have only been accessible in the last ten to fifteen
+years \cite{faulon2010}.
+
+So why is chemical information important and why should we care about
+its public availability -- and how does this relate to the field of
+computer science?
+
+Given that the art and science of drug discovery is based around small
+molecules, techniques and algorithms to handle, analyse and interpret
+chemical structure information are a core component of the drug
+discovery toolbox. Public availability of chemical information and
+tools to handle it are important as drug discovery research is slow,
+expensive and prone to failure. The more information available to each
+researcher, facilitating the development of novel treatments, the
+better the chances of avoiding the many causes of attrition. The
 relevance of computer science to this research becomes clear when we
@@
 chemical structure records, along with an even larger number of
-\emph{annotations} (comments on the structure that are database searchable), such as synonyms (different names for the same molecule), known targets (drugs with which the molecule is known to interact), mode of action (how the molecule reacts with others),
+\emph{annotations} (comments on the structure that are database
+searchable), such as synonyms (different names for the same molecule),
+known targets (drugs with which the molecule is known to interact),
+mode of action (how the molecule interacts with its targets),
 regulatory approval status and so on.  Simply considering the synonym
@@
 
-The design of data structures for this chemical information, as well
-as subsequent data mining of this information, are areas where
-expertise in computer science is urgently needed. (For a recent,
+The design of data structures to handle chemical structures and
+associated information, as well as subsequent data mining are areas
+where expertise in computer science is urgently needed. (For a recent,
 detailed introduction to cheminformatics, written for computer
@@
 Language for publications); \emph{storing data} (designing database
-schemas, devising ontologies) and \emph{mining data} (such as for
-bioactivity prediction, which might involve algorithms based on
-graphs).  However, the type of information being analyzed is different
-from biological information: while bioinformatics often deals with
-sequences, the domain of cheminformatics is chemical structures. In
-the case of the former, information can very frequently be represented
-as one-dimensional strings, which are relatively easy to handle
+schemas, devising ontologies) and \emph{mining data} (such as for the
+prediction of biological activities of compounds).  However, the type
+of information being analyzed is different from biological
+information: while bioinformatics often deals with sequences, the
+domain of cheminformatics is chemical structures. In the case of the
+former, information can very frequently be represented as
+one-dimensional strings, which are relatively easy to handle
 computationally. In the case of the latter however, chemical
 structures may possess rings, branches as well as multiple valid
-representations of the same structure (such as \emph{tautomers}, where hydrogen atoms can be positioned in different
-places of the molecules), potentially giving rise to
-ambiguities. Hence, chemical structures are often reported to be more
-difficult to standardize.
+representations of the same structure (such as \emph{tautomers}, where
+hydrogen atoms can be positioned in different places of the
+molecules), potentially giving rise to ambiguities. Hence, chemical
+structures are often reported to be more difficult to standardize and
+algorithms dealing with them must take this into account.
 
@@
 
-Although graphs implicitly encode the 3D structure of a molecule
-(when combined with knowledge about favored bond angles and
-distances), many different low energy 3D structures, or conformers,
-may be consistent with the same graph structure. In addition, the 3D
-structure may have a ``secondary'' geometrical arrangement of features
-(such as the presence of a right-handed helix) which cannnot be
-encoded in the graph. Thus we have 3D representations
+Although graphs implicitly encode the 3D structure of a molecule (when
+combined with knowledge about favored bond angles and distances), many
+different low energy 3D structures, or conformers, may be consistent
+with the same graph structure. In addition, the 3D structure may have
+a ``secondary'' geometrical arrangement of features (such as the
+presence of a right-handed helix) which cannot be encoded in the
+graph. Thus we have 3D representations
 (Fig.~\ref{figure:chemical-structures}C) that make the 3D arrangement
@@
 goes further and represents a molecular surface, encoding some
-property (such as lipophilicity, the ability to dissolve in fats) that varies over the surface.
+property (such as lipophilicity, the ability to dissolve in non-polar
+solvents) that varies over the surface.
 
@@
 is often a complex mixture of combinatorial and continuous
-problems. If we oversimplify the nature of chemical structures and
-their properties, we can not describe the world around us in
-sufficient detail. On the other hand, if we try to encode every detail
-(\emph{e.g.}, all possible 3D structures and their properties, all
-metabolites of a structure in humans in different tissues), the
-complexity of collecting, storing, and mining these data increases
-dramatically. For the field of cheminformatics to flourish, we need
+problems. 
+% If we oversimplify the nature of chemical structures and
+% their properties, we can not describe the world around us in
+% sufficient detail. On the other hand, if we try to encode every detail
+% (\emph{e.g.}, all possible 3D structures and their properties, all
+% metabolites of a structure in humans in different tissues), the
+% complexity of collecting, storing, and mining these data increases
+% dramatically. 
+For the field of cheminformatics to flourish, we need
 closer collaboration between chemists (or, more generally, life
@@
 support better chemical decision making by 1. storing and integrating
-data in maintainable ways, 2.  providing enough open standards and
-tools so software engineering concepts allow applications and data to
-be used across heterogeneous platforms, and 3. mining the many
-chemical property spaces in a time- and space-efficient way. We will
-discuss all three goals in the following sections.
+data in maintainable ways, 2.  providing sufficient open standards and
+tools allowing applications and data to be used across heterogeneous
+platforms, and 3. mining the many chemical property spaces in a time-
+and space-efficient way. We will discuss all three goals in the
+following sections.
 
@@
 Our motivation for this article is to highlight a variety of topics in
-cheminformatics where modern CS research could contribute, and
-to present the computational and algorithmic problems
-that are addressed in the field.
+cheminformatics, presenting the computational and algorithmic problems
+faced in the field and how modern CS research can contribute.
 
@@
 \bibliographystyle{abbrv}
-\bibliography{resubmission}
+\bibliography{paper}
 

### 2011-11-16 paper.tex
some more edits

@@
 its public availability -- and how does this relate to the field of
-computer science?
+computer science? We note that though cheminformatics has a broad
+applicability, ranging from agrochemical research to the design of
+novel materials, we have chosen to use drug discovery as the context
+of this article due to its central relevance for human well being.
 
@@
 cheminformatics, presenting the computational and algorithmic problems
-faced in the field and how modern CS research can contribute.
-
-While cheminformatics has a broad applicability, ranging from
-agrochemical research and drug discovery to the design of novel
-materials, we have chosen to use drug discovery as the context of this
-article due to its central relevance for human wellbeing. Given that
-the safety of drugs on the market is a key concern both for the
-general audience as well as researchers in pharmaceutical companies,
-the remainder of this article is structured around the concept of
-``risk minimization'' in drug discovery -- the question of how to
-minimize the chances of a small molecule failing, due to poor
-physical, chemical or biological properties, in the various research
-and development stages as a drug candidate, and how to successfully
-employ cheminformatics methods in these efforts. As an example, a
-molecule must be soluble and show a certain degree of bioavailability
-to be considered as drug candidate. In cases where these properties
-are poor, a cheminformatics approach can suggest replacement of
-certain functional groups, that maintain potency but improve the
-solubility and bioavailability.  Table~\ref{table:properties} (in the
-Appendix) summarizes a number of properties a drug candidate must
-satisfy to be considered therapeutically useful (such as being
-appropriately soluble, exhibiting minimal toxicity and being selective
-for the intended target) and sketches the role of cheminformatics at
-each stage of drug development.
+faced in the field and how modern CS research can contribute to their
+solution.
+
+As noted above, in this article we choose to focus on cheminformatics
+in the context of drug discovery and development. Given that this is a
+slow and expensive process, the concept of ``risk minimization'' plays
+an important role in this field. Thus, we structure the article around
+this idea -- how to minimize the chances of a small
+molecule failing, due to poor physical, chemical or biological
+properties, in the various research and development stages as a drug
+candidate, and how to successfully employ cheminformatics methods in
+these efforts. As an example, a molecule must be soluble and show a
+certain degree of bioavailability to be considered as drug
+candidate. In cases where these properties are poor, a cheminformatics
+approach can suggest replacement of certain functional groups, that
+maintain potency but improve the solubility and bioavailability.
+Table~\ref{table:properties} (in the Appendix) summarizes a number of
+properties a drug candidate must satisfy to be considered
+therapeutically useful (such as being appropriately soluble,
+exhibiting minimal toxicity and being selective for the intended
+target) and sketches the role of cheminformatics at each stage of drug
+development.
 %
@@
 There is considerable progress in open chemical and bioactivity
-databases like ChEMBL and NCBI-PubChem, which are both freely
+databases like ChEMBL and PubChem, which are both freely
 available and contain in the order of millions (ChEMBL) and tens of
 millions (PubChem) of data points each. Still, the integration of
-chemical databases with one other remains a challenge, not only due
+chemical databases with one another remains a challenge, not only due
 to the normalization problems in the chemical and bioactivity
@@
 hand as possible, we have to be able to search in complex data, for
-example multi-labelled molecular graphs, where graph labels can change from database to database. Hence, we need to be able to capture molecular graphs in a
-risk-bounded machine encoded way.
+example multi-labelled molecular graphs, where graph labels can change
+from database to database. Hence, we need to be able to capture
+molecular graphs in a risk-bounded machine encoded way.
 
@@
 molecule. When one considers the need to include other aspects such as
-protonation states (how the molecule changes when one proton is added to it) and tautomer states (how the molecule can change when a proton migrates from one part of the molecule to another), the challenge increases.
-The need for
-a unique representation that is invariant with respect to the atom
-ordering arises due to the fact that graph isomorphism (i.e., checking
-whether two molecules are the same) is an expensive operation. The
-first such ``unique representation algorithm,'' or \emph{canonicalization} algorithm, was described by Morgan
-\cite{Morgan1965}, allowing one to generate unique string
+protonation states (how the molecule changes when one proton is added
+to it) and tautomer states (how the molecule can change when a proton
+migrates from one part of the molecule to another), the challenge
+increases. The need for a unique representation that is invariant with
+respect to the atom ordering arises due to the fact that graph
+isomorphism (i.e., checking whether two molecules are the same) is an
+expensive operation. The first such ``unique representation
+algorithm,'' or \emph{canonicalization} algorithm, was described by
+Morgan \cite{Morgan1965}, allowing one to generate unique string
 representations of chemical graphs letting one compare structures via
@@
 structures, but it is still limited. For example, it cannot
-differentiate between certain types of stereoisomers (informally, molecules that are ``3D mirror images'' of each other):
+differentiate between certain types of stereoisomers (informally,
+molecules that are ``3D mirror images'' of each other):
 Figure~\ref{figure:cistrans} (in the Appendix) illustrates two
@@
 part of the structures in the database, while in similarity searching,
-the database is searched for structures similar (in some property
-space) to a provided search, or query, structure. Chemical search
-packages are often implemented and optimized for a given database
-technology; for example, the OrChem package is an open source chemical
-search package for the Oracle database \cite{rijnbeek2009}.
+the database is searched for structures similar (in some structure or
+property space) to a provided search, or query, structure. Chemical
+search packages are often implemented and optimized for a given
+database technology; for example, the OrChem package is an open source
+chemical search package for the Oracle database \cite{rijnbeek2009}.
 
@@
 and hashed keys. In the former, each bit position corresponds to a
-distinct substructure such as a \emph{functional group} (a connected set of atoms that affects the characteristics of the chemical reactions of the molecule). Examples
-include MACCS and PubChem keys. The latter class are designed such
-that substructural patterns are represented as strings and then hashed
-to a random bit position. As a result, a given position can encode
-multiple substructures. The advantage of such fingerprints is that
-they can cover an arbitrarily large collection of substructures such
-as ``paths of length $N$,'' or circular environments, and so on. Examples
-include ECFPs (Extended Connectivity Fingerprints, which use local topological information) and Daylight fingerprints (which are ``folded'' to optimize information density and screening speed).
+distinct substructure such as a \emph{functional group} (a connected
+set of atoms that affects the characteristics of the chemical
+reactions of the molecule). Examples include MACCS and PubChem
+keys. The latter class are designed such that substructural patterns
+are represented as strings and then hashed to a random bit
+position. As a result, a given position can encode multiple
+substructures. The advantage of such fingerprints is that they can
+cover an arbitrarily large collection of substructures such as ``paths
+of length $N$,'' or circular environments, and so on. Examples include
+ECFPs (Extended Connectivity Fingerprints, which use local topological
+information) and Daylight fingerprints (which are ``folded'' to
+optimize information density and screening speed).
 
@@
 fingerprints. While fast, heuristics have been developed
-\cite{Swamidass:2007ve} to further
-speed up similarity searches.
+\cite{Swamidass:2007ve} to further speed up similarity searches.
 
@@
 same or similar targets; they may be have similar biological function
-(say, anti-neoplastic, that is to say, it prevents the development of tumors) and so on. To assess such higher level
-characteristics, ontologies and semantic technologies have been
-employed, to first, support annotations of chemical systems and
-second, enable semantic analyses of chemical systems in conjunction
-with other biological systems (such as pathways). A number of
-challenges remain in this area - the use of ontologies allows one to
-build a semantic network of concepts that links molecules to other
-molecules as well as other objects such as targets, genes and
-pathways. Open questions include the following.  How does one evaluate similarities in such a heterogenous
-network \cite{couto2010}? How does one integrate chemical structure
-information with ontologies \cite{hastingsowled2010}? And more
-generally, how can we achieve automated inferencing in such
+(say, anti-neoplastic, that is to say, it prevents the development of
+tumors) and so on. To assess such higher level characteristics,
+ontologies and other semantic technologies have been employed, to first,
+support annotations of chemical systems and second, enable semantic
+analyses of chemical systems in conjunction with other biological
+systems (such as pathways). A number of challenges remain in this area
+- the use of ontologies allows one to build a semantic network of
+concepts that links molecules to other molecules as well as other
+objects such as targets, genes and pathways. Open questions include
+the following.  How does one evaluate similarities in such a
+heterogenous network \cite{couto2010}? How does one integrate chemical
+structure information with ontologies \cite{hastingsowled2010}? And
+more generally, how can we achieve automated inferencing in such
 heterogenous linked systems, allowing us to achieve a systems level
@@
 combinatorics were developed in the context of counting isomers of
-paraffin by Cayley, Polya and others. In the 1960's, Lederberg
+paraffin by Cayley, Polya and others. In the 1960's, Lederberg, 
 Dejerassi and others developed algorithms to enumerate structures
@@
 
-From a risk reduction point of view efficient methods to enumerate
+From a risk reduction point of view, efficient methods to enumerate
 structures allows us to explore new regions of chemical space (for
@@
 one-way molecular features but still correlate with physicochemical
-properties would be invaluable.
+properties would be valuable.
 
@@
 comprising almost a billion chemical structures~\cite{GDB} (but only
-considering molecules up to 13 heavy atoms and using only carbon, nitrogen, oxygen, sulfur
-and chlorine). Yet, the current enumeration software products do not
-generally produce stereoisomers or tautomers; these require specific
-enumeration procedures, which are still the subject of investigations
-by the cheminformatics community.
+considering molecules up to 13 heavy atoms and using only carbon,
+nitrogen, oxygen, sulfur and chlorine). Yet, the current enumeration
+software products do not generally produce stereoisomers or tautomers;
+these require specific enumeration procedures, which are still the
+subject of investigations by the cheminformatics community.
 
@@
 dynamics of metabolic networks, applying metabolic engineering and
-synthetic biology to produce heterologous compounds (compounds from different species) in microorganisms,
-all involve enumeration of reaction networks. As reviewed in Chapter
-11 in \cite{faulon2010} several network enumeration techniques have
-been developed. However, these techniques generally suffer from a
-combinatorial explosion of product compounds. One way to limit the
-number of compounds generated is to simulate the dynamics of the
-network while it is being constructed and remove compounds of low
-concentration. Following that idea, methods have been developed based
-on the Gillespie Stochastic Simulation Algorithm (SSA) to compute
-on-the-fly species concentrations. Chemical reaction network
-enumeration and sampling is an active field of research, particularly
-in the context of metabolism, either to study biodegradation, or to
-propose metabolic engineering strategies to biosynthesize compounds of
-commercial interest. The difficulty with metabolic network design is
-that in addition to network generation based on reactions, one also
-needs to verify that there are enzymatic events possible for enabling
-reactions catalyses (in other words, enzymes must be present to reduce the energy required for some reactions to take place). That additional task requires encompassing both
-chemical structure and protein sequence information and the
-development of tools that are at the interface between cheminformatics
-and bioinformatics.
+synthetic biology to produce heterologous compounds (compounds from
+different species) in microorganisms, all involve enumeration of
+reaction networks. As reviewed in Chapter 11 in \cite{faulon2010}
+several network enumeration techniques have been developed. However,
+these techniques generally suffer from a combinatorial explosion of
+product compounds. One way to limit the number of compounds generated
+is to simulate the dynamics of the network while it is being
+constructed and remove compounds of low concentration. Following that
+idea, methods have been developed based on the Gillespie Stochastic
+Simulation Algorithm (SSA) to compute on-the-fly species
+concentrations. Chemical reaction network enumeration and sampling is
+an active field of research, particularly in the context of
+metabolism, either to study biodegradation, or to propose metabolic
+engineering strategies to biosynthesize compounds of commercial
+interest. The difficulty with metabolic network design is that in
+addition to network generation based on reactions, one also needs to
+verify that there are enzymatic events possible for enabling reactions
+catalyses (in other words, enzymes must be present to reduce the
+energy required for some reactions to take place). That additional
+task requires encompassing both chemical structure and protein
+sequence information and the development of tools that are at the
+interface between cheminformatics and bioinformatics.
 
@@
 information on ligand-receptor interactions. As a result, techniques
-such as docking (which predicts how molecules fit together), pharmacophore modeling (which predicts the surface features of a molecule) and proteochemometric methods (which predict the interaction between small molecules and proteins)
-have been designed to take into account both ligand and receptor
-structures. The last method is an example of an extension of
-statistical QSAR methods to simultaneous model the receptor and
-ligand as first reported by Lapinsh \textit{et  al.}~\cite{lapinsh2001}.
+such as docking (which predicts how molecules fit together),
+pharmacophore modeling (a 3D approach to capturing protein-ligand
+interactions) and proteochemometric methods have been designed to take
+into account both ligand and receptor structures. The last method is
+an example of an extension of statistical QSAR methods to simultaneous
+model the receptor and ligand as first reported by Lapinsh \textit{et
+  al.}~\cite{lapinsh2001}.
 
 The first step in the predictive modeling of biological activities is
-to generate \emph{molecular descriptors} (a.k.a. \emph{features}) that are numerical
-representations of different structural features. Here multiple
-approaches exist. We can transform molecular graphs into a numerical
-vector representation of chemical structures \cite{todeschini2000}, ranging from counts of
-elements to eigenvalues of the Laplacian matrix.  Alternatively, we
-can compare molecular graphs directly, via \emph{kernel methods}, where a kernel on graphs $G$ and $G'$ provides a measure of how similar $G$ is to $G'$, or a kernel on a single graph compares measures similarities between the graph's nodes.
-In these
-cases, rather than compute vectorial representations, we directly
-operate on the graph representations.
+to generate \emph{molecular descriptors} (a.k.a. \emph{features}) that
+are numerical representations of different structural features. Here
+multiple approaches exist. We can transform molecular graphs into a
+numerical vector representation of chemical structures
+\cite{todeschini2000}, ranging from counts of elements to eigenvalues
+of the Laplacian matrix.  Alternatively, we can compare molecular
+graphs directly, via \emph{kernel methods}, where a kernel on graphs
+$G$ and $G'$ provides a measure of how similar $G$ is to $G'$, or a
+kernel on a single graph compares measures similarities between the
+graph's nodes. In these cases, rather than compute vectorial
+representations, we directly operate on the graph representations.
 
@@
 identify an appropriate kernel. This is an important challenge as the
-kernel should be selected to satisfy \emph{Mercer's condition} (a well-known mathematical property in machine learning that makes a set of observations easier to make predictions about), and this is
-not always possible with traditional cheminformatics based kernels
-(such as a multiple common substructure based kernel). These
-challenges can make kernel based methods prohibitive on very large
-datasets.
+kernel should be selected to satisfy \emph{Mercer's condition} (a
+well-known mathematical property in machine learning that makes a set
+of observations easier to make predictions about), and this is not
+always possible with traditional cheminformatics based kernels (such
+as a multiple common substructure based kernel). These challenges can
+make kernel based methods prohibitive on very large datasets.
 
@@
 running analytical experiments). Legal disputes would require manual
-searches through many paper lab notebooks and data synchornization was
+searches through many paper lab notebooks and data synchronization was
 painful, hampering large scale collaborations.
@@
 position of `if we get an ELN' to `when we get an
-ELN'.''~\cite{ELNstatus}. The growth of the use of ELNs has been
+ELN'.''~\cite{ELNstatus}. The growth of ELN usage has been
 driven by many factors including ease of use (rapid searches across
@@
 infrastructure for all parties involved, but also to provide the
-required private/public access toolbox. Furthermore, there are still
-many data sources not being indexed properly: chemistry patents are
-often cryptic and chemical image and text mining remains a challenge
-(albeit, currently being addressed in academia and industry
+required private/public interface. Furthermore, there are still many
+data sources that are not being indexed properly: chemistry patents
+are often cryptic and chemical image and text mining remains a
+challenge (albeit, currently being addressed in academia and industry
 \cite{Jessop:2011fk,Sayle:2009uq}). The closed CAS (Chemical Abstract
@@
 candidates. In contrast, much bioinformatics software is employed in
-upstream areas such as target identification (finding molecules that react well with the molecule under consideration).  While certainly
+upstream areas such as target identification (identifying the protein
+that a small molecule binds to).  While certainly
 important, this type of research is not as directly related to
@@
 knowledge is required to attack many problems in cheminformatics due
-to the underlying chemistry of the systems being studies. Indeed, many
+to the underlying chemistry of the systems being studied. Indeed, many
 issues faced by chemists are often detailed and subtle and do not
@@
 data, the barrier to entry for non-chemists and non-cheminformaticians
-is significantly lower than it was a decade ago. Many questions that benefit from computer
-science can now be comprehensively tackled: for a theorist,
-graph-theoretic questions of three-dimensional enumeration; for a
-database designer, more effective search algorithms and ontologies;
-for a practical programmer, expansion of the many open-source
-cheminformatics projects.  If more chemists think algorithmically, and
-more computer scientists think chemically we are much better
-positioned taking not only ``simple'' chemicals into account (only one
-identifier and isomer possible per molecule), but also their complex
-relationships, transformations, and combinatorial challenges.
+is significantly lower than it was a decade ago. Many questions that
+benefit from computer science can now be comprehensively tackled: for
+a theorist, graph-theoretic questions of three-dimensional
+enumeration; for a database designer, more effective search algorithms
+and ontologies; for a practical programmer, expansion of the many
+open-source cheminformatics projects.  If more chemists think
+algorithmically, and more computer scientists think chemically we are
+much better positioned taking not only ``simple'' chemicals into
+account (only one identifier and isomer possible per molecule), but
+also their complex relationships, transformations, and combinatorial
+challenges.
 

### 2011-11-17 OpenQuestions.tex
Added open questions to the repo

\documentclass{sig-alternate}

\usepackage{array}
\usepackage{pifont}
\usepackage{url}
\usepackage{graphicx}
\usepackage{multirow}

\newcommand{\none}{\ding{55}}
\newcommand{\least}{\ding{51}}
\newcommand{\little}{\ding{51}\ding{51}}
\newcommand{\lots}{\ding{51}\ding{51}\ding{51}}

\begin{document}
\pagenumbering{arabic}

\title{Open Questions for Computer Science and Cheminformatics}
\numberofauthors{9}
\author{
\alignauthor
Joerg Kurt Wegner\\
       \affaddr{Tibotec BVBA}\\
       \affaddr{Turnhoutseweg 30}\\
       \affaddr{2340 Beerse Turnhout, Belgium}\\
       \email{jwegner@its.jnj.com}
% 2nd. author
\alignauthor
Aaron Sterling\\
       \affaddr{Department of Computer Science}\\
       \affaddr{Iowa State University}\\
       \affaddr{Ames, Iowa, USA}\\
       \email{sterling@iastate.edu}
% 3rd author
\alignauthor
Rajarshi Guha\\
\affaddr{NIH Center for Translational Therapeutics}\\
\affaddr{9800 Medical Center Drive}\\
\affaddr{Rockville, MD 20850}\\
\email{guhar@mail.nih.gov}
}

\additionalauthors{Additional authors:
Andreas Bender (University of Cambridge, email: {\texttt{andreas.bender@cantab.net}}),
Jean-Loup Faulon (University of Evry, email: {\texttt{Jean-Loup.Faulon@issb.genopole.fr}}),
Janna Hastings (European Bioinformatics Institute, Cambridge, UK, email: {\texttt{janna.hastings@gmail.com}}),
Noel O'Boyle (University College Cork, Cork, Ireland, email: {\texttt{baoilleach@gmail.com}}),
John Overington (European Bioinformatics Institute, Cambridge, UK, email: {\texttt{jpo@ebi.ac.uk}}),
Herman Van Vlijmen (Tibotec, Beerse, Belgium, email: {\texttt{hvvlijme@its.jnj.com}}), and
Egon Willighagen (Karolinska Institutet, Stockholm, Sweden, email: {\texttt{egon.willighagen@ki.se}})
.}
\maketitle
%
We offer the following open questions to suggest concrete interdisciplinary research directions for computer scientists working with chemoinformaticians.  This list is by no means exhaustive.  We have focused on questions that, in our opinion, are answerable, and whose answers would be of great interest to the field.  The questions are grouped roughly by computer science subarea, and include both open theoretical problems and ``requests'' for open-source practical implementations.  We hope CS researchers new to cheminformatics find this useful.

\section*{Algorithmic graph theory}
\begin{enumerate}
\item \emph{Design an algorithm that approximately counts the number of (3D) conformers of a chemical formula.}

This question encompasses the open questions of \emph{approximately counting the number of stereoisomers, or the number of tautomers, of a chemical formula}.  Goldberg and Jerrum designed an algorithm which, given a chemical formula, would output an isomer of that formula chosen uniformly at random~\cite{RandomlySampling}.  Perhaps the main theoretical obstacle to overcome is that molecules tend to be \emph{chiral} (that is, they have a ``handedness'' or 3D orientation), whereas traditional graph theory (and the Goldberg/Jerrum algorithm) treats graphs with identical vertices and edges as isomorphic.  So part of the question could be rephrased as, ``Given a labeled vertex set, count the number of structures that are identical on that vertex set, except that they differ in their 3D orientation.''  Mathematical chemists have designed measures of molecular chirality~\cite{ChiralityMeasures}. A resolution of this problem may require connecting graph enumeration algorithms to knot theory or other topics in topology~\cite{TopologicalLook}.
%
\item \emph{Write code that approximately counts the number of conformers of a chemical formula.}

While this is related to the previous question, it is conceivable that one could answer this question without solving the theoretical problem.  For example, as we note in the main article, the commercial software MOLGEN can approximately enumerate structures that satisfy given spectral data, even though related theoretical enumeration problems appear wide open.
%
\item \emph{Design and implement efficient subgraph isomorphism algorithms for useful special cases of molecular graphs}.

The Subgraph Isomorphism Problem is known to be $\textsf{NP}$-complete (hence possibly harder than the Graph Isomorphism Problem, which is not known to be $\textsf{NP}$-complete).  Nevertheless, finding maximum common subgraphs to match chemical structures is of fundamental importance in chemistry; hence, much work has been invested in partial solutions to this problem.  Raymond and Willet reviewed the state of the art in 2002~\cite{MCSreview}.  As one possible way to attack this problem, we note the empirical fact that molecular graphs are of bounded degree, and are observed to have \emph{treewidth} $\leq 5$~\cite{treewidth}.  Bounded treewidth is at least theoretically useful~\cite{Epp-JGAA-99}, and it may be possible to improve on current open-source implementations whose isomorphism-checking routines are written for all graphs, instead of taking advantage of special properties of chemical graphs.

\item \emph{Searching within complex data types, e.g. molecules, for semantic web approaches}.

The key bonus of the semantic web is that different data sources can be readily integrated with each other. In the field
of Cheminformatics, we are not only interest in linking two molecules 
(this normalization problem for different protomers, tautomers, or special cases of isomerisms remains open), but we
are also interested in being able to search efficiently within molecules when being linked via semantic web approaches.
What could be algorithmic solutions for this?
\end{enumerate}

\section*{Cryptography}
One-way molecular featurization (???)

\section*{Data mining}
evaluation of similarities in a heterogeneous network.  What is a specific example here?

integrate chemical structure information with ontologies.  Specific example problem?

Classify molecular descriptors up to equivalence, and dependence on one another.

systems-level understanding of small molecules.  What does this mean, and what would a specific challenge problem be?

\begin{enumerate}

\item \emph{Efficient molecule browsing, e.g. on scaffold level}.

Chemical Abstract Services have a molecule browsing tool called SubScape, which allows to brows large-scale
chemical spaces efficiently. What could be large-scale solutions for doing this within (combined and aligned) 
public databases.
%
\item \emph{Large-scale browsing of molecular property spaces, e.g. on scaffold level, side-effect-level, ...}.

Certain molecules might have hundreds of biological activities, side-effects in humans (from clinical trials), or
many other properties attached to them. What are large-scale mining and visulization options, especially when thinking
about mining private and public datasources at the very same time?
%
\item \emph{Patent text mining (curation)}.

There are various tools for doing automatic text mining on chemical patents. Still, the overall acceptance rate is improvable
since many medicinal chemists are very concerned about the data quality of such efforts. What could be done to improve
the mining quality and to provide confidence level estimations for each molecule coming from patent mining? 
How can patent mining be used to create new drugs faster or to speed-up collaboration/licensing discussions?
\end{enumerate}

\section*{Machine learning}
I don't know.  Help?  Ideas: better kernelization techniques (something specific though), better algorithms to train X model about Y thing.
\begin{enumerate}
\item \emph{Large-scale vectorial versus kernels molecule similarity}

Vectorial molecule encodings serve as efficient approximations.
Sometimes non-vectorial molecular 3D shape or molecule kernel comparisons might be more suitabe to compare molecules, since
they might better correlate with biological activity, toxicity in humans, etc. One key problem is that molecular 3D shape 
or molecule kernel approaches require to compare two molecules (or their 3D conformational eplosions) directly.
This becomes prohibitively expensive when considering millions of molecules. What could be potential solutions for approximating boundary 
conditions or large-scale mining methods allowing within a second range to return all similar molecules to molecule X, especially
considering not just one vectorial encoding, but molecule kernels, ligand-protein similarities (for example from ligand-protein crystal structures), 
or chemogenomics similarities (all reported biological activities for one molecule).
%
\item \emph{Using multiple annotations for improving molecular mining/predictions (chemogenomics)}

As an example: Biological activities might not be independent of each other, but have a certain correlation between each other.
In Chemogenomics this is used for creating models of combining molecules with protein sequences, molecules with active sites of proteins,
molecules with biological read-outs of multiple assays. How can we optimize such highly complex mining scenarios, especially
when considering large-scale data sets with hundred of thousands molecules and thousands of biological activities? 
How can we combine, mine, and visualize categorial and continuous output variables, e.g. hydrophobicity of a molecule and toxicity in humans, 
by still being able to make conrete proposals to medicinal chemistry of which parts of the molecule needs to be changed to
optimize the effect on a certain set of multi-objective variables? Is analoging (creating vrey small modifications of a molecule and measuring its activities)
really the most efficient way forward? If we test molecules, should we test it in a single biological assay or in multiple biological assays, if multiple, which ones?
If a company does not have a biological assay within reach, which other partner could offer testing a molecule within two days (vendor matching based on licenses)?
\end{enumerate}

\section*{Software engineering}
\begin{enumerate}
\item \emph{Large-scale exchange and use of 3D conformational databases}

There exist a multitude of programs generation 3D conformations of molecules. Still, often input formats vary and require
to re-create 3D conformational spaces with other programs. This does not only complicate the maintenance scenario of such
databases, especially with the explosion of public databases, but also creates hugh space and time complexity issues when
searching within such databases. What could be better interfaces, maintenance, data structures, and private/public sharing
scenarios for conformational 3D databases? 
Many software vendors use different solutions for parallizing comput jobs: SGE, PVM, MPI, etc. 
Everyone knowing thh enterprise structure within companies might know that having more than one parallel processing framework 
is not easy? What could be better ways to streamline parallel processing structures for 
cheminformatics (and molecular modeling) algorithms? Is a cloud really an option? What about SaaS with secured data transfer?
\end{enumerate}

\section*{Enterprise software (KM,ELN)}
We know that the enterprise software and ELN market is still growing.
\begin{enumerate}
\item \emph{Public-private collaboration and security scenarios}

Having within an organization, e.g. a commercial company, single or a small number of established KM and ELN products. 
How can we improve the maintenance, leveraging, and collaboration with many external partners (each of them potentially
with another KM/ELN solution)? Which party is hosting which data in which data structure, and how can we ensure that only
pre-defined data entries (and a limited number of annotations, e.g. bioological activities) are visible to a partner.
How can this be organized for a multitude of partners? Cloud computing?
\end{enumerate}

\bibliographystyle{abbrv}
\bibliography{paper}
\end{document}

### 2011-11-17 OpenQuestions.tex
Minot change

@@
 of Cheminformatics, we are not only interest in linking two molecules 
-(this normalization problem for different protomers, tautomers, or special cases of isomerisms remains open), but we
+(the normalization problem for different protomers, tautomers, or special cases of isomerisms remains open), but we
 are also interested in being able to search efficiently within molecules when being linked via semantic web approaches.
@@
 \section*{Data mining}
-evaluation of similarities in a heterogeneous network.  What is a specific example here?
-
-integrate chemical structure information with ontologies.  Specific example problem?
-
-Classify molecular descriptors up to equivalence, and dependence on one another.
-
-systems-level understanding of small molecules.  What does this mean, and what would a specific challenge problem be?
+\begin{itemize}
+\item evaluation of similarities in a heterogeneous network.  What is a specific example here?
+\item integrate chemical structure information with ontologies.  Specific example problem?
+\item Classify molecular descriptors up to equivalence, and dependence on one another.
+\item systems-level understanding of small molecules.  What does this mean, and what would a specific challenge problem be?
+\end{itemize}
 

### 2011-11-17 OpenQuestions.tex
Minor rewrite

@@
 
-The key bonus of the semantic web is that different data sources can be readily integrated with each other. In the field
+One key concept of the linked data web, the semantic web, is that different data sources can be readily integrated with each other. Still, in the field
 of Cheminformatics, we are not only interest in linking two molecules 
-(the normalization problem for different protomers, tautomers, or special cases of isomerisms remains open), but we
+(the linking normalization problem for different protomers, tautomers, or special cases of isomerisms remain open), but we
 are also interested in being able to search efficiently within molecules when being linked via semantic web approaches.
@@
 \section*{Data mining}
-\begin{itemize}
-\item evaluation of similarities in a heterogeneous network.  What is a specific example here?
+\begin{enumerate}
+\item evaluation of similarities in a heterogeneous network (JKW: What are heterogenous networks?).  What is a specific example here?
 \item integrate chemical structure information with ontologies.  Specific example problem?
 \item Classify molecular descriptors up to equivalence, and dependence on one another.
-\item systems-level understanding of small molecules.  What does this mean, and what would a specific challenge problem be?
-\end{itemize}
-
-\begin{enumerate}
-
-\item \emph{Efficient molecule browsing, e.g. on scaffold level}.
-
-Chemical Abstract Services have a molecule browsing tool called SubScape, which allows to brows large-scale
-chemical spaces efficiently. What could be large-scale solutions for doing this within (combined and aligned) 
-public databases.
+\item systems-level understanding of small molecules (JKW: which systems? This will help being clearer on the challenge).  What does this mean, and what would a specific challenge problem be?
+%\item \emph{Efficient molecule browsing, e.g. on scaffold level}.
+%
+%Chemical Abstract Services have a molecule browsing tool called SubScape, which allows to browse large-scale
+%chemical spaces efficiently. What could be large-scale solutions for doing this within (combined and aligned) 
+%public databases.
 %
@@
 
-Certain molecules might have hundreds of biological activities, side-effects in humans (from clinical trials), or
-many other properties attached to them. What are large-scale mining and visulization options, especially when thinking
-about mining private and public datasources at the very same time?
+Certain molecules might have hundreds of biological activities, side-effects in humans (SIDER database \cite{Kuhn_Campillos_Letunic_Jensen_Bork_2010}), or
+many other properties attached to them. What are large-scale mining and visulization options? 
+How can we mine private and public data sources at the very same time?
 %
@@
 
-There are various tools for doing automatic text mining on chemical patents. Still, the overall acceptance rate is improvable
-since many medicinal chemists are very concerned about the data quality of such efforts. What could be done to improve
-the mining quality and to provide confidence level estimations for each molecule coming from patent mining? 
+There are various tools for doing automatic text mining on chemical patents. Still, the overall acceptance rate of chemical
+text mining is improvable, since many medicinal chemists are very concerned about the data quality of such efforts. 
+What could be done to improve the mining quality, curate the obtained data, and to provide confidence level estimations 
+for each molecule coming from patent mining? Do require image2structure and text2structure mining also data stores
+for ensuring a sufficient amount of confidence and data quality?
 How can patent mining be used to create new drugs faster or to speed-up collaboration/licensing discussions?
@@
 \section*{Machine learning}
-I don't know.  Help?  Ideas: better kernelization techniques (something specific though), better algorithms to train X model about Y thing.
 \begin{enumerate}
@@
 
-Vectorial molecule encodings serve as efficient approximations.
+Vectorial molecule encodings can serve as efficient approximations of molecules.
 Sometimes non-vectorial molecular 3D shape or molecule kernel comparisons might be more suitabe to compare molecules, since
-they might better correlate with biological activity, toxicity in humans, etc. One key problem is that molecular 3D shape 
-or molecule kernel approaches require to compare two molecules (or their 3D conformational eplosions) directly.
-This becomes prohibitively expensive when considering millions of molecules. What could be potential solutions for approximating boundary 
-conditions or large-scale mining methods allowing within a second range to return all similar molecules to molecule X, especially
-considering not just one vectorial encoding, but molecule kernels, ligand-protein similarities (for example from ligand-protein crystal structures), 
-or chemogenomics similarities (all reported biological activities for one molecule).
+they might better correlate with activities. One key problem is that non-vectorial encodings require to compare all molecules
+(or their 3D conformational eplosions) in a pair-wise manner.
+This becomes prohibitively expensive when considering millions of molecules.
+Can dyadic data approaches help \cite{Hochreiter:2006:SVM:1159508.1159516}? Other approximations or cascading flows?
 %
@@
 In Chemogenomics this is used for creating models of combining molecules with protein sequences, molecules with active sites of proteins,
-molecules with biological read-outs of multiple assays. How can we optimize such highly complex mining scenarios, especially
+or molecules with biological activities of multiple assays. How can we optimize such highly complex mining scenarios, especially
 when considering large-scale data sets with hundred of thousands molecules and thousands of biological activities? 
 How can we combine, mine, and visualize categorial and continuous output variables, e.g. hydrophobicity of a molecule and toxicity in humans, 
-by still being able to make conrete proposals to medicinal chemistry of which parts of the molecule needs to be changed to
-optimize the effect on a certain set of multi-objective variables? Is analoging (creating vrey small modifications of a molecule and measuring its activities)
+by still being able to make concrete proposals to medicinal chemistry? Is analoging (creating very small modifications of a molecule and measuring its activities)
 really the most efficient way forward? If we test molecules, should we test it in a single biological assay or in multiple biological assays, if multiple, which ones?
-If a company does not have a biological assay within reach, which other partner could offer testing a molecule within two days (vendor matching based on licenses)?
+If a company does not have a biological assay within reach, which other partner could offer testing a molecule within two days (vendor matching based on licenses or contracts)?
 \end{enumerate}
@@
 scenarios for conformational 3D databases? 
-Many software vendors use different solutions for parallizing comput jobs: SGE, PVM, MPI, etc. 
-Everyone knowing thh enterprise structure within companies might know that having more than one parallel processing framework 
-is not easy? What could be better ways to streamline parallel processing structures for 
+Many software vendors use different solutions for parallizing compute jobs: SGE, PVM, MPI, etc. 
+Everyone knowing the enterprise IT approval cycles might know that having more than one parallel processing framework 
+is not easy. What could be better ways to streamline parallel processing structures for 
 cheminformatics (and molecular modeling) algorithms? Is a cloud really an option? What about SaaS with secured data transfer?
+Can this also offer alternative licensing strategies for software suites in this domain?
 \end{enumerate}
@@
 
-Having within an organization, e.g. a commercial company, single or a small number of established KM and ELN products. 
+Let us assume an organization, e.g. a commercial company, has a single or a small number of established KM and ELN products. 
 How can we improve the maintenance, leveraging, and collaboration with many external partners (each of them potentially
-with another KM/ELN solution)? Which party is hosting which data in which data structure, and how can we ensure that only
-pre-defined data entries (and a limited number of annotations, e.g. bioological activities) are visible to a partner.
-How can this be organized for a multitude of partners? Cloud computing?
+with another KM/ELN solution)? Which party is hosting which data in which data structure (ontologies?), and 
+how can we ensure that only pre-defined data entries 
+(and a limited number of annotations, e.g. bioological activities) are visible to a partner.
+How can this be organized for a multitude of partners? Cloud computing, user management, encryption granularity and efficient security management?
 \end{enumerate}

### 2011-11-18 paper.tex

@@
 Jean-Loup Faulon (University of Evry, email: {\texttt{Jean-Loup.Faulon@issb.genopole.fr}}),
-Janna Hastings (European Bioinformatics Institute, Cambridge, UK, email: {\texttt{janna.hastings@gmail.com}}),
+Janna Hastings (European Bioinformatics Institute, Cambridge, UK, email: {\texttt{hastings@ebi.ac.uk}}),
 Noel O'Boyle (University College Cork, Cork, Ireland, email: {\texttt{baoilleach@gmail.com}}),
@@
   One of the most prominent success stories in all the sciences over
-  the last decade has been the advance of bioinformatics: The
+  the last decade has been the advance of bioinformatics: the
   interdisciplinary collaboration between computer scientists and
@@
   physicochemical properties of small molecules. Until recently, much
-  of the the data and some of the techniques employed in the
+  of the data and some of the techniques employed in the
   cheminformatics field have been closely guarded secrets of companies
-  whose financial success depended on being the first to produce the
-  new therapeutic molecules.  Only within the last decade -- and as an
+  whose financial success depended on being the first to produce
+  new therapeutic molecules.  Only within the last decade -- as an
   effect of a change in mindsets, government-mandated policy changes,
-  and, not least, because of chemists volunteering their time
+  and, not least, as a result of chemists volunteering their time
   for an Open Science ``movement'' -- have researchers gained access to
@@
 Novel technologies in the life sciences churn out information at an
-ever increasing rate -- public data stores such as the one at the
-European Bioinformatics Institute (EBI) contain on the order of 10
+ever increasing rate -- public data stores such as that of the
+European Bioinformatics Institute (EBI) contain in the order of 10
 petabytes of biological information.  While biological information has
@@
 the foundational algorithms of cheminformatics have been described
-since the 1950s, publicly available software implementing many of
-these algorithms have only been accessible in the last ten to fifteen
+since the 1950s, publicly available (open source) software implementing many of
+these algorithms has only been accessible in the last ten to fifteen
 years \cite{faulon2010}.
@@
 structures may possess rings, branches as well as multiple valid
-representations of the same structure (such as \emph{tautomers}, where
+representations of the same molecule (such as \emph{tautomers}, where
 hydrogen atoms can be positioned in different places of the
@@
   in information content from left to right and top to bottom. A
-  chemists traditional 2D depiction (A), a labeled graph (B), a 3D
+  chemist's traditional 2D depiction (A), a labeled graph (B), a 3D
   conformation (C) and a surface represention (D). All show differing
   (and valid) aspects of the same molecule and are suited for
-  differing purposes. The choice of representation is guided by the
+  different purposes. The choice of representation is guided by the
   intended use of the structure data.}

### 2011-11-18 rebuttal.tex
cleaned up rebuttal
Overall edit and specific changes to ontology section

@@
 We would like to thank the anonymous reviewers for their constructive
-comments. We have updated the main text in several places and have
+comments which have served to significantly improve our manuscript. 
+We have updated the main text in several places and have
 also included a supplemental document that provides a brief
@@
   already provided such overviews. Our aim was to present a brief
-  overview of \emph{topics} in cheminformatics, selected because we
+  overview of \emph{current topics} in cheminformatics, selected because we
   believed that they represent on going challenges for the field and
@@
   Access solutions to allow reasders to explore the problems and
-  topics we have discussed with a minimum of encumbrances}
+  topics we have discussed with a minimum of hindrances}
 
@@
   compress or drop other sections which we believe to be more
-  relevant. Second, we are of the view that dokcing, for example, is
+  relevant. Second, we are of the view that docking, for example, is
   not really a cheminformatics technique, though it makes extensive
-  use of cheminformatic concepts. Instead we place it under molecular
+  use of cheminformatic methods. Instead we place it under molecular
   modeling. We realize that this is a somewhat subjective
@@
   state-of-the-art and usefulness of ontologies in cheminformatics.}
-\response{The tetx has been updated on page 4 to make the discussion
-  of ontologies more focused, specifically identfying challenges in
-  linking chemical information to other data types and how ontologies
-  and semantic methods can be used ot address this.}
+\response{The text has been updated on page 4 to make the discussion
+  of ontologies more focused and the use cases clearer, specifically 
+  identifying challenges in cross-domain querying, classification 
+  and integration within the context of whole-systems biology, and the 
+  growing essental role that ontologies and semantic technologies are 
+  playing in this matter.}
 
@@
 \response{The section on fingerprints has been rewritten to be more
-  clearer as well as distinguish between the two braod classes of
-  fingerpritns. We have also mentioned the use of fingerprints in
-  similarity searchs and the Tanimoto coeffcient}
+  clearer as well as distinguish between the two broad classes of
+  fingerprints. We have also mentioned the use of fingerprints in
+  similarity searches and the Tanimoto coefficient}
 
@@
   the operation. In practice this is usually not the case, but still,
-  isomorphism algorithms do take longer than fingerprint
-  screening. Hence in practice performing isomorphsim tests for large
+  isomorphism algorithms do take much longer than fingerprint
+  screening. Hence in practice performing isomorphism tests for large
   databases is infeasible. We have included a reference that
@@
 \response{The text has been updated and simplified}
-
-\comment{p4, c1, l30: a citation is required regarding these
-  assertions of graph theory and combinatorics improvements.}
-\response{\textbf{What is he referring to?}}
 
@@
 correlation structure between the multiple y's could lead to }
+%%% THE ABOVE SENTENCE SEEMS INCOMPLETE %%%
 
@@
   source/open data, because this ensures that a reader will be able to
-  explore cheminformatics problems with a minimum of encumbrances}
+  explore cheminformatics problems with a minimum of hindrances}
 
@@
   restructured the Conclusions section to expand on the role of
-  cheminformatics in industry vis a vis bioinformatics, and why much
+  cheminformatics in industry versus bioinformatics, and why much
   of cheminformatics tools and data have been commercial, in contrast

### 2011-11-18 paper.tex
Overall edit and specific changes to ontology section

@@
 obtain accurate predictions from a subset of conformers? Or, can we
-obtain comparable accuracy by using a 2D representation? And so,
+obtain comparable accuracy by using a 2D representation? And if so,
 what type of labels are required? Currently, many of these questions
@@
 essential to enable many scientific applications. Yet, each one of
-those areas poses its unique computer science challenges.
+these areas poses its unique computer science challenges.
 
@@
 available and contain in the order of millions (ChEMBL) and tens of
-millions (PubChem) of data points each. Still, the integration of
+millions (PubChem) of data points each. Integration of these disparate
+data is essential for researchers to obtain the fullest possible 
+perspective on what is presently known, tracking ongoing advances in 
+science as these become available. Still, the integration of
 chemical databases with one another remains a challenge, not only due
@@
 landscape, but also due to sheer data volume. One problem is the
-annotation with the ``protein targets'' a molecule binds to -- here
+annotation of the ``protein targets'' a molecule binds to, as  
 dozens of protein and gene identifiers exist, generally without the
@@
 merging databases often a cumbersome exercise.
+% [JH comment] 
+% Is this above sentence really true? I thought there was general standardization on 
+% UniProt for proteins, GenBank for genes, and that mappings between the other
+% sorts of identifiers to these primary ones were actually well maintained?
+% As I understand it, the main problem with target representation is that
+% protein/gene targets are often unknown, and then it is necessary to map to
+% pathway, cell, tissue, biological process etc. --> representing the maximum
+% that is known about the mechanism of action for the molecule -- and THESE
+% data types are often not well standardized
+% [JH comment]
 
@@
 from database to database. Hence, we need to be able to capture
-molecular graphs in a risk-bounded machine encoded way.
+molecular graphs in a risk-bounded machine encoded way. %risk-bounded meaning here not clear
 
@@
 to it) and tautomer states (how the molecule can change when a proton
-migrates from one part of the molecule to another), the challenge
+migrates from one part of the molecule to another), the challenge  %This is repeating something stated earlier
 increases. The need for a unique representation that is invariant with
 respect to the atom ordering arises due to the fact that graph
-isomorphism (i.e., checking whether two molecules are the same) is an
+isomorphism (i.e., checking whether two structures represent the same molecule) is an
 expensive operation. The first such ``unique representation
@@
 structure-based identifiers became a pressing need, resulting in the
-IUPAC International Chemical Identifier (InChI) to be developed. The
+development of the IUPAC International Chemical Identifier (InChI). The
 InChI is a non-proprietary, structured textual identifier for chemical
@@
 search packages are often implemented and optimized for a given
-database technology; for example, the OrChem package is an open source
+database technology, for example, the OrChem package is an open source
 chemical search package for the Oracle database \cite{rijnbeek2009}.
@@
 
-\subsection{Linking Molecules to Other Data}
+\subsection{Situating Molecular Data in Biological Context}
 \label{sec:prof-ident}
@@
 The preceding section focuses on the choice of representation of
-chemical structures. With the profusion of databases that contain
-details of small molecules, an emerging problem is the integration of
-such databases. Such integration forms the basis of ``systems'' level
-investigations of biological systems. A current challenge in
-cheminformatics is how to identify ``equivalent'' objects across data
-sources. At one level, equivalent compounds can be identified via
-canonical structure representations (InChI). But there are other
-equivalences beyond structure. For example, compounds may bind to the
-same or similar targets; they may be have similar biological function
-(say, anti-neoplastic, that is to say, it prevents the development of
-tumors) and so on. To assess such higher level characteristics,
-ontologies and other semantic technologies have been employed, to first,
-support annotations of chemical systems and second, enable semantic
-analyses of chemical systems in conjunction with other biological
-systems (such as pathways). A number of challenges remain in this area
-- the use of ontologies allows one to build a semantic network of
-concepts that links molecules to other molecules as well as other
-objects such as targets, genes and pathways. Open questions include
-the following.  How does one evaluate similarities in such a
-heterogenous network \cite{couto2010}? How does one integrate chemical
-structure information with ontologies \cite{hastingsowled2010}? And
-more generally, how can we achieve automated inferencing in such
-heterogenous linked systems, allowing us to achieve a systems level
-understanding of small molecules \cite{Oprea:2007fk}.
+chemical structures. In the quest for the discovery of novel
+therapeutic agents, it is essential to study molecules as they are
+situated in a biological context, underlying the processes of life, 
+as metabolites, cofactors for enzymatic reactions, hormones, signalling
+molecules and environmental xenobiotics. This gives rise to the need
+to adequately represent and organise chemical data across biological
+databases such as those containing pathways, protein information or 
+biological samples. Integration and processing of data from such 
+disparate domains underlies ``systems'' level biological research.
+This gives rise to some additional challenges that standard chemical
+structural representation cannot address: firstly, the representation of
+\textit{classes of chemical entities} -- since, in many cases, a class of 
+compounds behave in a certain context in a biological system rather than
+a single molecule with a fully-specified structure -- and for similar
+reasons, chemical concepts such as \textit{groups} (parts of molecules); 
+secondly, the need to represent \textit{non-structural} groupings of 
+compounds of interest, since compounds may bind to the same or similar 
+targets or they may have  similar biological functions such as acting as 
+an anti-neoplastic -- preventing the development of tumors. With the 
+many different biological databases needing to refer to these different
+chemical entities in an organised and standardized 
+fashion to enable cross-database integration for a whole-systems 
+perspective, \textit{ontologies} and other semantic technologies 
+are being successfully employed to support annotation, classification and
+semantic cross-domain querying of chemical entities within
+biological systems. The most widely used ontology for biologically relevant
+chemical entities is ChEBI \cite{chebi2010}, containing around 26,500 
+entities in an interlinked semantic graph as of November 2011, and
+being actively used for annotation in dozens of biological databases. 
+Ontologies are based on logical languages of 
+varying levels of expressivity, accompanied by sophisticated reasoning
+algorithms. They can be used to construct various sorts of semantic
+similarity (e.g. similarity in function, similarity in application) 
+which complement traditional structural similarity methods \cite{couto2010}.
+An open challenge is integrating graph-based chemical
+structure information with the wider logic-based ontology information, which 
+will allow combined automated reasoning and querying over the different domains 
+encoded in a growing number of interlinked ontologies \cite{hastingsowled2010}. 
+Enabling automated inferencing in such heterogenous linked systems brings us closer to
+achieving a systems level understanding of bioactive chemistry \cite{Oprea:2007fk}.
 
@@
 paraffin by Cayley, Polya and others. In the 1960's, Lederberg, 
-Dejerassi and others developed algorithms to enumerate structures
+Djerassi and others developed algorithms to enumerate structures
 based on spectral data, leading to the development of DENDRAL, widely
@@
 rise to the observed spectra. Some commercial products such as MOLGEN
-(\url{http://molgen.de}) are able to perform this task in a rapid
-manner.
+(\url{http://molgen.de}) are able to perform this task rapidly.
 
@@
 While the complexity of molecular graph enumeration remains an open
-problem, an alternative to enumerating the space is to sample it
+problem, 
+% [JH comment]
+% Not sure what is meant here? Surely, the complexity of molecular graph enumeration 
+% is upper-bounded exponentially in the number of atoms? Additional complexity
+% arises from algorithms to select "good" candidates from the candidates?
+% Perhaps what is meant is not that the complexity itself is an open problem,
+% but finding methods for enumeration which are more efficient is an open problem?
+% Alternatively, if complexity in the CS sense is not what is meant here, maybe it will
+% help this audience if a different word is used. 
+% [JH comment]
+an alternative to enumerating the space is to sample it
 \cite{goldberg1999}. Sampling procedures based on the Metropolis or
-the Genetic Algorithms have been developed and used to elucidate
-natural compounds from NMR data. It should be noted that even in the
+Genetic Algorithms have been developed and used to elucidate
+compounds from NMR data. It should be noted that even in the
 face of the computational complexity of molecular enumeration,
@@
 these require specific enumeration procedures, which are still the
-subject of investigations by the cheminformatics community.
+subject of research by the cheminformatics community.
 
@@
 $G$ and $G'$ provides a measure of how similar $G$ is to $G'$, or a
-kernel on a single graph compares measures similarities between the
+kernel on a single graph compares measured similarities between the
 graph's nodes. In these cases, rather than compute vectorial
@@
 also some explanation as to why a molecule is predicted as toxic or
-non-toxic, for example. The former scenario is common in virtual
+non-toxic (for example)? The former scenario is common in virtual
 screening settings, where we require a high degree of accuracy and fast
@@
 One of the key challenges faced by predictive modeling is the fact
-that small molecules are not static and do not exist in
-isolation. Traditionally, predictive models have focused on a single
+that small molecules are not static and do not exist in isolation. 
+Traditionally, predictive models have focused on a single
 structure for a small molecule and have ignored the role of the
@@
 conformations. As a result, enhancing the accuracy of predictions
-ideally will require that the 3D geometries of that molecule be taking into account and, as
-far as possible, take into account the receptor
-simultaneously. Multi-conformer modeling has been addressed in the
+ideally will require that the 3D geometries of that molecule be taken 
+into account and, as far as possible, that the receptor also be taken
+into account simultaneously. Multi-conformer modeling has been addressed in the
 4D-QSAR methodology described by Hopfinger and co-workers
@@
 structural descriptors within a single model. Such an approach could
-lead to more robust predictions for panel assays (that is, which battery of tests would be most useful to perform). Finally, this
+lead to more robust predictions for panel assays (that is, which 
+battery of tests would be most useful to perform). Finally, this
 approach also better reflects clinically relevant compound
@@
 only for reducing the risk in current research, but also for enabling
-new collaboration/innovation opportunities with internal partners and
+new collaboration and innovation opportunities with internal partners and
 a rapidly growing number of external/public partners. In
@@
 tools. The overall architecture was too inflexible to
-adopt to fast changing data standards, scientists spent too much time
-with administrative work, chemical data quality was inconsistent, an
+adapt to fast-changing data standards, scientists spent too much time
+with administrative work, chemical data quality was inconsistent, and
 alignment with other working groups was inefficient (\emph{e.g.}, for
@@
 cost~\cite{WavingGoodbye}.  It is worth mentioning that chemistry is
-traditionally a field where many people have been used to huge
+traditionally a field where many people have been accustomed to huge
 bookshelves of reactions and compounds. Chemists changed to online
@@
 companies producing ELNs as of February 2011~\cite{ELNreview}, for use
-in chemistry, biology, quality control, and other research.  Another
+in chemistry, biology, quality control, and other research domains.  Another
 knowledge management tool specifically for chemists is the web-based
@@
 licenses) and Open Source.
+% [JH comment]
+% we should standardize our nomenclature for the phrase 'Open Source' in the
+% document. Here it is capitalized, earlier it has been uncapitalized 
+% ('open source'), and below it is hyphenated! (open-source)
+% [JH comment]
 
@@
 arguments for and against Open Source; we focus on Open Source
-tools simply because this ensures that a reader will be able to explore
-cheminformatics problems with a minimum of encumbrances.
+tools simply because this ensures that any interested reader will be able to explore
+cheminformatics problems with a minimum of hindrances.
 
@@
 there is ample scope for software engineering (algorithm
-implementation, code quality and analysis, build systems etc), both
+implementation, code quality and analysis, building systems and so on), both
 the CDK and Open Babel are open to new contributions. Both are driven
@@
 taking into account symmetry, and symmetry detection itself, which not
-merely derives from the chemical graph, but typically also from the 3D
+only derives from the chemical graph, but typically also from the 3D
 geometry of the chemical. It is important to realize that although
 chemical representations have increased in sophistication over the
-years, many chiral molecules cannot be satisfactorily handled by
+years, many chiral molecules or non-standard bonding cases such as 
+organometallic compounds cannot be satisfactorily handled by
 current representation systems. Perhaps a more challenging aspect is
@@
 non-static graphs, in order to take into account delocalisation and
-the tautomeric phenomena molecules can undergo in practice.
+the tautomeric phenomena molecules undergo in practice.
 
@@
 and software have a more direct impact on profits, in terms of
-identifying lead like compounds and improving properties of drug
+identifying lead-like compounds and improving the properties of drug
 candidates. In contrast, much bioinformatics software is employed in
@@
 that a small molecule binds to).  While certainly
-important, this type of research is not as directly related to
+important, this type of research is perhaps not as directly related to
 possible profits as an actual small molecule.
 
-Furthermore, chemical data (structure, activity information etc.) is
-harder to acquire and when done on large scales can be extremely
-intensive in terms of time and effort. Thus much of it has been
-proprietary. Whereas a lot of bioinformatics data (sequences for
+Furthermore, chemical data (structure, activity information etc.) has
+been harder to acquire and when done on large scales can be extremely
+intensive in terms of time and effort, whereas bioinformatics data (sequences for
 example) are much easier to obtain. While both fields have theoretical
@@
 biosystems implies that cheminformatics and bioinformatics will both
-be required to tackle future problems.
+be required to tackle future cross-domain integrative problems.
 
@@
 open-source cheminformatics projects.  If more chemists think
-algorithmically, and more computer scientists think chemically we are
-much better positioned taking not only ``simple'' chemicals into
+algorithmically, and more computer scientists think chemically we will be
+much better positioned to take not only ``simple'' chemicals into
 account (only one identifier and isomer possible per molecule), but
@@
 Northwestern University, supported in part by NSF grant
-CCF-1049899. A. Bender thanks Unilever for funding.
+CCF-1049899. A. Bender thanks Unilever for funding. J. Hastings
+thanks the EU project EU-OPENSCREEN for funding.
 
@@
 \item \textit{QSAR} - Quantitative Structure-Activity Relationships; methods to relate the structure and the
-    biological activity of a chemical structure in a quantitative manner.
+    biological activity of a chemical structure in a quantitative fashion.
 \item \textit{Reaction catalysis} - Catalysis is the change in rate of a chemical reaction due to the participation
@@
 \item \textit{SMARTS} - SMiles ARbitrary Target Specification \\(SMARTS) is a language for specifying substructure
-    patterns in molecules
+    patterns in molecules.
 \item \textit{Stereoisomer/stereoisomerism} - Stereoisomers are isomeric molecules that have the same molecular

### 2011-11-20 paper.tex

@@
   interdisciplinary collaboration between computer scientists and
-  molecular biologists led as one of its major accomplishments to the
-  sequencing of the human genome, where researchers connected
-  biological concepts to the theory of string algorithms. However,
+  molecular biologists that led as one of its major accomplishments to the
+  sequencing of the human genome. However,
   despite this great success, few computer scientists are familiar
@@
 the foundational algorithms of cheminformatics have been described
-since the 1950s, publicly available (open source) software implementing many of
+since the 1950s, publicly available software implementing many of
 these algorithms has only been accessible in the last ten to fifteen
@@
   (and valid) aspects of the same molecule and are suited for
-  different purposes. The choice of representation is guided by the
-  intended use of the structure data.}
+  different purposes, with the choice of representation guided by its
+  intended use.}
 \label{figure:chemical-structures}
@@
 information than others, they are all equally valid and are suited for
-different problems. Indeed, Fig.~\ref{figure:chemical-structures}
-represents a hierarchy of representations. Thus, when searching for
-substructures, a 2D representation is sufficient. When trying to
-rationalize why a molecule binds to a protein, a 3D structure is
-vital.
+different problems. Fig.~\ref{figure:chemical-structures} shows 
+a hierarchy of representations; when searching for
+substructures, a 2D representation suffices, but when exploring 
+protein binding, a 3D structure is vital.
 
@@
 perspective on what is presently known, tracking ongoing advances in 
-science as these become available. Still, the integration of
-chemical databases with one another remains a challenge, not only due
+science as these become available. Still, integration of data
+across chemical databases remains a challenge, not only due
 to the normalization problems in the chemical and bioactivity
-landscape, but also due to sheer data volume. One problem is the
-annotation of the ``protein targets'' a molecule binds to, as  
-dozens of protein and gene identifiers exist, generally without the
-availability of a one-to-one mapping from each ontology, which makes
-merging databases often a cumbersome exercise.
-% [JH comment] 
-% Is this above sentence really true? I thought there was general standardization on 
-% UniProt for proteins, GenBank for genes, and that mappings between the other
-% sorts of identifiers to these primary ones were actually well maintained?
-% As I understand it, the main problem with target representation is that
-% protein/gene targets are often unknown, and then it is necessary to map to
-% pathway, cell, tissue, biological process etc. --> representing the maximum
-% that is known about the mechanism of action for the molecule -- and THESE
-% data types are often not well standardized
-% [JH comment]
-
-Given the requirement to ``make sense'' of as much chemical data at
-hand as possible, we have to be able to search in complex data, for
+landscape, but also due to sheer data volume. 
+Given the requirement to ``make sense'' of as much chemical data 
+as possible, we have to be able to search in complex data, for
 example multi-labelled molecular graphs, where graph labels can change
-from database to database. Hence, we need to be able to capture
-molecular graphs in a risk-bounded machine encoded way. %risk-bounded meaning here not clear
+from database to database. Hence, molecular graphs must be captured 
+in a machine-readable fashion.
 
 A fundamental challenge in cheminformatics is the unique
-representation of chemical structures in a machine readable format. A
+representation of chemical structures in a machine-readable format. A
 (chemical) graph can be encoded in multiple ways, depending on how one
 orders the nodes, resulting in multiple representations of the same
-molecule. When one considers the need to include other aspects such as
-protonation states (how the molecule changes when one proton is added
+molecule. Considering protonation states 
+(how the molecule changes when one proton is added
 to it) and tautomer states (how the molecule can change when a proton
-migrates from one part of the molecule to another), the challenge  %This is repeating something stated earlier
+migrates from one part of the molecule to another), the challenge  
 increases. The need for a unique representation that is invariant with
@@
 
-\subsection{Situating Molecular Data in Biological Context}
+\subsection{Molecules in their Biological Context}
 \label{sec:prof-ident}
@@
 will allow combined automated reasoning and querying over the different domains 
-encoded in a growing number of interlinked ontologies \cite{hastingsowled2010}. 
+encoded in a growing number of interlinked ontologies.% \cite{hastingsowled2010}. 
 Enabling automated inferencing in such heterogenous linked systems brings us closer to
@@
 
-While the complexity of molecular graph enumeration remains an open
-problem, 
-% [JH comment]
-% Not sure what is meant here? Surely, the complexity of molecular graph enumeration 
-% is upper-bounded exponentially in the number of atoms? Additional complexity
-% arises from algorithms to select "good" candidates from the candidates?
-% Perhaps what is meant is not that the complexity itself is an open problem,
-% but finding methods for enumeration which are more efficient is an open problem?
-% Alternatively, if complexity in the CS sense is not what is meant here, maybe it will
-% help this audience if a different word is used. 
-% [JH comment]
+While molecular graph enumeration remains a challenge, 
 an alternative to enumerating the space is to sample it
-\cite{goldberg1999}. Sampling procedures based on the Metropolis or
+\cite{goldberg1999}. Sampling procedures based on Metropolis or
 Genetic Algorithms have been developed and used to elucidate
 compounds from NMR data. It should be noted that even in the
-face of the computational complexity of molecular enumeration,
+face of the complexity of molecular enumeration,
 computational chemists have developed and successfully used
 enumeration tools to generate large chemical libraries such as GDB-13,
-comprising almost a billion chemical structures~\cite{GDB} (but only
+comprising almost a billion chemical structures~\cite{GDB} (only
 considering molecules up to 13 heavy atoms and using only carbon,
@@
 
-Another relevant application of enumeration methods is in the
+Another relevant application of enumeration methods is the
 \emph{generation of chemical reaction networks}. The problem here
@@
 engineering strategies to biosynthesize compounds of commercial
-interest. The difficulty with metabolic network design is that in
+interest. With metabolic network design, a difficulty is that in
 addition to network generation based on reactions, one also needs to
 verify that there are enzymatic events possible for enabling reactions
-catalyses (in other words, enzymes must be present to reduce the
+catalyses (enzymes must be present to reduce the
 energy required for some reactions to take place). That additional
-task requires encompassing both chemical structure and protein
-sequence information and the development of tools that are at the
+task requires encompassing both chemical structures and protein
+sequences and the development of tools that are at the
 interface between cheminformatics and bioinformatics.
@@
 structure. Together with the \emph{similar property principle}
-\cite{Johnson:1990qf} (structurally similar molecules will exhibit
+\cite{Johnson:1990qf} (structurally similar molecules exhibit
 similar properties), the goal of any modeling approach is to capture
-and characterize the correlations between structural features and the
+and characterize correlations between structural features and
 observed biological activities. Simultaneously, such approaches must
-also describe the likelihood of error when using such
+also describe the likelihood of error when using the
 models for decision making.  A variety of approaches can be employed
@@
 the region of the input (i.e., chemical structures) that can be
-reliably predicted, usually deliniated using some form of similarity
+reliably predicted, usually delineated by similarity
 to the training set.
@@
 into account both ligand and receptor structures. The last method is
-an example of an extension of statistical QSAR methods to simultaneous
-model the receptor and ligand as first reported by Lapinsh \textit{et
+an extension of statistical QSAR methods to simultaneously
+model the receptor and ligand, as first reported by Lapinsh \textit{et
  al.}~\cite{lapinsh2001}.
@@
 requires one to identify a subset of \emph{relevant} (to the property
-being modeled) descriptors. This is the feature selection problem and
-is well discussed in the data mining literature. A kernel approach
-does not require feature selection; but now, one faces the problem of
-evaluating a data set in a pairwise fashion. Furthermore, one must now
+being modeled) descriptors: the feature selection problem 
+well discussed in the data mining literature. A kernel approach
+does not require feature selection, but one faces the problem of
+evaluating a data set in a pairwise fashion, and one must
 identify an appropriate kernel. This is an important challenge as the
@@
 Traditionally, predictive models have focused on a single
-structure for a small molecule and have ignored the role of the
-receptor (in receptor-mediated scenarios). Yet, small molecules can
+structure for a small molecule and ignored the receptor. Yet, small molecules can
 exist in multiple tautomeric forms and usually in multiple
@@
 ideally will require that the 3D geometries of that molecule be taken 
-into account and, as far as possible, that the receptor also be taken
-into account simultaneously. Multi-conformer modeling has been addressed in the
-4D-QSAR methodology described by Hopfinger and co-workers
+into account and, as far as possible, that the receptor also be considered. 
+Multi-conformer modeling has been addressed in the
+4D-QSAR methodology described by Hopfinger \textit{et al.}
 \cite{Albuquerque:1998ys}.  Recent techniques such as
@@
 knowledge management tool specifically for chemists is the web-based
-service Reaxys, founded in 2009, whose company goal is to provide ``a
+service Reaxys, founded in 2009, which aims to provide ``a
 single, fully integrated chemical workflow solution.''
@@
 Due to the importance of cheminformatics to the pharmaceutical
-industry in particular, numerous commercial vendors provide libraries,
+industry, numerous commercial vendors provide libraries,
 applications and database cartridges to handle a wide variety of
-needs including Accelrys, BioSolveIT, ChemAxon, Chemical Computing Group,
+needs including Accelrys, BioSolveIT, Chem-Axon, Chemical Computing Group,
 Daylight, Open Eye, Schrodinger, Tripos, and Xemistry. More recently
@@
 arguments for and against Open Source; we focus on Open Source
-tools simply because this ensures that any interested reader will be able to explore
+tools because this ensures that an interested reader will be able to explore
 cheminformatics problems with a minimum of hindrances.
@@
 
-It is important to note that certain challenges faced by nearly all
+Certain challenges faced by nearly all
 cheminformatics toolkits stem from the graph representation of
@@
 organometallic compounds cannot be satisfactorily handled by
-current representation systems. Perhaps a more challenging aspect is
+current representation systems. A more challenging aspect is
 that the chemical representation of a molecule really needs to capture
@@
 \emph{benchmarking of algorithms}. The latter is especially relevant
-for the plethora of data mining techniques that are employed in
+for the plethora of data mining techniques employed in
 cheminformatics. At this point, the nature of Open Data focuses
@@
 there is a distinct lack of textual data (such as journal articles)
-that are ``Open'' in nature. While PubMed abstracts serve as a proxy
+that are ``Open''. While PubMed abstracts serve as a proxy
 for journal articles, text mining methods in cheminformatics are
@@
 data and tools have been freely available since the inception of the
-field. One could argue that the outcomes from cheminformatics methods
+field. This disparity could be attributed to the fact that 
+the outcomes from cheminformatics methods
 and software have a more direct impact on profits, in terms of
 identifying lead-like compounds and improving the properties of drug
-candidates. In contrast, much bioinformatics software is employed in
-upstream areas such as target identification (identifying the protein
-that a small molecule binds to).  While certainly
-important, this type of research is perhaps not as directly related to
+candidates, while bioinformatics software is employed in
+upstream areas such as target identification, 
+perhaps not as directly related to
 possible profits as an actual small molecule.
 
-Furthermore, chemical data (structure, activity information etc.) has
-been harder to acquire and when done on large scales can be extremely
-intensive in terms of time and effort, whereas bioinformatics data (sequences for
-example) are much easier to obtain. While both fields have theoretical
+%Furthermore, chemical data (structure, activity information etc.) has
+%been harder to acquire and when done on large scales can be extremely
+%intensive in terms of time and effort, whereas bioinformatics data (sequences for
+%example) are much easier to obtain. 
+While both fields have theoretical
 components, it is possible that the free availability of large amounts

### 2011-11-20 OpenQuestions.tex
Minor rewrite
Another challenge was added

@@
 (the linking normalization problem for different protomers, tautomers, or special cases of isomerisms remain open), but we
-are also interested in being able to search efficiently within molecules when being linked via semantic web approaches.
+are also interested in being able to search efficiently within molecules when being linked via semantic web approaches. Typical
+searches will require being able to apply substructure or similarity searches.
 What could be algorithmic solutions for this?
@@
 %
+\item \emph{Dynamic similarity search on instant binary vectors}.
+
+Binary feature vectors are a common practice for chemical similarity searches. The typical process starts with 1. creating
+binary substructure (or other feature vectors) \cite{citeulike:8530538}, 2. creating fixed-length binary vectors of typically
+1024 bits for reducing space requirements and speeding up further similarity searches (by loosing some accuracy), 3. creating 
+further pre-computations for speeding up threshold based similarity searches \cite{doi:10.1021/ci800076s}.\\
+Still, if we are interested to employ dynamic changes in the similarity encodings, e.g. using only a set of binary features,
+then previously done hashing or pre-computations might need to be redone efficiently on the fly. Finally, the major goal
+is to employ similarity searches \cite{doi:10.1021/ci200235e} on a scale of multiple million entries and more optimizations and
+benchmarking studies are urgently required, e.g. using GPUs
+\cite{doi:10.1021/ci1004948}, optimizing pair-wise similarity calculations \cite{MINF:MINF201100050}.
+%
 \item \emph{Large-scale browsing of molecular property spaces, e.g. on scaffold level, side-effect-level, ...}.
@@
 %
-\item \emph{Patent text mining (curation)}.
+\item \emph{Chemical image/text mining in patents (curation)}.
 
@@
 how can we ensure that only pre-defined data entries 
-(and a limited number of annotations, e.g. bioological activities) are visible to a partner.
+(and a limited number of annotations, e.g. biological activities) are visible to a partner.
 How can this be organized for a multitude of partners? Cloud computing, user management, encryption granularity and efficient security management?

### 2011-11-21 OpenQuestions.tex
Added in Joerges edits

@@
 
-Binary feature vectors are a common practice for chemical similarity searches. The typical process starts with 1. creating
-binary substructure (or other feature vectors) \cite{citeulike:8530538}, 2. creating fixed-length binary vectors of typically
-1024 bits for reducing space requirements and speeding up further similarity searches (by loosing some accuracy), 3. creating 
-further pre-computations for speeding up threshold based similarity searches \cite{doi:10.1021/ci800076s}.\\
-Still, if we are interested to employ dynamic changes in the similarity encodings, e.g. using only a set of binary features,
-then previously done hashing or pre-computations might need to be redone efficiently on the fly. Finally, the major goal
-is to employ similarity searches \cite{doi:10.1021/ci200235e} on a scale of multiple million entries and more optimizations and
+  Binary feature vectors are a common practice for chemical similarity
+  searches. The typical process starts with 1. creating binary
+  substructure (or other feature vectors) \cite{citeulike:8530538},
+  2. creating fixed-length binary vectors of typically 1024 bits for
+  reducing space requirements and speeding up further similarity
+  searches (by loosing some accuracy), 3. creating further
+  pre-computations for speeding up threshold based similarity searches
+  \cite{doi:10.1021/ci800076s}.\\  Still, if we are interested to employ
+  dynamic changes in the similarity encodings, e.g. using only a set
+  of binary features, then previously done hashing or pre-computations
+  might need to be redone efficiently on the fly. Finally, the major
+  goal is to employ similarity searches \cite{doi:10.1021/ci200235e}
+  on a scale of multiple million entries and more optimizations and
   benchmarking studies are urgently required, e.g. using GPUs
-\cite{doi:10.1021/ci1004948}, optimizing pair-wise similarity calculations \cite{MINF:MINF201100050}.
+  \cite{doi:10.1021/ci1004948}, optimizing pair-wise similarity
+  calculations \cite{MINF:MINF201100050}.
 %
+
 \item \emph{Large-scale browsing of molecular property spaces, e.g. on scaffold level, side-effect-level, ...}.

### 2011-11-21 OpenQuestions.tex
added a hadoop challenge

@@
 
-There exist a multitude of programs generation 3D conformations of molecules. Still, often input formats vary and require
-to re-create 3D conformational spaces with other programs. This does not only complicate the maintenance scenario of such
-databases, especially with the explosion of public databases, but also creates hugh space and time complexity issues when
-searching within such databases. What could be better interfaces, maintenance, data structures, and private/public sharing
-scenarios for conformational 3D databases? 
-Many software vendors use different solutions for parallizing compute jobs: SGE, PVM, MPI, etc. 
-Everyone knowing the enterprise IT approval cycles might know that having more than one parallel processing framework 
-is not easy. What could be better ways to streamline parallel processing structures for 
-cheminformatics (and molecular modeling) algorithms? Is a cloud really an option? What about SaaS with secured data transfer?
-Can this also offer alternative licensing strategies for software suites in this domain?
+  There exist a multitude of programs generation 3D conformations of
+  molecules. Still, often input formats vary and require to re-create
+  3D conformational spaces with other programs. This does not only
+  complicate the maintenance scenario of such databases, especially
+  with the explosion of public databases, but also creates hugh space
+  and time complexity issues when searching within such
+  databases. What could be better interfaces, maintenance, data
+  structures, and private/public sharing scenarios for conformational
+  3D databases?
+
+\item \emph{Parallelization of cheminformatics problems \& algorithms}
+Many cheminformatics tasks apply algorithms over large input files or
+across many molecules. A simple way to parallelize this is to chunk
+the input data and let individual threads/node process each chunk. A
+trivial solution is to manually chunk the input and submit a series of
+jobs to a scheduling system.
+
+Can we employ modern frameworks like Hadoop to perform this? Can we
+develop a generalized framework that includes the requisite
+cheminformatics tools that allows users to seamless distribute jobs
+over Hadoop and other map reduce systems?
+
+Going further, are there cheminformatics problems that can make use of
+the map/reduce paradigm at the algorithmic level? More generally, are
+cheminformatics algorithms amenable to parallelization (that does not
+involve chunking the input data)
+
+\item \emph{Licensing in a parallel world} Many software vendors use
+  different solutions for parallizing compute jobs: SGE, PVM, MPI,
+  etc.  Is a cloud really an option?  What about SaaS with secured data
+  transfer?  Can this also offer alternative licensing strategies for
+  software suites in this domain?
+
 \end{enumerate}

### 2011-11-21 OpenQuestions.tex
some more edits

@@
 
-Certain molecules might have hundreds of biological activities, side-effects in humans (SIDER database \cite{Kuhn_Campillos_Letunic_Jensen_Bork_2010}), or
-many other properties attached to them. What are large-scale mining and visulization options? 
-How can we mine private and public data sources at the very same time?
+  Certain molecules might have hundreds of biological activities,
+  side-effects in humans (SIDER database
+  \cite{Kuhn_Campillos_Letunic_Jensen_Bork_2010}), or many other
+  properties attached to them. What are large-scale mining and
+  visulization options?  How can we mine private and public data
+  sources at the very same time?
 %
@@
 
-There are various tools for doing automatic text mining on chemical patents. Still, the overall acceptance rate of chemical
-text mining is improvable, since many medicinal chemists are very concerned about the data quality of such efforts. 
-What could be done to improve the mining quality, curate the obtained data, and to provide confidence level estimations 
-for each molecule coming from patent mining? Do require image2structure and text2structure mining also data stores
-for ensuring a sufficient amount of confidence and data quality?
-How can patent mining be used to create new drugs faster or to speed-up collaboration/licensing discussions?
+  There are various tools for doing automatic text mining on chemical
+  patents. Still, the overall acceptance rate of chemical text mining
+  is improvable, since many medicinal chemists are very concerned
+  about the data quality of such efforts.  What could be done to
+  improve the mining quality, curate the obtained data, and to provide
+  confidence level estimations for each molecule coming from patent
+  mining? Do require image2structure and text2structure mining also
+  data stores for ensuring a sufficient amount of confidence and data
+  quality?  How can patent mining be used to create new drugs faster
+  or to speed-up collaboration/licensing discussions?
 \end{enumerate}
@@
 
-Vectorial molecule encodings can serve as efficient approximations of molecules.
-Sometimes non-vectorial molecular 3D shape or molecule kernel comparisons might be more suitabe to compare molecules, since
-they might better correlate with activities. One key problem is that non-vectorial encodings require to compare all molecules
-(or their 3D conformational eplosions) in a pair-wise manner.
-This becomes prohibitively expensive when considering millions of molecules.
-Can dyadic data approaches help \cite{Hochreiter:2006:SVM:1159508.1159516}? Other approximations or cascading flows?
+  Vectorial molecule encodings can serve as efficient approximations
+  of molecules.  Sometimes non-vectorial molecular 3D shape or
+  molecule kernel comparisons might be more suitabe to compare
+  molecules, since they might better correlate with activities. One
+  key problem is that non-vectorial encodings require to compare all
+  molecules (or their 3D conformational eplosions) in a pair-wise
+  manner.  This becomes prohibitively expensive when considering
+  millions of molecules.  Can dyadic data approaches help
+  \cite{Hochreiter:2006:SVM:1159508.1159516}? Other approximations or
+  cascading flows?
 %
@@
 
-As an example: Biological activities might not be independent of each other, but have a certain correlation between each other.
-In Chemogenomics this is used for creating models of combining molecules with protein sequences, molecules with active sites of proteins,
-or molecules with biological activities of multiple assays. How can we optimize such highly complex mining scenarios, especially
-when considering large-scale data sets with hundred of thousands molecules and thousands of biological activities? 
-How can we combine, mine, and visualize categorial and continuous output variables, e.g. hydrophobicity of a molecule and toxicity in humans, 
-by still being able to make concrete proposals to medicinal chemistry? Is analoging (creating very small modifications of a molecule and measuring its activities)
-really the most efficient way forward? If we test molecules, should we test it in a single biological assay or in multiple biological assays, if multiple, which ones?
-If a company does not have a biological assay within reach, which other partner could offer testing a molecule within two days (vendor matching based on licenses or contracts)?
+  As an example: Biological activities might not be independent of
+  each other, but have a certain correlation between each other.  In
+  Chemogenomics this is used for creating models of combining
+  molecules with protein sequences, molecules with active sites of
+  proteins, or molecules with biological activities of multiple
+  assays. How can we optimize such highly complex mining scenarios,
+  especially when considering large-scale data sets with hundred of
+  thousands molecules and thousands of biological activities?  How can
+  we combine, mine, and visualize categorial and continuous output
+  variables, e.g. hydrophobicity of a molecule and toxicity in humans,
+  by still being able to make concrete proposals to medicinal
+  chemistry? Is analoging (creating very small modifications of a
+  molecule and measuring its activities) really the most efficient way
+  forward? If we test molecules, should we test it in a single
+  biological assay or in multiple biological assays, if multiple,
+  which ones?  If a company does not have a biological assay within
+  reach, which other partner could offer testing a molecule within two
+  days (vendor matching based on licenses or contracts)?
 \end{enumerate}
@@
 develop a generalized framework that includes the requisite
-cheminformatics tools that allows users to seamless distribute jobs
+cheminformatics tools that allows users to seamlessly distribute jobs
 over Hadoop and other map reduce systems?
@@
 cheminformatics algorithms amenable to parallelization (that does not
-involve chunking the input data)
+involve chunking the input data)? Are they worth the effort?
 

### 2011-11-21 OpenQuestions.tex
some clean up of open questions

@@
 
-\section*{Data mining}
+\section*{Data mining \& machine learning}
 \begin{enumerate}
@@
   or to speed-up collaboration/licensing discussions?
-\end{enumerate}
 
-\section*{Machine learning}
-\begin{enumerate}
 \item \emph{Large-scale vectorial versus kernels molecule similarity}
@@
 
-\section*{Software engineering}
+\section*{Databases \& software engineering}
 \begin{enumerate}
-\item \emph{Large-scale exchange and use of 3D conformational databases}
+\item \emph{Large scale conformational databases}
 
-  There exist a multitude of programs generation 3D conformations of
-  molecules. Still, often input formats vary and require to re-create
-  3D conformational spaces with other programs. This does not only
-  complicate the maintenance scenario of such databases, especially
-  with the explosion of public databases, but also creates hugh space
-  and time complexity issues when searching within such
-  databases. What could be better interfaces, maintenance, data
-  structures, and private/public sharing scenarios for conformational
-  3D databases?
+  While many programs are availabel to generate 3D conformations, on
+  the fly generation for large collections can be time
+  consuming. Rather, can we store massive conformer collections and
+  support 3D searches over them? Beyond generating the conformers
+  themselves, this problem has several aspects covering database
+  design, parallel systems and software engineering:
+  \begin{itemize}
+  \item How many conformers are required to provide \emph{sufficient} coverage?
+  \item What representation will be used to store conformers and run
+    queries? How does the choice of representation affect the type of
+    queries we can run?
+  \item Most 3D similarity approaches either employ a vectorial
+    representation or a volumetric representation. How can these be
+    efficiently indexed?
+  \item What type of parallel infrastructure can be used to speed up queries?
+  \end{itemize}
 
-\item \emph{Parallelization of cheminformatics problems \& algorithms}
-Many cheminformatics tasks apply algorithms over large input files or
-across many molecules. A simple way to parallelize this is to chunk
-the input data and let individual threads/node process each chunk. A
-trivial solution is to manually chunk the input and submit a series of
-jobs to a scheduling system.
+\item \emph{Map/Reduce in cheminformatics}
+  Many cheminformatics tasks apply algorithms over large input files
+  or across many molecules (similarity and substructure search,
+  docking, filtering). A simple way to parallelize this is to chunk
+  the input data and let individual threads/nodes process each chunk. A
+  trivial solution is to manually chunk the input and submit a series
+  of jobs to a scheduling system. Using the Hadoop ecosystem as an
+  example, we can ask various questions:
+  \begin{itemize}
+  \item Can we employ modern frameworks like Hadoop to support
+    embarassingly parallel cases (filtering $10^7$ molecule libraries)?
+  \item Can we develop a generalized framework that includes the
+    requisite cheminformatics tools that allows users to seamlessly
+    distribute jobs over Hadoop and other map reduce systems?
+  \item Is the latency involved in Hadoop based datastores (HBase)
+    worth the ability to handle massive molecule collections and run
+    M/R queries across them? 
+  \item Going further, are there cheminformatics problems that can make use
+  of the map/reduce paradigm at the algorithmic level?
+  \end{itemize}
+\end{enumerate}
 
-Can we employ modern frameworks like Hadoop to perform this? Can we
-develop a generalized framework that includes the requisite
-cheminformatics tools that allows users to seamlessly distribute jobs
-over Hadoop and other map reduce systems?
+\section*{Enterprise software (KM,ELN)}
+We know that the enterprise software and ELN market is still growing.
+\begin{enumerate}
+\item \emph{Public-private collaboration and security scenarios}
 
-Going further, are there cheminformatics problems that can make use of
-the map/reduce paradigm at the algorithmic level? More generally, are
-cheminformatics algorithms amenable to parallelization (that does not
-involve chunking the input data)? Are they worth the effort?
+  Let us assume an organization, e.g. a commercial company, has a
+  single or a small number of established KM and ELN products.  How
+  can we improve the maintenance, leveraging, and collaboration with
+  many external partners (each of them potentially with another KM/ELN
+  solution)? Which party is hosting which data in which data structure
+  (ontologies?), and how can we ensure that only pre-defined data
+  entries (and a limited number of annotations, e.g. biological
+  activities) are visible to a partner.  How can this be organized for
+  a multitude of partners? Cloud computing, user management,
+  encryption granularity and efficient security management?
 
@@
 
-\section*{Enterprise software (KM,ELN)}
-We know that the enterprise software and ELN market is still growing.
-\begin{enumerate}
-\item \emph{Public-private collaboration and security scenarios}
-
-Let us assume an organization, e.g. a commercial company, has a single or a small number of established KM and ELN products. 
-How can we improve the maintenance, leveraging, and collaboration with many external partners (each of them potentially
-with another KM/ELN solution)? Which party is hosting which data in which data structure (ontologies?), and 
-how can we ensure that only pre-defined data entries 
-(and a limited number of annotations, e.g. biological activities) are visible to a partner.
-How can this be organized for a multitude of partners? Cloud computing, user management, encryption granularity and efficient security management?
-\end{enumerate}
-
 \bibliographystyle{abbrv}

### 2011-11-21 rebuttal.tex

@@
 regression models, one could argue that taking into account the
-correlation structure between the multiple y's could lead to }
-%%% THE ABOVE SENTENCE SEEMS INCOMPLETE %%%
+correlation structure between the multiple y's could lead to improved
+models in scenarios where a molecule has affinities for multiple
+related targets. We admit that the answer is not clear at this point
+and hence suggest that this could a topic for research }
 

### 2011-11-21 OpenQuestions.tex

@@
 \begin{enumerate}
-\item evaluation of similarities in a heterogeneous network (JKW: What are heterogenous networks?).  What is a specific example here?
-\item integrate chemical structure information with ontologies.  Specific example problem?
-\item Classify molecular descriptors up to equivalence, and dependence on one another.
-\item systems-level understanding of small molecules (JKW: which systems? This will help being clearer on the challenge).  What does this mean, and what would a specific challenge problem be?
+
+\item \emph{Small molecules \& phenotypic data}
+
+Phenotypic (where one takes images of cells and then analyzes them
+extract numerical features from the image) screens are increasingly
+common. These types of screens lead to very large, very high
+dimensional datasets.
+\begin{itemize}
+\item What methods are suitable to perform feature selection and
+  modeling of such data
+\item Does phenotype-derived data lead to to ``better'' models for
+  small molecules compared to the usual structure-derived data?
+\end{itemize}
+
+\item \emph{Inverse QSAR}
+
+(statistical) QSAR is an example of traditional data mining where one
+correlates a set of independent variables to a dependent
+variable. This lets one predict properties of a new molecule based on
+its structural features and those of a training set. The inverse QSAR
+problem is such that given a molecule structure representation
+(usually a descriptor vector), what are the possible input structures
+that satisfy the representation. This can be made more complex, by
+asking that the input structures also satisfy aexperimental property
+range.
+\begin{itemize}
+\item What methodologies can be devised to address this problem?
+\item How does a given methodology allow us to translate the
+  descriptor space to a minimal region of the input space?
+\item Are there internal features of a modeling algorithm (say
+  hyperplanes in a SVM approach) that lets us simplify the problem?
+\end{itemize}
+
 %\item \emph{Efficient molecule browsing, e.g. on scaffold level}.
@@
 
-\item \emph{Large-scale browsing of molecular property spaces, e.g. on scaffold level, side-effect-level, ...}.
-
-  Certain molecules might have hundreds of biological activities,
-  side-effects in humans (SIDER database
-  \cite{Kuhn_Campillos_Letunic_Jensen_Bork_2010}), or many other
-  properties attached to them. What are large-scale mining and
-  visulization options?  How can we mine private and public data
-  sources at the very same time?
-%
 \item \emph{Chemical image/text mining in patents (curation)}.
@@
   is improvable, since many medicinal chemists are very concerned
-  about the data quality of such efforts.  What could be done to
-  improve the mining quality, curate the obtained data, and to provide
-  confidence level estimations for each molecule coming from patent
-  mining? Do require image2structure and text2structure mining also
-  data stores for ensuring a sufficient amount of confidence and data
-  quality?  How can patent mining be used to create new drugs faster
-  or to speed-up collaboration/licensing discussions?
+  about the data quality of such efforts.
+
+  \begin{itemize}
+  \item What could be done to improve the mining quality, curate the
+    obtained data, and to provide confidence level estimations for
+    each molecule coming from patent mining?
+  \item Do require image2structure and text2structure mining also data
+    stores for ensuring a sufficient amount of confidence and data
+    quality?
+  \item How can patent mining be used to create new drugs faster or to
+    speed-up collaboration/licensing discussions?
+ \end{itemize}
 
@@
 \begin{enumerate}
-\item \emph{Large scale conformational databases}
+\item \emph{Large scale conformational databases}.
 
@@
 
+\item \emph{Database indexing schemes for chemical representations}.
+
+Databases of chemical structures are ubiqituous, employing a variety
+of RDBMS products. Structure based queries (exact match, substructure,
+similarity) are dependent on the chemical representation - most
+solutions employ a linear string based form (SMILES, InChI) and
+depending on the nature queries to be supported, some are preferred
+over others. While one can perform linear scans, indexing is key to
+efficient query performance.
+\begin{itemize}
+\item  Since many queries use binary fingerprints as pre-screens, what
+  types of indexing schemes can be designed to support queries on
+  binary vectors?
+\item If we choose to support 3D searches (shape, pharmacophore) what
+  indexing scheme will allow us to perform these types of queries
+  efficiently?
+\end{itemize}
+
 \item \emph{Map/Reduce in cheminformatics}
+
   Many cheminformatics tasks apply algorithms over large input files

### 2011-11-21 paper.tex

@@
 From the above it becomes clear that the nature of chemical problems
-is often a complex mixture of combinatorial and continuous
-problems. 
-% If we oversimplify the nature of chemical structures and
-% their properties, we can not describe the world around us in
-% sufficient detail. On the other hand, if we try to encode every detail
-% (\emph{e.g.}, all possible 3D structures and their properties, all
-% metabolites of a structure in humans in different tissues), the
-% complexity of collecting, storing, and mining these data increases
-% dramatically. 
-For the field of cheminformatics to flourish, we need
-closer collaboration between chemists (or, more generally, life
-scientists) and computer scientists -- where the former need to be
-able to pose their problems in a way relevant for practical
-applications; and where the latter are able to devise ways of
-capturing, storing, and analyzing chemical data which achieve optimal
-balances between the different risks in generalizable encodings and
-complexity (space and time). The goals of cheminformatics are to
-support better chemical decision making by 1. storing and integrating
-data in maintainable ways, 2.  providing sufficient open standards and
-tools allowing applications and data to be used across heterogeneous
-platforms, and 3. mining the many chemical property spaces in a time-
-and space-efficient way. We will discuss all three goals in the
-following sections.
+is often a complex mixture of combinatorial and continuous problems.
+For the field of cheminformatics to flourish, we need closer
+collaboration between chemists (or, more generally, life scientists)
+and computer scientists -- where the former need to be able to pose
+their problems in a way relevant for practical applications; and where
+the latter are able to devise ways of capturing, storing, and
+analyzing chemical data which achieve optimal balances between the
+different risks in generalizable encodings and complexity (space and
+time). The goals of cheminformatics are to support better chemical
+decision making by 1. storing and integrating data in maintainable
+ways, 2.  providing sufficient open standards and tools allowing
+applications and data to be used across heterogeneous platforms, and
+3. mining the many chemical property spaces in a time- and
+space-efficient way. We will discuss all three goals in the following
+sections.
 
@@
 search packages are often implemented and optimized for a given
-database technology, for example, the OrChem package is an open source
+database technology, for example, the OrChem package is an Open Source
 chemical search package for the Oracle database \cite{rijnbeek2009}.
@@
 
-The preceding section focuses on the choice of representation of
-chemical structures. In the quest for the discovery of novel
-therapeutic agents, it is essential to study molecules as they are
-situated in a biological context, underlying the processes of life, 
-as metabolites, cofactors for enzymatic reactions, hormones, signalling
-molecules and environmental xenobiotics. This gives rise to the need
-to adequately represent and organise chemical data across biological
-databases such as those containing pathways, protein information or 
-biological samples. Integration and processing of data from such 
-disparate domains underlies ``systems'' level biological research.
-This gives rise to some additional challenges that standard chemical
-structural representation cannot address: firstly, the representation of
-\textit{classes of chemical entities} -- since, in many cases, a class of 
-compounds behave in a certain context in a biological system rather than
-a single molecule with a fully-specified structure -- and for similar
-reasons, chemical concepts such as \textit{groups} (parts of molecules); 
-secondly, the need to represent \textit{non-structural} groupings of 
-compounds of interest, since compounds may bind to the same or similar 
-targets or they may have  similar biological functions such as acting as 
-an anti-neoplastic -- preventing the development of tumors. With the 
-many different biological databases needing to refer to these different
-chemical entities in an organised and standardized 
+In the quest for the discovery of novel therapeutic agents, it is
+essential to study molecules as they are situated in a biological
+context, as metabolites, cofactors for enzymatic reactions, hormones
+and signalling molecules. Thus we need to adequately represent and
+organise chemical data across biological databases such as those
+containing pathways, protein information or biological
+samples. Integration and processing of data from such disparate
+domains underlies ``systems'' level biological research.  This leads
+to additional challenges that standard chemical structural
+representation cannot address: firstly, the representation of
+\textit{classes of chemical entities} -- since, in many cases, a class
+of compounds behave in a certain context in a biological system rather
+than a single molecule with a fully-specified structure -- and for
+similar reasons, chemical concepts such as \textit{groups} (parts of
+molecules); secondly, the need to represent \textit{non-structural}
+groupings of compounds of interest, since compounds may bind to the
+same or similar targets or they may have similar biological functions
+such as acting as an anti-neoplastic -- preventing the development of
+tumors. With the many different biological databases needing to refer
+to these different chemical entities in an organised and standardized
 fashion to enable cross-database integration for a whole-systems
-perspective, \textit{ontologies} and other semantic technologies 
-are being successfully employed to support annotation, classification and
-semantic cross-domain querying of chemical entities within
-biological systems. The most widely used ontology for biologically relevant
+perspective, \textit{ontologies} and other semantic technologies are
+being successfully employed to support annotation, classification and
+semantic cross-domain querying of chemical entities within biological
+systems. The most widely used ontology for biologically relevant
 chemical entities is ChEBI \cite{chebi2010}, containing around 26,500
@@
 being actively used for annotation in dozens of biological databases.
-Ontologies are based on logical languages of 
-varying levels of expressivity, accompanied by sophisticated reasoning
-algorithms. They can be used to construct various sorts of semantic
-similarity (e.g. similarity in function, similarity in application) 
-which complement traditional structural similarity methods \cite{couto2010}.
-An open challenge is integrating graph-based chemical
-structure information with the wider logic-based ontology information, which 
-will allow combined automated reasoning and querying over the different domains 
-encoded in a growing number of interlinked ontologies.% \cite{hastingsowled2010}. 
-Enabling automated inferencing in such heterogenous linked systems brings us closer to
-achieving a systems level understanding of bioactive chemistry \cite{Oprea:2007fk}.
+Ontologies are based on logical languages of varying levels of
+expressivity, accompanied by sophisticated reasoning algorithms. They
+can be used to construct various sorts of semantic similarity
+(e.g. similarity in function, similarity in application) which
+complement traditional structural similarity methods \cite{couto2010}.
+An open challenge is integrating graph-based chemical structure
+information with the wider logic-based ontology information, which
+will allow combined automated reasoning and querying over the
+different domains encoded in a growing number of interlinked
+ontologies. % \cite{hastingsowled2010}.
+Enabling automated inferencing in such heterogenous linked systems
+brings us closer to achieving a systems level understanding of small
+molecule activity \cite{Oprea:2007fk}.
 
@@
 licenses) and Open Source.
-% [JH comment]
-% we should standardize our nomenclature for the phrase 'Open Source' in the
-% document. Here it is capitalized, earlier it has been uncapitalized 
-% ('open source'), and below it is hyphenated! (open-source)
-% [JH comment]
 
@@
 
-The extent of open-source software in cheminformatics has somewhat
+The extent of Open Source software in cheminformatics has somewhat
 lagged behind related fields such as bioinformatics, partly due to the
@@
 software. However, in recent years there has been a rise in the amount
-of open-source cheminformatics software. One contribution to this has
+of Open Source cheminformatics software. One contribution to this has
 been the Blue Obelisk group~\cite{BlueObelisk2011}
 (\url{http://blueobelisk.org}) of chemists, who have worked together
-to create interoperable open-source chemistry software (among other
-activities). Now there are a number of Open-Source chemical toolkits
+to create interoperable Open Source chemistry software (among other
+activities). Now there are a number of Open Source chemical toolkits
 available including the Chemistry Development Kit
@@
 formats. Table~\ref{tab:features} (in the Appendix) compares features
-offered by the open-source toolkits currently available.  Given that
+offered by the Open Source toolkits currently available.  Given that
 there is ample scope for software engineering (algorithm
@@
 
-Certain challenges faced by nearly all
-cheminformatics toolkits stem from the graph representation of
-chemistry. This representation, while succinct and amenable to
-computation, is only an approximation of reality -- and edge cases
-abound. Existing areas of interest are enumeration of colored graphs,
-taking into account symmetry, and symmetry detection itself, which not
-only derives from the chemical graph, but typically also from the 3D
-geometry of the chemical. It is important to realize that although
-chemical representations have increased in sophistication over the
-years, many chiral molecules or non-standard bonding cases such as 
-organometallic compounds cannot be satisfactorily handled by
-current representation systems. A more challenging aspect is
-that the chemical representation of a molecule really needs to capture
-non-static graphs, in order to take into account delocalisation and
-the tautomeric phenomena molecules undergo in practice.
+Certain challenges faced by nearly all cheminformatics toolkits stem
+from the graph representation of chemistry. This representation, while
+succinct and amenable to computation, is only an approximation of
+reality -- and edge cases abound. Existing areas of interest are
+enumeration of colored graphs, taking into account symmetry, and
+symmetry detection itself, which not only derives from the chemical
+graph, but typically also from the 3D geometry of the chemical. It is
+important to realize that although chemical representations have
+increased in sophistication over the years, many chiral molecules or
+non-standard bonding cases such as organometallic compounds cannot be
+satisfactorily handled by current representation systems. A more
+challenging aspect is that the chemical representation of a molecule
+really needs to capture non-static graphs, in order to take into
+account delocalisation and the tautomeric phenomena molecules undergo
+in practice.
 
@@
 data and tools have been freely available since the inception of the
-field. This disparity could be attributed to the fact that 
-the outcomes from cheminformatics methods
-and software have a more direct impact on profits, in terms of
-identifying lead-like compounds and improving the properties of drug
-candidates, while bioinformatics software is employed in
-upstream areas such as target identification, 
-perhaps not as directly related to
-possible profits as an actual small molecule.
-
-%Furthermore, chemical data (structure, activity information etc.) has
-%been harder to acquire and when done on large scales can be extremely
-%intensive in terms of time and effort, whereas bioinformatics data (sequences for
-%example) are much easier to obtain. 
-While both fields have theoretical
-components, it is possible that the free availability of large amounts
-of data in bioinformatics drove the development of publicly
-available tools to handle it. In contrast, the proprietary nature of
-chemical data would imply that tools to process and analyze that data
-were of interest primarily to the owners. As a result, the incentive
-to make cheminformatics software public has been low and since the
-users of it were primarily industrial, commercial software was the
-norm.
+field. This disparity could be attributed to the fact that the
+outcomes from cheminformatics methods and software have a more direct
+impact on profits, in terms of identifying lead-like compounds and
+improving the properties of drug candidates, while bioinformatics
+software is employed in upstream areas such as target identification,
+perhaps not as directly related to possible profits as an actual small
+molecule. Furthermore, chemical data (structure, activity information
+etc.) has been harder to acquire and when done on large scales can be
+extremely intensive in terms of time and effort, whereas
+bioinformatics data (e.g., sequences) are much easier to obtain.
+
+While both fields have theoretical components, it is possible that the
+free availability of large amounts of data in bioinformatics drove the
+development of publicly available tools to handle it. In contrast, the
+proprietary nature of chemical data would imply that tools to process
+and analyze that data were of interest primarily to the owners. As a
+result, the incentive to make cheminformatics software public has been
+low and since the users of it were primarily industrial, commercial
+software was the norm.
 
@@
 As we enter the second decade of the 21st century, the cheminformatics
-landscape is rapidly shifting.  There now exist free-of-charge
-open-source cheminformatics toolkits, applications and open databases
-with tens of millions of compounds and containing valuable
-experimental bioactivity data. Modern experimental techniques such as
-high throughput sequencing have made the generation of large amounts
-of structure-activity data much more accessible.  We believe the
+landscape is rapidly shifting.  There now exist free-of-charge Open
+Source cheminformatics toolkits, applications and open databases with
+tens of millions of compounds and containing valuable experimental
+bioactivity data. Modern experimental techniques such as high
+throughput sequencing have made the generation of large amounts of
+structure-activity data much more accessible.  We believe the
 prospects for academic and industrial collaboration between chemists
-and computer scientists are bright, and we would like to encourage
-more computer scientists to participate in cheminformatics research.
+and computer scientists are bright, and encourage more computer
+scientists to participate in cheminformatics research.
 
@@
 to the underlying chemistry of the systems being studied. Indeed, many
-issues faced by chemists are often detailed and subtle and do not
-admit as clean an abstraction as ``algorithms on a string'', the way
-many bioinformatics algorithms can be abstracted into a computer
-science framework.  As a result, while one can certainly make
-significant contributions to cheminformatics by just considering
-structures as graphs, one is limited to somewhat abstract
-problems. Keeping in mind that cheminformatics is fundamentally a
-practical field that serves to help and advance experimental
-chemistry, the key challenges require an understanding of the
-underlying chemical systems. 
-%One could certainly study for a degree in chemistry. Alternatively one could collaborate with a chemist or cheminformatics practitioner.
+issues faced by chemists do not admit as clean an abstraction as
+``algorithms on a string'', the way many bioinformatics algorithms can
+be abstracted into a computer science framework.  As a result, while
+one can certainly make significant contributions to cheminformatics by
+just considering structures as graphs, one is limited to somewhat
+abstract problems. Keeping in mind that cheminformatics is
+fundamentally a practical field that serves to help and advance
+experimental chemistry, the key challenges require an understanding of
+the underlying chemical systems.
 
@@
 and ontologies; for a practical programmer, expansion of the many
-open-source cheminformatics projects.  If more chemists think
+Open Source cheminformatics projects.  If more chemists think
 algorithmically, and more computer scientists think chemically we will be
@@
   \end{tabular}
-  \caption{Cheminformatics features provided by open-source toolkits.}
+  \caption{Cheminformatics features provided by Open Source toolkits.}
   \label{tab:features}

### 2011-11-21 OpenQuestions.tex
Added some more comments and fixed typos.

@@
 
-\item \emph{Inverse QSAR}
+\item \emph{Inverse QSAR (or de-novo design)}
 
@@
 its structural features and those of a training set. The inverse QSAR
-problem is such that given a molecule structure representation
+problem, also de-novo design, is such that given a molecule structure representation
 (usually a descriptor vector), what are the possible input structures
-that satisfy the representation. This can be made more complex, by
-asking that the input structures also satisfy aexperimental property
+that satisfy the representation. This process might be considered as
+a chemical design of experiments. It is clearly a non-continuous optimization problem, since
+not all chemical molecules might be accessible, and cost/risk to create molecules phsically is another
+critical factor. This can be made more complex, by
+asking that the input structures also satisfy an experimental property
 range.
@@
 \item What methodologies can be devised to address this problem?
+\item How can we ensure combinatorically created suggestions make use of the large 
+chemical structure, chemical building block, and chemical reaction databases by suggesting
+chemically feasible molecules (not just virtual accessible ones)? How can we improve 
+synthetic accessibility predictions \cite{Boda_Seidel_Gasteiger_2007}?
 \item How does a given methodology allow us to translate the
@@
   benchmarking studies are urgently required, e.g. using GPUs
-  \cite{doi:10.1021/ci1004948}, optimizing pair-wise similarity
+  \cite{doi:10.1021/ci1004948}, or optimizing pair-wise similarity
   calculations \cite{MINF:MINF201100050}.
@@
   key problem is that non-vectorial encodings require to compare all
-  molecules (or their 3D conformational eplosions) in a pair-wise
+  molecules (or their 3D conformational explosions) in a pair-wise
   manner.  This becomes prohibitively expensive when considering

### 2011-11-21 rebuttal.tex

@@
 \response{We have updated the text on page 1 to be more explicit in
-  terms of what annotations are and provide a concete example of the
+  terms of what annotations are and provide a concrete example of the
   sizes involved.}
@@
   technologies can play in this area. We have, however, chosen to not
-  address the issue of translational medicine. While there have been
-  some cases where cheminformatics have played a role in translational
-  medicine, the bulk of cheminformatics lies upstream of this
-  topic. Hence, to keep the article focused we have chosen to not
-  include this topic.}
+  address the issue of translational medicine in the main article. 
+  Still, we have added a paragraph plus references to the historic 
+  supplement.}
 
@@
   tried-and-tested methods in an over-arching review such as this.}
-\response{While it is true that proteochemometric methodologies are
-  not as well known as say docking or pharmacophore modeling, we
-  choose to include them since they are an example of a (statistical)
-  QSAR technique that does explicitly include ligand and receptor features}
+\response{It is true that proteochemometric methodologies
+  might not be as well known as say docking or pharmacophore modeling.
+  Though, what is considered as proteochemometric, is in docking known as 
+  'interaction fingerprint', e.g. for Kinase databases, and in the field of HIV the
+  term 'phenotypic modeling' has a long tradition.
+  We choose, along with multiple reviews, that the term proteochemometrics
+  is worth being mentioned, since it unifies the (statistical)
+  mining/QSAR techniques that do explicitly include multiple ligand 
+  activities and multiple receptor features.\\
+  Finally, we also agree with reviewer number one that proteochemometric modeling
+  plays an important role in translational medicine, e.g. HIV-HAART theraphy. 
+  A parapgraph and references about translational medicine was added to the 
+  historic supplment.}
 
@@
   century, but that freely accessible toolkits implementing them are
-  relatively recent}
+  relatively recent, as well as public and open benchmark data sets.}
 

### 2011-11-23 rebuttal.tex
Fix line endings... someone has the incorrect settings...see http://help.github.com/line-endings/
Mainly trivial edits to rebuttal. Addings periods, fixing spellings, and so forth.

@@
   milestones. Furthermore, we have updated the text to note commercial
-  vendors and solutions; but we still stress on Open Source and Open
-  Access solutions to allow reasders to explore the problems and
-  topics we have discussed with a minimum of hindrances}
+  vendors and solutions; but we still emphasise Open Source and Open
+  Access solutions as these allow readers to explore the problems and
+  topics we have discussed with a minimum of hindrances.}
 
@@
   plays an important role in translational medicine, e.g. HIV-HAART theraphy. 
-  A parapgraph and references about translational medicine was added to the 
-  historic supplment.}
+  A paragraph and references about translational medicine was added to the 
+  historic supplement.}
 
@@
 \response{We agree that this toolkit should have been included. We
-  have updated the text on page 7 to take note of RDKit, though we
-  still primarily focus on the CDK and OpenBabel. However, Table 2
-  includes RDKit in the feature comparison}
+  have updated the text on page 7 as well as Table 2 to take note of RDKit, 
+  along with the newer Indigo toolkit.}
 
@@
   most certainly be aware}
-\response{The text has been updated}
+\response{The text has been updated.}
 
@@
   technique. We have also provide relevant references to key topics in
-  this area}
+  this area.}
 
@@
   purposes. The text on page 2 has been updated to better represent
-  this view}
+  this view.}
 
@@
 \response{We have updated the discussion of descriptors to be more
-  focused}
+  focused.}
 
@@
   than bioinformatics. Primarily, because there is no obvious
-  reference that claims this. But alse because, depending on what one
+  reference that claims this. But also because, depending on what one
   considers bioinformatics and cheminformatics, one field can be
   identified as being older or younger than the other. We feel that
-  claims on age do not strengthen the paper and have thus removed it}
+  claims on age do not strengthen the paper and have thus removed it.}
 
@@
 \response{We have the updated the text on page 1 to attenuate this
-  statement, stressing on the fact that much of the data in
+  statement, stressing the fact that much of the data in
   cheminformatics is proprietary, whereas only some of the techniques (such
@@
   supplemental history document, we make a distinction between
-  computational chemistry and cheminformatics}
+  computational chemistry and cheminformatics.}
 
@@
 \response{The text has been updated to use the term ``therapeutic
-  molecule''}
+  molecule''.}
 
@@
 \response{The text has been updated to provide some examples of
-  properties that would characterize a therapeutically useful small molecule}
+  properties that would characterize a therapeutically useful small molecule.}
 
@@
   fingerprints. We have also mentioned the use of fingerprints in
-  similarity searches and the Tanimoto coefficient}
+  similarity searches and the Tanimoto coefficient.}
 
@@
 \comment{P4, c2, l17: citation needed on the size of chemistry space.}
-\response{The appropriate reference has been added}
+\response{The appropriate reference has been added.}
 
@@
   mentioned previously in this manuscript.}
-\response{The text has been updated and simplified}
+\response{The text has been updated and simplified.}
 
@@
   molecules, which is nearly a billion, not a trillion.}
-\response{The text has been updated to use the correct number}
+\response{The text has been updated to use the correct number.}
 
@@
   the similarity property principle and also include the relevant
-  reference}
+  reference.}
 
@@
 this is perhaps the most widely applied method in the field.}
-\response{We have included the Na\"{i}ve Bayes as well}
+\response{We have included the Na\"{i}ve Bayes as well.}
 
@@
   required here.}
-\response{The relevant reference has been added}
+\response{The relevant reference has been added.}
 
@@
 related targets. We admit that the answer is not clear at this point
-and hence suggest that this could a topic for research }
+and hence suggest that this could a topic for research. }
 
@@
   inclusion of this reference would be somewhat irrelevant. We have
-  included this reference in the supplemental history}
+  included this reference in the supplemental history.}
 
@@
   algorithmically? I would prefer inverted commas for neither.}
-\response{Quotes have been removed}
+\response{Quotes have been removed.}
 
 \comment{P8, c1, l10: ELN's should be ELNs.}
-\response{Corrected}
+\response{Corrected.}
 
@@
   source/open data, because this ensures that a reader will be able to
-  explore cheminformatics problems with a minimum of hindrances}
+  explore cheminformatics problems with a minimum of hindrances.}
 
@@
   of cheminformatics tools and data have been commercial, in contrast
-  to freely available tools and data in bioinformatics}
+  to freely available tools and data in bioinformatics.}
 

### 2011-11-23 paper.tex
Fix line endings... someone has the incorrect settings...see http://help.github.com/line-endings/

@@
   \begin{tabular}{lcccc}  \hline
-    \textbf{Feature} & \textbf{CDK} & \textbf{OpenBabel} &
-    \textbf{RDKit} & \textbf{Indigo}\textsuperscript{1} \tabularnewline \hline
+    \textbf{Feature} & \textbf{CDK} & \textbf{Open Babel} &
+    \textbf{RDKit} & \textbf{Indigo} \tabularnewline \hline
     License & LGPL & GPL & new BSD & GPL or commercial \tabularnewline
-    Language & Java & C++ & C++ & C++ / Python \tabularnewline
-    SLOC\textsuperscript{2} & 188,554 & 194,358 & 173,219 & ? \tabularnewline
-    Structure-based Fingerprints & \lots\textsuperscript{3} & \lots & \lots & yes \tabularnewline
+    Language & Java & C++ & C++/Python & C/C++ \tabularnewline
+    SLOC\textsuperscript{1} & 188,554 & 194,358 & 173,219 & 144,080 \tabularnewline
+    Structure-based fingerprints & \lots\textsuperscript{2} & \lots & \lots & \lots \tabularnewline
     %\ \ \ Substructure & \lots & \lots & \lots \tabularnewline
-    Support for multiple chemical file formats & \little & \lots & \least & \little \tabularnewline
+    Support for multiple chemical file formats & \little & \lots & \least & \least \tabularnewline
     %Aromaticity models & \least & \least & \least \tabularnewline
     %Stereochemistry & \least & \little & \lots \tabularnewline
-    Canonical labeling of graphs & \lots & \lots & \lots & yes \tabularnewline
-    Calculation of molecular descriptors & \lots & \least & \lots & no \tabularnewline
+    Canonical labeling of graphs & \lots & \lots & \lots & \lots \tabularnewline
+    Calculation of molecular descriptors & \lots & \least & \lots & \none \tabularnewline
     %2D coordinate generation & \lots & \none & \lots & no \tabularnewline
-    3D coordinate generation & \least & \lots & \lots & no \tabularnewline
+    3D coordinate generation & \least & \lots & \lots & \none \tabularnewline
     2D depictions & \lots & \little & \lots & \little \tabularnewline
-    Generation of conformational isomers & \none & \least & \least & no \tabularnewline
+    Generation of conformational isomers & \none & \least & \least & \none \tabularnewline
     %Rigid alignment & \lots & \lots & \lots \tabularnewline
-    SMARTS searching/matching & \lots & \lots & \lots & yes \tabularnewline
+    SMARTS searching/matching & \lots & \lots & \lots & \lots \tabularnewline
     %Pharmacophore searching & \little & \none & \lots \tabularnewline
-    Chemical reaction enumeration & no & no & no & yes \tabularnewline
+    Chemical reaction enumeration & \none & \none & \none & \lots \tabularnewline
     \hline
-    \multicolumn{5}{l}{1. Indigo is a new toolkit (published late 2010), and we have not investigated its features in depth.} \\
-    \multicolumn{5}{l}{2. Source Lines of Code as measured by the tool \emph{sloccount} (\url{http://www.dwheeler.com/sloccount/}).} \\
+    \multicolumn{5}{l}{1. Source Lines of Code as measured by the tool \emph{sloccount} (\url{http://www.dwheeler.com/sloccount/}).} \\
     \multicolumn{5}{l}{The count includes all source files, in any language, for the
   project. In addition, only the trunk} \\
-    \multicolumn{5}{l}{for each project was considered.  We did not count Indigo's lines of code.} \\
-    \multicolumn{5}{l}{3. More checks indicates more complete implementation of a feature.} \\
+    \multicolumn{5}{l}{for each project was considered.} \\
+    \multicolumn{5}{l}{2. More checks indicates more complete implementation of a feature.} \\
   \end{tabular}

### 2011-11-24 rebuttal.tex

@@
   term 'phenotypic modeling' has a long tradition.
-  We choose, along with multiple reviews, that the term proteochemometrics
-  is worth being mentioned, since it unifies the (statistical)
-  mining/QSAR techniques that do explicitly include multiple ligand 
+  We choose, along with multiple reviews and the books written by 
+  Qing Yan (Pharmacogenomics) and Kubinyi/M\"{u}ller (Chemogenomics), 
+  that the term proteochemometrics is worth being mentioned, since it unifies the 
+  (statistical) mining/QSAR techniques that do explicitly include multiple ligand 
   activities and multiple receptor features.\\
   Finally, we also agree with reviewer number one that proteochemometric modeling
-  plays an important role in translational medicine, e.g. HIV-HAART theraphy. 
+  plays an important role in translational medicine, e.g. HIV-HAART therapy. 
   A paragraph and references about translational medicine was added to the 

### 2011-11-28 OpenQuestions.tex
included JVDs suggestions

@@
 
-This question encompasses the open questions of \emph{approximately counting the number of stereoisomers, or the number of tautomers, of a chemical formula}.
-Goldberg and Jerrum designed an algorithm which, given a chemical formula, would output an isomer of that formula chosen uniformly at random~\cite{RandomlySampling}.
-Perhaps the main theoretical obstacle to overcome is that molecules tend to be \emph{chiral} (that is, they have a ``handedness'' or 3D orientation), whereas traditional graph theory (and the Goldberg/Jerrum algorithm) treats graphs with identical vertices and edges as isomorphic.
-So part of the question could be rephrased as, ``Given a labeled vertex set, count the number of structures that are identical on that vertex set, except that they differ in their 3D orientation.''  Mathematical chemists have designed measures of molecular chirality~\cite{ChiralityMeasures}.
-A resolution of this problem may require connecting graph enumeration algorithms to knot theory or other topics in topology~\cite{TopologicalLook}.
+  This question encompasses the open questions of \emph{approximately
+    counting the number of stereoisomers, or the number of tautomers,
+    of a chemical formula}.  Goldberg and Jerrum designed an algorithm
+  which, given a chemical formula, would output an isomer of that
+  formula chosen uniformly at random~\cite{RandomlySampling}.  Perhaps
+  the main theoretical obstacle to overcome is that molecules tend to
+  be \emph{chiral} (that is, they have a ``handedness'' or 3D
+  orientation), whereas traditional graph theory (and the
+  Goldberg/Jerrum algorithm) treats graphs with identical vertices and
+  edges as isomorphic.  So part of the question could be rephrased as,
+  ``Given a labeled vertex set, count the number of structures that
+  are identical on that vertex set, except that they differ in their
+  3D orientation.''  Mathematical chemists have designed measures of
+  molecular chirality~\cite{ChiralityMeasures}. A resolution of this
+  problem may require connecting graph enumeration algorithms to knot
+  theory or other topics in topology~\cite{TopologicalLook}.
 %
-\item \emph{Write code that approximately counts the number of conformers of a chemical formula.}
+\item \emph{Design and implement efficient subgraph isomorphism
+    algorithms for useful special cases of molecular graphs}.
 
-While this is related to the previous question, it is conceivable that one could answer this question without solving the theoretical problem.
-For example, as we note in the main article, the commercial software MOLGEN can approximately enumerate structures that satisfy given spectral data, even though related theoretical enumeration problems appear wide open.
-%
-\item \emph{Design and implement efficient subgraph isomorphism algorithms for useful special cases of molecular graphs}.
-
-The Subgraph Isomorphism Problem is known to be $\textsf{NP}$-complete (hence possibly harder than the Graph Isomorphism Problem, which is not known to be $\textsf{NP}$-complete).
-Nevertheless, finding maximum common subgraphs to match chemical structures is of fundamental importance in chemistry; hence, much work has been invested in partial solutions to this problem.
-Raymond and Willet reviewed the state of the art in 2002~\cite{MCSreview}.
-As one possible way to attack this problem, we note the empirical fact that molecular graphs are of bounded degree, and are observed to have \emph{treewidth} $\leq 5$~\cite{treewidth}.
-Bounded treewidth is at least theoretically useful~\cite{Epp-JGAA-99}, and it may be possible to improve on current open-source implementations whose isomorphism-checking routines are written for all graphs, instead of taking advantage of special properties of chemical graphs.
+  The Subgraph Isomorphism Problem is known to be
+  $\textsf{NP}$-complete (hence possibly harder than the Graph
+  Isomorphism Problem, which is not known to be
+  $\textsf{NP}$-complete).  Nevertheless, finding maximum common
+  subgraphs to match chemical structures is of fundamental importance
+  in chemistry; hence, much work has been invested in partial
+  solutions to this problem.  Raymond and Willet reviewed the state of
+  the art in 2002~\cite{MCSreview}.  As one possible way to attack
+  this problem, we note the empirical fact that molecular graphs are
+  of bounded degree, and are observed to have \emph{treewidth} $\leq
+  5$~\cite{treewidth}.  Bounded treewidth is at least theoretically
+  useful~\cite{Epp-JGAA-99}, and it may be possible to improve on
+  current open-source implementations whose isomorphism-checking
+  routines are written for all graphs, instead of taking advantage of
+  special properties of chemical graphs.
 
@@
 
-One key concept of the linked data web, the semantic web, is that different data sources can be readily integrated with each other. Still, in the field
-of Cheminformatics, we are not only interest in linking two molecules 
-(the linking normalization problem for different protomers, tautomers, or special cases of isomerisms remain open), but we
-are also interested in being able to search efficiently within molecules when being linked via semantic web approaches. Typical
-searches will require being able to apply substructure or similarity searches.
-What could be algorithmic solutions for this?
+  One key concept of the linked data web, the semantic web, is that
+  different data sources can be readily integrated with each
+  other. Still, in the field of Cheminformatics, we are not only
+  interest in linking two molecules (the linking normalization problem
+  for different protomers, tautomers, or special cases of isomerisms
+  remain open), but we are also interested in being able to search
+  efficiently within molecules when being linked via semantic web
+  approaches. Typical searches will require being able to apply
+  substructure or similarity searches.  What could be algorithmic
+  solutions for this?
 \end{enumerate}
@@
 \begin{enumerate}
+
+\item \emph{Real time substructure searching in massive chemical
+    databases}.
+
+  Chemical databases have grown tremendously in size. A common task in
+  such databases is substructure searching. However most cases of very
+  large structure databases do not support substructure in real time
+  (something that would allow applications such as
+  ``type-ahead''). How can one enable such rapid substructure searches
+  on massive ($> 10^9$ molecules) collections of structures? This
+  problem has various aspects:
+  \begin{itemize}
+  \item What type of indexing schemes will support rapid substructure searches?
+  \item How can other molecular properties be included in substructure
+    searchs that allow ranking of results, all the while maintaining
+    real-time response (i.e. $< 1$ sec)?
+  \item What type of database architectures are required for this
+    scale of structure searching?
+  \end{itemize}
+
 \item \emph{Large scale conformational databases}.

### 2011-11-28 supplement_history.tex
Update rebuttal. Added ref and updated history
Added some more comments and fixed typos.
Adding book references for Chemo-/Pharmacogenomics and relation to translational medicine.
included JVDs suggestions

@@
 overview; rather, we point the reader to more detailed references and
-to highlight some historic milestones.  In addition, we also discuss
+highlight some historic milestones.  In addition, we also discuss
 some terms as they relate to cheminformatics.  It is thus useful to
@@
     scientific discipline.}'' Johann Gasteiger, 2002
+\item ``\textit{The set of approaches to computer-aided drug design
+    that do not rely on the 3D structure of the [protein] target}''
+  John Van Drie, 2011 (personal communication)
 \end{itemize}
@@
   broadly such as in drug repurposing
-  efforts \cite{Dudley:2011fk,Swamidass:2011uq}.
+  efforts \cite{Dudley:2011fk,Swamidass:2011uq}. Related topics and
+  books are also Pharmacogenomics \cite{yan2008pharmacogenomics} and
+  Chemogenomics \cite{kubinyi2004chemogenomics}.
 \end{itemize}
@@
 first chemical graphs occured in publications of Brown in 1864 and
-Cayley in 1874 \cite{bonchevrouvrey1991,brown2009}.
+Cayley in 1874 \cite{bonchevrouvrey1991,brown2009}. The notion of
+mathematical chemistry in general was discussed by Helm in 1897
+\cite{Helm:1897ys} and the reader is referred to Balaban
+\cite{Balaban:2005zr} for a historical overview of this field.
 

### 2012-02-06 resubmit/cheminformatics_CACM_revision.tex
added resubmission to new dir;

--- 2011-11-23 paper.tex
@@
 
-\title{Cheminformatics: The Computer Science of Chemical Discovery}
+\title{Cheminformatics, the Computer Science of Chemical Discovery, Turning Open-Source}
 \numberofauthors{9}
@@
 Rajarshi Guha\\
-\affaddr{NIH Center for Translational Therapeutics}\\
+\affaddr{NIH Center for Advancing Translational Science}\\
 \affaddr{9800 Medical Center Drive}\\
@@
   discovery of novel drug candidates to the optimization of certain
-  physicochemical properties of small molecules. Until recently, much
-  of the data and some of the techniques employed in the
-  cheminformatics field have been closely guarded secrets of companies
-  whose financial success depended on being the first to produce
-  new therapeutic molecules.  Only within the last decade -- as an
+  physicochemical properties of small molecules.
+While researchers in both academia and industry have made important contributions to this field since the 1960s (if not before), new and exciting research opportunities now exist because of an ``opening'' of data and software that began around the year 2000.
+Since that time -- as an
   effect of a change in mindsets, government-mandated policy changes,
   and, not least, as a result of chemists volunteering their time
-  for an Open Science ``movement'' -- have researchers gained access to
+  for an Open Science ``movement'' -- researchers have gained access to
   freely available software packages and databases of tens of millions
   of chemicals. Academic chemists now confront a variety of unsolved
-  algorithmic problems that could not have been tackled a decade ago,
+  algorithmic problems that might have been prohibitive to tackle a decade ago,
   but whose solutions are critical to research ranging from
@@
 
-Novel technologies in the life sciences churn out information at an
+Novel technologies in the life sciences produce information at an
 ever increasing rate -- public data stores such as that of the
@@
 molecules, techniques and algorithms to handle, analyse and interpret
-chemical structure information are a core component of the drug
-discovery toolbox. Public availability of chemical information and
+chemical structure information are a core component of the methods of drug
+discovery. Public availability of chemical information and
 tools to handle it are important as drug discovery research is slow,
@@
 much implicit information (such as unlabeled nodes represent carbons
-with valencies satisified by hydrogens) to a fully labeled graph,
+with valencies satisfied by hydrogens) to a fully labeled graph,
 where all information is explicit
@@
 labeled graph representations, to discover patterns in large-scale
-datasets \cite{horst2009}.
+datasets \cite{horst2009}; this information can also be determined by simulations such as molecular dynamics.
 
@@
   in information content from left to right and top to bottom. A
-  chemist's traditional 2D depiction (A), a labeled graph (B), a 3D
-  conformation (C) and a surface represention (D). All show differing
+  chemist's traditional 2D depiction (top left), a labeled graph (top right), a 3D
+  conformation (bottom left) and a surface representation (bottom right). All show differing
   (and valid) aspects of the same molecule and are suited for
@@
 
-Our motivation for this article is to highlight a variety of topics in
+This main article will highlight current topics in
 cheminformatics, presenting the computational and algorithmic problems
 faced in the field and how modern CS research can contribute to their
-solution.
-
-As noted above, in this article we choose to focus on cheminformatics
-in the context of drug discovery and development. Given that this is a
-slow and expensive process, the concept of ``risk minimization'' plays
-an important role in this field. Thus, we structure the article around
-this idea -- how to minimize the chances of a small
+solution.  We have also provided two appendices: a brief history of cheminformatics, and a detailed list of open problems.
+
+We structure the main article around ``risk minimization'' of drug discovery (a slow and expensive process)---how to minimize the chances of a small
 molecule failing, due to poor physical, chemical or biological
@@
 these efforts. As an example, a molecule must be soluble and show a
-certain degree of bioavailability to be considered as drug
+certain degree of bioavailability to be considered as a drug
 candidate. In cases where these properties are poor, a cheminformatics
@@
 string comparisons. The SMILES (Simplified Molecular Input Line
-Specification) format \cite{Weininger:1988kx} is an example of a
+Specification) format as defined by Weininger 1988\cite{Weininger:1988kx} is an example of a
 representation that can be canonicalized. Since the original
@@
 tumors. With the many different biological databases needing to refer
-to these different chemical entities in an organised and standardized
+to these different chemical entities in an organized and standardized
 fashion to enable cross-database integration for a whole-systems
@@
 methods to search (based on structural or property similarity) through
-the implicit enumerated space. A number of commerical vendors such as
+the implicit enumerated space. A number of commercial vendors such as
 Digitial Chemistry and ChemAxon provide toolkits to handle these
@@
 molecular structures are thus undesirable. A number of workers have
-investigated the problem of identify such one-way molecular
+investigated the problem of identifying such one-way molecular
 descriptors as well as methods to reconstruct molecules from
@@
 such as docking (which predicts how molecules fit together),
-pharmacophore modeling (a 3D approach to capturing protein-ligand
+structure-based pharmacophore modeling (a 3D approach to capturing protein-ligand
 interactions) and proteochemometric methods have been designed to take
@@
 kernel on a single graph compares measured similarities between the
-graph's nodes. In these cases, rather than compute vectorial
+graph's nodes. In these cases, rather than compute vector
 representations, we directly operate on the graph representations.
 
-Both methods have advantages and disadvantages. The vectorial approach
+Both methods have advantages and disadvantages. The vector approach
 requires one to identify a subset of \emph{relevant} (to the property
-being modeled) descriptors: the feature selection problem 
+being modeled) descriptors: the feature selection problem is
 well discussed in the data mining literature. A kernel approach
@@
 predictions, but do not really care why one molecule is active
-and another inactive. These types of models can be black boxes (neural
+and another inactive. These types of models can be black boxes (e.g., neural
 networks) and algorithmic models \cite{Breiman:2001fk} (random
@@
 ``Enterprise 2.0,'' required delivering on the promise to increase
-interconnectivity, and to improve collaboration opportunities.  The
+interconnectivity, and to improve collaboration opportunities.
+These opportunities were made legally possible in the USA because of FDA regulation Title 21 CFR Part 11, which permits the use of electronic signatures in chemical documents.
+The
 leading players in this area include CambridgeSoft's E-Notebook and
@@
 bioactivity space has to improve quality and interconnectivity to
-compare with it. Efforts like ChEMBL and PubChem-BioAssay are on the
-right track. Still, improving data quality and standards between
+compare with it. SciFinder is another chemical abstract service, with a version, SciFinder Scholar, marketed to universities.  Open-source efforts like ChEMBL and PubChem-BioAssay are on the
+right track, though, unlike some commercial tools, these efforts do not yet abstract reactions. Still, improving data quality and standards between
 public and closed sources will be absolutely critical for ensuring
@@
 languages). Such toolkits are used, for example, to read/write
-chemical file formats, manipulate chemical structures, measure
+chemical files in various formats, manipulate chemical structures, measure
 similarity of molecules, search for substructures, and generate 2D
-diagrams. The underlying algorithms used include graph algorithms
+depictions. The underlying algorithms used include graph algorithms
 (e.g. maximal common substructure, canonical labelling of graphs),
@@
 issues faced by chemists do not admit as clean an abstraction as
-``algorithms on a string'', the way many bioinformatics algorithms can
-be abstracted into a computer science framework.  As a result, while
+``algorithms on a string,'' the way many bioinformatics algorithms can
+be abstracted into a computer science framework.  Hence, while
 one can certainly make significant contributions to cheminformatics by
@@
 We would like to thank Danny Verbinnen for sharing his insights into
-ELNs.  This article was written while A. Sterling was visiting the
-Department of Electrical Engineering and Computer Science,
-Northwestern University, supported in part by NSF grant
-CCF-1049899. A. Bender thanks Unilever for funding. J. Hastings
-thanks the EU project EU-OPENSCREEN for funding.
+ELNs and John Van Drie for insights into cheminformatics.  This
+article was written while A. Sterling was visiting the Department of
+Electrical Engineering and Computer Science, Northwestern University,
+supported in part by NSF grant CCF-1049899. A. Bender thanks Unilever
+for funding. J. Hastings thanks the EU project EU-OPENSCREEN for
+funding.
 
 \bibliographystyle{abbrv}
-\bibliography{paper}
+\bibliography{resubmission}
 

### 2012-02-06 resubmit/cheminformatics_CACM_revision_appendix.tex
added resubmission to new dir;

--- 2011-11-28 OpenQuestions.tex
@@
 
-\usepackage{array}
-\usepackage{pifont}
-\usepackage{url}
-\usepackage{graphicx}
-\usepackage{multirow}
+\begin{document}
 
-\newcommand{\none}{\ding{55}}
-\newcommand{\least}{\ding{51}}
-\newcommand{\little}{\ding{51}\ding{51}}
-\newcommand{\lots}{\ding{51}\ding{51}\ding{51}}
-
-\begin{document}
-\pagenumbering{arabic}
-
-\title{Open Questions for Computer Science and Cheminformatics}
+\title{Cheminformatics: The Computer Science of Chemical Discovery: Appendix}
 \numberofauthors{9}
@@
 .}
+\date{25 June 2011}
 \maketitle
+
+\appendix
+%
+\section{History of Cheminformatics}
+%
+\subsection{Definitions and References}
+The aim of this brief review of the history of cheminformatics is to
+put the content of the main article into a broader perspective. If we
+consider all information, analysis, and \emph{in silico} optimization
+of a molecule as ``cheminformatics,'' then the field is very large,
+since a molecule plays the central role in many related
+disciplines. Clearly, this article is not meant to be an exhaustive
+overview; rather, we point the reader to more detailed references and
+highlight some historic milestones.  In addition, we also discuss
+some terms as they relate to cheminformatics.  It is thus useful to
+first provide some definitions of cheminformatics itself, given that
+they also shed some light on the founding principles of the field. For
+example, we have the following definitions:
+\begin{itemize}
+\item ``\textit{The mixing of information resources to transform data
+    into information, and information into knowledge, for the intended
+    purpose of making better decisions faster in the arena of drug
+    lead identification and optimization.}'' Frank K. Brown, 1998.
+\item ``\textit{[Chemoinformatics involves]... the computer
+    manipulation of two- or three-dimensional chemical structures and
+    excludes textual information. This distinguishes the term from
+    chemical information, largely a discipline of chemical librarians
+    and does not include the development of computational methods.}''
+  Peter Willett, 2002.
+\item ``\textit{\ldots the application of informatics to solve chemical
+    problems.}'' and ``\textit{\ldots chemoinformatics makes the point that
+    you're using one scientific discipline to understand another
+    scientific discipline.}'' Johann Gasteiger, 2002
+\item ``\textit{The set of approaches to computer-aided drug design
+    that do not rely on the 3D structure of the [protein] target}''
+  John Van Drie, 2011 (personal communication)
+\end{itemize}
+More generally, Chen \cite{Chen2006} and Brown \cite{brown2009} cited various
+definitions of pioneers in the field with Brown concluding that
+``'\textit{The differences in definitions (of the term
+  cheminformatics) are largely a result of the types of analyses that
+  particular scientists practice and no single definition is intended
+  to be all-encompassing},'' as is evident from the three definitions above.
+It is also useful to clarify some other terms and how
+cheminformatics might relate to them:
+\begin{itemize}
+\item \textbf{Quantum chemistry} (QM, where ``M'' stands for ``mechanics'') is generally associated with
+  theoretical chemistry or chemical physics. Briefly, it focuses on
+  the description of chemical systems starting from first principles
+  and is based on the Schr\"{o}dinger wave equation and frameworks
+  built on top of it. The use of quantum mechanics allows one to
+  describe molecules in terms of electron density, and hence calculate
+  molecular properties (spectra, energies, etc.) in a physically
+  accurate manner.  The results of quantum chemistry calculations are
+  used in molecular modeling (energy force field parameters) and
+  cheminformatics (atom partial charges, or full molecule properties,
+  e.g., molecule polarizabilities).
+\item \textbf{Molecular modeling} or \textbf{Computational chemistry}
+  can be considered an approximation of quantum chemistry and aims to
+  evaluate many of the properties considered by QM methods. The key
+  difference is that these methods employ a variety of
+  parameterizations (usually, but not always, derived from QM studies),
+  allowing one to evaluate energies for small molecules very rapidly
+  as well as to handle larger systems (such as proteins) that are too
+  time consuming for QM methods. A variety of techniques including
+  semi-empirical methods, molecular mechanics and docking belong to
+  this category. While it is true that many of these methods involve
+  cheminformatics concepts (choice of tautomer, partial charge
+  assignments, etc.), the methodologies are fundamentally physical in
+  nature. We note that computational chemistry has close ties to
+  topics such as x-ray crystallography and NMR structure elucidation.
+\item \textbf{Chemometrics} is a focused application of statistical
+  methods to problems in analytical chemistry, such as infrared, mass
+  and NMR spectroscopy.
+\item \textbf{Translational research/medicine} aims to make more
+  direct links between academic (or early stage) research and clinical
+  practice in an effort to speed up the results of research into
+  actual treatments. When applied to medical research this usually
+  implies efforts to speed up the process by which a compounds goes
+  from the lead stage to the clinical trial stage. Naturally,
+  cheminformatics plays a role at various stages of this process (lead
+  identification, lead optimization, ADMET modeling) as well as more
+  broadly such as in drug repurposing
+  efforts \cite{Dudley:2011fk,Swamidass:2011uq}. Related topics and
+  books are also Pharmacogenomics \cite{yan2008pharmacogenomics} and
+  Chemogenomics \cite{kubinyi2004chemogenomics}.
+\end{itemize}
+Obviously, we cannot cover all domains that involve or refer to
+cheminformatics concepts. However, there are a number of publications
+that cover cheminformatics topics in different ways.  Probably the
+most comprehensive cheminformatic treatise was published by Gasteiger
+\cite{Gasteiger2003} in 2003.  More concise, introductory texts
+followed from Gasteiger \& Engel \cite{gasteigerengel2003} and Leach
+\& Gillet \cite{leachgillet2007}.  The books of Bajorath
+\cite{Bajorath2004} and Oprea \cite{oprea2005} focus on the
+applications of cheminformatics in drug design and span a number of
+topics. The first books discussing chemical graph theory and
+algorithms were published in 1989 by Zupan \cite{zupan1989}, 1991 by
+Bonchev and Rouvrey \cite{bonchevrouvrey1991,bonchevrouvrey2003} and
+1992 by Trinastic \cite{Trinajstic1992}.  More recently, Faulon \&
+Bender \cite{faulon2010} have edited a collection focusing on
+algorithms in cheminformatics, including various applications of graph
+theory.  Other books focus more on specific topcs such as mathematical
+challenges \cite{mathchallenges1995}, molecular diversity
+\cite{moleculardiversity1999}, factor analysis \cite{Malinowski2002},
+evolutionary algorithms \cite{clark2000} and molecular descriptors
+\cite{todeschini2000}.
+
+\subsection{Historical Milestones}
+In this section we highlight the development of a variety of
+cheminformatics concepts and techniques over time. The main journal for this field
+(\textit{Journal of Chemical Documentation} was founded in
+1961. Interestingly, the journal was renamed to \textit{Journal of
+  Chemical Information and Computer Sciences} in 1975 to reflect the
+tight connection between chemical information and computer
+science. The term
+``chemoinformatics'' was coined by Brown in
+1998\cite{brown1998}.
+Many authors, including us in this article, drop the ``o'' and use the word ``cheminformatics''; the words can be used interchangeably.
+The Chemical Abstract Service (CAS) of the American Chemical
+Society (ACS) formed a research department in 1955 and starting from
+1965 they provided their Chemical Registry System, and in 1968 they
+made the first computer-readable file of all abstracted documents
+available \cite{Chen2006}.
+
+From a more technical viewpoint, the first chemical graphs were drawn
+by the Scottish chemist Willam Cullen in 1758
+\cite{bonchevrouvrey1991}, who initially called them affinity graphs.  Many fundamental principles of graph theory and
+combinatorics were developed in the context of counting isomers of
+paraffin by Cayley~\cite{CayleyTrees}, Polya~\cite{PolyaEnumeration} and others.
+After developing the concept of bonds between atoms (Couper, 1858) the
+first chemical graphs occured in publications of Brown in 1864 and
+Cayley in 1874 \cite{bonchevrouvrey1991,brown2009}. The notion of
+mathematical chemistry in general was discussed by Helm in 1897
+\cite{Helm:1897ys} and the reader is referred to Balaban
+\cite{Balaban:2005zr} for a historical overview of this field.
+
+The year 1946 may be regarded as the birth year of cheminformatics
+\cite{Chen2006}: \textit{``In 1946 King et al.\cite{kct1946} published
+  an article illustrating the use of IBM's business accounting
+  machines in carrying out the construction of the rotational spectra
+  of asymmetric rotors by the evaluation of mathematical equations for
+  line position and line intensity''}. One could argue that this is
+more of a computational chemistry application.
+
+From an informatics perspective, the first record of managing chemical
+information goes back to the chemical literature being indexed since
+the year 1771.  In 1881 the first edition of the \textit{Beilstein's
+  Handbuch der Organischen Chemie} encyclopedia was published
+\cite{polanski2009} registering 1500 chemical compounds.  Many key
+cheminformatics algorithms appeared in the early to mid 20th century
+(though in some cases, the underlying mathematics was known much
+earlier). Ray and Kirsch published in 1957 the first substructure
+searching algorithm to support retrieval of computerized structure
+records \cite{RayKirsch1957}. Subsequent developments improved
+substructure searches, notably the use of pre-screening via fragments
+\cite{Adamson:1973fk,Feldman:1975uq} in the 1970's. One of the key
+algorithms for molecular graph canonicalization is the Morgan
+algorithm \cite{Morgan1965}, described in 1965.
+
+In 1965 the DENDRAL expert system started with the aim to
+automatically determine the structure of an unknown chemical compound
+from the corresponding mass spectrum \cite{Gray1986}. The system is
+also often cited as one of the earliest artificial intelligence
+and expert systems \cite{Chen2006}.
+
+The first system supporting chemical synthesis planning was OCSS
+(Organic Chemical Simulation of Syntheses) developed by Corey and
+Wipke in 1969 \cite{CoreyWipke1969} which in turn was an
+implementation of theories described by Vleduts in 1963
+\cite{Vleduts:1963kx}.
+
+QSAR is one of the most well known applications of predictive
+cheminformatics and originates with the work of Hansch
+\cite{Hansch:1962vn} in 1962. While Hansch's approach focused on
+property-activity relationships, true structure-activity relationships
+originated with the work of Free and Wilson in 1964
+\cite{Free:1964ys}. Subsequent work in this area led to 3D-QSAR
+techniques including CoMFA \cite{Cramer:1988zr} and
+CoMSIA\cite{Klebe:1994ly}.
+
+In 1978 Gasteiger and Marsili published a fast algorithm to calculate
+partial charges in organic molecules by a Partial Equalization of
+Orbital Electronegativities (PEOE) \cite{gm78}, which even today
+is considered one of the gold-standards to calculate partial charges of atoms.
+
+Instead of converting molecular graphs to a vector representation for
+applying machine learning methods, it is also possible to use direct
+molecular graph mining methods \cite{okada2006}. Some of the first
+published methods use neural networks \cite{kireev1995}, inductive
+logic programming \cite{yh02a}, or graph kernels \cite{kti03}.
+%
+\section{Open Questions for Computer Science and Cheminformatics}
 %
@@
 
-\section*{Algorithmic graph theory}
+\subsection*{Algorithmic graph theory}
 \begin{enumerate}
@@
 
-\section*{Cryptography}
-One-way molecular featurization (???)
+\subsection*{Cryptography}
 
-\section*{Data mining \& machine learning}
+Currently, companies simply cannot submit any chemical queries over web services, because it is not possible to ensure the web-service is not keeping a log, and not possible to prevent the service from accidently publishing a compound after five years of research (thus killing a drug from entering the market).
+There are some papers that address cryptographic problems in chemistry (for example,~\cite{Digital_Watermarking_Chemicals}), but the state of the art is unsatisfactory.
+How might one encrypt molecules so they could only be used over the web as a black box?
+Beyond that, how might one attach security levels ensuring that only very specific and certified algorithms would be allowed on the molecule, or licensing, collaboration conditions, etc.?
+
+\subsection*{Data mining \& machine learning}
 \begin{enumerate}
@@
 
-Phenotypic (where one takes images of cells and then analyzes them
-extract numerical features from the image) screens are increasingly
+Phenotypic screens (where one takes images of cells and then analyzes them,
+extracting numerical features from the image) are increasingly
 common. These types of screens lead to very large, very high
@@
 \item What methods are suitable to perform feature selection and
-  modeling of such data
-\item Does phenotype-derived data lead to to ``better'' models for
+  modeling of such data?
+\item Do phenotype-derived data lead to ``better'' models for
   small molecules compared to the usual structure-derived data?
@@
 
-(statistical) QSAR is an example of traditional data mining where one
+(Statistical) QSAR is an example of traditional data mining where one
 correlates a set of independent variables to a dependent
 variable. This lets one predict properties of a new molecule based on
-its structural features and those of a training set. The inverse QSAR
-problem, also de-novo design, is such that given a molecule structure representation
+its structural features and those of a training set. The \emph{inverse QSAR
+problem}, also called \emph{de-novo design}, is: given a molecule structure representation
 (usually a descriptor vector), what are the possible input structures
@@
 a chemical design of experiments. It is clearly a non-continuous optimization problem, since
-not all chemical molecules might be accessible, and cost/risk to create molecules phsically is another
+not all chemical molecules might be accessible, and cost/risk to create molecules physically is another
 critical factor. This can be made more complex, by
@@
 \item What methodologies can be devised to address this problem?
-\item How can we ensure combinatorically created suggestions make use of the large 
+\item How can we ensure combinatorially created suggestions make use of the large
 chemical structure, chemical building block, and chemical reaction databases by suggesting
-chemically feasible molecules (not just virtual accessible ones)? How can we improve 
+chemically feasible molecules (not just virtually accessible ones)? How can we improve
 synthetic accessibility predictions \cite{Boda_Seidel_Gasteiger_2007}?
@@
 \item Are there internal features of a modeling algorithm (say
-  hyperplanes in a SVM approach) that lets us simplify the problem?
+  hyperplanes in a SVM approach) that let us simplify the problem?
 \end{itemize}
 
-%\item \emph{Efficient molecule browsing, e.g. on scaffold level}.
-%
-%Chemical Abstract Services have a molecule browsing tool called SubScape, which allows to browse large-scale
-%chemical spaces efficiently. What could be large-scale solutions for doing this within (combined and aligned) 
-%public databases.
-%
 \item \emph{Dynamic similarity search on instant binary vectors}.
@@
   reducing space requirements and speeding up further similarity
-  searches (by loosing some accuracy), 3. creating further
-  pre-computations for speeding up threshold based similarity searches
-  \cite{doi:10.1021/ci800076s}.\\  Still, if we are interested to employ
+  searches (by losing some accuracy), 3. creating further
+  pre-computations for speeding up threshold-based similarity searches
+  \cite{doi:10.1021/ci800076s}.\\  Still, if we would like to employ
   dynamic changes in the similarity encodings, e.g. using only a set
-  of binary features, then previously done hashing or pre-computations
+  of binary features, then pre-computations like hashing
   might need to be redone efficiently on the fly. Finally, the major
@@
   is improvable, since many medicinal chemists are very concerned
-  about the data quality of such efforts.
+  about the data quality of such efforts.\cite{chemicalEntityRecognition,textMiningForDrugs}
 
   \begin{itemize}
-  \item What could be done to improve the mining quality, curate the
+  \item What could be done to improve the mining quality, to curate the
     obtained data, and to provide confidence level estimations for
     each molecule coming from patent mining?
-  \item Do require image2structure and text2structure mining also data
+  \item Do image2structure and text2structure mining also require data
     stores for ensuring a sufficient amount of confidence and data
@@
   of molecules.  Sometimes non-vectorial molecular 3D shape or
-  molecule kernel comparisons might be more suitabe to compare
+  molecule kernel comparisons might be more suitable to compare
   molecules, since they might better correlate with activities. One
-  key problem is that non-vectorial encodings require to compare all
+  key problem is that non-vectorial encodings require comparison of all
   molecules (or their 3D conformational explosions) in a pair-wise
@@
   millions of molecules.  Can dyadic data approaches help
-  \cite{Hochreiter:2006:SVM:1159508.1159516}? Other approximations or
+  \cite{Hochreiter:2006:SVM:1159508.1159516}? What about other approximations or
   cascading flows?
@@
   each other, but have a certain correlation between each other.  In
-  Chemogenomics this is used for creating models of combining
+  Chemogenomics this is used to create models of combining
   molecules with protein sequences, molecules with active sites of
@@
   assays. How can we optimize such highly complex mining scenarios,
-  especially when considering large-scale data sets with hundred of
-  thousands molecules and thousands of biological activities?  How can
+  especially when considering large-scale data sets with hundreds of
+  thousands of molecules and thousands of biological activities?  How can
   we combine, mine, and visualize categorial and continuous output
@@
   forward? If we test molecules, should we test it in a single
-  biological assay or in multiple biological assays, if multiple,
+  biological assay or in multiple biological assays, and, if multiple,
   which ones?  If a company does not have a biological assay within
@@
 
-\section*{Databases \& software engineering}
+\subsection*{Databases \& software engineering}
 \begin{enumerate}
@@
 
-  While many programs are availabel to generate 3D conformations, on
+  While many programs are available to generate 3D conformations, on
   the fly generation for large collections can be time
@@
 Databases of chemical structures are ubiqituous, employing a variety
-of RDBMS products. Structure based queries (exact match, substructure,
-similarity) are dependent on the chemical representation - most
+of RDBMS products. Structure-based queries (exact match, substructure,
+similarity) are dependent on the chemical representation---most
 solutions employ a linear string based form (SMILES, InChI) and
@@
     distribute jobs over Hadoop and other map reduce systems?
-  \item Is the latency involved in Hadoop based datastores (HBase)
+  \item Is the latency involved in Hadoop-based datastores (HBase)
     worth the ability to handle massive molecule collections and run
@@
 
-\section*{Enterprise software (KM,ELN)}
+\subsection*{Enterprise software (KM,ELN)}
 We know that the enterprise software and ELN market is still growing.
@@
   can we improve the maintenance, leveraging, and collaboration with
-  many external partners (each of them potentially with another KM/ELN
+  many external partners (each of them potentially with a distinct KM/ELN
   solution)? Which party is hosting which data in which data structure
@@
   entries (and a limited number of annotations, e.g. biological
-  activities) are visible to a partner.  How can this be organized for
-  a multitude of partners? Cloud computing, user management,
+  activities) are visible to a partner?  How can this be organized for
+  a multitude of partners?  How best to incorporate cloud computing, user management,
   encryption granularity and efficient security management?
@@
 \item \emph{Licensing in a parallel world} Many software vendors use
-  different solutions for parallizing compute jobs: SGE, PVM, MPI,
+  different solutions for parallelizing computing jobs: SGE, PVM, MPI,
   etc.  Is a cloud really an option?  What about SaaS with secured data
@@
 \end{enumerate}
+%
+\bibliographystyle{abbrv}
+\bibliography{resubmission}
 
-\bibliographystyle{abbrv}
-\bibliography{paper}
 \end{document}

### 2012-02-06 resubmit/cheminformatics_CACM_revision.tex
Finished off second round

@@
   in information content from left to right and top to bottom. A
-  chemist's traditional 2D depiction (top left), a labeled graph (top right), a 3D
+  chemist's traditional 2D depiction (top left), a labeled graph (top
+  right; 'A' corresponds to aromatic atoms, 'O' to oxygens and 'On' to
+  oxygens or nitrogens. Unlabeled nodes have an implicit label of 'C'), a 3D
   conformation (bottom left) and a surface representation (bottom right). All show differing
@@
 different low energy 3D structures, or conformers, may be consistent
-with the same graph structure. In addition, the 3D structure may have
+with the same graph structure.
+
+In addition, the 3D structure may have
 a ``secondary'' geometrical arrangement of features (such as the
@@
 information than others, they are all equally valid and are suited for
-different problems. Fig.~\ref{figure:chemical-structures} shows
-a hierarchy of representations; when searching for
-substructures, a 2D representation suffices, but when exploring
-protein binding, a 3D structure is vital.
+different problems. Fig.~\ref{figure:chemical-structures} shows a
+hierarchy of representations; when searching for substructures, a 2D
+representation suffices, but when exploring protein binding, a 3D
+structure is vital. Figure \ref{fig:structure-classes} in the Appendix
+highlights yet another subtlety where the number and types of atoms
+and their connectivity may not suffice to uniquely define the
+structure.
 
@@
 what type of labels are required? Currently, many of these questions
-are answered using trial and error.
+are answered using trial and error - in that an objective function
+(usually root mean square error or percentage correct) is defined and
+descriptors and modeling approaches are iteratively adapted to
+optimize the objective function . In other words, there is no
+underlying theory to explain or suggest optimal approaches.
 
@@
 
-We structure the main article around ``risk minimization'' of drug discovery (a slow and expensive process)---how to minimize the chances of a small
-molecule failing, due to poor physical, chemical or biological
-properties, in the various research and development stages as a drug
-candidate, and how to successfully employ cheminformatics methods in
-these efforts. As an example, a molecule must be soluble and show a
-certain degree of bioavailability to be considered as a drug
-candidate. In cases where these properties are poor, a cheminformatics
-approach can suggest replacement of certain functional groups, that
-maintain potency but improve the solubility and bioavailability.
-Table~\ref{table:properties} (in the Appendix) summarizes a number of
-properties a drug candidate must satisfy to be considered
-therapeutically useful (such as being appropriately soluble,
-exhibiting minimal toxicity and being selective for the intended
-target) and sketches the role of cheminformatics at each stage of drug
-development.
+We structure the main article around ``risk minimization'' of drug
+discovery (a slow and expensive process)---how to minimize the chances
+of a small molecule failing, due to poor physical, chemical or
+biological properties, in the various research and development stages
+as a drug candidate, and how to successfully employ cheminformatics
+methods in these efforts. As an example, a molecule must be soluble
+and show a certain degree of bioavailability to be considered as a
+drug candidate. In cases where these properties are poor, a
+cheminformatics approach can suggest replacement of certain functional
+groups, that maintain potency but improve the solubility and
+bioavailability.  Table~\ref{table:properties} (in the Appendix)
+summarizes a number of properties a drug candidate must satisfy to be
+considered therapeutically useful (such as being appropriately
+soluble, exhibiting minimal toxicity and being selective for the
+intended target) and sketches the role of cheminformatics at each
+stage of drug development.
 %
@@
 
+\subsection{Activity Mining \& Prediction}
+\label{sec:activity-mining}
+
+The basis of predictive modeling in cheminformatics is that the
+biological activity of a molecule is a function of its chemical
+structure. Together with the \emph{similar property principle}
+\cite{Johnson:1990qf} (structurally similar molecules exhibit
+similar properties), the goal of any modeling approach is to capture
+and characterize correlations between structural features and
+observed biological activities. Simultaneously, such approaches must
+also describe the likelihood of error when using the
+models for decision making.  A variety of approaches can be employed
+to assess the error in (or conversely, the reliability or confidence of) a
+prediction, ranging from statistical approaches to more
+empirical approaches such as defining an \emph{applicability domain} -
+the region of the input (i.e., chemical structures) that can be
+reliably predicted, usually delineated by similarity
+to the training set.
+
+In many cases, the activity of a small molecule is due to its
+interaction with a receptor. Traditionally, QSAR \cite{Hansch:1962vn,
+ Free:1964ys} approaches do not take into account receptor features,
+focusing only on small molecule features, and therefore lose valuable
+information on ligand-receptor interactions. As a result, techniques
+such as docking (which predicts how molecules fit together),
+structure-based pharmacophore modeling (a 3D approach to capturing protein-ligand
+interactions) and proteochemometric methods have been designed to take
+into account both ligand and receptor structures. The last method is
+an extension of statistical QSAR methods to simultaneously
+model the receptor and ligand, as first reported by Lapinsh \textit{et
+ al.}~\cite{lapinsh2001}.
+
+The first step in the predictive modeling of biological activities is
+to generate \emph{molecular descriptors} (a.k.a. \emph{features}) that
+are numerical representations of different structural features. For
+example, labeled graphs (Figure 1B), and their associated
+characterizations are easily accessible to computer scientists, yet
+such features miss significant physicochemical features (surface
+properties such as charge, flexibility and so on). At the same time,
+it can be difficult to objectively quantify many chemical aspects of a
+molecule, such that the resultant descriptors are suitable for
+predictive modeling.  Hence, the choice of a chemical descriptor
+should by no means be treated as a ‘solved’ problem. For a more
+detailed discussion the reader is referred to more comprehensive
+textbooks \cite{todeschini2000,faulon2010}.
+
+As noted above, molecular graphs can be transformed to numerical
+vector representations ranging from counts of elements to eigenvalues
+of the Laplacian matrix.  Alternatively, we can compare molecular
+graphs directly, via \emph{kernel methods}, where a kernel on graphs
+$G$ and $G'$ provides a measure of how similar $G$ is to $G'$, or a
+kernel on a single graph compares measured similarities between the
+graph's nodes. In these cases, rather than compute vector
+representations, we directly operate on the graph representations.
+
+Both methods have advantages and disadvantages. The vector approach
+requires one to identify a subset of \emph{relevant} (to the property
+being modeled) descriptors: the feature selection problem is
+well discussed in the data mining literature. A kernel approach
+does not require feature selection, but one faces the problem of
+evaluating a data set in a pairwise fashion, and one must
+identify an appropriate kernel. This is an important challenge as the
+kernel should be selected to satisfy \emph{Mercer's condition} (a
+well-known mathematical property in machine learning that makes a set
+of observations easier to make predictions about), and this is not
+always possible with traditional cheminformatics based kernels (such
+as a multiple common substructure based kernel). These challenges can
+make kernel based methods prohibitive on very large datasets.
+
+Having settled on a numerical representation and a possible class of
+model types, one must address the goal of the model. Are we looking
+for pure predictive power, with no interest in any explanatory
+features? Or are we more interested in decent predictive power, but
+also some explanation as to why a molecule is predicted as toxic or
+non-toxic (for example)? The former scenario is common in virtual
+screening settings, where we require a high degree of accuracy and
+fast predictions, but do not really care why one molecule is active
+and another inactive. These types of models can be black boxes (e.g.,
+neural networks) and algorithmic models \cite{Breiman:2001fk} (random
+forests). The latter scenario is more common in exploratory and
+optimization settings, where we hope that the output of the model will
+guide us in chemical modifications to improve the property being
+modeled.  In this case, we need to understand the effect of a certain
+structural feature on the observed potency and expect that the model
+will provide insight. These types of models are generally
+distributional in nature (such as linear regression, Partial Least
+Squares and Na\"{i'}ve Bayes) though some algorithmic approaches such
+as random forests can provide insight.
+
+Having chosen a modeling approach, a vital question is that of model
+reliability which is closely tied to the concept of model
+applicability -- the question of when the prediction of the model for
+a new object is reliable. This issue has grown in importance with the
+increasing use of predictive models in regulatory settings. A
+misprediction (due to the model not being applicable to a certain class
+of inputs) can have significant financial repercussions. A variety of
+methods have been developed that attempt to characterize the domain of
+a model. These methods not only determine whether models are
+applicable, but can also be used for deciding if additional biological
+experiments are required for reducing the prediction error on certain
+compound classes.
+
+One of the key challenges faced by predictive modeling is the fact
+that small molecules are not static and do not exist in isolation.
+Traditionally, predictive models have focused on a single structure
+for a small molecule and ignored the receptor. Yet, small molecules
+can exist in multiple tautomeric forms and usually in multiple
+conformations. As a result, enhancing the accuracy of predictions
+ideally will require that the 3D geometries of that molecule be taken
+into account and, as far as possible, that the receptor also be
+considered. While some relevant low-energy conformers of small
+molecules may be accessible from crystallographic databases, this is
+not always the case. Though it is now possible to generate reasonable
+low energy conformations \emph{ab initio} it must be noted that the
+``biologically relevant'' conformation might differ from the lowest
+energy conformation of the molecule considered in isolation,
+necessitating the need for conformational search.  Multi-conformer
+modeling has been addressed in the 4D-QSAR methodology described by
+Hopfinger \textit{et al.}  \cite{Albuquerque:1998ys}.  Recent
+techniques such as \emph{multiple-instance learning} could also be
+applied to the multi-conformer problem.
+
+With the advent of high-throughput screening technologies, large
+libraries of compounds can be screened against multiple targets in an
+efficient manner. Such panel assays provide a broad, systems-level
+view of small molecule activities.  Models developed on such data
+afford us the opportunity to identify targets, characterize off-target
+effects and so on. However, most approaches to this problem tend to
+develop multiple individual models \cite{Chen:2010zr}, leading to
+multiple, independent predictions for a given input molecule.
+Instead, one could imagine an approach that takes into account the
+covariance structure of the multiple observed activities and the
+structural descriptors within a single model. Such an approach could
+lead to more robust predictions for panel assays (that is, which
+battery of tests would be most useful to perform). Finally, this
+approach also better reflects clinically relevant compound
+profiles~\cite{kuhn2010} and the ``personalized medicine'' concept
+(\emph{e.g.}, drug-drug interaction profiles).
+
 \subsection{Expanding Chemical Space}
@@
 interface between cheminformatics and bioinformatics.
-
-\subsection{Activity Mining \& Prediction}
-\label{sec:activity-mining}
-
-The basis of predictive modeling in cheminformatics is that the
-biological activity of a molecule is a function of its chemical
-structure. Together with the \emph{similar property principle}
-\cite{Johnson:1990qf} (structurally similar molecules exhibit
-similar properties), the goal of any modeling approach is to capture
-and characterize correlations between structural features and
-observed biological activities. Simultaneously, such approaches must
-also describe the likelihood of error when using the
-models for decision making.  A variety of approaches can be employed
-to assess the error in (or conversely, the reliability or confidence of) a
-prediction, ranging from statistical approaches to more
-empirical approaches such as defining an \emph{applicability domain} -
-the region of the input (i.e., chemical structures) that can be
-reliably predicted, usually delineated by similarity
-to the training set.
-
-In many cases, the activity of a small molecule is due to its
-interaction with a receptor. Traditionally, QSAR \cite{Hansch:1962vn,
- Free:1964ys} approaches do not take into account receptor features,
-focusing only on small molecule features, and therefore lose valuable
-information on ligand-receptor interactions. As a result, techniques
-such as docking (which predicts how molecules fit together),
-structure-based pharmacophore modeling (a 3D approach to capturing protein-ligand
-interactions) and proteochemometric methods have been designed to take
-into account both ligand and receptor structures. The last method is
-an extension of statistical QSAR methods to simultaneously
-model the receptor and ligand, as first reported by Lapinsh \textit{et
- al.}~\cite{lapinsh2001}.
-
-The first step in the predictive modeling of biological activities is
-to generate \emph{molecular descriptors} (a.k.a. \emph{features}) that
-are numerical representations of different structural features. Here
-multiple approaches exist. We can transform molecular graphs into a
-numerical vector representation of chemical structures
-\cite{todeschini2000}, ranging from counts of elements to eigenvalues
-of the Laplacian matrix.  Alternatively, we can compare molecular
-graphs directly, via \emph{kernel methods}, where a kernel on graphs
-$G$ and $G'$ provides a measure of how similar $G$ is to $G'$, or a
-kernel on a single graph compares measured similarities between the
-graph's nodes. In these cases, rather than compute vector
-representations, we directly operate on the graph representations.
-
-Both methods have advantages and disadvantages. The vector approach
-requires one to identify a subset of \emph{relevant} (to the property
-being modeled) descriptors: the feature selection problem is
-well discussed in the data mining literature. A kernel approach
-does not require feature selection, but one faces the problem of
-evaluating a data set in a pairwise fashion, and one must
-identify an appropriate kernel. This is an important challenge as the
-kernel should be selected to satisfy \emph{Mercer's condition} (a
-well-known mathematical property in machine learning that makes a set
-of observations easier to make predictions about), and this is not
-always possible with traditional cheminformatics based kernels (such
-as a multiple common substructure based kernel). These challenges can
-make kernel based methods prohibitive on very large datasets.
-
-Having settled on a numerical representation and a possible class of
-model types, one must address the goal of the model. Are we
-looking for pure predictive power, with no interest in any explanatory
-features? Or are we more interested in decent predictive power, but
-also some explanation as to why a molecule is predicted as toxic or
-non-toxic (for example)? The former scenario is common in virtual
-screening settings, where we require a high degree of accuracy and fast
-predictions, but do not really care why one molecule is active
-and another inactive. These types of models can be black boxes (e.g., neural
-networks) and algorithmic models \cite{Breiman:2001fk} (random
-forests). The latter scenario is more common in exploratory and
-optimization settings, where we hope that the output of the model will
-guide us in chemical modifications to improve the property being
-modeled.  In this case, we need to understand the effect of a
-certain structural feature on the observed potency and expect that the
-model will provide insight. These types of models are generally
-distributional in nature (such as linear regression and Partial Least Squares) though
-some algorithmic approaches such as random forests can provide insight.
-
-Having chosen a modeling approach, a vital question is that of model
-reliability which is closely tied to the concept of model
-applicability -- the question of when the prediction of the model for
-a new object is reliable. This issue has grown in importance with the
-increasing use of predictive models in regulatory settings. A
-misprediction (due to the model not being applicable to a certain class
-of inputs) can have significant financial repercussions. A variety of
-methods have been developed that attempt to characterize the domain of
-a model. These methods not only determine whether models are
-applicable, but can also be used for deciding if additional biological
-experiments are required for reducing the prediction error on certain
-compound classes.
-
-One of the key challenges faced by predictive modeling is the fact
-that small molecules are not static and do not exist in isolation.
-Traditionally, predictive models have focused on a single
-structure for a small molecule and ignored the receptor. Yet, small molecules can
-exist in multiple tautomeric forms and usually in multiple
-conformations. As a result, enhancing the accuracy of predictions
-ideally will require that the 3D geometries of that molecule be taken
-into account and, as far as possible, that the receptor also be considered.
-Multi-conformer modeling has been addressed in the
-4D-QSAR methodology described by Hopfinger \textit{et al.}
-\cite{Albuquerque:1998ys}.  Recent techniques such as
-\emph{multiple-instance learning} could also be applied to the
-multi-conformer problem.
-
-With the advent of high-throughput screening technologies, large
-libraries of compounds can be screened against multiple targets in an
-efficient manner. Such panel assays provide a broad, systems-level
-view of small molecule activities.  Models developed on such data
-afford us the opportunity to identify targets, characterize off-target
-effects and so on. However, most approaches to this problem tend to
-develop multiple individual models \cite{Chen:2010zr}, leading to
-multiple, independent predictions for a given input molecule.
-Instead, one could imagine an approach that takes into account the
-covariance structure of the multiple observed activities and the
-structural descriptors within a single model. Such an approach could
-lead to more robust predictions for panel assays (that is, which
-battery of tests would be most useful to perform). Finally, this
-approach also better reflects clinically relevant compound
-profiles~\cite{kuhn2010} and the ``personalized medicine'' concept
-(\emph{e.g.}, drug-drug interaction profiles).
 
@@
 %
+\begin{figure}
+\centering
+\includegraphics[width=\linewidth]{cacm_tautomers.png}
+\caption{An depiction of the various classes of chemical structures
+  (shown using a 2D representation). This figure serves to highlight
+  the danger of assuming that a molecule is \emph{uniquely} defined by
+the number and types of atoms and even simple connectivity between them.}
+\label{fig:structure-classes}
+\end{figure}
+%
+
 \end{document}

### 2012-02-07 resubmit/cheminformatics_CACM_revision.tex

@@
-\documentclass{sig-alternate}
+\documentclass{../sig-alternate}
 
@@
   interdisciplinary collaboration between computer scientists and
-  molecular biologists that led as one of its major accomplishments to the
-  sequencing of the human genome. However,
-  despite this great success, few computer scientists are familiar
+  molecular biologists that led to the
+  sequencing of the human genome and other accomplishments. However,
+  few computer scientists are familiar
   with a related discipline: cheminformatics, the use of computers to
@@
   properties.  Cheminformatics has wide applicability, from the
-  discovery of novel drug candidates to the optimization of certain
-  physicochemical properties of small molecules.
-While researchers in both academia and industry have made important contributions to this field since the 1960s (if not before), new and exciting research opportunities now exist because of an ``opening'' of data and software that began around the year 2000.
-Since that time -- as an
-  effect of a change in mindsets, government-mandated policy changes,
-  and, not least, as a result of chemists volunteering their time
-  for an Open Science ``movement'' -- researchers have gained access to
-  freely available software packages and databases of tens of millions
-  of chemicals. Academic chemists now confront a variety of unsolved
-  algorithmic problems that might have been prohibitive to tackle a decade ago,
-  but whose solutions are critical to research ranging from
+  drug discovery to agrochemicals and materials design. 
+  While researchers in both academia and industry have made important 
+  contributions to this field for decades, new and 
+  exciting collaborative opportunities have arisen from an ``opening'' of 
+  data and software as an effect of changing mindsets,
+  policy changes, 
+  and chemists volunteering time
+  for ``Open Science''.  Researchers have gained access to
+  freely available open source software packages and open databases of tens of millions
+  of chemicals, allowing academic chemists to confront a variety of 
+  algorithmic problems whose solutions will be critical to address 
+  current challenges ranging from
   determining the behavior of small molecules in biological pathways,
   to finding therapies for rare and neglected diseases.
+  In this paper, we give a broad overview of the field of cheminformatics with a
+  focus on open questions and challenges. 
 \end{abstract}
@@
 Novel technologies in the life sciences produce information at an
-ever increasing rate -- public data stores such as that of the
-European Bioinformatics Institute (EBI) contain in the order of 10
-petabytes of biological information.  While biological information has
-been publicly available for decades, the same has not been true for
-chemical information, until recently. It was not until 2004 that a
+ever increasing rate, with public data stores such as that of the
+European Bioinformatics Institute (EBI) containing in the order of 10
+petabytes of biological information.  Until recently, the same 
+had not been true for
+chemical information, but in 2004 a
 large public small molecule structure repository (PubChem) became
-freely available. This was followed by other databases. While much of
+freely available, and this was soon followed by other databases. 
+Similarly, while many of
 the foundational algorithms of cheminformatics have been described
-since the 1950s, publicly available software implementing many of
-these algorithms has only been accessible in the last ten to fifteen
+since the 1950s, open source software implementing many of
+these algorithms have only become accessible in the last ten to fifteen
 years \cite{faulon2010}.
@@
 its public availability -- and how does this relate to the field of
-computer science? We note that though cheminformatics has a broad
-applicability, ranging from agrochemical research to the design of
-novel materials, we have chosen to use drug discovery as the context
-of this article due to its central relevance for human well being.
-
-Given that the art and science of drug discovery is based around small
-molecules, techniques and algorithms to handle, analyse and interpret
-chemical structure information are a core component of the methods of drug
-discovery. Public availability of chemical information and
-tools to handle it are important as drug discovery research is slow,
-expensive and prone to failure. The more information available to each
-researcher, facilitating the development of novel treatments, the
-better the chances of avoiding the many causes of attrition. The
-relevance of computer science to this research becomes clear when we
-consider the quantity and types of data available -- currently a
-single data base such as PubChem contains more than 34 million
-chemical structure records, along with an even larger number of
-\emph{annotations} (comments on the structure that are database
-searchable), such as synonyms (different names for the same molecule),
+computer science? Cheminformatics is widely used in fields 
+from agrochemical research to the design of
+novel materials -- but we have chosen to use drug discovery as the context
+of this article due to its central relevance for human well-being.
+The art and science of drug discovery centers around small
+molecules, with a core component of methods involving 
+techniques and algorithms to handle, analyse and interpret
+chemical structure information. 
+Unfortunately, drug discovery research is slow, expensive and prone to failure. 
+Public availability of chemical information and
+tools to handle it are important, as the more information available to each
+researcher, the better the chances of avoiding the many causes of attrition. 
+Computer science research is highly relevant to managing the \textit{volume} 
+and \textit{complexity} of chemical information. 
+A single database such as PubChem contains more than 34 million
+chemical structure records), along with an even larger number of
+\emph{annotations}, such as synonyms (different names for the same molecule; 
+PubChem stores 50.1 million synonym for 19.6 million compounds),
 known targets (drugs with which the molecule is known to interact),
 mode of action (how the molecule interacts with its targets),
-regulatory approval status and so on.  Simply considering the synonym
-list, PubChem stores 50.1 million synonym for 19.6 million compounds.
-
-The design of data structures to handle chemical structures and
-associated information, as well as subsequent data mining are areas
-where expertise in computer science is urgently needed. (For a recent,
-detailed introduction to cheminformatics, written for computer
-scientists, see \cite{brown2009}).
-
-Cheminformatics comprises different areas, which broadly speaking can
-be divided into three fields: \emph{capturing data} (using lab
-notebooks or potentially using formats such as the Chemical Markup
-Language for publications); \emph{storing data} (designing database
-schemas, devising ontologies) and \emph{mining data} (such as for the
-prediction of biological activities of compounds).  However, the type
-of information being analyzed is different from biological
-information: while bioinformatics often deals with sequences, the
-domain of cheminformatics is chemical structures. In the case of the
-former, information can very frequently be represented as
+regulatory approval status and much more.
+
+While bioinformatics often deals with sequences, the
+domain of cheminformatics is chemical structures. 
+In the case of the former, information can frequently be represented as
 one-dimensional strings, which are relatively easy to handle
-computationally. In the case of the latter however, chemical
-structures may possess rings, branches as well as multiple valid
-representations of the same molecule (such as \emph{tautomers}, where
-hydrogen atoms can be positioned in different places of the
-molecules), potentially giving rise to ambiguities. Hence, chemical
-structures are often reported to be more difficult to standardize and
+computationally. In the case of the latter, however, chemical
+structures are complex graphs that may possess rings and branches, 
+and there may be multiple valid representations of the same molecule 
+that need to be considered for particular algorithms 
+(such as \emph{tautomers}, where hydrogen atoms are positioned at 
+different places in the molecule). Hence, chemical
+structures are more difficult to standardize and
 algorithms dealing with them must take this into account.
-
 To illustrate the difficulty of standardizing chemical structures,
 different representations of the same chemical structure are shown in
-Figure~\ref{figure:chemical-structures}. The representations range
-from a graph (Fig.~\ref{figure:chemical-structures}A) that contains
-much implicit information (such as unlabeled nodes represent carbons
-with valencies satisfied by hydrogens) to a fully labeled graph,
-where all information is explicit
-(Fig.~\ref{figure:chemical-structures}B). The former is what is
+Figure~\ref{figure:chemical-structures}. The top left graph contains
+much implicit information, since unlabeled nodes represent carbons
+with valencies satisfied by hydrogens, while the top right graph is 
+fully labeled. The former is what is 
 usually exchanged informally between chemists, but to actually mine
@@
 required.  For example, graph mining algorithms can be applied to
-labeled graph representations, to discover patterns in large-scale
-datasets \cite{horst2009}; this information can also be determined by simulations such as molecular dynamics.
+labeled graph representations to discover patterns in large-scale
+datasets \cite{horst2009}; this information can also be determined 
+by simulations such as molecular dynamics.
 
@@
   in information content from left to right and top to bottom. A
-  chemist's traditional 2D depiction (top left), a labeled graph (top
-  right; 'A' corresponds to aromatic atoms, 'O' to oxygens and 'On' to
-  oxygens or nitrogens. Unlabeled nodes have an implicit label of 'C'), a 3D
-  conformation (bottom left) and a surface representation (bottom right). All show differing
-  (and valid) aspects of the same molecule and are suited for
-  different purposes, with the choice of representation guided by its
+  chemist's traditional 2D depiction (top left); a labeled graph (top
+  right -- `A' corresponds to aromatic atoms, `O' to oxygens and `On' to
+  oxygens or nitrogens, unlabeled nodes have an implicit label of `C'; a 3D
+  conformation (bottom left) and a surface representation (bottom right). All show differing,
+  valid aspects of the same molecule, and are suited for
+  different purposes, with representation choice guided by
   intended use.}
@@
 different low energy 3D structures, or conformers, may be consistent
-with the same graph structure.
-
-In addition, the 3D structure may have
+with the same graph structure. In addition, the 3D structure may have
 a ``secondary'' geometrical arrangement of features (such as the
@@
 graph. Thus we have 3D representations
-(Fig.~\ref{figure:chemical-structures}C) that make the 3D arrangement
-of atoms and bonds explicit. Fig.~\ref{figure:chemical-structures}D
-goes further and represents a molecular surface, encoding some
-property (such as lipophilicity, the ability to dissolve in non-polar
-solvents) that varies over the surface.
-
+(Fig.~\ref{figure:chemical-structures}, bottom) that make the 3D arrangement
+of atoms and bonds explicit. The representation at the bottom right 
+goes further, showing a molecular surface and some
+property that varies over the surface -- such as lipophilicity, the ability to dissolve in non-polar
+solvents.
 Even though some structure representations contain more explicit
-information than others, they are all equally valid and are suited for
-different problems. Fig.~\ref{figure:chemical-structures} shows a
-hierarchy of representations; when searching for substructures, a 2D
+information than others, they are all equally valid,  suited for
+different problems. When searching for substructures, a 2D
 representation suffices, but when exploring protein binding, a 3D
 structure is vital. Figure \ref{fig:structure-classes} in the Appendix
-highlights yet another subtlety where the number and types of atoms
-and their connectivity may not suffice to uniquely define the
+highlights another subtlety where the number and types of atoms
+and their connectivity may not suffice to uniquely define a
 structure.
 
-Given that a fundamental principle of the field is that \emph{similar
-molecules exhibit similar properties} \cite{Johnson:1990qf}, the
+A fundamental principle of cheminformatics is that \emph{similar
+molecules exhibit similar properties} \cite{Johnson:1990qf}. The
 choice of representation is key in determining how such similarities
 are evaluated, and thus the effectiveness of subsequent analyses. But
-even after one has selected a representation one faces the challenge
-of balancing computational costs with the utility of the
+there is a further challenge in balancing computational costs with 
+the utility of a
 representation. For example, a full 3D description of a molecule,
-taking into account all possible conformers would let us accurately
-predict many properties. But the size of this representation and the
+taking into account all possible conformers, would allow accurate
+prediction of many properties. But the size of this representation and the
 time required to evaluate it would be prohibitive. Instead, can we
@@
 what type of labels are required? Currently, many of these questions
-are answered using trial and error - in that an objective function
-(usually root mean square error or percentage correct) is defined and
-descriptors and modeling approaches are iteratively adapted to
-optimize the objective function . In other words, there is no
-underlying theory to explain or suggest optimal approaches.
-
-From the above it becomes clear that the nature of chemical problems
-is often a complex mixture of combinatorial and continuous problems.
-For the field of cheminformatics to flourish, we need closer
-collaboration between chemists (or, more generally, life scientists)
+are answered using trial and error through definition of an objective function
+(usually root mean square error or percentage correct) and
+iterative adaptation of descriptors and modeling approaches to
+optimize the objective function. There is no
+unifying theory to explain or suggest optimal approaches in all cases.
+
+The goals of cheminformatics are to support better chemical
+decision making by (1) storing and integrating data in maintainable
+ways, (2) providing sufficient open standards and tools allowing
+applications and data to be used across heterogeneous platforms, and
+(3) mining the many chemical property spaces in a time- and
+space-efficient way. For the field of cheminformatics 
+to flourish, we need closer
+collaboration between chemists 
 and computer scientists -- where the former need to be able to pose
-their problems in a way relevant for practical applications; and where
+their problems in a way relevant for practical applications, and where
 the latter are able to devise ways of capturing, storing, and
-analyzing chemical data which achieve optimal balances between the
-different risks in generalizable encodings and complexity (space and
-time). The goals of cheminformatics are to support better chemical
-decision making by 1. storing and integrating data in maintainable
-ways, 2.  providing sufficient open standards and tools allowing
-applications and data to be used across heterogeneous platforms, and
-3. mining the many chemical property spaces in a time- and
-space-efficient way. We will discuss all three goals in the following
+analyzing chemical data which achieve an optimal balance between space,
+performance and complexity. For a recent,
+detailed introduction to cheminformatics, written for computer
+scientists, see \cite{brown2009}.
+Cheminformatics comprises different areas, which can broadly
+be divided into three fields: \emph{capturing data} using lab
+notebooks or potentially using formats such as the Chemical Markup
+Language for publications; \emph{storing data} -- designing database
+schemas, devising ontologies -- and \emph{mining data} such as for the
+prediction of biological activities of compounds.  
+We will discuss all of these aspects in the following
 sections.
@@
 This main article will highlight current topics in
-cheminformatics, presenting the computational and algorithmic problems
-faced in the field and how modern CS research can contribute to their
-solution.  We have also provided two appendices: a brief history of cheminformatics, and a detailed list of open problems.
-
-We structure the main article around ``risk minimization'' of drug
-discovery (a slow and expensive process)---how to minimize the chances
+cheminformatics, presenting computational and algorithmic problems
+and indicating how modern CS research can contribute to their
+solution.  We have also provided two appendices: a brief history of 
+cheminformatics, and a list of open problems.
+
+We use as running example cheminformatics methods for 
+``risk minimization'' in drug
+discovery -- minimizing the chances
 of a small molecule failing, due to poor physical, chemical or
-biological properties, in the various research and development stages
-as a drug candidate, and how to successfully employ cheminformatics
-methods in these efforts. As an example, a molecule must be soluble
+biological properties, during the various research 
+and development stages
+as a drug candidate. As an example, a molecule must be soluble
 and show a certain degree of bioavailability to be considered as a
@@
 cheminformatics approach can suggest replacement of certain functional
-groups, that maintain potency but improve the solubility and
+groups (connected
+sets of atoms that affect the characteristics of the chemical
+reactions of the molecule) 
+to maintain potency but improve the solubility and
 bioavailability.  Table~\ref{table:properties} (in the Appendix)
 summarizes a number of properties a drug candidate must satisfy to be
-considered therapeutically useful (such as being appropriately
-soluble, exhibiting minimal toxicity and being selective for the
-intended target) and sketches the role of cheminformatics at each
+considered therapeutically useful 
+and sketches the role of cheminformatics at each
 stage of drug development.
@@
 essential to enable many scientific applications. Yet, each one of
-these areas poses its unique computer science challenges.
-
-There is considerable progress in open chemical and bioactivity
-databases like ChEMBL and PubChem, which are both freely
+these areas poses computer science challenges.
+
+Chemical and bioactivity
+databases such as ChEMBL and PubChem are freely
 available and contain in the order of millions (ChEMBL) and tens of
-millions (PubChem) of data points each. Integration of these disparate
+millions (PubChem) of data points. Integration of these disparate
 data is essential for researchers to obtain the fullest possible
 perspective on what is presently known, tracking ongoing advances in
-science as these become available. Still, integration of data
-across chemical databases remains a challenge, not only due
-to the normalization problems in the chemical and bioactivity
-landscape, but also due to sheer data volume.
-Given the requirement to ``make sense'' of as much chemical data
-as possible, we have to be able to search in complex data, for
-example multi-labelled molecular graphs, where graph labels can change
-from database to database. Hence, molecular graphs must be captured
-in a machine-readable fashion.
-
-A fundamental challenge in cheminformatics is the unique
-representation of chemical structures in a machine-readable format. A
-(chemical) graph can be encoded in multiple ways, depending on how one
-orders the nodes, resulting in multiple representations of the same
+science as these become available. Integration of data
+across chemical databases is a challenge due to sheer data volume and
+due to the difficulties in normalization of chemical and bioactivity
+data. Molecular graphs must be captured
+in a machine-readable fashion. It is also necessary to be able to 
+search in these data, for example multi-labeled graphs, where graph labels can change
+from database to database. Not only can labels differ between databases, 
+but a chemical graph can be encoded in multiple ways, depending on how 
+the nodes are ordered, resulting in multiple representations of the same
 molecule. Considering protonation states 
@@
 string comparisons. The SMILES (Simplified Molecular Input Line
-Specification) format as defined by Weininger 1988\cite{Weininger:1988kx} is an example of a
+Specification) format as defined by Weininger in 1988 
+\cite{Weininger:1988kx} is an example of a
 representation that can be canonicalized. Since the original
@@
 \emph{substructure searching} and \emph{similarity searching}, which
-are required to generalize from the data points contained in a
-particular database to other, related molecules. In substructure
+are required to generalize from the search molecule 
+to other, related molecules. In substructure
 searching, the database is searched for a specified wholly contained
-part of the structures in the database, while in similarity searching,
-the database is searched for structures similar (in some structure or
-property space) to a provided search, or query, structure. Chemical
+part of the search structure, while in similarity searching,
+structures are retrieved that are similar (in some structure or
+property space) to the provided search structure. Chemical
 search packages are often implemented and optimized for a given
@@
 bitmap. Fingerprints fall broadly into two categories: structure keys
-and hashed keys. In the former, each bit position corresponds to a
-distinct substructure such as a \emph{functional group} (a connected
-set of atoms that affects the characteristics of the chemical
-reactions of the molecule). Examples include MACCS and PubChem
-keys. The latter class are designed such that substructural patterns
+and hashed keys. In structure keys, each bit position corresponds to a
+distinct substructure such as a functional group. Examples include MACCS and PubChem
+keys. Hashed keys are those in which substructural patterns
 are represented as strings and then hashed to a random bit
@@
 cover an arbitrarily large collection of substructures such as ``paths
-of length $N$,'' or circular environments, and so on. Examples include
+of length $N$,'' or circular environments. Examples include
 ECFPs (Extended Connectivity Fingerprints, which use local topological
@@
 optimize information density and screening speed).
-
 Given a binary fingerprint, we can first ``pre-screen'' a database, to
-ignore molecules that cannot possibly match the query. This is
-achieved by requiring that all bits in a query fingerprint must also
+ignore molecules that cannot possibly match the query,
+ by requiring that all bits in a query fingerprint must also
 be present in the target fingerprint. Since the target fingerprints
-are pre-computed, comparing binary strings can be performed extremely
+are pre-computed, this check can be performed extremely
 rapidly on modern hardware. As a result, we apply the actual
@@
 screen. Fingerprints can also be used to rapidly search databases for
-similar molecules, in which case a similarity metric such as the
-Tanimoto coefficient is used to compare the query and target
-fingerprints. While fast, heuristics have been developed
+similar molecules, by using a similarity metric such as the
+Tanimoto coefficient to compare the query and target
+fingerprints. Additional heuristics have been developed
 \cite{Swamidass:2007ve} to further speed up similarity searches.
@@
 systems. The most widely used ontology for biologically relevant
-chemical entities is ChEBI \cite{chebi2010}, containing around 26,500
-entities in an interlinked semantic graph as of November 2011, and
-being actively used for annotation in dozens of biological databases.
+chemical entities is ChEBI \cite{chebi2010}, which contains around 27,100
+entities in an interlinked semantic graph as of February 2012, and
+is actively used for annotation in dozens of biological databases.
 Ontologies are based on logical languages of varying levels of
@@
 different domains encoded in a growing number of interlinked
-ontologies. % \cite{hastingsowled2010}.
+ontologies. 
 Enabling automated inferencing in such heterogenous linked systems
@@
 structure. Together with the \emph{similar property principle}
-\cite{Johnson:1990qf} (structurally similar molecules exhibit
-similar properties), the goal of any modeling approach is to capture
+\cite{Johnson:1990qf} mentioned earlier, 
+the goal of any modeling approach is to capture
 and characterize correlations between structural features and
@@
 empirical approaches such as defining an \emph{applicability domain} -
-the region of the input (i.e., chemical structures) that can be
+the region of the input that can be
 reliably predicted, usually delineated by similarity
@@
 are numerical representations of different structural features. For
-example, labeled graphs (Figure 1B), and their associated
+example, labeled graphs and their associated
 characterizations are easily accessible to computer scientists, yet
-such features miss significant physicochemical features (surface
+such features miss significant physicochemical features (
 properties such as charge, flexibility and so on). At the same time,
@@
 predictive modeling.  Hence, the choice of a chemical descriptor
-should by no means be treated as a ‘solved’ problem. For a more
+should by no means be treated as a ``solved'' problem. For a more
 detailed discussion the reader is referred to more comprehensive
@@
 representations, we directly operate on the graph representations.
-
 Both methods have advantages and disadvantages. The vector approach
@@
 does not require feature selection, but one faces the problem of
-evaluating a data set in a pairwise fashion, and one must
+evaluating a data set in a pairwise fashion, and must
 identify an appropriate kernel. This is an important challenge as the
@@
 of observations easier to make predictions about), and this is not
-always possible with traditional cheminformatics based kernels (such
-as a multiple common substructure based kernel). These challenges can
-make kernel based methods prohibitive on very large datasets.
+always possible with traditional cheminformatics based kernels such as those based on multiple common substructures. 
+These challenges can
+make kernel based methods prohibitive on larger datasets.
 
@@
 and another inactive. These types of models can be black boxes (e.g.,
-neural networks) and algorithmic models \cite{Breiman:2001fk} (random
+neural networks) and algorithmic models \cite{Breiman:2001fk} (e.g., random
 forests). The latter scenario is more common in exploratory and
@@
 distributional in nature (such as linear regression, Partial Least
-Squares and Na\"{i'}ve Bayes) though some algorithmic approaches such
-as random forests can provide insight.
+Squares and Na\"{i'}ve Bayes) though some algorithmic approaches
+ can provide insight.
 
@@
 for a small molecule and ignored the receptor. Yet, small molecules
-can exist in multiple tautomeric forms and usually in multiple
-conformations. As a result, enhancing the accuracy of predictions
-ideally will require that the 3D geometries of that molecule be taken
-into account and, as far as possible, that the receptor also be
-considered. While some relevant low-energy conformers of small
-molecules may be accessible from crystallographic databases, this is
+can exist in multiple tautomeric forms conformations. 
+Enhancing the accuracy of predictions
+will ideally require that the 3D geometries of that molecule be taken
+into account and the receptor be considered as far as possible.  
+While some relevant low-energy conformers of small
+molecules may be accessible in crystallographic databases, this is
 not always the case. Though it is now possible to generate reasonable
-low energy conformations \emph{ab initio} it must be noted that the
+low energy conformations \emph{ab initio}, the
 ``biologically relevant'' conformation might differ from the lowest
@@
 profiles~\cite{kuhn2010} and the ``personalized medicine'' concept
-(\emph{e.g.}, drug-drug interaction profiles).
+(e.g. drug-drug interaction profiles).
 
@@
 Enumerating molecules is a combinatorial problem that has fascinated
-chemists, computer scientists and mathematicians alike, for more than
+chemists, computer scientists and mathematicians alike for more than
 a century. Indeed, many fundamental principles of graph theory and
-combinatorics were developed in the context of counting isomers of
-paraffin by Cayley, Polya and others. In the 1960's, Lederberg,
+combinatorics were developed by Cayley, Polya and others 
+in the context of counting isomers of
+paraffin. In the 1960's, Lederberg,
 Djerassi and others developed algorithms to enumerate structures
@@
 just 30 heavy atoms, the size of this space is on the order of
-$10^{60}$ \cite{Bohacek:1996ve}. Any method that purports to enumerate
-molecules will inevitably face a combinatorial explosion if
+$10^{60}$ \cite{Bohacek:1996ve}. Any enumeration method 
+will face a combinatorial explosion if
 implemented na\"{i}vely.
@@
 especially relevant for identifying metabolites, small molecules
-that are the byproducts of metabolic processes and which provide
+that are the byproducts of metabolic processes and thus provide
 insight into the biological state (diseased, fasting, etc.) of an
-organism. Ideally, a chemist would gather spectral data (mass, LC/MS, etc.),
-and the algorithm would provide a list of structures that can give
+organism. A chemist gathers 
+spectral data (NMR, mass, LC/MS, etc.) 
+and an algorithm would ideally provide a list of structures that can give
 rise to the observed spectra. Some commercial products such as MOLGEN
@@
 addition to network generation based on reactions, one also needs to
-verify that there are enzymatic events possible for enabling reactions
-catalyses (enzymes must be present to reduce the
+verify that there are possible enzymatic events to enable reaction
+catalysis (enzymes must be present to reduce the 
 energy required for some reactions to take place). That additional
@@
 new collaboration and innovation opportunities with internal partners and
-a rapidly growing number of external/public partners. In
-cheminformatics and chemistry, scientists already switched years
-ago from paper lab-notebooks to online collaboration platforms, called
+a rapidly growing number of external (public) partners. In
+cheminformatics and chemistry, scientists switched years ago
+from paper lab-notebooks to online collaboration platforms, called
 ELNs (Electronic Lab Notebooks). So, what were the drivers behind
-chemists, and companies, adopting ``Enterprise 2.0,'' a social online
+chemists and companies adopting ``Enterprise 2.0,'' a social online
 collaboration culture?
@@
 Before the year 2000, many chemists were still using paper
-lab-notebooks, multiple compound registration, and search
+lab-notebooks and multiple compound registration and search
 tools. The overall architecture was too inflexible to
@@
 with administrative work, chemical data quality was inconsistent, and
-alignment with other working groups was inefficient (\emph{e.g.}, for
+alignment with other working groups was inefficient (e.g., for
 running analytical experiments). Legal disputes would require manual
 searches through many paper lab notebooks and data synchronization was
-painful, hampering large scale collaborations.
+painful, hindering large scale collaborations.
 
@@
 perhaps driven by the early adoption by chemists of computer
-technologies \ldots The pharmaceutical industry shifted from a
+technologies [\ldots] The pharmaceutical industry shifted from a
 position of `if we get an ELN' to `when we get an
@@
 data approvals and quality control information). Given that any experiment could
-become the core of a patent filing, ELNs allow organizations to define
+become the core of a patent, ELNs allow organizations to define
 and implement audit trails in a standardized and efficient fashion.
@@
 
-However, the future still leaves many challenges ahead. Using
-external/public databases with chemical and bioactivity data remains a
+However, the future still holds many challenges. Using
+external (public) databases with chemical and bioactivity data remains a
 challenge due to differences in identifiers, synchronization, curation
-\& error correcting mechanisms and efficient substructure and
-similarity search within complex data types (compounds). A
+and error correcting mechanisms, and efficient substructure and
+similarity search within complex data types. A
 collaboration with external parties, e.g. contract research, poses
@@
 infrastructure for all parties involved, but also to provide the
-required private/public interface. Furthermore, there are still many
+required private--public interface. Furthermore, there are still many
 data sources that are not being indexed properly: chemistry patents
 are often cryptic and chemical image and text mining remains a
-challenge (albeit, currently being addressed in academia and industry
+challenge (albeit being addressed in academic and industrial research
 \cite{Jessop:2011fk,Sayle:2009uq}). The closed CAS (Chemical Abstract
-Service) is a highly trusted source, and public chemistry and
-bioactivity space has to improve quality and interconnectivity to
-compare with it. SciFinder is another chemical abstract service, with a version, SciFinder Scholar, marketed to universities.  Open-source efforts like ChEMBL and PubChem-BioAssay are on the
-right track, though, unlike some commercial tools, these efforts do not yet abstract reactions. Still, improving data quality and standards between
+Service) is a highly trusted source, and public chemical and
+bioactivity databases have to improve their quality and interconnectivity to
+compete. 
+SciFinder is another chemical abstract service, with a version, SciFinder Scholar, 
+marketed to universities.  Open-source efforts like ChEMBL and PubChem-BioAssay 
+are on the
+right track, though, unlike some commercial tools, these efforts do not yet 
+abstract reactions. Still, improving data quality and standards between
 public and closed sources will be absolutely critical for ensuring
-constant growth, usage, and collaboration scenarios between private
+constant growth, usage, and collaboration between private
 and public parties.
@@
 The extent of Open Source software in cheminformatics has somewhat
-lagged behind related fields such as bioinformatics, partly due to the
-existence of a lucrative pharmaceutical market for proprietary
-software. However, in recent years there has been a rise in the amount
-of Open Source cheminformatics software. One contribution to this has
+lagged behind bioinformatics. However, in recent years there 
+has been a rise in Open Source cheminformatics software. 
+One contribution to this has
 been the Blue Obelisk group~\cite{BlueObelisk2011}
@@
 to create interoperable Open Source chemistry software (among other
-activities). Now there are a number of Open Source chemical toolkits
-available including the Chemistry Development Kit
+activities), driving forward a number of Open Source chemical toolkits
+including the Chemistry Development Kit
 (CDK)~\cite{steinbeck2003}, Open Babel~\cite{openbabel2011}, RDKit and
@@
 (e.g. converting coordinates in various systems, 3D structure
-generation). As well as being of direct use to practising
-cheminformaticians in their day-to-day work, many applications have
+generation). As well as being subject to programmatic use by
+cheminformaticians, many applications have
 been developed that rely on these toolkits to handle chemical data;
@@
 Certain challenges faced by nearly all cheminformatics toolkits stem
-from the graph representation of chemistry. This representation, while
+from the graph representation of chemicals. This representation, while
 succinct and amenable to computation, is only an approximation of
@@
 challenging aspect is that the chemical representation of a molecule
-really needs to capture non-static graphs, in order to take into
+would need to capture non-static graphs in order to take into
 account delocalisation and the tautomeric phenomena molecules undergo
-in practice.
+in real biological contexts.
 
@@
 software is employed in upstream areas such as target identification,
-perhaps not as directly related to possible profits as an actual small
+perhaps not as directly related to possible profits as a candidate small
 molecule. Furthermore, chemical data (structure, activity information
@@
 extremely intensive in terms of time and effort, whereas
-bioinformatics data (e.g., sequences) are much easier to obtain.
+bioinformatics data (e.g., sequences) have become much easier to obtain.
 
@@
 
-There is an admittedly steep learning curve, as significant domain
+There is an admitted learning curve, as significant domain
 knowledge is required to attack many problems in cheminformatics due
@@
 also their complex relationships, transformations, and combinatorial
-challenges.
+challenges, bringing cheminformatics closer to the goal of supporting 
+primary translational research in the wider context of chemical biology.
 
@@
 for funding. J. Hastings thanks the EU project EU-OPENSCREEN for
-funding.
+funding support.
 
@@
 
+\newpage
 \appendix

### 2012-03-18 resubmit2/cheminformatics_CACM_revision.tex
Put in the start of the third and final revision

--- 2012-02-07 resubmit/cheminformatics_CACM_revision.tex
@@
   right -- `A' corresponds to aromatic atoms, `O' to oxygens and `On' to
-  oxygens or nitrogens, unlabeled nodes have an implicit label of `C'; a 3D
+  oxygens or nitrogens, unlabeled nodes have an implicit label of `C'); a 3D
   conformation (bottom left) and a surface representation (bottom right). All show differing,
@@
 for a small molecule and ignored the receptor. Yet, small molecules
-can exist in multiple tautomeric forms conformations. 
+can exist in multiple tautomeric forms and conformations. 
 Enhancing the accuracy of predictions

### 2012-03-18 resubmit2/cheminformatics_second_CACM_revision.tex
Put in the start of the third and final revision

--- 2012-03-18 resubmit2/cheminformatics_CACM_revision.tex
@@
-\documentclass{../sig-alternate}
+\documentclass{sig-alternate}
 
@@
 \usepackage{multirow}
+
+\usepackage[numbers,sort&compress]{natbib}
 
@@
 \emph{annotations}, such as synonyms (different names for the same molecule;
-PubChem stores 50.1 million synonym for 19.6 million compounds),
+PubChem stores 50.1 million synonyms for 19.6 million compounds),
 known targets (drugs with which the molecule is known to interact),
@@
 Chemical and bioactivity
-databases such as ChEMBL and PubChem are freely
-available and contain in the order of millions (ChEMBL) and tens of
+databases such as ChEMBLdb and PubChem are freely
+available and contain in the order of millions (ChEMBLdb) and tens of
 millions (PubChem) of data points. Integration of these disparate
@@
 representations of chemical graphs letting one compare structures via
-string comparisons. The SMILES (Simplified Molecular Input Line
-Specification) format as defined by Weininger in 1988 
+string comparisons. The SMILES (Simplified Molecular-Input Line-Entry
+System) format as defined by Weininger in 1988
 \cite{Weininger:1988kx} is an example of a
@@
 across a full chemical database of thousands or even millions of
-structures is infeasible \cite{Weininger:2011ly}. Speedups can be
+structures is unfeasible \cite{Weininger:2011ly}. Speedups can be
 obtained via the use of heuristics such as structural
@@
 cover an arbitrarily large collection of substructures such as ``paths
-of length $N$,'' or circular environments. Examples include
+of length $N$,'' or circular environments. Examples include Daylight fingerprints (which are ``folded'' to
+optimize information density and screening speed) and
 ECFPs (Extended Connectivity Fingerprints, which use local topological
-information) and Daylight fingerprints (which are ``folded'' to
-optimize information density and screening speed).
+information).
 Given a binary fingerprint, we can first ``pre-screen'' a database, to
@@
 \bibliographystyle{abbrv}
-\bibliography{resubmission}
+\bibliography{CACM_second_revision}
 

### 2012-03-19 resubmit2/cheminformatics_second_CACM_revision.tex
final set of revisions

@@
 are evaluated, and thus the effectiveness of subsequent analyses. But
-there is a further challenge in balancing computational costs with
-the utility of a
-representation. For example, a full 3D description of a molecule,
-taking into account all possible conformers, would allow accurate
-prediction of many properties. But the size of this representation and the
-time required to evaluate it would be prohibitive. Instead, can we
-obtain accurate predictions from a subset of conformers? Or, can we
-obtain comparable accuracy by using a 2D representation? And if so,
-what type of labels are required? Currently, many of these questions
-are answered using trial and error through definition of an objective function
-(usually root mean square error or percentage correct) and
-iterative adaptation of descriptors and modeling approaches to
-optimize the objective function. There is no
-unifying theory to explain or suggest optimal approaches in all cases.
+there is a further challenge in balancing computational costs with the
+utility of a representation. For example, a full 3D description of a
+molecule, taking into account all possible conformers, would allow
+more accurate prediction of many properties (though this is also dependent
+on qood quality force-fields, algorithms and so on). But the size of this
+representation and the time required to evaluate it would be
+prohibitive. Instead, can we obtain accurate predictions from a subset
+of conformers? Or, can we obtain comparable accuracy by using a 2D
+representation? And if so, what type of labels are required?
+Currently, many of these questions are answered using trial and error
+through definition of an objective function (usually root mean square
+error or percentage correct) and iterative adaptation of descriptors
+and modeling approaches to optimize the objective function. There is
+no unifying theory to explain or suggest optimal approaches in all
+cases.
 
@@
 structures is unfeasible \cite{Weininger:2011ly}. Speedups can be
-obtained via the use of heuristics such as structural
-\emph{fingerprints}. Fingerprints encode characteristic features of a
-given chemical structure, usually in a fixed-length
-bitmap. Fingerprints fall broadly into two categories: structure keys
-and hashed keys. In structure keys, each bit position corresponds to a
-distinct substructure such as a functional group. Examples include MACCS and PubChem
-keys. Hashed keys are those in which substructural patterns
+obtained via the use of structural \emph{fingerprint}
+filters. Fingerprints encode characteristic features of a given
+chemical structure, usually in a fixed-length bitmap. Fingerprints
+fall broadly into two categories: structure keys and hashed keys. In
+structure keys, each bit position corresponds to a distinct
+substructure such as a functional group. Examples include MACCS and
+PubChem keys. Hashed keys are those in which substructural patterns
 are represented as strings and then hashed to a random bit
@@
 cover an arbitrarily large collection of substructures such as ``paths
-of length $N$,'' or circular environments. Examples include Daylight fingerprints (which are ``folded'' to
-optimize information density and screening speed) and
-ECFPs (Extended Connectivity Fingerprints, which use local topological
-information).
-Given a binary fingerprint, we can first ``pre-screen'' a database, to
-ignore molecules that cannot possibly match the query,
- by requiring that all bits in a query fingerprint must also
-be present in the target fingerprint. Since the target fingerprints
-are pre-computed, this check can be performed extremely
-rapidly on modern hardware. As a result, we apply the actual
+of length $N$,'' or circular environments. Examples include Daylight
+fingerprints (which are ``folded'' to optimize information density and
+screening speed) and ECFPs (Extended Connectivity Fingerprints, which
+use local topological information).  Given a binary fingerprint, we
+can first ``pre-screen'' a database, to ignore molecules that cannot
+possibly match the query, by requiring that all bits in a query
+fingerprint must also be present in the target fingerprint. Since the
+target fingerprints are pre-computed, this check can be performed
+extremely rapidly on modern hardware. As a result, we apply the actual
 isomorphism test only on those molecules that pass the
 screen. Fingerprints can also be used to rapidly search databases for
-similar molecules, by using a similarity metric such as the
-Tanimoto coefficient to compare the query and target
-fingerprints. Additional heuristics have been developed
-\cite{Swamidass:2007ve} to further speed up similarity searches.
+similar molecules, by using a similarity metric such as the Tanimoto
+coefficient to compare the query and target fingerprints. Additional
+heuristics have been developed \cite{Swamidass:2007ve} to further
+speed up similarity searches.
 
@@
 interaction with a receptor. Traditionally, QSAR \cite{Hansch:1962vn,
- Free:1964ys} approaches do not take into account receptor features,
-focusing only on small molecule features, and therefore lose valuable
-information on ligand-receptor interactions. As a result, techniques
-such as docking (which predicts how molecules fit together),
-structure-based pharmacophore modeling (a 3D approach to capturing protein-ligand
+  Free:1964ys} approaches do not consider receptor features, focusing
+only on small molecule features, thus losing valuable information on
+ligand-receptor interactions. As a result, techniques such as docking
+(which predicts how molecules fit together), structure-based
+pharmacophore modeling (a 3D approach to capturing protein-ligand
 interactions) and proteochemometric methods have been designed to take
 into account both ligand and receptor structures. The last method is
-an extension of statistical QSAR methods to simultaneously
-model the receptor and ligand, as first reported by Lapinsh \textit{et
+an extension of statistical QSAR methods to simultaneously model the
+receptor and ligand, as first reported by Lapinsh \textit{et
   al.}~\cite{lapinsh2001}.
 
-The first step in the predictive modeling of biological activities is
-to generate \emph{molecular descriptors} (a.k.a. \emph{features}) that
-are numerical representations of different structural features. For
-example, labeled graphs and their associated
-characterizations are easily accessible to computer scientists, yet
-such features miss significant physicochemical features (
-properties such as charge, flexibility and so on). At the same time,
-it can be difficult to objectively quantify many chemical aspects of a
-molecule, such that the resultant descriptors are suitable for
-predictive modeling.  Hence, the choice of a chemical descriptor
-should by no means be treated as a ``solved'' problem. For a more
-detailed discussion the reader is referred to more comprehensive
-textbooks \cite{todeschini2000,faulon2010}.
-
-As noted above, molecular graphs can be transformed to numerical
-vector representations ranging from counts of elements to eigenvalues
-of the Laplacian matrix.  Alternatively, we can compare molecular
-graphs directly, via \emph{kernel methods}, where a kernel on graphs
-$G$ and $G'$ provides a measure of how similar $G$ is to $G'$, or a
-kernel on a single graph compares measured similarities between the
-graph's nodes. In these cases, rather than compute vector
-representations, we directly operate on the graph representations.
-Both methods have advantages and disadvantages. The vector approach
-requires one to identify a subset of \emph{relevant} (to the property
-being modeled) descriptors: the feature selection problem is
-well discussed in the data mining literature. A kernel approach
-does not require feature selection, but one faces the problem of
-evaluating a data set in a pairwise fashion, and must
-identify an appropriate kernel. This is an important challenge as the
-kernel should be selected to satisfy \emph{Mercer's condition} (a
-well-known mathematical property in machine learning that makes a set
-of observations easier to make predictions about), and this is not
-always possible with traditional cheminformatics based kernels such as those based on multiple common substructures.
-These challenges can
-make kernel based methods prohibitive on larger datasets.
+The first step in predicting biological activities is to generate
+\emph{molecular descriptors} (a.k.a. \emph{features}) that are
+numerical representations of structural features. For example, labeled
+graphs and their associated characterizations are easily accessible to
+computer scientists, yet such features miss significant
+physicochemical features (properties such as surface distributions or
+3D pharmacophores). At the same time, it can be difficult to
+objectively quantify many chemical aspects of a molecule, such that
+the resultant descriptors are suitable for predictive modeling.
+Hence, the choice of a chemical descriptor should by no means be
+treated as a ``solved'' problem. For a more detailed discussion the
+reader is referred to more comprehensive textbooks
+\cite{todeschini2000,faulon2010}.
+
+Molecular graphs can be transformed to numerical vector
+representations ranging from counts of elements to eigenvalues of the
+Laplacian matrix.  Alternatively, molecular graphs can be compared 
+directly via \emph{kernel methods}, where a kernel on graphs $G$ and
+$G'$ provides a measure of how similar $G$ is to $G'$, or a kernel on
+a single graph compares measured similarities between the graph's
+nodes. In these cases, rather than compute vector representations, we
+directly operate on the graph representations.  Both methods have
+advantages and disadvantages. The vector approach requires one to
+identify a subset of \emph{relevant} (to the property being modeled)
+descriptors: the feature selection problem is well discussed in the
+data mining literature. A kernel approach does not require feature
+selection, but one faces the problem of evaluating a data set in a
+pairwise fashion, and must identify an appropriate kernel. This is an
+important challenge as the kernel should be selected to satisfy
+\emph{Mercer's condition} (a well-known mathematical property in
+machine learning that makes a set of observations easier to make
+predictions about), and this is not always possible with traditional
+cheminformatics based kernels such as those based on multiple common
+substructures.  These challenges can make kernel based methods
+prohibitive on larger datasets.
 
@@
 for a small molecule and ignored the receptor. Yet, small molecules
-can exist in multiple tautomeric forms and conformations.
-Enhancing the accuracy of predictions
-will ideally require that the 3D geometries of that molecule be taken
-into account and the receptor be considered as far as possible.
-While some relevant low-energy conformers of small
-molecules may be accessible in crystallographic databases, this is
-not always the case. Though it is now possible to generate reasonable
-low energy conformations \emph{ab initio}, the
-``biologically relevant'' conformation might differ from the lowest
-energy conformation of the molecule considered in isolation,
-necessitating the need for conformational search.  Multi-conformer
-modeling has been addressed in the 4D-QSAR methodology described by
-Hopfinger \textit{et al.}  \cite{Albuquerque:1998ys}.  Recent
-techniques such as \emph{multiple-instance learning} could also be
-applied to the multi-conformer problem.
+can exist in multiple tautomeric forms and conformations.  Enhancing
+the accuracy of predictions will ideally require that the 3D
+geometries of that molecule be taken into account and the receptor be
+considered as far as possible.  Though it is now possible to generate
+reasonable low energy conformations \emph{ab initio}, the
+``biologically relevant'' conformation might differ significantly (in
+terms of energetics) from the lowest energy conformation of the
+molecule considered in isolation, necessitating the need for
+conformational search.  Multi-conformer modeling has been addressed in
+the 4D-QSAR methodology described by Hopfinger \textit{et al.}
+\cite{Albuquerque:1998ys}.  Recent techniques such as
+\emph{multiple-instance learning} could also be applied to the
+multi-conformer problem.
 
@@
 Enumerating molecules is a combinatorial problem that has fascinated
-chemists, computer scientists and mathematicians alike for more than
-a century. Indeed, many fundamental principles of graph theory and
-combinatorics were developed by Cayley, Polya and others
-in the context of counting isomers of
-paraffin. In the 1960's, Lederberg,
-Djerassi and others developed algorithms to enumerate structures
-based on spectral data, leading to the development of DENDRAL, widely
+chemists, computer scientists and mathematicians alike for more than a
+century. Indeed, many fundamental principles of graph theory and
+combinatorics were developed by Cayley, Polya and others in the
+context of counting isomers of paraffin. In the 1960's, Lederberg,
+Djerassi and others developed algorithms to enumerate structures based
+on spectral data, leading to the development of DENDRAL, widely
 considered as the first expert system \cite{DENDRAL}.
@@
 just 30 heavy atoms, the size of this space is on the order of
-$10^{60}$ \cite{Bohacek:1996ve}. Any enumeration method
-will face a combinatorial explosion if
-implemented na\"{i}vely.
+$10^{60}$ \cite{Bohacek:1996ve}. Any enumeration method will face a
+combinatorial explosion if implemented na\"{i}vely.
 
@@
 structures based on spectral data \cite{Kind:2010zr}. This is
-especially relevant for identifying metabolites, small molecules
-that are the byproducts of metabolic processes and thus provide
-insight into the biological state (diseased, fasting, etc.) of an
-organism. A chemist gathers
-spectral data (NMR, mass, LC/MS, etc.)
-and an algorithm would ideally provide a list of structures that can give
+especially relevant for identifying metabolites, small molecules that
+are the byproducts of metabolic processes and thus provide insight
+into the biological state (diseased, fasting, etc.) of an organism. A
+chemist gathers spectral data (NMR, mass, LC/MS, etc.)  and an
+algorithm would ideally provide a list of structures that can give
 rise to the observed spectra. Some commercial products such as MOLGEN
@@
 Structure enumeration plays a fundamental role in \emph{molecular
-  design} -- the design of compounds (drugs, for instance) that
-optimize some physical, chemical, or biological property or activity
+  design} -- the design of compounds that optimize some physical,
+chemical, or biological property or activity
 \cite{Schneider:2005uq}. A key challenge in this area is to combine
@@
 reaction networks. As reviewed in Chapter 11 in \cite{faulon2010}
-several network enumeration techniques have been developed. However,
-these techniques generally suffer from a combinatorial explosion of
-product compounds. One way to limit the number of compounds generated
-is to simulate the dynamics of the network while it is being
-constructed and remove compounds of low concentration. Following that
-idea, methods have been developed based on the Gillespie Stochastic
-Simulation Algorithm (SSA) to compute on-the-fly species
-concentrations. Chemical reaction network enumeration and sampling is
-an active field of research, particularly in the context of
-metabolism, either to study biodegradation, or to propose metabolic
-engineering strategies to biosynthesize compounds of commercial
-interest. With metabolic network design, a difficulty is that in
-addition to network generation based on reactions, one also needs to
-verify that there are possible enzymatic events to enable reaction
-catalysis (enzymes must be present to reduce the
-energy required for some reactions to take place). That additional
-task requires encompassing both chemical structures and protein
-sequences and the development of tools that are at the
+several network enumeration techniques have been developed but they
+generally suffer from a combinatorial explosion of product
+compounds. One way to limit the number of compounds generated is to
+simulate the dynamics of the network while it is being constructed and
+remove compounds of low concentration. Following that idea, methods
+have been developed based on the Gillespie Stochastic Simulation
+Algorithm (SSA) to compute on-the-fly species concentrations. Chemical
+reaction network enumeration and sampling is an active field of
+research, particularly in the context of metabolism, either to study
+biodegradation, or to propose metabolic engineering strategies to
+biosynthesize compounds of commercial interest. With metabolic network
+design, a difficulty is that in addition to network generation based
+on reactions, one also needs to verify that there are possible
+enzymatic events to enable reaction catalysis (enzymes must be present
+to reduce the energy required for some reactions to take place). That
+additional task requires encompassing both chemical structures and
+protein sequences and the development of tools that are at the
 interface between cheminformatics and bioinformatics.
@@
 
-However, the future still holds many challenges. Using
-external (public) databases with chemical and bioactivity data remains a
+However, the future still holds many challenges. Using external
+(public) databases with chemical and bioactivity data remains a
 challenge due to differences in identifiers, synchronization, curation
 and error correcting mechanisms, and efficient substructure and
-similarity search within complex data types. A
-collaboration with external parties, e.g. contract research, poses
-other problems including compound duplication and efficient data
-synchronization. If external partners are small they often do not even
-have sufficient IT resources themselves and thus rely on external
-services. Cloud services could help not only to provide a service
-infrastructure for all parties involved, but also to provide the
-required private--public interface. Furthermore, there are still many
-data sources that are not being indexed properly: chemistry patents
-are often cryptic and chemical image and text mining remains a
-challenge (albeit being addressed in academic and industrial research
+similarity search within complex data types. A collaboration with
+external parties, e.g. contract research, poses other problems
+including compound duplication and efficient data synchronization. If
+external partners are small they often do not even have sufficient IT
+resources themselves and thus rely on external services. Cloud
+services could help not only to provide a service infrastructure for
+all parties involved, but also to provide the required private--public
+interface. Furthermore, there are still many data sources that are not
+being indexed properly: chemistry patents are often cryptic and
+chemical image and text mining remains a challenge (albeit being
+addressed in academic and industrial research
 \cite{Jessop:2011fk,Sayle:2009uq}). The closed CAS (Chemical Abstract
 Service) is a highly trusted source, and public chemical and
-bioactivity databases have to improve their quality and interconnectivity to
-compete.
-SciFinder is another chemical abstract service, with a version, SciFinder Scholar,
-marketed to universities.  Open-source efforts like ChEMBL and PubChem-BioAssay
-are on the
-right track, though, unlike some commercial tools, these efforts do not yet
-abstract reactions. Still, improving data quality and standards between
-public and closed sources will be absolutely critical for ensuring
-constant growth, usage, and collaboration between private
-and public parties.
+bioactivity databases have to improve their quality and
+interconnectivity to compete.  SciFinder is another chemical abstract
+service, with a version, SciFinder Scholar, marketed to universities.
+Open-source efforts like ChEMBL and PubChem-BioAssay are on the right
+track, though, unlike some commercial tools, these efforts do not yet
+abstract reactions. Still, improving data quality and standards
+between public and closed sources will be absolutely critical for
+ensuring constant growth, usage, and collaboration between private and
+public parties.
 
@@
 for the plethora of data mining techniques employed in
-cheminformatics. At this point, the nature of Open Data focuses
-primarily on structure and activity data types. On the other hand,
-there is a distinct lack of textual data (such as journal articles)
-that are ``Open''. While PubMed abstracts serve as a proxy
-for journal articles, text mining methods in cheminformatics are
-hindered by not being able to mine the full text of many scientific
-publications.  Interestingly, patent information is publicly
-accessible (cf. Google Patents) and represents an excellent resource
-to support these types of efforts.
+cheminformatics. Currently Open Data focuses primarily on structure
+and activity data types and there is a distinct lack of textual data
+(such as journal articles) that are ``Open''. While PubMed abstracts
+serve as a proxy for journal articles, text mining methods in
+cheminformatics are hindered by not being able to mine the full text
+of many scientific publications.  Interestingly, patent information is
+publicly accessible and can be used resource to support these types of
+efforts. Of course, Open Data does not explicitly address data quality
+issues nor the problems with integrating data sources. In many cases,
+manual curation is the only option to maintain high quality databases.
 
@@
 There is an admitted learning curve, as significant domain
-knowledge is required to attack many problems in cheminformatics due
-to the underlying chemistry of the systems being studied. Indeed, many
-issues faced by chemists do not admit as clean an abstraction as
-``algorithms on a string,'' the way many bioinformatics algorithms can
-be abstracted into a computer science framework.  Hence, while
-one can certainly make significant contributions to cheminformatics by
-just considering structures as graphs, one is limited to somewhat
-abstract problems. Keeping in mind that cheminformatics is
-fundamentally a practical field that serves to help and advance
-experimental chemistry, the key challenges require an understanding of
-the underlying chemical systems.
+knowledge is required to attack many problems in cheminformatics 
+%%due to the underlying chemistry of the systems being studied.
+Indeed, many issues faced by chemists do not admit as clean an
+abstraction as ``algorithms on a string,'' the way many bioinformatics
+algorithms can be abstracted into a computer science framework.
+Hence, while one can certainly make significant contributions to
+cheminformatics by just considering structures as graphs, one is
+limited to somewhat abstract problems. Keeping in mind that
+cheminformatics is fundamentally a practical field that serves to help
+and advance experimental chemistry, the key challenges require an
+understanding of the underlying chemical systems.
 
@@
 \bibliographystyle{abbrv}
-\bibliography{CACM_second_revision}
+\bibliography{resubmission}
 
